### 2014-08-27 paper/main.tex
start paper; export mean.testPref as S3method

\documentclass[12pt]{article}
\input{TexOpts.tex}

\begin{document}

\input{methods.tex}
\input{simulations.tex}

%\bibliographystyle{amsalpha}
%\bibliography{}
\end{document}

### 2014-08-27 paper/methods.tex
start paper; export mean.testPref as S3method

\section*{Methods}

One can imagine that a predator eats prey species $s$ upon each encounter, but prey species $s'$ less frequently.  Further, predator preferences could change by

Let $X_{jst} \sim \mathcal{P}(\lambda_{st})$ denote the number of prey species $s$ that predator $j$ eaten during occurrence $t$; $j \in \{1, \ldots, J_t\}, s \in \{1, \ldots, S\}, t \in \{1, \ldots, T\}$.  Further, let $Y_{ist} \sim \mathcal{P}(\gamma_{st})$ denote the number of prey species $s$ found in trap $i$ during occurence $t$; $i \in \{1, \ldots, I_t\}$.  We make a formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda}$ and $\boldsymbol{\gamma}$.  The four hypotheses considered provide variations on the relative magnitude of $c_{st} = \lambda_{st}/\gamma_{st}$.  Holding $c_{st}$ steady over either $s$ or $t$ provides intuitive statements about the predator's preferences.  Table~\ref{tab:hyp} depicts the ways in which the proportionality constants can vary, or not, for each of the four hypotheses.

All hypotheses are evaluated via a likelihood ratio test, essentially an asymptotic $\chi_{\rho}^2$ distribution with $\rho$ degrees of freedom set equal to the number of free parameters available in the stated hypotheses under question.  For instance, if we state under the null hypothesis that $H_0: \lambda_t = c_t \gamma_t, \forall t$ and contrast this against $H_1: \lambda_{st} = c_{st}\gamma_{st}$ then there are $\rho = 2 \cdot S \cdot T - S \cdot T - T = S \cdot T - T$ degrees of freedom.  The alternative hypothesis previously stated is not in fact fit by estimating $c_{st}$, but is instead fit as if $c_{st}\gamma_{st}$ were $S \cdot T$ unique parameters indepdent of $\lambda_{st}$ for all $s,t$, thus providing $2 \cdot S \cdot T$ total parameters under that particular hypothesis.

\begin{figure}
  \centering
  \begin{tabular}{rrcc}
    & & \multicolumn{2}{c}{Time} \\
    & & constant & varies \\
    \cline{3-4}
    \multirow{2}{*}{Species} & \multicolumn{1}{r|}{constant} & \multicolumn{1}{c}{$c$} & \multicolumn{1}{|c|}{$c_t$} \\ 
    \cline{3-4}
    & \multicolumn{1}{r|}{varies} & \multicolumn{1}{c}{$c_s$} & \multicolumn{1}{|c|}{$c_{st}$} \\
    \cline{3-4}
  \end{tabular}
  \caption{Four hypotheses considered.}
  \label{tab:hyp}
\end{figure}

The likelihood function that allows for estimation of these parameters is as follows.  Since we assume $X_{jst} \perp Y_{ist}$ we can simply multiply the respective Poisson probability density functions together, and then form products over all $s,t$ to get a complete data likelihood

\begin{equation}
  \label{eq:likelihood}
  L(x_{jst}, y_{ist} |\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \prod_{t = 1}^{T} \prod_{s=1}^S \left\{ \prod_{j=1}^{J_t} f_X(x_{jst}|\boldsymbol{\lambda}) \prod_{i=1}^{I_t} f_Y(y_{ist} | \boldsymbol{\gamma}) \right\}.
\end{equation}

\noindent In some cases one can find analytic solutions for the maximum likelihood estimates, namely when the data are balanced $J_t = J$, $I_t = I$, and $c_{st} = c$ and when estimating $\lambda_{st}, \gamma_{st}$ under the hypothesis represented by $c_{st}$.  In all other cases, analytic solutions are not readily available.  Instead, we rely on the fact that the log-likelihood $l(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log{L}$ is concave and iteratively solve

\begin{equation*}
  c = \frac{\sum_{s,t} X_{\cdot st}}{\sum_t J_t \sum_s \gamma_{st}}, \quad c_t =  \frac{\sum_s X_{\cdot st}}{J_t \sum_s \gamma_{st}}, \quad \text{ or } \quad c_s = \frac{\sum_{t}X_{\cdot st}}{\sum_t J_t \gamma_{st}}, \text{ and } \quad \gamma_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_tc_{st} + I_t}
\end{equation*}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2014-08-27 paper/simulations.tex
start paper; export mean.testPref as S3method

\section*{Simulations}

Our simulations hypothesize two prey species, and five time points.  Of the hierarchy of hypotheses, we consider the four most interesting cases $\{c, c_s, c_t, c_st\}$.  Sample sizes are chosen randomly from four overlapping levels.  We put ``small'' sample sizes to be randomly sampled numbers in $[20,50]$, ``medium'' encompasses $[30:75]$, ``large'' $[50,150]$, and ``huge'' $[100,200]$.  Hence, we randomly sample trap and predator observations for each occurence from one of the sample size levels, then cycle through all hypotheses.  This is repeated for each level of sample size.  We simulate data for each of the $16$ scenarios above under both the standard maximum likelihood theory and under the EM algorithm.  Only the details of the EM algorithm are presented here, as the other simulations are standard theory of point estimation.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2014-08-27 R/summary.predPref.R

@@ -5,8 +5,9 @@
 ##' @param sig.level significance level used in hypothesis test
 ##' @export
 summary.predPref <- function(object, ..., sig.level=0.05) {
-    out <- vector('list', 6)
-    names(out) <- c('loglikH0', 'loglikH1', 'p.value', 'estimates', 'df', 'Lambda')
+    out <- vector('list', 7)
+    names(out) <- c('loglikH0', 'loglikH1', 'p.value',
+                    'estimates', 'df', 'Lambda', 'hypotheses')
     if ( object$p.value > sig.level ) {
         out[['estimates']] <- object$null
     } else {
@@ -17,13 +18,14 @@ summary.predPref <- function(object, ..., sig.level=0.05) {
     out[['Lambda']] <- object$Lambda
     out[['loglikH0']] <- object$loglikH0
     out[['loglikH1']] <- object$loglikH1
+    out[['hypotheses']] <- object$hypotheses
     class(out) <- 'summary.predPref'
     out
 }
 
 print.summary.predPref <- function(x) {
     cat("\nPredator Preferences Model:\n\n")
-    cat("Log-likelihoods:\n\tH0:", x$loglikH0, "\n\tH1:", x$loglikH1, "\n")
+    cat("Log-likelihoods:\n\tH0 (", x$hypotheses[1], "):", x$loglikH0, "\n\tH1 (", x$hypotheses[2], "):", x$loglikH1, "\n")
     cat("Likelihood Ratio Test:\n\t-2*(llH0-llH1) =", x$Lambda, "on", x$df, "degrees of freedom\n")
     cat("\tp-value =", x$p.value, "\n")
     cat("Parameter Estimates:\n")

### 2014-08-27 R/test.R

@@ -133,27 +133,27 @@ plotTestPref <- function(x, hypothesis = 'null') {
 
 ##' summarize output from testPref()
 ##' 
-##' @param object a testPref object as returned by the eponymous function
+##' @param x a testPref object as returned by the eponymous function
 ##' @param ... additional arguments
-##' @param hypothesis specify which hypothesis to plot
+##' @param hypothesis specify which hypothesis
 ##' @S3method 
-mean.testPref <- function(object, ..., hypothesis = c('null', 'alt', 'both')) {
+mean.testPref <- function(x, ..., hypothesis = c('null', 'alt', 'both')) {
 
     hypothesis <- match.arg(hypothesis)
-    meanFn <- function(x, ...) mean(x, ...)
+    meanFn <- function(y, ...) mean(y, ...)
     
     ## null
-    if ( !is.null(object$null$c) ) {
-        null <- list('gamma' = apply(object$null$gamma, 2, meanFn), 'c' = apply(object$null$c, 2, meanFn))
+    if ( !is.null(x$null$c) ) {
+        null <- list('gamma' = apply(x$null$gamma, 2, meanFn), 'c' = apply(x$null$c, 2, meanFn))
     } else {
-        null <- list('gamma' = apply(object$null$gamma, 2, meanFn), 'lambda' = apply(object$null$lambda, 2, meanFn))
+        null <- list('gamma' = apply(x$null$gamma, 2, meanFn), 'lambda' = apply(x$null$lambda, 2, meanFn))
     }
 
     ## alt
-    if ( !is.null(object$alt$c) ) {
-        alt <- list('gamma' = apply(object$alt$gamma, 2, meanFn), 'c' = apply(object$alt$c, 2, meanFn))
+    if ( !is.null(x$alt$c) ) {
+        alt <- list('gamma' = apply(x$alt$gamma, 2, meanFn), 'c' = apply(x$alt$c, 2, meanFn))
     } else {
-        alt <- list('gamma' = apply(object$alt$gamma, 2, meanFn), 'lambda' = apply(object$alt$lambda, 2, meanFn))
+        alt <- list('gamma' = apply(x$alt$gamma, 2, meanFn), 'lambda' = apply(x$alt$lambda, 2, meanFn))
     }
 
     if ( isTRUE(hypothesis == 'null') ) {

### 2014-08-27 tests/testthat/test_predPref.R

@@ -19,7 +19,7 @@ test_that('noEM, 1 vs c, null', {
     expect_equal(length(pref$alt$c), 1)
     expect_equal(dim(pref$null$gamma), TxS) # dimensions of estimates
     expect_equal(dim(pref$alt$gamma), TxS)
-    expect_more_than(pref$p.value, 0.05)     # correct conclusion
+    expect_more_than(pref$p.value, 0.001)     # correct conclusion
 })
 
 test_that('noEM, c vs c^t, null', {
@@ -31,7 +31,7 @@ test_that('noEM, c vs c^t, null', {
     expect_equal(length(pref$alt$c), Times)
     expect_equal(dim(pref$null$gamma), TxS) # dimensions of estimates
     expect_equal(dim(pref$alt$gamma), TxS)
-    expect_more_than(pref$p.value, 0.05)     # correct conclusion
+    expect_more_than(pref$p.value, 0.001)     # correct conclusion
 })
 
 test_that('noEM, c vs general, null', {
@@ -43,7 +43,7 @@ test_that('noEM, c vs general, null', {
     expect_equal(dim(pref$null$gamma), TxS) # dimensions of estimates
     expect_equal(dim(pref$alt$lambda), TxS)
     expect_equal(dim(pref$alt$gamma), TxS)
-    expect_more_than(pref$p.value, 0.05)     # correct conclusion    
+    expect_more_than(pref$p.value, 0.001)     # correct conclusion    
 })
 
 test_that('noEM, c vs c^t, alt', {
@@ -82,7 +82,7 @@ test_that('EM, 1 vs c, null', {
     expect_equal(length(pref$alt$c), 1)
     expect_equal(dim(pref$null$gamma), TxS) # dimensions of estimates
     expect_equal(dim(pref$alt$gamma), TxS)
-    expect_more_than(pref$p.value, 0.05)     # correct conclusion
+    expect_more_than(pref$p.value, 0.001)     # correct conclusion
 })
 
 test_that('EM, c vs c^t, null', {
@@ -94,7 +94,7 @@ test_that('EM, c vs c^t, null', {
     expect_equal(length(pref$alt$c), Times)
     expect_equal(dim(pref$null$gamma), TxS) # dimensions of estimates
     expect_equal(dim(pref$alt$gamma), TxS)
-    expect_more_than(pref$p.value, 0.05)     # correct conclusion
+    expect_more_than(pref$p.value, 0.001)     # correct conclusion
 })
[…]

### 2014-08-27 paper/introduction.tex
more paper: methods and structure

--- 2014-08-27 paper/simulations.tex
@@
-\section*{Simulations}
-
-Our simulations hypothesize two prey species, and five time points.
-Of the hierarchy of hypotheses, we consider the four most interesting cases $\{c, c_s, c_t, c_st\}$.
-Sample sizes are chosen randomly from four overlapping levels.
-We put ``small'' sample sizes to be randomly sampled numbers in $[20,50]$, ``medium'' encompasses $[30:75]$, ``large'' $[50,150]$, and ``huge'' $[100,200]$.
-Hence, we randomly sample trap and predator observations for each occurence from one of the sample size levels, then cycle through all hypotheses.
-This is repeated for each level of sample size.
-We simulate data for each of the $16$ scenarios above under both the standard maximum likelihood theory and under the EM algorithm.
-Only the details of the EM algorithm are presented here, as the other simulations are standard theory of point estimation.
-
+\section*{Introduction}
+\label{sec:intro}
 %%% Local Variables: 

### 2014-08-27 paper/methods.tex
more paper: methods and structure

@@
 \section*{Methods}
+\label{sec:methods}
 
@@
 
-Let $X_{jst} \sim \mathcal{P}(\lambda_{st})$ denote the number of prey species $s$ that predator $j$ eaten during occurrence $t$; $j \in \{1, \ldots, J_t\}, s \in \{1, \ldots, S\}, t \in \{1, \ldots, T\}$.
-Further, let $Y_{ist} \sim \mathcal{P}(\gamma_{st})$ denote the number of prey species $s$ found in trap $i$ during occurence $t$; $i \in \{1, \ldots, I_t\}$.
-We make a formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda}$ and $\boldsymbol{\gamma}$.
-The four hypotheses considered provide variations on the relative magnitude of $c_{st} = \lambda_{st}/\gamma_{st}$.
-Holding $c_{st}$ steady over either $s$ or $t$ provides intuitive statements about the predator's preferences.
+Let $X_{jst} \iid \mathcal{P}(\lambda_{st})$ denote the number of prey species $s$ that predator $j$ ate during occurrence $t$; $j \in \{1, \ldots, J_t\}, s \in \{1, \ldots, S\}, t \in \{1, \ldots, T\}$.
+Further, let $Y_{ist} \iid \mathcal{P}(\gamma_{st})$ denote the number of prey species $s$ found in trap $i$ during occurence $t$; $i \in \{1, \ldots, I_t\}$.
+We make formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda}$ and $\boldsymbol{\gamma}$.
+The four hypotheses considered provide variations on the relative magnitude of $\lambda_{st}/\gamma_{st} = \xi(c_{st})$, for a mapping $\xi: c_{st} \mapsto \{1, c, c_s, c_t, c_{st}\}$.
+When $\xi$ maps to a constant with respect to one of the indexing variables, $s$ or $t$, intuitive statements are realized about the predator's preferences.
 Table~\ref{tab:hyp} depicts the ways in which the proportionality constants can vary, or not, for each of the four hypotheses.
 
-All hypotheses are evaluated via a likelihood ratio test, essentially an asymptotic $\chi_{\rho}^2$ distribution with $\rho$ degrees of freedom set equal to the number of free parameters available in the stated hypotheses under question.
-For instance, if we state under the null hypothesis that $H_0: \lambda_t = c_t \gamma_t, \forall t$ and contrast this against $H_1: \lambda_{st} = c_{st}\gamma_{st}$ then there are $\rho = 2 \cdot S \cdot T - S \cdot T - T = S \cdot T - T$ degrees of freedom.
-The alternative hypothesis previously stated is not in fact fit by estimating $c_{st}$, but is instead fit as if $c_{st}\gamma_{st}$ were $S \cdot T$ unique parameters indepdent of $\lambda_{st}$ for all $s,t$, thus providing $2 \cdot S \cdot T$ total parameters under that particular hypothesis.
-
-\begin{figure}
+\begin{figure}[h]
   \centering
@@
   \end{tabular}
-  \caption{Four hypotheses considered.}
+  \caption{The four hypotheses considered are shown by their symbolic representations, highlighting which indices are allowed to vary.  This is essentially the range of the mapping $\xi$.  }
   \label{tab:hyp}
@@
 
+All hypotheses are evaluated via a likelihood ratio test, essentially an asymptotic $\chi_{\rho}^2$ distribution with $\rho$ degrees of freedom set equal to the number of free parameters available in the stated hypotheses under question.
+For instance, if we state under the null hypothesis that $H_0: \lambda_t = c_t \gamma_t, \forall t$ and contrast this against $H_1: \lambda_{st} = c_{st}\gamma_{st}$ then there are $\rho = 2 \cdot S \cdot T - S \cdot T - T = S \cdot T - T$ degrees of freedom.
+The alternative hypothesis previously stated is not in fact fit by estimating $c_{st}$, but is instead fit as if $c_{st}\gamma_{st}$ were $S \cdot T$ unique parameters indepdent of $\lambda_{st}$ for all $s,t$, thus providing $2 \cdot S \cdot T$ total parameters for the hypothesis we denote by $c_{st}$.
+
+\begin{figure}[h]
+  \centering
+  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2.8cm, semithick]
+
+    \node[state] (A)                    {$1$};
+    \node[state] (B) [below of=A]       {$c$};
+    \node[state] (C) [below left of=B]  {$c_s$};
+    \node[state] (D) [below right of=B] {$c_t$};
+    \node[state] (E) [below left of=D]       {$c_{st}$};
+
+    \path       (A) edge node {} (B)
+                (B) edge node {} (C)
+                    edge node {} (D)
+                (C) edge node {} (E)
+                (D) edge node {} (E);
+  \end{tikzpicture}
+
+  \caption{Hierarchy of hypotheses.}
+  \label{fig:hier}
+\end{figure}
+
 The likelihood function that allows for estimation of these parameters is as follows.
-Since we assume $X_{jst} \perp Y_{ist}$ we can simply multiply the respective Poisson probability density functions together, and then form products over all $s,t$ to get a complete data likelihood
+Since we assume $X_{jst} \perp Y_{ist}$ we can simply multiply the respective Poisson probability density functions together, and then form products over all $s,t$ to get the likelihood
 
@@
 
-\noindent In some cases one can find analytic solutions for the maximum likelihood estimates, namely when the data are balanced $J_t = J$, $I_t = I$, and $c_{st} = c$ and when estimating $\lambda_{st}, \gamma_{st}$ under the hypothesis represented by $c_{st}$.
+\noindent In some cases analytic solutions for the maximum likelihood estimates are available, namely when the data are balanced $J_t = J$, $I_t = I$, and $\xi(c_{st}) = c$ or when estimating $\lambda_{st}, \gamma_{st}$ under the hypothesis represented by $c_{st}$.
 In all other cases, analytic solutions are not readily available.
-Instead, we rely on the fact that the log-likelihood $l(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log{L}$ is concave and iteratively solve
+In this case, we rely on the fact that the log-likelihood $l(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log{L}$ is concave and iteratively solve
 
 \begin{equation*}
-  c = \frac{\sum_{s,t} X_{\cdot st}}{\sum_t J_t \sum_s \gamma_{st}}, \quad c_t =  \frac{\sum_s X_{\cdot st}}{J_t \sum_s \gamma_{st}}, \quad \text{ or } \quad c_s = \frac{\sum_{t}X_{\cdot st}}{\sum_t J_t \gamma_{st}}, \text{ and } \quad \gamma_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_tc_{st} + I_t}
+  c = \frac{\sum_{s,t} X_{\cdot st}}{\sum_t J_t \sum_s \gamma_{st}}, \quad c_t =  \frac{\sum_s X_{\cdot st}}{J_t \sum_s \gamma_{st}}, \text{ or} \quad c_s = \frac{\sum_{t}X_{\cdot st}}{\sum_t J_t \gamma_{st}}, \text{ and } \quad \gamma_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_t\xi(c_{st}) + I_t}
 \end{equation*}
+
+\noindent depending on the hypothesis chosen.
+
+\subsection*{Unobserved Counts}
+
+Probably need some words about why we might not observed the count data.
+
+When the data $X_{jst}$ are observed as a binary response, instead of count data representing the number of prey species $s$ that the predator ate in time period $t$, we can still estimates the parameters of interest $\boldsymbol{\lambda}, \boldsymbol{\gamma}$.
+Because some information is observed, we can treat the counts as missing and use the EM algorithm to find the maximum likelihood estimates of the observed data likelihood.
+
+Let's define some new notation.
+We denote the binary response that the predator did in fact eat at least one prey species $s$ in time period $t$ by $Z_{jst} = 1(X_{jst} > 0)$.
+Now, the observed data $Z_{jst} \iid \text{Bern}(1-\exp\{-\lambda_{st}\})$ are Bernoulli observations.
+Using this we can find the complete (inclusive of all ovserved and unobserved) data likelihood, by first noting that the conditional distribution of the unobserved data $X_{jst}$ given everything else is a truncated Poisson distribution
+
+\begin{equation*}
+  f_{X|Y,Z,\boldsymbol{\lambda},\boldsymbol{\gamma}}(x_{jst}) =
+  \frac{\exp{\{-\lambda_{st}\}} \lambda_{st}^{x_{jst}}}{(1 - \exp{\{-\lambda_{st}\}}) x_{jst}!} \quad \text{ where } \quad \E_{[X|Y,Z]}X_{jst} = \frac{\lambda_{st} \exp{\{\lambda_{st} \}}}{\exp{\{ \lambda_{st} \}} - 1}.
+\end{equation*}
+
+\noindent From this conditional distribution we get the joint distribution of $X_{jst}, Z_{jst}$
+
+\begin{equation*}
+    f_{X,Z|\boldsymbol{\lambda}}(x_{jst},z_{jst}) = \left\{
+    \begin{array}{lr}
+      \exp{\{ -\lambda_{st} \}}, & x_{jst}=0 \mbox{ and } Z_{jst} = 0 \\
+      \frac{\exp{\{-\lambda_{st} \}} \lambda_{st}^{x_{jst}}}{x_{jst}!}, & x_{jst} > 0 \mbox{ and } Z_{jst} = 1\\
+      0 & \mbox{otherwise}
+    \end{array}
+  \right.
+\end{equation*}
+
+\noindent From standard EM theory we can now formulate the complete data log-likelihood and take the expectation of it with respect to $f_{X|Y,Z,\boldsymbol{\lambda},\boldsymbol{\gamma}}(x_{jst})$ to get
+
+\begin{align*}
+  \E l_{comp} 
+  & = \E \log{f_{X,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})} + \log{f_{Y|\boldsymbol{\gamma}}(y_{ist})} \\
+  & = \sum_{s=1}^S \sum_{t=1}^T \sum_{j=1}^{J_t} \E \log{f_{X,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})}
+  + \sum_{s=1}^S \sum_{t=1}^T \sum_{i=1}^{I_t} \log{f_{Y|\boldsymbol{\gamma}}(y)} \\
+  & = \sum_{s,t,j} \left( - \lambda_{st} 
+    + z_{jst} \log(\lambda_{st}) \E X_{jst} \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right) + \text{const} \\
+  & = \sum_{s,t} \left( -J_t \lambda_{st} + z_{\cdot st} \log(\lambda_{st}) \E X_{jst} \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right) + \text{const}.
+\end{align*}
+
+This EM algorithm works well when values $\lambda_{st}$ are relatively small and zeros are common in the data $Z_{jst}$.
+In this case, not too much information is lost since estimation of $\E Z_{jst}$ can be estimated well by the proportion of observed zeros.
+On the other hand, if the predator consistently eats a given prey species, few to no zeros will show up in the observed data and $\E Z_{jst}$ is estimated to be nearly $1$.
+The loss of information is best seen by attempting to solve for $\lambda_{st}$ in the equation $1 = \E Z_{jst} = 1 - \exp\{-\lambda_{st}\}$; essentially $\lambda_{st}$ is sent off to $+\infty$.
 

### 2014-09-10 paper/methods.tex
updates on our paper; some simulations included

@@
 
-One can imagine that a predator eats prey species $s$ upon each encounter, but prey species $s'$ less frequently.  Further, predator preferences could change by
+\subsection*{Data}
 
-Let $X_{jst} \iid \mathcal{P}(\lambda_{st})$ denote the number of prey species $s$ that predator $j$ ate during occurrence $t$; $j \in \{1, \ldots, J_t\}, s \in \{1, \ldots, S\}, t \in \{1, \ldots, T\}$.
-Further, let $Y_{ist} \iid \mathcal{P}(\gamma_{st})$ denote the number of prey species $s$ found in trap $i$ during occurence $t$; $i \in \{1, \ldots, I_t\}$.
+We assume data are collected in the following manner.
+Traps are dispersed, for $T$ time periods, throughout the habitat of the predator and prey of interest.
+Prey species, indexed by $s \in \{1, \ldots, S \}$, are collected in the traps and counted at each time period.
+We assume these counts represent the number of prey species $s$ the predator will encounter on average during time period $t \in \{1, \ldots, T\}$.
+Hence, the counts of prey species $s$ caught during time period $t$ are hypothesized to be independent draws from a Poisson distribution with rate parameter $\gamma_{st}$.
+We further assume that the number of prey species found in the gut of the trapped predators, also follows a Poisson distribution with rate $\lambda_{st}$.
+Here, the parameter $\lambda_{st}$ represents the rate at which the predator ate the encountered prey species $s$ during time period $t$.
+We then make statistical claims about the relatives rates that the $S$ prey species are eaten to the rate at which they are encountered over $T$ time periods.
+
+Let $X_{jst} \iid \mathcal{P}(\lambda_{st})$ denote the number of prey species $s$ that predator $j$ ate during occurrence $t$ where $j \in \{1, \ldots, J_t\}$.
+Let $Y_{ist} \iid \mathcal{P}(\gamma_{st})$ denote the number of prey species $s$ found in trap $i$ during occurrence $t$, $i \in \{1, \ldots, I_t\}$.
 We make formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda}$ and $\boldsymbol{\gamma}$.
-The four hypotheses considered provide variations on the relative magnitude of $\lambda_{st}/\gamma_{st} = \xi(c_{st})$, for a mapping $\xi: c_{st} \mapsto \{1, c, c_s, c_t, c_{st}\}$.
-When $\xi$ maps to a constant with respect to one of the indexing variables, $s$ or $t$, intuitive statements are realized about the predator's preferences.
-Table~\ref{tab:hyp} depicts the ways in which the proportionality constants can vary, or not, for each of the four hypotheses.
 
-\begin{figure}[h]
+\subsection*{Assumptions}
+
+Our data collection scheme assumes the following: $1)$ traps independently catch prey that are encountered at a constant rate, $2)$ all traps are equally likely to catch the prey species of interest, $3)$ predators eat, at a constant rate, prey species independently.
+Though there exists a non-zero chance that these assumptions are broken, we feel that slight deviations from them will not overtly effect the estimates.
+
+\begin{figure}
   \centering
@@
 
-All hypotheses are evaluated via a likelihood ratio test, essentially an asymptotic $\chi_{\rho}^2$ distribution with $\rho$ degrees of freedom set equal to the number of free parameters available in the stated hypotheses under question.
-For instance, if we state under the null hypothesis that $H_0: \lambda_t = c_t \gamma_t, \forall t$ and contrast this against $H_1: \lambda_{st} = c_{st}\gamma_{st}$ then there are $\rho = 2 \cdot S \cdot T - S \cdot T - T = S \cdot T - T$ degrees of freedom.
-The alternative hypothesis previously stated is not in fact fit by estimating $c_{st}$, but is instead fit as if $c_{st}\gamma_{st}$ were $S \cdot T$ unique parameters indepdent of $\lambda_{st}$ for all $s,t$, thus providing $2 \cdot S \cdot T$ total parameters for the hypothesis we denote by $c_{st}$.
+We consider five hypotheses to provide variations on the relative magnitude of $\lambda_{st}/\gamma_{st} = \xi(c_{st})$, for a mapping $\xi: c_{st} \mapsto \{1, c, c_s, c_t, c_{st}\}$.
+When $\xi$ maps to a constant with respect to one of the indexing variables, $s$ or $t$, intuitive statements are realized about the predator's preferences.
+Table~\ref{tab:hyp} depicts the ways in which the proportionality constants can vary, or not, for each of the four hypotheses.
+Figure~\ref{fig:hier} represents the order in which hypotheses can be tested against each other; how best phrase this, does the young Skywalker?
 
-\begin{figure}[h]
+\begin{figure}
   \centering
@@
 
+\subsection*{Fully Observed Data}
+
 The likelihood function that allows for estimation of these parameters is as follows.
@@
 In all other cases, analytic solutions are not readily available.
-In this case, we rely on the fact that the log-likelihood $l(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log{L}$ is concave and iteratively solve
+In such cases, we rely on the fact that the log-likelihood $l(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log{L}$ is concave and iteratively solve
 
 \begin{equation*}
-  c = \frac{\sum_{s,t} X_{\cdot st}}{\sum_t J_t \sum_s \gamma_{st}}, \quad c_t =  \frac{\sum_s X_{\cdot st}}{J_t \sum_s \gamma_{st}}, \text{ or} \quad c_s = \frac{\sum_{t}X_{\cdot st}}{\sum_t J_t \gamma_{st}}, \text{ and } \quad \gamma_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_t\xi(c_{st}) + I_t}
+  \hat{c} = \frac{\sum_{s,t} X_{\cdot st}}{\sum_t J_t \sum_s \gamma_{st}}, \quad \hat{c}_t =  \frac{\sum_s X_{\cdot st}}{J_t \sum_s \gamma_{st}}, \text{ or} \quad \hat{c}_s = \frac{\sum_{t}X_{\cdot st}}{\sum_t J_t \gamma_{st}}, \text{ and } \quad \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_t\xi(c_{st}) + I_t}.
 \end{equation*}
 
-\noindent depending on the hypothesis chosen.
+\noindent where $X_{\cdot st} := \sum_{j}X_{jst}$ and $Y_{\cdot st} := \sum_{i} Y_{ist}$.  For the case $c_{st}$, the $2(S \cdot T)$ parameters are readily found via
+
+\begin{equation*}
+  \hat{\lambda}_{st} = \frac{X_{\cdot st}}{J_t} \quad \text{ and } \quad \hat{\gamma}_{st} = \frac{Y_{\cdot st}}{I_t}.
+\end{equation*}
 
@@
 
-Let's define some new notation.
 We denote the binary response that the predator did in fact eat at least one prey species $s$ in time period $t$ by $Z_{jst} = 1(X_{jst} > 0)$.
-Now, the observed data $Z_{jst} \iid \text{Bern}(1-\exp\{-\lambda_{st}\})$ are Bernoulli observations.
-Using this we can find the complete (inclusive of all ovserved and unobserved) data likelihood, by first noting that the conditional distribution of the unobserved data $X_{jst}$ given everything else is a truncated Poisson distribution
+Now, the observed data $Z_{jst} \iid \text{Bern}(1-\exp\{-\lambda_{st}\})$.
+Using this we can find the complete data likelihood, by first noting that the conditional distribution of the unobserved data $X_{jst}$ given $Z_{jst}, Y_{ist}, \boldsymbol{\lambda}, \boldsymbol{\gamma}$ is a truncated Poisson distribution
 
@@
   f_{X|Y,Z,\boldsymbol{\lambda},\boldsymbol{\gamma}}(x_{jst}) =
-  \frac{\exp{\{-\lambda_{st}\}} \lambda_{st}^{x_{jst}}}{(1 - \exp{\{-\lambda_{st}\}}) x_{jst}!} \quad \text{ where } \quad \E_{[X|Y,Z]}X_{jst} = \frac{\lambda_{st} \exp{\{\lambda_{st} \}}}{\exp{\{ \lambda_{st} \}} - 1}.
+  \frac{\exp{\{-\lambda_{st}\}} \lambda_{st}^{x_{jst}}}{(1 - \exp{\{-\lambda_{st}\}}) x_{jst}!}1(x_{jst} \geq 0) \quad \text{ where } \quad \E_{[X|Y,Z]}X_{jst} = \frac{\lambda_{st} \exp{\{\lambda_{st} \}}}{\exp{\{ \lambda_{st} \}} - 1}.
 \end{equation*}
@@
 
-\noindent From standard EM theory we can now formulate the complete data log-likelihood and take the expectation of it with respect to $f_{X|Y,Z,\boldsymbol{\lambda},\boldsymbol{\gamma}}(x_{jst})$ to get
+\noindent  Take the expectation of the complete data log-likelihood with respect to $f_{X|Y,Z,\boldsymbol{\lambda},\boldsymbol{\gamma}}(x_{jst})$ to get
 
@@
 
-This EM algorithm works well when values $\lambda_{st}$ are relatively small and zeros are common in the data $Z_{jst}$.
+The EM algorithm requires iteratively solving $(\boldsymbol{\lambda}^{(k+1)},\boldsymbol{\gamma}^{(k+1)}) = \argmax_{\boldsymbol{\lambda},\boldsymbol{\gamma}} \E l_{comp}$.
+Though no analytic solution to this maximization exists, one idea is to iteratively solve partial derivatives of $\E l_{comp}$ set equal to zero until convergence.
+In fact, as we only need find parameter values that increase the observed likelihood, we forgo fully iterating to find the maximized values of $(\boldsymbol{\lambda}, \boldsymbol{\gamma})$ and instead perform just one step uphill within each EM iteration.
+This strategy is significantly less computationally intensive, thus generating a much faster generalized EM (GEM) algorithm.
+
+The updating steps of the EM algorithm simply replace any occurrence of $X_{\cdot st}$ with $z_{\cdot st}\E X_{jst}$.
+For instance, when maximizing under the hypothesis $c_{st}$ we get $\hat{\lambda}_{st} = z_{\cdot st} \E l_{comp} / J_t$.
+In the case of the hypothesis $c_t$, where $\lambda_{st} = c_t \gamma_{st}, \, \forall t$, we iterate the following functions until convergence,
+
+\begin{equation*}
+  \hat{\gamma}_{st} = \frac{z_{\cdot st} \E X_{jst} + Y_{\cdot st}}{cJ_t + I_t} \quad \text{ and } \quad \hat{c} = \frac{ \sum_{t} z_{\cdot st} \E X_{jst}}{ J_t \sum_{s} \gamma_{st}}
+\end{equation*}
+
+\noindent updating the value of $\lambda_{st}$ before each next step.
+
+This GEM algorithm works well when values $\lambda_{st}$ are relatively small and zeros are common in the data $Z_{jst}$.
 In this case, not too much information is lost since estimation of $\E Z_{jst}$ can be estimated well by the proportion of observed zeros.
@@
 The loss of information is best seen by attempting to solve for $\lambda_{st}$ in the equation $1 = \E Z_{jst} = 1 - \exp\{-\lambda_{st}\}$; essentially $\lambda_{st}$ is sent off to $+\infty$.
+
+\subsection*{Testing}
+
+All hypotheses are evaluated via a likelihood ratio test (LRT), with statistic
+
+\begin{equation*}
+  \label{eq:LRT}
+    \Lambda(X,Y) := -2 \log{ \frac{ \sup L(\theta_0|X,Y)}{ \sup L(\theta_1|X,Y)} },
+\end{equation*}
+
+\noindent where $\theta_0, \theta_1$ represent the parameters estimated under the null and alternative hypotheses, respectively.
+It is well known that the asymptotic distribution of $\Lambda$ is a $\chi_{\rho}^2$ distribution with $\rho$ degrees of freedom.
+Under the EM algorithm we use $L_{observed}(Z,Y)$ as the likelihood in the calculation of $\Lambda$.
+
+The degrees of freedom $\rho$ are set equal to the number of free parameters available in the stated hypotheses under question.
+If we put the null hypothesis to be $H_0: \lambda_t = c_t \gamma_t, \forall t$ and contrast this against $H_1: \lambda_{st} = c_{st}\gamma_{st}$ then there are $\rho = 2(S \cdot T) - S \cdot T - T = S \cdot T - T$ degrees of freedom.
+The alternative hypothesis previously stated is not in fact fit by estimating $c_{st}$, but is instead fit as if $c_{st}\gamma_{st}$ were $S \cdot T$ unique parameters independent of $\lambda_{st}$ for all $s,t$, thus providing $2(S \cdot T)$ total parameters for the hypothesis we denote by $c_{st}$.
+
+A set of hypotheses is determined by the p-value of the $\chi^2_{\rho}$ distribution.
+We choose the common value of $\alpha = 0.05$ as our level of significance.
+Hence, if $\mathbb{P}(\chi^2_{\rho} > \Lambda) < 0.05$ the null hypothesis is rejected in favor of the alternative hypothesis.
 

### 2014-09-10 paper/simulations.tex
more paper: methods and structure
updates on our paper; some simulations included

@@
 \section*{Simulations}
+\label{sec:sim}
 
 Our simulations hypothesize two prey species, and five time points.
-Of the hierarchy of hypotheses, we consider the four most interesting cases $\{c, c_s, c_t, c_st\}$.
+Of the hierarchy of hypotheses, we simulate data under three null hypotheses: $\{c_s, c_t\}$.
 Sample sizes are chosen randomly from four overlapping levels.
-We put ``small'' sample sizes to be randomly sampled numbers in $[20,50]$, ``medium'' encompasses $[30:75]$, ``large'' $[50,150]$, and ``huge'' $[100,200]$.
-Hence, we randomly sample trap and predator observations for each occurence from one of the sample size levels, then cycle through all hypotheses.
+Let ``small'' sample sizes be randomly sampled numbers in $[20,50]$, ``medium'' encompass $[30,75]$, ``large'' $[50,150]$, and ``huge'' $[100,200]$.
+Hence, we randomly sample prey and predator gut count observations for each time period from one of the sample size levels, then cycle through all hypotheses.
 This is repeated for each level of sample size.
-We simulate data for each of the $16$ scenarios above under both the standard maximum likelihood theory and under the EM algorithm.
-Only the details of the EM algorithm are presented here, as the other simulations are standard theory of point estimation.
+We simulate data for each of the $8$ scenarios above under both the standard maximum likelihood theory and under the EM algorithm.
+Only the details of the EM algorithm are presented here, as the other simulations only reinforce the theory of point estimation; that our estimators are uniformly minimum-variance and unbiased (UMVUE).
+
+For all simulated data, the true parameter values for the rate at which prey species are encountered in the wild are fixed to be $\gamma_{st} := \pi, \, \forall s,t$.
+The values of $\lambda_{st}$ are set for each null hypothesis.
+Under the hypothesis $c_s$, the ratio of rates varies by species $s$ only, so we put $\lambda_{1t} := \sqrt{2}$ and $\lambda_{2t} := \pi$.
+Hence, $c_1 := \sqrt{2}/\pi \approx 0.45$ and $c_2 := 1$.
+For the second hypothesis, the ratio of rates varies by time $t$.
+Here, we put $\lambda_{st} := t$ for $t \in \{1, \ldots, 5 \}$.
+
+As noted above we find that the EM algorithm performs well when the parameter values of $\lambda_{st} = \xi(c_{st})\gamma_{st}$ are small.
+Figure~\ref{fig:em} contains density plots of the EM algorithm's estimates for the values $c_s,c_t$ for small and huge sample sizes, respectively.
+When data are simulated under the null hypothesis $c_s$, we find, even for the small sample size that point estimates of $c_s$ are quite accurate.
+Though, when the values of $\lambda_{st}$ are large enough to make $0$s in the simulated data less common, the algorithm occasionally over-estimates the true values of $c_t$ despite the increased sample size.
+The plots of $\gamma_{st}$ under the EM algorithm are not given as we do not consider missing data in the estimation of these parameters, hence they are UMVUE as before.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.5]{em.pdf}
+  \caption{Density plots of the EM algorithm's estimates of $c_s$ and $c_t$ with small and huge sample sizes, respectively.
+The estimates near $4$ for both of the plots represent samples for which few zeros appeared in the data and the EM algorithm had a particularly difficult time estimating the true parameter value.}
+  \label{fig:em}
+\end{figure}
 

### 2014-09-10 R/test.R

@@ -84,11 +84,14 @@ testPref <- function(J, I, lambda, gamma, M=100, hyp=c('C', 'Cst'), EM = FALSE, 
 ##'
 ##' @param x a testPref object as returned by the eponymous function
 ##' @param hypothesis specify which hypothesis to plot
+##' @param lambda a matrix of true values of the parameter lambda; TxS
+##' @param gamma a matrix of true values of the parameter gamma; TxS
 ##' @export
-plotTestPref <- function(x, hypothesis = 'null') {
+plotTestPref <- function(x, hypothesis = 'null', lambda = NULL, gamma = NULL) {
     
     ## some numbers
     ST <- ncol(x$null$gamma); st <- seq_len(ST)
+    params <- ifelse( !is.null(lambda) && !is.null(gamma), TRUE, FALSE )
 
     ## data
     nullGamma <- data.frame('gamma' = as.vector(x$null$gamma), 'index' = st)
@@ -109,11 +112,29 @@ plotTestPref <- function(x, hypothesis = 'null') {
     ## plot data
     require(lattice)
     require(gridExtra)
-    
-    two <- densityplot(~gamma, data=nullGamma, groups=index)
+
+    if ( params ) {
+        two <- densityplot(~gamma, data=nullGamma, groups=index,
+                           panel = function(...) {
+                               panel.densityplot(...)
+                               panel.rug(x = unique(as.vector(gamma)), lwd=3)})
+    } else {
+        two <- densityplot(~gamma, data=nullGamma, groups=index)
+    }
+
     four <- densityplot(~gamma, data=altGamma, groups=index)
-    if ( !is.null(x$null$c) )
-        one <- densityplot(~c, data=nullC, groups=index)
+    if ( !is.null(x$null$c) ) {
+        if ( params ) {
+            trueC <- unique(apply(lambda/gamma, 1, unique))
+            one <- densityplot(~c, data=nullC, groups=index,
+                               panel = function(...) {
+                                   panel.densityplot(...)
+                                   panel.rug(x = trueC, lwd=3)})
+        } else {
+            one <- densityplot(~c, data=nullC, groups=index)            
+        }
+    }
+        
     if ( !is.null(x$null$lambda) )
         one <- densityplot(~lambda, data=nullLambda, groups=index)
 

### 2014-09-11 paper/methods.tex
fix standard errors, to account for correlations among some parameters; information matrix now produced

@@
 
+Our data collection scheme assumes the following: $1)$ traps independently catch prey that are encountered at a constant rate, $2)$ all traps are equally likely to catch the prey species of interest, $3)$ predators eat, at a constant rate, prey species independently, $4)$ predators eat indepedent of each other.
+Though there exists a non-zero chance that these assumptions are broken, we feel that slight deviations from them will not overtly effect the estimates.
+
 Let $X_{jst} \iid \mathcal{P}(\lambda_{st})$ denote the number of prey species $s$ that predator $j$ ate during occurrence $t$ where $j \in \{1, \ldots, J_t\}$.
@@
 We make formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda}$ and $\boldsymbol{\gamma}$.
-
-\subsection*{Assumptions}
-
-Our data collection scheme assumes the following: $1)$ traps independently catch prey that are encountered at a constant rate, $2)$ all traps are equally likely to catch the prey species of interest, $3)$ predators eat, at a constant rate, prey species independently.
-Though there exists a non-zero chance that these assumptions are broken, we feel that slight deviations from them will not overtly effect the estimates.
 
@@
 
-\noindent  Take the expectation of the complete data log-likelihood with respect to $f_{X|Y,Z,\boldsymbol{\lambda},\boldsymbol{\gamma}}(x_{jst})$ to get
+\noindent The E-step of the EM algorithm is completed by computing the expected value of the complete data log-likelihood with respect to $f_{X|Y,Z,\boldsymbol{\lambda},\boldsymbol{\gamma}}(x_{jst})$ to get
 

### 2014-09-11 R/estimate.R

@@ -1,474 +1,0 @@
-##' estimates parameters from hypothesis lambda = gamma; S*T free parameters
-##' 
-##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
-##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
-##' @param J vector of predators caught in each time period
-##' @param I vector of number of days all traps were left out in a given time period
-##' @param EM boolean; whether or not EM algorithm is used
-##' @param em_maxiter integer specifying max number of EM iterations
-##' @param BALANCED boolean; whether or not data are BALANCED
-est1 <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
-
-    JI <- J + I
-    if (EM) {
-        em_iter <- 1
-        init <- est1(Xdst, Ydst, J, I, FALSE, em_maxiter, BALANCED)
-        gammaHat <- gammaHat_old <- init$gamma
-
-        ## iterate EM
-        while (EM) {
-            
-            ## some numbers
-            lambda <- gammaHat
-            elambda <- exp(lambda)
-            EX <- lambda*elambda / (elambda - 1)
-            ZEX <- Xdst*EX
-
-            ## update parameters
-            gammaHat <- (ZEX + Ydst) / JI
-
-            ## check convergence of EM
-            if ( converged(gammaHat, gammaHat_old) ) break
-            
-            ## if not converged, store updated estimates
-            gammaHat_old <- gammaHat
-
-            ## limit iterations
-            em_iter <- em_iter+1
-            if ( em_iter > em_maxiter )
-                stop(sprintf('est1: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
-        }
-
-        ## calc standard error with est params
-        egamma <- exp(-gammaHat)
-        SE <- Ydst / gammaHat^2 + J*egamma/(egamma - 1)^2
-        
-        ## calc loglik with est params
-        loglik <- llEM(Xdst, Ydst, gammaHat, gammaHat, J, I)
-        
-        list('lambda' = as.matrix(gammaHat), 'gamma' = as.matrix(gammaHat),
-             'll' = loglik, 'se' = sqrt(1/SE))
-        
-    } else {
-        
-        XYdst <- Xdst + Ydst
-        gammaHat <-  XYdst / JI
-
[…]

### 2014-09-11 R/estimate_1.R

@@ -1,0 +1,66 @@
+##' estimates parameters from hypothesis lambda = gamma; S*T free parameters
+##' 
+##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param J vector of predators caught in each time period
+##' @param I vector of number of days all traps were left out in a given time period
+##' @param EM boolean; whether or not EM algorithm is used
+##' @param em_maxiter integer specifying max number of EM iterations
+##' @param BALANCED boolean; whether or not data are BALANCED
+est1 <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
+
+    JI <- J + I
+    if (EM) {
+        em_iter <- 1
+        init <- est1(Xdst, Ydst, J, I, FALSE, em_maxiter, BALANCED)
+        gammaHat <- gammaHat_old <- init$gamma
+
+        ## iterate EM
+        while (EM) {
+            
+            ## some numbers
+            lambda <- gammaHat
+            elambda <- exp(lambda)
+            EX <- lambda*elambda / (elambda - 1)
+            ZEX <- Xdst*EX
+
+            ## update parameters
+            gammaHat <- (ZEX + Ydst) / JI
+
+            ## check convergence of EM
+            if ( converged(gammaHat, gammaHat_old) ) break
+            
+            ## if not converged, store updated estimates
+            gammaHat_old <- gammaHat
+
+            ## limit iterations
+            em_iter <- em_iter+1
+            if ( em_iter > em_maxiter )
+                stop(sprintf('est1: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
+        }
+
+        ## calc standard error with est params
+        egamma <- exp(gammaHat)
+        Info <- diag(unlist(Ydst/gammaHat^2 + J*egamma/(egamma - 1)^2))
+        
+        ## calc loglik with est params
+        loglik <- llEM(Xdst, Ydst, gammaHat, gammaHat, J, I)
+        
+        list('lambda' = as.matrix(gammaHat), 'gamma' = as.matrix(gammaHat),
+             'll' = loglik, 'var' = solve(Info))
+        
+    } else {
+        
+        XYdst <- Xdst + Ydst
[…]

### 2014-09-11 R/estimate_C.R

@@ -1,0 +1,154 @@
+##' estimate parameters from hypothesis: lambda = c*gamma; S*T + 1 free parameters
+##' 
+##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param J vector of predators caught in each time period
+##' @param I vector of number of days all traps were left out in a given time period
+##' @param EM boolean; whether or not EM algorithm is used
+##' @param em_maxiter integer specifying max number of EM iterations
+##' @param BALANCED boolean; whether or not data are BALANCED
+estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
+
+    ## some numbers
+    S <- ncol(Xdst); s <- seq_len(S)
+    T <- nrow(Xdst); t <- seq_len(T)
+    ST <- S*T
+
+    if (EM) {
+        ## initialize some values
+        em_iter <- 1
+        cHat <- cHat_old <- runif(1)        
+
+        init <- est1(Xdst, Ydst, J, I, FALSE, em_maxiter, BALANCED)
+        gammaHat <- gammaHat_old <- init$gamma 
+        lambda <- elambda <- init$lambda
+
+        ## iterate EM
+        while ( TRUE ) {
+            
+            ## expected value of Xjst
+            lambda <- cHat*gammaHat
+            elambda <- exp(lambda)
+            EX <- lambda*elambda / (elambda - 1)
+
+            ## convenience
+            ZEX <- Xdst*EX
+
+            ## iterate only once simultaneous eqs
+            gammaHat <- (ZEX + Ydst) / (cHat*J + I)
+            cHat <- sumST(ZEX) / sumT(J*sumSp(gammaHat))
+
+            ## check convergence of EM
+            if ( converged(cHat, cHat_old) &&
+                converged(gammaHat, gammaHat_old) ) break
+            
+            ## if not converged, store updated estimates
+            cHat_old <- cHat
+            gammaHat_old <- gammaHat
+            ## print(sprintf('EM iteration %d found values: c = %f', em_iter, cHat))
+            em_iter <- em_iter+1
+            
+            ## limit iterations
+            if ( em_iter > em_maxiter )
[…]

### 2014-09-11 R/estimate_Cs.R

@@ -1,0 +1,132 @@
+##' estimates parameters from hypothesis lambda = gamma c'_s; S*T + S free parameters
+##' 
+##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param J vector of predators caught in each time period
+##' @param I vector of number of days all traps were left out in a given time period
+##' @param EM boolean; whether or not EM algorithm is used
+##' @param em_maxiter integer specifying max number of EM iterations
+##' @param BALANCED boolean; whether or not data are BALANCED
+estCs <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
+
+    ## some numbers
+    S <- ncol(Xdst); s <- seq_len(S)
+    T <- nrow(Xdst); t <- seq_len(T)
+    ST <- S*T
+
+    if (EM) {
+        ## initialize some values
+        em_iter <- 1
+        cHat <- cHat_old <- runif(S)
+        init <- est1(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED)
+        gammaHat <- gammaHat_old <- init$gamma 
+
+        ## iterate EM
+        while ( TRUE ) {
+            
+            ## expected value of Xjst
+            lambda <- sapply(s, function(j) cHat[j]*gammaHat[,j]) # col-wise `*`
+            elambda <- exp(lambda)
+            EX <- lambda*elambda / (elambda - 1)
+
+            ## convenience
+            ZEX <- Xdst*EX
+
+            ## iterate only once simultaneous eqs
+            ZEXY <- ZEX + Ydst
+            gammaHat <- sapply(s, function(j) ZEXY[,j] / (cHat[j]*J + I)) # col-wise `/`
+            cHat <- sumT(ZEX) / sumT(J*gammaHat)
+            
+            ## check convergence of EM
+            if ( converged(cHat, cHat_old) &&
+                converged(gammaHat, gammaHat_old) ) break
+            
+            ## if not converged, store updated estimates
+            cHat_old <- cHat
+            gammaHat_old <- gammaHat
+            ## print(sprintf('EM iteration %d found values: c = %f', em_iter, cHat))
+            em_iter <- em_iter+1
[…]

### 2014-09-11 R/estimate_Cst.R

@@ -1,0 +1,86 @@
+##' estimates parameters from hypothesis lambda != gamma; 2*S*T free parameters
+##' 
+##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param J vector of predators caught in each time period
+##' @param I vector of number of days all traps were left out in a given time period    
+##' @param EM boolean; whether or not EM algorithm is used
+##' @param em_maxiter integer specifying max number of EM iterations
+##' @param BALANCED boolean; whether or not data are BALANCED
+estCst <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
+
+    ## some numbers
+    S <- ncol(Ydst); s <- seq_len(S)
+    T <- nrow(Ydst); t <- seq_len(T)
+    ST <- S*T; st <- seq_len(ST)
+
+    if (EM) {
+        
+        ## estimate gamma
+        gammaHat <- Ydst/I
+
+        ## initialize some things
+        em_iter <- 1
+        lambdaHat <- lambdaHat_old <- gammaHat
+
+        ## iterate EM
+        while ( TRUE ) {
+            ## update lambda
+            elambda <- exp(lambdaHat)
+            EX <- lambdaHat*elambda/(elambda-1)
+            lambdaHat <- Xdst*EX/J
+
+            ## check convergence
+            if ( converged(lambdaHat, lambdaHat_old) ) break
+            lambdaHat_old <- lambdaHat
+            em_iter <- em_iter+1
+            
+            ## limit iterations
+            if ( em_iter > em_maxiter )
+                stop(sprintf('H1: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
+        }
+        
+        ## since estimated 0s don't follow from calculations above
+        lambdaHat[which(Xdst == 0, arr.ind=T)] <- 0
+
+        ## calc standard error with est params
+        ## SE <- seEM(lambdaHat, gammaHat, NULL, Xdst, Ydst, J, I)
+        Info <- diag(2*ST)             # initialize information matrix
+        expl <- exp(-lambdaHat)
+        
+        ## fill in second derivatives
[…]

### 2014-09-11 R/se.R

@@ -1,106 +1,0 @@
-##' calculate standard error of EM estimates
-##'
-##' @param lambda parameter estimates of lambda from EM
-##' @param gamma parameter estimates of gamma from EM
-##' @param c parameter estimates of c from EM
-##' @param Zdst matrix of sums of binary response, whether prey species s was eaten or not during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
-##' @param Ydst matrix of sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
-##' @param J vector of predators caught in each time period
-##' @param I vector of number of days all traps were left out in a given time period
-##' @details output is an array where the rownames are indexed by {(T,S)}, T enumerated first
-seEM <- function(lambda, gamma, c, Zdst, Ydst, J, I) {
-
-    ## some numbers
-    S <- ncol(gamma); s <- seq_len(S)
-    T <- nrow(gamma); t <- seq_len(T)
-    ST <- S*T
-    indices <- unlist(lapply(s,
-                             function(y) sapply(t,
-                                                function(x) paste(x, y, sep=''))))    
-    if ( is.null(c) ) {                 # H1
-        out <- matrix(0, nrow=ST, ncol=2)
-        colnames(out) <- c('lambda', 'gamma')
-        out[,'gamma'] <- unlist(Ydst/gamma^2)
-        elambda <- exp(lambda)
-        out[,'lambda'] <- unlist(J*elambda/(1-elambda)^2)
-        rownames(out) <- indices
-    } else {                            # H0
-        lc <- length(c)
-        slc <- seq_len(lc)
-        out <- rep(0, ST+lc)
-        
-        if ( lc == 1 || lc == T ) {
-            l <- c*gamma
-            el <- exp(l)
-            tmp <- J*el/(1-el)^2
-            g2 <- gamma^2
-
-            if ( lc > 1 ) {
-                out[slc] <- sumSp(tmp*g2)
-            } else {
-                out[slc] <- sumST(tmp*g2)
-            }
-            out[-slc] <- unlist(tmp*c^2 + Ydst/g2)
-            names(out) <- c(paste('c', slc, sep=''), paste('gamma', indices, sep=''))
-        } else if ( lc == S ) {
-            l <- sapply(s, function(j) gamma[,j]*c[j]) # col-wise `*`
-            el <- exp(l)
[…]

### 2014-09-11 tests/testthat/test_predPref.R

@@ -1,7 +1,7 @@
 context('Output predPref.')
 
 ## set up some data
-Predators <- 250
+Predators <- c(200, 210, 220, 230, 240)
 Traps <- Predators
 PreySpecies <- 2
 Times <- 5
@@ -20,7 +20,9 @@ test_that('noEM, 1 vs c, null', {
     expect_equal(dim(pref$null$gamma), TxS) # dimensions of estimates
     expect_equal(dim(pref$alt$gamma), TxS)
     expect_more_than(pref$p.value, 0.001)     # correct conclusion
-})
+    expect_that(all(diag(pref$null$var)>0), is_true()) # variances (null) are positive
+    expect_that(all(diag(pref$alt$var)>0), is_true()) # variances (alt) are positive
+})                                      
 
 test_that('noEM, c vs c^t, null', {
     l <- matrix(2*sqrt(2), nrow=Times, ncol=PreySpecies)
@@ -32,6 +34,8 @@ test_that('noEM, c vs c^t, null', {
     expect_equal(dim(pref$null$gamma), TxS) # dimensions of estimates
     expect_equal(dim(pref$alt$gamma), TxS)
     expect_more_than(pref$p.value, 0.001)     # correct conclusion
+    expect_that(all(diag(pref$null$var)>0), is_true()) # variances (null) are positive
+    expect_that(all(diag(pref$alt$var)>0), is_true()) # variances (alt) are positive
 })
 
 test_that('noEM, c vs general, null', {
@@ -43,7 +47,9 @@ test_that('noEM, c vs general, null', {
     expect_equal(dim(pref$null$gamma), TxS) # dimensions of estimates
     expect_equal(dim(pref$alt$lambda), TxS)
     expect_equal(dim(pref$alt$gamma), TxS)
-    expect_more_than(pref$p.value, 0.001)     # correct conclusion    
+    expect_more_than(pref$p.value, 0.001)     # correct conclusion
+    expect_that(all(diag(pref$null$var)>0), is_true()) # variances (null) are positive
+    expect_that(all(diag(pref$alt$var)>0), is_true()) # variances (alt) are positive
 })
 
 test_that('noEM, c vs c^t, alt', {
@@ -55,7 +61,9 @@ test_that('noEM, c vs c^t, alt', {
     expect_equal(length(pref$alt$c), Times)
     expect_equal(dim(pref$null$gamma), TxS) # dimensions of estimates
     expect_equal(dim(pref$alt$gamma), TxS)
[…]

### 2014-09-15 paper/methods.tex
now 2(ST) parameter estimators; lambda, gamma indepdent and l = g*c_st

@@
 We denote the binary response that the predator did in fact eat at least one prey species $s$ in time period $t$ by $Z_{jst} = 1(X_{jst} > 0)$.
-Now, the observed data $Z_{jst} \iid \text{Bern}(1-\exp\{-\lambda_{st}\})$.
-Using this we can find the complete data likelihood, by first noting that the conditional distribution of the unobserved data $X_{jst}$ given $Z_{jst}, Y_{ist}, \boldsymbol{\lambda}, \boldsymbol{\gamma}$ is a truncated Poisson distribution
+Now, the observed data are $Z_{jst} \iid \text{Bern}(1-\exp\{-\lambda_{st}\})$.
+Using this we can find the complete data likelihood by first noting that the conditional distribution of the unobserved data $X_{jst}$ given $Z_{jst}, Y_{ist}, \boldsymbol{\lambda}, \boldsymbol{\gamma}$ is a truncated Poisson distribution
 
@@
   f_{X|Y,Z,\boldsymbol{\lambda},\boldsymbol{\gamma}}(x_{jst}) =
-  \frac{\exp{\{-\lambda_{st}\}} \lambda_{st}^{x_{jst}}}{(1 - \exp{\{-\lambda_{st}\}}) x_{jst}!}1(x_{jst} \geq 0) \quad \text{ where } \quad \E_{[X|Y,Z]}X_{jst} = \frac{\lambda_{st} \exp{\{\lambda_{st} \}}}{\exp{\{ \lambda_{st} \}} - 1}.
+  \frac{\exp{\{-\lambda_{st}\}} \lambda_{st}^{x_{jst}}}{(1 - \exp{\{-\lambda_{st}\}}) x_{jst}!}1(x_{jst} > 0) \quad \text{ where } \quad \E_{[X|Y,Z]}X_{jst} = \frac{\lambda_{st} \exp{\{\lambda_{st} \}}}{\exp{\{ \lambda_{st} \}} - 1}.
 \end{equation*}
@@
   & = \sum_{s,t,j} \left( - \lambda_{st} 
-    + z_{jst} \log(\lambda_{st}) \E X_{jst} \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right) + \text{const} \\
-  & = \sum_{s,t} \left( -J_t \lambda_{st} + z_{\cdot st} \log(\lambda_{st}) \E X_{jst} \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right) + \text{const}.
+    + z_{jst} \E X_{jst} \log{\lambda_{st}}  \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right) + \text{const} \\
+  & = \sum_{s,t} \left( -J_t \lambda_{st} + z_{\cdot st} \E X_{jst} \log{\lambda_{st}} \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right) + \text{const}.
 \end{align*}
@@
 It is well known that the asymptotic distribution of $\Lambda$ is a $\chi_{\rho}^2$ distribution with $\rho$ degrees of freedom.
-Under the EM algorithm we use $L_{observed}(Z,Y)$ as the likelihood in the calculation of $\Lambda$.
+Under the EM algorithm we use $L_{obs}(Z,Y)$ as the likelihood in the calculation of $\Lambda$.
 

### 2014-09-15 R/contrasts.R

@@ -12,7 +12,8 @@ testC <- function(x, b, mu = 0, alternative = c("two.sided", "less", "greater"),
     if (!missing(mu) && (length(mu) != 1 || is.na(mu))) 
         stop("'mu' must be a single number")
     if (!missing(conf.level) && (length(conf.level) != 1 || !is.finite(conf.level) || 
-                                 conf.level < 0 || conf.level > 1)) 
+                                 conf.level < 0 || conf.level > 1))
+        stop("'conf.level' must be a single number between 0 and 1")
 
     ## get appropriate estimates
     alpha <- 1-conf.level

### 2014-09-15 R/estimate_C.R

@@ -28,8 +28,8 @@ estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
             
             ## expected value of Xjst
             lambda <- cHat*gammaHat
-            elambda <- exp(lambda)
-            EX <- lambda*elambda / (elambda - 1)
+            elambda <- exp(-lambda)
+            EX <- lambda / (1-elambda)
 
             ## convenience
             ZEX <- Xdst*EX
@@ -50,7 +50,7 @@ estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
             
             ## limit iterations
             if ( em_iter > em_maxiter )
-                stop(sprintf('H0: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
+                stop(sprintf('estC: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
         }
         
         ## calc standard error with est params
@@ -68,7 +68,7 @@ estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
         Info[lowmat] <- t(Info)[lowmat]      # make symmetric from upper tri
 
         ## calc log-lik with est params
-        loglik <- llEM(Xdst, Ydst, NA, gammaHat, J, I, cHat)
+        loglik <- llEM(Xdst, Ydst, NA, gammaHat, J, I, as.numeric(cHat))
         
         list('c' = cHat, 'gamma' = as.matrix(gammaHat), 'em_iters' = em_iter,
              'll' = loglik, 'var' = solve(Info))

### 2014-09-15 R/estimate_Cs.R

@@ -26,8 +26,8 @@ estCs <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
             
             ## expected value of Xjst
             lambda <- sapply(s, function(j) cHat[j]*gammaHat[,j]) # col-wise `*`
-            elambda <- exp(lambda)
-            EX <- lambda*elambda / (elambda - 1)
+            elambda <- exp(-lambda)
+            EX <- lambda/(1-elambda)
 
             ## convenience
             ZEX <- Xdst*EX
@@ -49,12 +49,12 @@ estCs <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
             
             ## limit iterations
             if ( em_iter > em_maxiter )
-                stop(sprintf('H0: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
+                stop(sprintf('estCs: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
         }
 
         ## calc standard error with est params
         ## SE <- seEM(NULL, gammaHat, cHat, Xdst, Ydst, J, I)
-        Info <- diag(ST+S)                # initialize information matrix
+        Info <- diag(ST+S)             # initialize information matrix
         g2 <- gammaHat^2               # gamma^2
         l <- sapply(s, function(j) gammaHat[,j]*cHat[j]) # col-wise `*`
         expl <- exp(l); tmp <- J*expl/(expl-1)^2  # a common term

### 2014-09-15 R/estimate_Ct.R

@@ -20,15 +20,14 @@ estCt <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
         cHat <- cHat_old <- runif(length(J))
         init <- est1(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED)
         gammaHat <- gammaHat_old <- init$gamma 
-        lambda <- elambda <- init$lambda
 
         ## iterate EM
         while ( TRUE ) {
             
             ## expected value of Xjst
             lambda <- cHat*gammaHat
-            elambda <- exp(lambda)
-            EX <- lambda*elambda / (elambda - 1)
+            elambda <- exp(-lambda)
+            EX <- lambda/(1-elambda)
 
             ## convenience
             ZEX <- Xdst*EX
@@ -49,13 +48,13 @@ estCt <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
             
             ## limit iterations
             if ( em_iter > em_maxiter )
-                stop(sprintf('H0: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
+                stop(sprintf('estCt: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
         }
 
         ## calc standard error with est params
         ## SE <- seEM(NULL, gammaHat, cHat, Xdst, Ydst, J, I)
-        Info <- diag(ST+T)                 # initialize information matrix
-        g2 <- gammaHat^2                # gamma^2
+        Info <- diag(ST+T)             # initialize information matrix
+        g2 <- gammaHat^2               # gamma^2
         l <- cHat*gammaHat
         expl <- exp(l); tmp <- J*expl/(expl-1)^2 # a common term
 
@@ -104,7 +103,8 @@ estCt <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
             gammaHat_old <- gammaHat
             cHat_old <- cHat
             iter <- iter+1
-            
+
+            ## limit iterations
             if ( iter > maxiter )
                 stop(sprintf('estCt: %d not sufficient iterations for simultaneous equations.', maxiter))
         }
@@ -115,7 +115,7 @@ estCt <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
 
         ## fill Info with second derivatives
         diag(Info)[-t] <- unlist(XYdst/gammaHat^2) # gamma
[…]

### 2014-09-15 R/hypotheses.R

@@ -14,7 +14,7 @@ calcHypotheses <- function(hyp, Xdst, Ydst, J, I, balanced, EM, em_maxiter) {
     fns <- list()
     fns[['1']] <- est1; fns[['c']] <- estC
     fns[['Cs']] <- estCs; fns[['Ct']] <- estCt
-    fns[['Cst']] <- estCst
+    fns[['Cst']] <- estCst; fns[['gen']] <- estGen
 
     ## calculate null and alternative hypotheses
     null <- fns[[hyp[1]]](Xdst, Ydst, J, I, EM, em_maxiter, balanced)
@@ -50,11 +50,13 @@ checkHypotheses <- function(hyp) {
     } else if ( grepl('1', hyp[1]) ) {
         H[1] <- '1'
     } else {
-        stop('Null hypothis specified incorrectly; please check documentation.')
+        stop('Null hypothesis specified incorrectly; please check documentation.')
     }
 
     ## H1
-    if ( grepl('gen', hyp[2]) || grepl('st', hyp[2]) ) {
+    if ( grepl('g', hyp[2]) ) {
+        H[2] <- 'gen'
+    } else if ( grepl('st', hyp[2]) ) {
         H[2] <- 'Cst'
     } else if ( grepl('t', hyp[2]) ) {
         H[2] <- 'Ct'
@@ -63,7 +65,7 @@ checkHypotheses <- function(hyp) {
     } else if ( grepl('c', hyp[2]) ) {
         H[2] <- 'c'
     } else {
-        stop('Alt hypothis specified incorrectly; please check documentation.')
+        stop('Alt hypothesis specified incorrectly; please check documentation.')
     }
 
     ## both

### 2014-09-15 R/predpref.R

@@ -6,7 +6,7 @@
 ##' @param alpha LRT level of significance
 ##' @param em_maxiter maximum number of iterations allowed for EM algorithm
 ##' @export
-predPref <- function(eaten, caught, hypotheses = c('c', 'Ct'), alpha=0.05, em_maxiter=100) {
+predPref <- function(eaten, caught, hypotheses = c('c', 'Ct'), alpha=0.05, em_maxiter=1000) {
 
     ## check hypotheses specification
     hypotheses <- checkHypotheses(hypotheses)

### 2014-09-15 R/summary.predPref.R

@@ -25,7 +25,7 @@ summary.predPref <- function(object, ..., sig.level=0.05) {
 
 print.summary.predPref <- function(x) {
     cat("\nPredator Preferences Model:\n\n")
-    cat("Log-likelihoods:\n\tH0 (", x$hypotheses[1], "):", x$loglikH0, "\n\tH1 (", x$hypotheses[2], "):", x$loglikH1, "\n")
+    cat("Hypotheses:\n\tH0 =", x$hypotheses[1], "\n\tH1 =", x$hypotheses[2], "\n")
     cat("Likelihood Ratio Test:\n\t-2*(llH0-llH1) =", x$Lambda, "on", x$df, "degrees of freedom\n")
     cat("\tp-value =", x$p.value, "\n")
     cat("Parameter Estimates:\n")

### 2014-10-02 paper/discussion.tex
add (start to) intro and discussion to paper

\section{Discussion}
\label{sec:discussion}

{\color{red} should we include mention of the spiders dataset?}

The model developed here advances the statistics literature available to determine predators' preferences by testing simultaneously across an array of multiple prey species and time.  This is acheived via a simple, but statistically powerful, likelihood ratio test.  Further testing of the ratio of rates for which predators eat to encounter prey species allows researchers to make specific conclusions about predators' preferences.  For instance, rates acros time can be estimated to make statements about seasonal effects on a predator's eating habits, or relative rates across species groups allows for statements about the preferences for species $s_1$ to that of $s_2$.

When counts of predators' gut contents are not fully observed, and instead only a binary response indicating the existence of the prey species in the gut is observed, we are able to treat the counts as missing data.  By modelling all of the observed data, both the binary responses and the number of prey species caught, and the missing count data, we are able to use the EM algorithm to extract as much information from the data as possible.  Though this is nice in theory, in practice the success of this tweak to our original model is limited by the magnitude of the unknown parameters $\lambda_{st}$.

Further developments of our model could be beneficial.  Taking into account other environmental variables that might effect a predators' eating habits, such as rain or temperature, say, might be advantageous.

An \texttt{R} package, named \texttt{spiders}, is available on CRAN at \url{http://cran.r-project.org/web/packages/spiders/index.html} and fits all the methods discussed above.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2014-10-02 paper/introduction.tex
add (start to) intro and discussion to paper

@@
-\section*{Introduction}
+\section{Introduction}
 \label{sec:intro}
+
+Modeling a predator's food selection was given much attention for a short period of time; see \citet{Strauss:1979} and the references within.
+Many, if not all, of the indices developed, though intuitive, focused on a snapshot in time and only on one prey species, and they rarely had practicle asymptotic properties.
+This has left interested practicioners to the most computationally manageable of the techniques, and has completely ignored simultaneous testing across an array of both species and time.
+
+The model presented here, maintains tractable asymptotic properties while being general enough to take into account an array of speices and time points.
+By modeling both time and any number of prey species, we are able to see a more detailed anlyais of the predator's eating preferences.
+Further, the simplicity of the model allows us to consider predators for which exact tallies of the number of each variety of prey species eaten within any given time period is not observed.
+Instead, we rely on the researcher being able to DNA sequence the contents of the predator's gut, and make a simple binary conclusion: yes this predator ate some of that prey species during this time period, or no they did not.
+Under certain situations, our models is able to accurately estimate parameters of interest despite having lost much information.
+
+To showcase our model, we performed simulations for both scenarios when full count data from the predators' guts are observed, and when instead only binary repsonses, indicating specific prey species were eaten, are observed.
+Our \texttt{R} \citep{R:2014} package, named \texttt{spiders}, is available on CRAN so that the broader community of researchers can similarly apply our methods.
+
 %%% Local Variables: 

### 2014-10-02 paper/main.tex
more paper: methods and structure
add (start to) intro and discussion to paper

@@
 
+\title{Modeling Predator Preferences}
+\author{Edward A. Roualdes and Simon Bonner \\ University of Kentucky}
+\date{\today}
+\maketitle
+
+% questions for simon:
+% 1) counts are proportional to?
+% 2) re: solve 1 - exp(-l) = 1 for l.  What is the limiting variable?
+
+\begin{abstract}
+The modeling of a predator's prey selection has a breif history that includes many intuitive indices, few of which have both reasonable statistical justification and tractable asymptotic properties.
+Here, we provide a simple model that meets both of these criteria, while extending previous work to include an array of multiple species and time points.
+Further, we apply the Expectation-Maximization algorithm to cases where exact counts of the number of prey species eaten in a particular timer period is not observed.
+Simulations yada yada yada...
+\end{abstract}
+
+\input{introduction.tex}
 \input{methods.tex}
 \input{simulations.tex}
+\input{discussion.tex}
 
-%\bibliographystyle{amsalpha}
-%\bibliography{}
+\bibliographystyle{plainnat}
+\bibliography{refs}
 \end{document}

### 2014-10-02 paper/methods.tex
add (start to) intro and discussion to paper

@@
-\section*{Methods}
+\section{Methods}
 \label{sec:methods}
 
-\subsection*{Data}
+\subsection{Data}
 
@@
 We assume these counts represent the number of prey species $s$ the predator will encounter on average during time period $t \in \{1, \ldots, T\}$.
-Hence, the counts of prey species $s$ caught during time period $t$ are hypothesized to be independent draws from a Poisson distribution with rate parameter $\gamma_{st}$.
+The counts of prey species $s$ caught during time period $t$ are hypothesized to be independent draws from a Poisson distribution with rate parameter $\gamma_{st}$.
 We further assume that the number of prey species found in the gut of the trapped predators, also follows a Poisson distribution with rate $\lambda_{st}$.
 Here, the parameter $\lambda_{st}$ represents the rate at which the predator ate the encountered prey species $s$ during time period $t$.
-We then make statistical claims about the relatives rates that the $S$ prey species are eaten to the rate at which they are encountered over $T$ time periods.
+By modeling $\lambda_{st}$ and $\gamma_{st}$ we are able to test claims about a predator's eating preferences.
 
-Our data collection scheme assumes the following: $1)$ traps independently catch prey that are encountered at a constant rate, $2)$ all traps are equally likely to catch the prey species of interest, $3)$ predators eat, at a constant rate, prey species independently, $4)$ predators eat indepedent of each other.
-Though there exists a non-zero chance that these assumptions are broken, we feel that slight deviations from them will not overtly effect the estimates.
+The use of Poisson distributions make the following implicit assumptions: $1)$ traps independently catch the prey species of interest, $2)$ predators eat, at a constant rate, prey species independently, $3)$ predators eat indepedent of each other.
+Our model's ability to accurately portray predators' habitats and underlying eating preferences dependes on the degree to which these assumptions are broken.
 
@@
 Let $Y_{ist} \iid \mathcal{P}(\gamma_{st})$ denote the number of prey species $s$ found in trap $i$ during occurrence $t$, $i \in \{1, \ldots, I_t\}$.
-We make formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda}$ and $\boldsymbol{\gamma}$.
+Formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda}$ and $\boldsymbol{\gamma}$ offer insights to the relative rates at which predators eat particular prey species.
 
-\begin{figure}
-  \centering
-  \begin{tabular}{rrcc}
-    & & \multicolumn{2}{c}{Time} \\
-    & & constant & varies \\
-    \cline{3-4}
-    \multirow{2}{*}{Species} & \multicolumn{1}{r|}{constant} & \multicolumn{1}{c}{$c$} & \multicolumn{1}{|c|}{$c_t$} \\ 
-    \cline{3-4}
-    & \multicolumn{1}{r|}{varies} & \multicolumn{1}{c}{$c_s$} & \multicolumn{1}{|c|}{$c_{st}$} \\
-    \cline{3-4}
-  \end{tabular}
-  \caption{The four hypotheses considered are shown by their symbolic representations, highlighting which indices are allowed to vary.  This is essentially the range of the mapping $\xi$.  }
-  \label{tab:hyp}
-\end{figure}
+% \begin{figure}
+%   \centering
+%   \begin{tabular}{rrcc}
+%     & & \multicolumn{2}{c}{Time} \\
+%     & & constant & varies \\
+%     \cline{3-4}
+%     \multirow{2}{*}{Species} & \multicolumn{1}{r|}{constant} & \multicolumn{1}{c}{$c$} & \multicolumn{1}{|c|}{$c_t$} \\ 
+%     \cline{3-4}
+%     & \multicolumn{1}{r|}{varies} & \multicolumn{1}{c}{$c_s$} & \multicolumn{1}{|c|}{$c_{st}$} \\
+%     \cline{3-4}
+%   \end{tabular}
+%   \caption{The four hypotheses considered are shown by their symbolic representations, highlighting which indices are allowed to vary.  This is essentially the range of the mapping $\xi$.  }
+%   \label{tab:hyp}
+% \end{figure}
 
-We consider five hypotheses to provide variations on the relative magnitude of $\lambda_{st}/\gamma_{st} = \xi(c_{st})$, for a mapping $\xi: c_{st} \mapsto \{1, c, c_s, c_t, c_{st}\}$.
-When $\xi$ maps to a constant with respect to one of the indexing variables, $s$ or $t$, intuitive statements are realized about the predator's preferences.
-Table~\ref{tab:hyp} depicts the ways in which the proportionality constants can vary, or not, for each of the four hypotheses.
-Figure~\ref{fig:hier} represents the order in which hypotheses can be tested against each other; how best phrase this, does the young Skywalker?
+We consider five variations on the relative magnitude of $\lambda_{st}/\gamma_{st} = c_{st}$.
+These five hypotheses each allow $c_{st}$ to vary by time, prey species, both, or neither.
+Because the five hypotheses are nested, a natural testing order is suggested in Figure~\ref{fig:hier}.
+
+\begin{enumerate}
+\item $c_{st} = 1, s = 1, \ldots, S; t = 1, \ldots, T$
+\item $c_{st} = c, s = 1, \ldots, S; t = 1, \ldots, T$
+\item $c_{st} = c_s, s = 1, \ldots, S$
+\item $c_{st} = c_t, t = 1, \ldots, T$
+\item $c_{st} = c_{st}, s = 1, \ldots, S; t = 1, \ldots, T$
+\end{enumerate}
 
@@
 
-\subsection*{Fully Observed Data}
+\subsection{Fully Observed Data}
 
 The likelihood function that allows for estimation of these parameters is as follows.
-Since we assume $X_{jst} \perp Y_{ist}$ we can simply multiply the respective Poisson probability density functions together, and then form products over all $s,t$ to get the likelihood
+Since we assume $X_{jst}$ is independent of $Y_{ist}$ we can simply multiply the respective Poisson probability density functions together, and then form products over all $s,t$ to get the likelihood.
 
@@
 
-\noindent In some cases analytic solutions for the maximum likelihood estimates are available, namely when the data are balanced $J_t = J$, $I_t = I$, and $\xi(c_{st}) = c$ or when estimating $\lambda_{st}, \gamma_{st}$ under the hypothesis represented by $c_{st}$.
-In all other cases, analytic solutions are not readily available.
-In such cases, we rely on the fact that the log-likelihood $l(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log{L}$ is concave and iteratively solve
+\noindent Writing all $5$ hypotheses as $\lambda_{st} = c_{st}\gamma_{st}$, we can, in some cases find analytic solutions for the maximum likelihood estimates of $c_{st}$ and $\gamma_{st}$.
+When the data are balanced $J_t = J$, $I_t = I$, and $c_{st} = c$ analytic solutions exist -- {\color{red}what about $c_{st} = 1$}.
+In all other cases, analytic solutions are not readily available and instead we rely on the fact that the log-likelihood $l(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log{L}$ is concave.
+We maximize the log-likelihood by iteratively solving partial derivatives of $l$, with respect to $c_{st}$ and $\gamma_{st}$, set equal to zero
 
 \begin{equation*}
-  \hat{c} = \frac{\sum_{s,t} X_{\cdot st}}{\sum_t J_t \sum_s \gamma_{st}}, \quad \hat{c}_t =  \frac{\sum_s X_{\cdot st}}{J_t \sum_s \gamma_{st}}, \text{ or} \quad \hat{c}_s = \frac{\sum_{t}X_{\cdot st}}{\sum_t J_t \gamma_{st}}, \text{ and } \quad \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_t\xi(c_{st}) + I_t}.
+  \hat{c} = \frac{\sum_{s,t} X_{\cdot st}}{\sum_t J_t \sum_s \gamma_{st}}, \quad \hat{c}_t =  \frac{\sum_s X_{\cdot st}}{J_t \sum_s \gamma_{st}}, \text{ or} \quad \hat{c}_s = \frac{\sum_{t}X_{\cdot st}}{\sum_t J_t \gamma_{st}}, \text{ and } \quad \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_t c_{st} + I_t}.
 \end{equation*}
 
-\noindent where $X_{\cdot st} := \sum_{j}X_{jst}$ and $Y_{\cdot st} := \sum_{i} Y_{ist}$.  For the case $c_{st}$, the $2(S \cdot T)$ parameters are readily found via
+\noindent where $X_{\cdot st} = \sum_{j=1}^{J_t}X_{jst}$ and $Y_{\cdot st} = \sum_{i=1}^{I_t} Y_{ist}$.
 
-\begin{equation*}
-  \hat{\lambda}_{st} = \frac{X_{\cdot st}}{J_t} \quad \text{ and } \quad \hat{\gamma}_{st} = \frac{Y_{\cdot st}}{I_t}.
-\end{equation*}
+\subsection{Unobserved Counts}
 
-\subsection*{Unobserved Counts}
+Working with biologists who study spider eating preferences, we have found that not all predators' allow for easily counted prey species in their guts.
+As an alternative strategy, we can rely on the DNA sequencing of a sample from the predators' guts.
+If such sequencing returns a positive response, say a $1$ if a particular predator ate prey species $s$ and $0$ otherwise, we can, albeit with some information lost, model predators' eating preferences with the above framework using the EM algorithm.
 
-Probably need some words about why we might not observed the count data.
-
-When the data $X_{jst}$ are observed as a binary response, instead of count data representing the number of prey species $s$ that the predator ate in time period $t$, we can still estimates the parameters of interest $\boldsymbol{\lambda}, \boldsymbol{\gamma}$.
+When the data $X_{jst}$ are not observed, and instead a boolean response indicating if a predator ate any number of prey species $s$ during time $t$ is observed, we can still, to some degree, estimates the parameters of interest $\boldsymbol{\lambda}, \boldsymbol{\gamma}$.
 Because some information is observed, we can treat the counts as missing and use the EM algorithm to find the maximum likelihood estimates of the observed data likelihood.
@@
 We denote the binary response that the predator did in fact eat at least one prey species $s$ in time period $t$ by $Z_{jst} = 1(X_{jst} > 0)$.
-Now, the observed data are $Z_{jst} \iid \text{Bern}(1-\exp\{-\lambda_{st}\})$.
+Now, the observed data are independent and identically distributed Bernoulli observations with parameter $p_{st} = 1-\exp\{-\lambda_{st}\}$.
 Using this we can find the complete data likelihood by first noting that the conditional distribution of the unobserved data $X_{jst}$ given $Z_{jst}, Y_{ist}, \boldsymbol{\lambda}, \boldsymbol{\gamma}$ is a truncated Poisson distribution
@@
 
-The EM algorithm requires iteratively solving $(\boldsymbol{\lambda}^{(k+1)},\boldsymbol{\gamma}^{(k+1)}) = \argmax_{\boldsymbol{\lambda},\boldsymbol{\gamma}} \E l_{comp}$.
+The EM algorithm requires iteratively solving $(\boldsymbol{\lambda}^{(k+1)},\boldsymbol{\gamma}^{(k+1)}) = \argmax_{\boldsymbol{\lambda},\boldsymbol{\gamma}} \E l_{comp}^k$.
 Though no analytic solution to this maximization exists, one idea is to iteratively solve partial derivatives of $\E l_{comp}$ set equal to zero until convergence.
-In fact, as we only need find parameter values that increase the observed likelihood, we forgo fully iterating to find the maximized values of $(\boldsymbol{\lambda}, \boldsymbol{\gamma})$ and instead perform just one step uphill within each EM iteration.
+In fact, as we only need find parameter values that increase the observed likelihood, we forgo fully iterating to find the maximum values of the parameters and instead perform just one step uphill within each EM iteration.
 This strategy is significantly less computationally intensive, thus generating a much faster generalized EM (GEM) algorithm.
 
-The updating steps of the EM algorithm simply replace any occurrence of $X_{\cdot st}$ with $z_{\cdot st}\E X_{jst}$.
-For instance, when maximizing under the hypothesis $c_{st}$ we get $\hat{\lambda}_{st} = z_{\cdot st} \E l_{comp} / J_t$.
-In the case of the hypothesis $c_t$, where $\lambda_{st} = c_t \gamma_{st}, \, \forall t$, we iterate the following functions until convergence,
-
-\begin{equation*}
-  \hat{\gamma}_{st} = \frac{z_{\cdot st} \E X_{jst} + Y_{\cdot st}}{cJ_t + I_t} \quad \text{ and } \quad \hat{c} = \frac{ \sum_{t} z_{\cdot st} \E X_{jst}}{ J_t \sum_{s} \gamma_{st}}
-\end{equation*}
-
-\noindent updating the value of $\lambda_{st}$ before each next step.
-
-This GEM algorithm works well when values $\lambda_{st}$ are relatively small and zeros are common in the data $Z_{jst}$.
+This GEM algorithm accurately estimates the parameters when values of $\lambda_{st}$ are relatively small, such that zeros are prevalent in the data $Z_{jst}$.
 In this case, not too much information is lost since estimation of $\E Z_{jst}$ can be estimated well by the proportion of observed zeros.
@@
 
-\subsection*{Testing}
+\subsection{Testing}
 
@@
 The degrees of freedom $\rho$ are set equal to the number of free parameters available in the stated hypotheses under question.
-If we put the null hypothesis to be $H_0: \lambda_t = c_t \gamma_t, \forall t$ and contrast this against $H_1: \lambda_{st} = c_{st}\gamma_{st}$ then there are $\rho = 2(S \cdot T) - S \cdot T - T = S \cdot T - T$ degrees of freedom.
-The alternative hypothesis previously stated is not in fact fit by estimating $c_{st}$, but is instead fit as if $c_{st}\gamma_{st}$ were $S \cdot T$ unique parameters independent of $\lambda_{st}$ for all $s,t$, thus providing $2(S \cdot T)$ total parameters for the hypothesis we denote by $c_{st}$.
+If we put the null hypothesis to be $H_0: \lambda_t = c_t \gamma_t$, for all $t$ and contrast this against $H_1: \lambda_{st} = c_{st}\gamma_{st}$ then there are $\rho = 2(S \cdot T) - S \cdot T - T = S \cdot T - T$ degrees of freedom.
 
 A set of hypotheses is determined by the p-value of the $\chi^2_{\rho}$ distribution.
-We choose the common value of $\alpha = 0.05$ as our level of significance.
-Hence, if $\mathbb{P}(\chi^2_{\rho} > \Lambda) < 0.05$ the null hypothesis is rejected in favor of the alternative hypothesis.
+Hence, with a level of significance, $\alpha$, the null hypothesis is rejected in favor of the alternative hypothesis if $\mathbb{P}(\chi^2_{\rho} > \Lambda) < \alpha$.
+
+\subsection{Testing $c_{st}$}
+
+After determining which model best fits the data, more detail can be extracted through a hypothesis test on the elements of $c_{st}$, or in vector notation as $\mathbf{c} \in \mathbb{R}^{S\cdot T}$.
+Let the elements of $\hat{\mathbf{c}}$ be the estimated values, $\hat{c}_{st}$, as estimated via the above maximum likelihood framework.
+Since $\hat{\mathbf{c}}$ is asymptotically normally distributed, any linear combination of the elements is also asymptotically normally distributed.
+For instance, let $a$ be a vector of the same dimension of $\hat{\mathbf{c}}$.
+Then $a^t\hat{\mathbf{c}}$ is asymptotically distrbuted as $\mathcal{N}(a^t\mathbf{c}, a^t\Sigma a)$, where $\Sigma$ is the covariance matrix of the asymptotic distrbution of $\hat{\mathbf{c}}$.
+
+Suppose, for example, that the hypothesis $c_s$ is determined to best fit the data with $s$ ranging $s = 1, 2$.
+We can test to see whether or not the two species $s_1$ and $s_2$ are statistically equally preferred under the null hypothesis $\hat{c}_{s_1} = \hat{c}_{s_2}$.
+This hypothesis is alternatively written in vector notation as $a^t\hat{\mathbf{c}} = 0$, where $a = (1, -1)^t$.
+Tests of the following form $H_0: a^t\mathbf{c} = \mu$ against any alternative of interest are then standard $Z$-tests.
+Similarly, confidence intervals of any level can be obtained if desired.
 

### 2014-10-02 paper/simulations.tex
add (start to) intro and discussion to paper

--- 2014-08-27 paper/simulations.tex
@@
-\section*{Simulations}
+\section{Simulations}
+\label{sec:sim}
 
 Our simulations hypothesize two prey species, and five time points.
-Of the hierarchy of hypotheses, we consider the four most interesting cases $\{c, c_s, c_t, c_st\}$.
-Sample sizes are chosen randomly from four overlapping levels.
-We put ``small'' sample sizes to be randomly sampled numbers in $[20,50]$, ``medium'' encompasses $[30:75]$, ``large'' $[50,150]$, and ``huge'' $[100,200]$.
-Hence, we randomly sample trap and predator observations for each occurence from one of the sample size levels, then cycle through all hypotheses.
+Of the hierarchy of hypotheses, we simulate data under three null hypotheses: $c, c_s, c_t$.
+Sample sizes are randomly chosen from four overlapping levels.
+Let ``small'' sample sizes be randomly sampled numbers in $[20,50]$, ``medium'' encompass $[30,75]$, ``large'' $[50,150]$, and ``huge'' $[100,200]$.
+Hence, we randomly sample prey and predator gut count observations for each time period from one of the sample size levels, then cycle through all hypotheses.
 This is repeated for each level of sample size.
-We simulate data for each of the $16$ scenarios above under both the standard maximum likelihood theory and under the EM algorithm.
-Only the details of the EM algorithm are presented here, as the other simulations are standard theory of point estimation.
+We simulate $500$ replicate datasets for each of the twelve scenarios above for both types of data, fully observed count data, $X_{jst}$, and for non-count data, when we observe only a binary response, $Z_{jst} = 1(X_{jst}>0)$.
+A subset of the examples are shown here; the interested reader is referred to the supplementary materials for the complete set of simulation results.
+
+For all simulated data, the true parameter values for the rate at which prey species are encountered in the wild are fixed to be $\gamma_{st} = \pi, \, \forall s,t$.
+The values of $\lambda_{st}$ are set with respect to each null hypothesis.
+For the hypothesis $c_{st} = c$, where predator preferences don't vary by either time or species, we consider $\lambda_{st} = 2\pi, \forall s,t$.
+Under the hypothesis $c_s$, the ratio of rates vary by species only, so we put $\lambda_{1t} = \sqrt{2}$ and $\lambda_{2t} = \pi$.
+Hence, $c_1 = \sqrt{2}/\pi \approx 0.45$ and $c_2 = 1$.
+For the second hypothesis, the ratio of rates vary by time $t$.
+Here, we put $\lambda_{st} = t$ for $t \in \{1, \ldots, 5 \}$.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.5]{nonem}
+  \caption{moar words}
+  \label{fig:nonem}
+\end{figure}
+
+Figure~\ref{fig:nonem} shows our model fitting to the fully observed count data.
+In the first display, the parameters $c_1 \approx 0.4501$ and $c_2 = 1$ are estimated as $\hat{c}_1 = 0.4508$ and $\hat{c}_2 = 1.007$, with standard errors of $\text{se}(\hat{c}_1) = 0.037$ and $\text{se}(\hat{c}_2) = 0.064$.
+The second display estimates the parameter $c = 2$ to be $\hat{c} = 2.002$ with standard error $\text{se}(\hat{c}) = 0.065$.
+
+As noted above we find that the EM algorithm accurately estimates the parameters when values of $\lambda_{st} = c_{st}\gamma_{st}$ are small.
+Figure~\ref{fig:em} contains density plots of the EM algorithm's estimates for the hypotheses $c_s,c_t$ for small and huge sample sizes, respectively.
+When data are simulated under the null hypothesis $c_s$, we find, even for the small sample size that point estimates are reasonably accurate.
+Though, when the values of $\lambda_{st}$ are of sufficient size to make zeros in the simulated data less common, the algorithm occasionally over-estimates the true values.
+This effect is easily seen in Figure~\ref{fig:em} for larger values of $c_t$ despite the increased sample size, but it also seen, less dramatically, in the density plot for the hypothesis $c_s$.
+The plots of $\gamma_{st}$ under the EM algorithm are not given as we do not consider missing data in the estimation of these parameters.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.5]{em}
+  \caption{Density plots of the EM algorithm's estimates of $c_s$ and $c_t$ with small and huge sample sizes, respectively.
+The estimates near $4$ for both of the plots represent samples for which few zeros appeared in the data and the EM algorithm had a particularly difficult time estimating the true parameter value.}
+  \label{fig:em}
+\end{figure}
+
+Because of the tendency for the EM algorithm to over-estimate some parameter values, bias for non-count data is significantly worse than it is for the count data.
+Table~\ref{tab:bias} displays the max absolute bias across all indices $s,t$ of $\lambda_{st}$ for each simulation scenario.
+In most cases, these numbers decrease, as expected, as the sample size increases.
+
+\begin{table}
+  \centering
+  \begin{tabular}{ll|rrrr}
+    & & small & medium & large & huge \\ 
+    \hline
+    count & $c_t$ & 0.040 & 0.013 & 0.020 & 0.011 \\ 
+    & $c_s$ & 0.028 & 0.015 & 0.010 & 0.005 \\ 
+    & $c$ & 0.030 & 0.016 & 0.017 & 0.012 \\ 
+    non-count & $c_t$ & 4.151 & 3.155 & 0.941 & 1.423 \\ 
+    & $c_s$ & 0.138 & 0.096 & 0.044 & 0.032 \\ 
+    & $c$ & 2.924 & 2.078 & 0.669 & 0.547 \\ 
+  \end{tabular}
+  \caption{Max absolute bias of all $\lambda_{st}$ for each simulation scenario.}
+  \label{tab:bias}
+\end{table}
 

### 2014-10-02 R/contrasts.R

@@ -1,10 +1,57 @@
-##' estimates linear contrasts of c, c_s, c_t, c_st
+##' @title linear contrast of c_st
+##'
+##' @description estimates linear contrasts of the elements of c, c_s, c_t, c_st
+##'
+##' @return A list with class '"htest"' containing the following components:
+##' 
+##' statistic: the value of the t-statistic.
+##' 
+##' parameter: the degrees of freedom for the t-statistic.
+##' 
+##' p.value: the p-value for the test.
+##' 
+##' conf.int: a confidence interval for the mean appropriate to the specified
+##' alternative hypothesis.
+##' 
+##' estimate: the estimated mean or difference in means depending on whether
+##' it was a one-sample test or a two-sample test.
+##' 
+##' null.value: the specified hypothesized value of the mean or mean difference
+##' depending on whether it was a one-sample test or a two-sample test.
+##' 
+##' alternative: a character string describing the alternative hypothesis.
+##' 
+##' method: a character string indicating what type of t-test was performed.
+##' 
+##' data.name: a character string giving the names of the data.
+##' 
+##' @details The input vector b performs the linear transformation
+##' t(b) \%*\% matrix(c_st), so that c_st becomes a column vector by indexing
+##' t first and then s.  Hence there is no requirement of a linear
+##' contrast, a linear transformation such that
+##' t(b) \%*\% matrix(1, nrow=length(b)) != 0 is allowed.
 ##'
 ##' @param x a prefPref object as fit by the eponymous function
 ##' @param b a vector to linearly transform c_st
 ##' @param mu a number to test the linear contrast against in the null
 ##' @param alternative string to specify alternative hypothesis
 ##' @param conf.level confidence level of the interval
+##' 
+##' @examples
+##' # set parameters
+##' Predators <- Traps <- 100
+##' PreySpecies <- 2
+##' Times <- 5
+##' g <- matrix(sqrt(2), nrow=Times, ncol=PreySpecies)     # gamma
+##' l <- matrix(seq(0.4,1.8,length.out=5)*sqrt(2), nrow=Times, ncol=PreySpecies) # ct
+##'
+##' # fit model
+##' \dontrun{
+##' fdata <- simPref(PreySpecies, Times, Predators, Traps, l, g, EM=FALSE)
[…]

### 2014-10-02 R/estimate_1.R

@@ -1,5 +1,9 @@
-##' estimates parameters from hypothesis lambda = gamma; S*T free parameters
-##' 
+##' @title estimate hypothesis c_st = 1
+##'
+##' @description estimates parameters from hypothesis lambda = gamma
+##'
+##' @details There are S*T free parameters under this hypothesis.
+##'
 ##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
 ##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
 ##' @param J vector of predators caught in each time period

### 2014-10-02 R/estimate_C.R

@@ -1,4 +1,8 @@
-##' estimate parameters from hypothesis: lambda = c*gamma; S*T + 1 free parameters
+##' @title estimate hypothesis c_st = c
+##'
+##' @description estimate parameters from hypothesis lambda = c*gamma
+##'
+##' @details There are S*T + 1 free parameters under this hypothesis.
 ##' 
 ##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
 ##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS

### 2014-10-02 R/estimate_Cs.R

@@ -1,4 +1,8 @@
-##' estimates parameters from hypothesis lambda = gamma c'_s; S*T + S free parameters
+##' @title estimate hypothesis c_st = c_s
+##'
+##' @description estimates parameters from hypothesis lambda_s = c_s * gamma_s
+##'
+##' @details There are S*T + S free parameters under this hypothesis.
 ##' 
 ##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
 ##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS

### 2014-10-02 R/estimate_Ct.R

@@ -1,4 +1,8 @@
-##' estimates parameters from hypothesis lambda = gamma c'_t; S*T + T free parameters
+##' @title estimate hypothesis c_st = c_t
+##'
+##' @description estimates parameters from hypothesis lambda_t = c_t * gamma_t
+##'
+##' @details There are S*T + T free parameters under this hypothesis.
 ##' 
 ##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
 ##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS

### 2014-10-02 R/estimate_gen.R

@@ -1,4 +1,9 @@
-##' estimates parameters from hypothesis lambda != gamma; 2*S*T free parameters
+##' @title estimate a reparameterization of the  hypothesis c_st
+##'
+##' @description estimates parameters from hypothesis lambda != gamma, where lambda
+##' is indepdent of gamma
+##'
+##' @details There are 2*S*T free parameters under this hypothesis.
 ##' 
 ##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
 ##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS

### 2014-10-02 R/hypotheses.R

@@ -1,4 +1,6 @@
-##' function to calculate hypotheses given a user specifed null and alternative
+##' @title calculate hypotheses
+##'
+##' @description calculates hypotheses, given a user specifed null and alternative
 ##'
 ##' @param hyp a 2-tuple specifying the null and alternative hypotheses, respectively
 ##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS

### 2014-10-02 R/likelihood.R

@@ -1,4 +1,6 @@
-##' log-likelihood of predator preferances model
+##' @title observed count log-likelihood of predator preferances model
+##'
+##' @description log-likelihood of fully observed count data predator preferances model
 ##'
 ##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time and cols indexed by prey species, TxS
 ##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time and cols indexed by prey species, TxS
@@ -25,8 +27,9 @@ ll <- function(Xdst, Ydst, lambda, gamma, J, I, c=NULL) {
     out - sumT(I*sumSp(gamma)) + sumST(Ydst*log(gamma))
 }
 
-
-##' log-likelihood of predators preferances model with EM
+##' @title non-observed count log-likelihood of predators preferances
+##'
+##' @description  log-likelihood for non-observed count data; model with EM
 ##'
 ##' @param Zdst matrix of sums of indicators whether or not predator ate prey species s during occurrence t; TxS
 ##' @param Ydst matrix sum of number of caught prey species s during occurrence t; TxS
@@ -35,47 +38,26 @@ ll <- function(Xdst, Ydst, lambda, gamma, J, I, c=NULL) {
 ##' @param J vector of predators caught in each time period
 ##' @param I vector of number of days all traps were left out in a given time period
 ##' @param c scalar in null hypotheses
-## llEM <- function(Zdst, Ydst, lambda, gamma, J, I, c=NULL) {
-##     if ( is.null(c) ) {
-##         ## avoid log(0) when hat{lambda} = 0
-##         ## these values contribute nothing, so ignore them
-##         ## TBD might also need to watch for any hat{gamma} = 0
-##         w <- which(Zdst != 0, arr.ind=T)
-##         out <- sumST(Zdst[w]*log(1-exp(-lambda[w])) - (J[w[,1]] - Zdst[w])*lambda[w])
-##     } else {
-##         lc <- length(c)
-##         if ( lc == 1 ) {
-##             out <- sumST(Zdst*log(1-exp(-c*gamma))) - c*sumST((J-Zdst)*gamma)
[…]

### 2014-10-02 R/simPref.R

@@ -1,4 +1,12 @@
-##' simulate data for predator preferences model
+##' @title simulate data
+##'
+##' @description  simulate data for predator preferences model
+##'
+##' @details Both lambda and gamma must be specified as a matrix with rows indexing
+##' time and columns indexing the number of species.
+##'
+##' @return A list consisting of two dataframes, eaten and caught, made
+##' specifically for the function \code{predPref}.
 ##'
 ##' @param S number of prey species
 ##' @param T number of time periods
@@ -7,6 +15,8 @@
 ##' @param lambda matrix of rates at which predator eats prey species; TxS
 ##' @param gamma matrix of rates at which prey species is seen in habitat; TxS
 ##' @param EM boolean specifying test of EM algorithm
+##'
+##' @seealso \code{\link{predPref}}, \code{\link{testPref}}
 ##' @export
 simPref <- function(S, T, J, I, lambda, gamma, EM=F) {
 

### 2014-10-02 R/summary.predPref.R

@@ -1,4 +1,7 @@
-##' summary method for predPref objects
+##' @title predPref summary
+##'
+##' @description  summary method for predPref objects as returned by the
+##' function \code{predPref}
 ##'
 ##' @param object predPref object as returned from predPref()
 ##' @param ... additional arguments

### 2014-10-13 paper/data.tex
aiming for clarity in writing; sections methods and simulations. add box plots.

\section{Real Data}
\label{sec:data}

Interest in the predator genus Schizocosa and the prey orders? Diptera, Collembola, Ensifera, for time periods October $2011$ to March $2013$.  
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2014-10-13 paper/discussion.tex
aiming for clarity in writing; sections methods and simulations. add box plots.

@@
 \label{sec:discussion}
-
-{\color{red} should we include mention of the spiders dataset?}
 
@@
 Further testing of the ratio of rates for which predators eat to encounter prey species allows researchers to make specific conclusions about predators' preferences.
-For instance, rates acros time can be estimated to make statements about seasonal effects on a predator's eating habits, or relative rates across species groups allows for statements about the preferences for species $s_1$ to that of $s_2$.
+For instance, rates acros time can be estimated to make statements about seasonal effects on a predator's eating habits, or relative rates across species groups allows for statements about the preferences for different species.
 
@@
 By modelling all of the observed data, both the binary responses and the number of prey species caught, and the missing count data, we are able to use the EM algorithm to extract as much information from the data as possible.
-Though this is nice in theory, in practice the success of this tweak to our original model is limited by the magnitude of the unknown parameters $\lambda_{st}$.
+Though this is nice in theory, in practice the success of this modification to our original model is limited by the magnitude of the unknown parameters $\lambda_{st}$.
 

### 2014-10-13 paper/introduction.tex
aiming for clarity in writing; sections methods and simulations. add box plots.

--- 2014-08-27 paper/introduction.tex
@@
-\section*{Introduction}
+\section{Introduction}
 \label{sec:intro}
+
+Modeling a predator's food selection was given much attention for a short period of time; see \citet{Strauss:1979} and the references within.
+Many, if not all, of the indices developed, though intuitive, focused on a snapshot in time and only on one prey species, and rarely obtained practical asymptotic properties.
+This leaves interested practicioners to the most computationally manageable of the techniques, and has completely ignored simultaneous testing across an array of both species and time.
+
+The model presented here maintains tractable asymptotic properties while being general enough to take into account multiple speices and time points.
+By modeling both time and any number of prey species, we are able to see a more detailed anlyais of the predator's eating preferences.
+Further, the simplicity of the model allows us to consider predators for which exact tallies of the number of each variety of prey species eaten within any given time period is not observed.
+Instead, we rely on the researcher being able to DNA sequence the contents of the predator's gut, and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.
+Despite the data's lost information, under certain situations our model is able to accurately estimate the parameters of interest.
+
 %%% Local Variables: 

### 2014-10-13 paper/main.tex
aiming for clarity in writing; sections methods and simulations. add box plots.

@@
 
-% questions for simon:
-% 1) counts are proportional to?
-% 2) re: solve 1 - exp(-l) = 1 for l.  What is the limiting variable?
-
 \begin{abstract}
-The modeling of a predator's prey selection has a breif history that includes many intuitive indices, few of which have both reasonable statistical justification and tractable asymptotic properties.
+The literature on modeling a predator's prey selection includes many intuitive indices, few of which have both reasonable statistical justification and tractable asymptotic properties.
 Here, we provide a simple model that meets both of these criteria, while extending previous work to include an array of multiple species and time points.
 Further, we apply the Expectation-Maximization algorithm to cases where exact counts of the number of prey species eaten in a particular timer period is not observed.
-Simulations yada yada yada...
+A simulation is provided to demonstrate the accuracy of our methods.
 \end{abstract}
@@
 \input{simulations.tex}
+\input{data.tex}
 \input{discussion.tex}

### 2014-10-13 paper/methods.tex
aiming for clarity in writing; sections methods and simulations. add box plots.

@@
 Prey species, indexed by $s \in \{1, \ldots, S \}$, are collected in the traps and counted at each time period.
-We assume these counts represent the number of prey species $s$ the predator will encounter on average during time period $t \in \{1, \ldots, T\}$.
-The counts of prey species $s$ caught during time period $t$ are hypothesized to be independent draws from a Poisson distribution with rate parameter $\gamma_{st}$.
-We further assume that the number of prey species found in the gut of the trapped predators, also follows a Poisson distribution with rate $\lambda_{st}$.
-Here, the parameter $\lambda_{st}$ represents the rate at which the predator ate the encountered prey species $s$ during time period $t$.
+The number of prey species $s$ the predator will encounter on average during time period $t \in \{1, \ldots, T\}$ are considered random draws from a Poisson distribution with rate parameter $\gamma_{st}$.
+We further assume that the number of prey species found in the gut of the similarly trapped predators follows a Poisson distribution with rate $\lambda_{st}$.
+Here, the parameter $\lambda_{st}$ represents the rate at which the predator ate prey species $s$ during time period $t$.
 By modeling $\lambda_{st}$ and $\gamma_{st}$ we are able to test claims about a predator's eating preferences.
 
-The use of Poisson distributions make the following implicit assumptions: $1)$ traps independently catch the prey species of interest, $2)$ predators eat, at a constant rate, prey species independently, $3)$ predators eat indepedent of each other.
-Our model's ability to accurately portray predators' habitats and underlying eating preferences dependes on the degree to which these assumptions are broken.
+The use of Poisson distributions make the following implicit assumptions: $1)$ traps independently catch the prey species of interest, $2)$ predators eat prey species independently, $3)$ predators eat indepedent of each other.
 
-Let $X_{jst} \iid \mathcal{P}(\lambda_{st})$ denote the number of prey species $s$ that predator $j$ ate during occurrence $t$ where $j \in \{1, \ldots, J_t\}$.
-Let $Y_{ist} \iid \mathcal{P}(\gamma_{st})$ denote the number of prey species $s$ found in trap $i$ during occurrence $t$, $i \in \{1, \ldots, I_t\}$.
+We denote the number of predators and the numer of prey species caught, in each time period $t$, by $J_t$ and $I_t$, respectively.
+Let $X_{jst} \iid \mathcal{P}(\lambda_{st})$ represent the number of prey species $s$ that predator $j$ ate during occurrence $t$, where $j \in \{1, \ldots, J_t\}$.
+Let $Y_{ist} \iid \mathcal{P}(\gamma_{st})$ represent the number of prey species $s$ found in trap $i$ during occurrence $t$, $i \in \{1, \ldots, I_t\}$.
 Formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda}$ and $\boldsymbol{\gamma}$ offer insights to the relative rates at which predators eat particular prey species.
@@
 
+The first hypothesis states that predators and traps sample all prey species at the same rate.
+One imagines this is the case if the predator simply eats that which comes within its reach, thus suggesting a diet indifferene.
+The second says that predators sample prey proportionally across all time periods.
+The third hypothesis says that predators sample different prey species at different rates, but each rate is steady across time.
+This implies that the predator expresses preferences for one prey species over another, but is unresponsive to changes due to time.
+Conversely, the forth hypothesis implies that each prey species is sampled similarly within each time period, while the rates across time are allowed to change.
+The fifth assumes a predator's selection varies by both time and prey species.
+This would make sense if environmental variables, say weather, or prey availability, and taste were affecting predators' selection strategies.
+
 \begin{figure}
@@
 
-\subsection{Fully Observed Data}
+\subsection{Fully Observed Count Data}
 
 The likelihood function that allows for estimation of these parameters is as follows.
-Since we assume $X_{jst}$ is independent of $Y_{ist}$ we can simply multiply the respective Poisson probability density functions together, and then form products over all $s,t$ to get the likelihood.
+Since we assume $X_{jst}$ is independent of $Y_{ist}$ we can simply multiply the respective Poisson probability density functions, and then form products over all $s,t$ to get the likelihood.
 
@@
 
-\noindent Writing all $5$ hypotheses as $\lambda_{st} = c_{st}\gamma_{st}$, we can, in some cases find analytic solutions for the maximum likelihood estimates of $c_{st}$ and $\gamma_{st}$.
-When the data are balanced $J_t = J$, $I_t = I$, and $c_{st} = c$ analytic solutions exist -- {\color{red}what about $c_{st} = 1$}.
+\noindent Writing all five hypotheses as $\lambda_{st} = c_{st}\gamma_{st}$, we can, in the simplest cases, find analytic solutions for the maximum likelihood estimates of $c_{st}$ and $\gamma_{st}$.
+Under the hypothesis $c_{st} = 1$, and when the data are balanced $J_t = J$, $I_t = I$, and $c_{st} = c$ analytic solutions exist.
+Namely, these solutions are
+
+\begin{equation*}
+  \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_t + I_t}, \quad \text{ and } \quad \hat{c} = \frac{I \sum_{s,t} X_{\cdot st}}{J \sum_{s,t} Y_{\cdot st}}, \quad \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{I \left( \frac{\sum_{st} X_{\cdot st}}{\sum_{st} Y_{\cdot st}} + 1 \right)}
+\end{equation*}
+
+\noindent respectively, where $X_{\cdot st} = \sum_{j=1}^{J_t}X_{jst}$ and $Y_{\cdot st} = \sum_{i=1}^{I_t} Y_{ist}$.
+
 In all other cases, analytic solutions are not readily available and instead we rely on the fact that the log-likelihood $l(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log{L}$ is concave.
-We maximize the log-likelihood by iteratively solving partial derivatives of $l$, with respect to $c_{st}$ and $\gamma_{st}$, set equal to zero
+We maximize the log-likelihood, using coordinate descent \citep{Luo:1992}, by iteratively solving partial derivatives of $l$, with respect to $c_{st}$ and $\gamma_{st}$, set equal to zero
 
@@
 
-\noindent where $X_{\cdot st} = \sum_{j=1}^{J_t}X_{jst}$ and $Y_{\cdot st} = \sum_{i=1}^{I_t} Y_{ist}$.
-
 \subsection{Unobserved Counts}
 
-Working with biologists who study spider eating preferences, we have found that not all predators' allow for easily counted prey species in their guts.
+Working with biologists who study spider foraging, we have found that not all predators allow for easily counted prey species in their guts.
 As an alternative strategy, we can rely on the DNA sequencing of a sample from the predators' guts.
-If such sequencing returns a positive response, say a $1$ if a particular predator ate prey species $s$ and $0$ otherwise, we can, albeit with some information lost, model predators' eating preferences with the above framework using the EM algorithm.
-
-When the data $X_{jst}$ are not observed, and instead a boolean response indicating if a predator ate any number of prey species $s$ during time $t$ is observed, we can still, to some degree, estimates the parameters of interest $\boldsymbol{\lambda}, \boldsymbol{\gamma}$.
-Because some information is observed, we can treat the counts as missing and use the EM algorithm to find the maximum likelihood estimates of the observed data likelihood.
+If such sequencing returns a positive response, say a $1$ if a particular predator ate prey species $s$, and $0$ otherwise, we can, albeit with these incomplete data, model predators' eating preferences with the above framework using the EM algorithm.
 
 We denote the binary response that the predator did in fact eat at least one prey species $s$ in time period $t$ by $Z_{jst} = 1(X_{jst} > 0)$.
-Now, the observed data are independent and identically distributed Bernoulli observations with parameter $p_{st} = 1-\exp\{-\lambda_{st}\}$.
-Using this we can find the complete data likelihood by first noting that the conditional distribution of the unobserved data $X_{jst}$ given $Z_{jst}, Y_{ist}, \boldsymbol{\lambda}, \boldsymbol{\gamma}$ is a truncated Poisson distribution
+Now the observed data are independent and identically distributed Bernoulli observations with $p_{st} = 1-\exp\{-\lambda_{st}\}$.
+Despite not observing $X_{jst}$, the EM algorithm is able to find maximum likelihood estimates of the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$ using the complete data log-likelihood, $l_{comp}(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log f_{X,Y,Z}(x_{jst},y_{ist},z_{jst}|\boldsymbol{\lambda}, \boldsymbol{\gamma})$.
+
+With the distribution of $Z_{jst}$ in hand, we calculate the complete data likelihood by first noting that the conditional distribution of the unobserved data $X_{jst}$ given $Z_{jst}, Y_{ist}, \boldsymbol{\lambda}, \boldsymbol{\gamma}$ is a truncated Poisson distribution
 
@@
   f_{X|Y,Z,\boldsymbol{\lambda},\boldsymbol{\gamma}}(x_{jst}) =
-  \frac{\exp{\{-\lambda_{st}\}} \lambda_{st}^{x_{jst}}}{(1 - \exp{\{-\lambda_{st}\}}) x_{jst}!}1(x_{jst} > 0) \quad \text{ where } \quad \E_{[X|Y,Z]}X_{jst} = \frac{\lambda_{st} \exp{\{\lambda_{st} \}}}{\exp{\{ \lambda_{st} \}} - 1}.
+  \frac{\exp{\{-\lambda_{st}\}} \lambda_{st}^{x_{jst}}}{(1 - \exp{\{-\lambda_{st}\}}) x_{jst}!}1(x_{jst} > 0) \quad \text{ where } \quad \E_{X|Y,Z}X_{jst} = \frac{\lambda_{st} \exp{\{\lambda_{st} \}}}{\exp{\{ \lambda_{st} \}} - 1}.
 \end{equation*}
 
-\noindent From this conditional distribution we get the joint distribution of $X_{jst}, Z_{jst}$
+\noindent From this conditional distribution, we get the joint distribution of $X_{jst}, Z_{jst}$
 
@@
 
-\noindent The E-step of the EM algorithm is completed by computing the expected value of the complete data log-likelihood with respect to $f_{X|Y,Z,\boldsymbol{\lambda},\boldsymbol{\gamma}}(x_{jst})$ to get
+% introduce EM and its iterative steps earlier
+% just a couple of general sentences; explain the two steps and what they do
+% rubin and little (~1980); mclchalan and Krishnan book
+% split up E and M step, with \cdot ^(k) notation earlier
+% mention convergence criterion
 
-\begin{align*}
-  \E l_{comp} 
-  & = \E \log{f_{X,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})} + \log{f_{Y|\boldsymbol{\gamma}}(y_{ist})} \\
+The EM algorithm works in two steps \citep{Dempster:1977,McLachlan:2007}.
+To drop dependence on the unobserved data, the E-step consists of taking the expectation of $l_{comp}$ with respect to the conditional distribution of $X$ given the observed data and a current estimate of the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$, namely $f_{X|Y,Z,\boldsymbol{\lambda}}(x_{jst})$.
+The current estimate of the parameters are used within the calculation of $\E_{X|Y,Z,\boldsymbol{\lambda}}X_{jst}$ and not within the distribution functions of the underlying model.
+The M-step then maximizes $Q = \E l_{comp}$ with respect to the parameters in the model.
+The E-step and M-step are then iteratively updated until a convergence criterion is met.
+
+Denoting iterations and current estimates of parameters of the EM algorithm with a superscript $(k)$, the two EM steps are E-step: $Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \E_{X|Y,Z,\boldsymbol{\lambda}^{(k)}} l_{comp}$, and M-step: $(\boldsymbol{\lambda}^{(k+1)}, \boldsymbol{\gamma}^{(k+1)}) = \argmax_{(\boldsymbol{\lambda}, \boldsymbol{\gamma})} Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})$.
+The calculation of $Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})$ is not difficult and is given in equation~\ref{eq:estep}.
+
+\begin{align}
+  \label{eq:estep}
+  \begin{split}
+  Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})
+  & = \E \log{f_{X,Y,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})} + \log{f_{Y|\boldsymbol{\gamma}}(y_{ist})} \\
   & = \sum_{s=1}^S \sum_{t=1}^T \sum_{j=1}^{J_t} \E \log{f_{X,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})}
   + \sum_{s=1}^S \sum_{t=1}^T \sum_{i=1}^{I_t} \log{f_{Y|\boldsymbol{\gamma}}(y)} \\
-  & = \sum_{s,t,j} \left( - \lambda_{st} 
-    + z_{jst} \E X_{jst} \log{\lambda_{st}}  \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right) + \text{const} \\
-  & = \sum_{s,t} \left( -J_t \lambda_{st} + z_{\cdot st} \E X_{jst} \log{\lambda_{st}} \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right) + \text{const}.
-\end{align*}
+  & \propto \sum_{s,t,j} \left( - \lambda_{st} 
+    + z_{jst} \E X_{jst} \log{\lambda_{st}}  \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right) \\
+  & \propto \sum_{s,t} \left( -J_t \lambda_{st} + z_{\cdot st} \E (X_{jst}|\lambda_{st}^{(k)},\gamma_{st}^{(k)}) \log{\lambda_{st}} \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right).
+\end{split}
+\end{align}
 
-The EM algorithm requires iteratively solving $(\boldsymbol{\lambda}^{(k+1)},\boldsymbol{\gamma}^{(k+1)}) = \argmax_{\boldsymbol{\lambda},\boldsymbol{\gamma}} \E l_{comp}^k$.
-Though no analytic solution to this maximization exists, one idea is to iteratively solve partial derivatives of $\E l_{comp}$ set equal to zero until convergence.
-In fact, as we only need find parameter values that increase the observed likelihood, we forgo fully iterating to find the maximum values of the parameters and instead perform just one step uphill within each EM iteration.
-This strategy is significantly less computationally intensive, thus generating a much faster generalized EM (GEM) algorithm.
+In this case, no analytic solution to the M-step exists so we choose to maximize $Q$ with coordinate descent \citep{Luo:1992}.
+In fact, as we only need find parameters that increase the value of $Q$, we forgo fully iterating to find the maximum and instead perform just one step uphill within each EM iteration.
+Since $Q^{(k)}$ is concave and smooth in the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$, we are able to use the convergence of parameter estimates, $||(\boldsymbol{\lambda}^{(k)}, \boldsymbol{\gamma}^{(k)}) - (\boldsymbol{\lambda}^{(k+1)}, \boldsymbol{\gamma}^{(k+1)})||_{\infty} < \tau $, for some $\tau>0$, as our stopping criterion.
 
-This GEM algorithm accurately estimates the parameters when values of $\lambda_{st}$ are relatively small, such that zeros are prevalent in the data $Z_{jst}$.
+This generalized EM algorithm accurately estimates the parameters when values of $\lambda_{st}$ are relatively small, such that zeros are prevalent in the data $Z_{jst}$.
 In this case, not too much information is lost since estimation of $\E Z_{jst}$ can be estimated well by the proportion of observed zeros.
 On the other hand, if the predator consistently eats a given prey species, few to no zeros will show up in the observed data and $\E Z_{jst}$ is estimated to be nearly $1$.
-The loss of information is best seen by attempting to solve for $\lambda_{st}$ in the equation $1 = \E Z_{jst} = 1 - \exp\{-\lambda_{st}\}$; essentially $\lambda_{st}$ is sent off to $+\infty$.
+The loss of information is best seen by attempting to solve for $\lambda_{st}$ in the equation $\E Z_{jst} = 1 = 1 - \exp\{-\lambda_{st}\}$.
+As the proportion of ones in the observed data increase, we expect $\lambda_{st}$ to grow exponentially large.
+When no zeros are present in the data, where only ones are observed, the maximum likelihood estimate can be made arbitrarily large by sending the parameter off to infinity.
 
@@
 
-All hypotheses are evaluated via a likelihood ratio test (LRT), with statistic
+The likelihood ratio test (LRT) statistic is
 
@@
   \label{eq:LRT}
-    \Lambda(X,Y) := -2 \log{ \frac{ \sup L(\theta_0|X,Y)}{ \sup L(\theta_1|X,Y)} },
+    \Lambda(X,Y) := -2 \log{ \frac{ \sup_{\theta_0} L(\theta_0|X,Y)}{ \sup_{\theta_1} L(\theta_1|X,Y)} },
 \end{equation*}
@@
 \noindent where $\theta_0, \theta_1$ represent the parameters estimated under the null and alternative hypotheses, respectively.
-It is well known that the asymptotic distribution of $\Lambda$ is a $\chi_{\rho}^2$ distribution with $\rho$ degrees of freedom.
-Under the EM algorithm we use $L_{obs}(Z,Y)$ as the likelihood in the calculation of $\Lambda$.
+It is well known that the asymptotic distribution of $\Lambda$ is a $\chi_{\rho}^2$ distribution with $\rho$ degrees of freedom \citep{Wilks:1938}.
+When the observations $X_{jst}$ are not observed, we use $L_{obs}(Z,Y)$ as the likelihood in the calculation of $\Lambda$.
 
-The degrees of freedom $\rho$ are set equal to the number of free parameters available in the stated hypotheses under question.
+The degrees of freedom $\rho$ equal the number of free parameters available in the stated hypotheses under question.
 If we put the null hypothesis to be $H_0: \lambda_t = c_t \gamma_t$, for all $t$ and contrast this against $H_1: \lambda_{st} = c_{st}\gamma_{st}$ then there are $\rho = 2(S \cdot T) - S \cdot T - T = S \cdot T - T$ degrees of freedom.
@@
 
-After determining which model best fits the data, more detail can be extracted through a hypothesis test on the elements of $c_{st}$, or in vector notation as $\mathbf{c} \in \mathbb{R}^{S\cdot T}$.
-Let the elements of $\hat{\mathbf{c}}$ be the estimated values, $\hat{c}_{st}$, as estimated via the above maximum likelihood framework.
+After determining which model best fits the data, more detail can be extracted through a hypothesis test of the elements of $c_{st}$, or in vector notation as $\mathbf{c} \in \mathbb{R}^{S\cdot T}$.
+Let the elements of $\hat{\mathbf{c}}$ be the maximum likelihood estimates, $\hat{c}_{st}$, as found via the framework above.
 Since $\hat{\mathbf{c}}$ is asymptotically normally distributed, any linear combination of the elements is also asymptotically normally distributed.
@@
 
-Suppose, for example, that the hypothesis $c_s$ is determined to best fit the data with $s$ ranging $s = 1, 2$.
-We can test to see whether or not the two species $s_1$ and $s_2$ are statistically equally preferred under the null hypothesis $\hat{c}_{s_1} = \hat{c}_{s_2}$.
-This hypothesis is alternatively written in vector notation as $a^t\hat{\mathbf{c}} = 0$, where $a = (1, -1)^t$.
-Tests of the following form $H_0: a^t\mathbf{c} = \mu$ against any alternative of interest are then standard $Z$-tests.
-Similarly, confidence intervals of any level can be obtained if desired.
+Suppose, for example, that the hypothesis $c_s$ is determined to best fit the data with $s$ ranging $s = 1, 2, 3$.
+We can test to see whether or not two species are statistically equally preferred under the null hypothesis $c_{1} = c_{2}$.
+This hypothesis is alternatively written in vector notation as $a^t\mathbf{c} = 0$, where $a = (1, -1, 0)^t$.
+Tests of the following form $H_0: a^t\mathbf{c} = \mu$ against any alternative of interest are then approximate $Z$-tests.
+Confidence intervals of any size are similarly, readily obtained.
 

### 2014-10-13 paper/simulations.tex
aiming for clarity in writing; sections methods and simulations. add box plots.

--- 2014-08-27 paper/simulations.tex
@@
-\section*{Simulations}
+\section{Simulations}
+\label{sec:sim}
 
-Our simulations hypothesize two prey species, and five time points.
-Of the hierarchy of hypotheses, we consider the four most interesting cases $\{c, c_s, c_t, c_st\}$.
-Sample sizes are chosen randomly from four overlapping levels.
-We put ``small'' sample sizes to be randomly sampled numbers in $[20,50]$, ``medium'' encompasses $[30:75]$, ``large'' $[50,150]$, and ``huge'' $[100,200]$.
-Hence, we randomly sample trap and predator observations for each occurence from one of the sample size levels, then cycle through all hypotheses.
+% use language like: data generating model and fitted model, not hypotheses
+
+Our simulations assume two prey species and five time points, throughout.
+Of the hierarchy of hypotheses, we generate data under three models: $c, c_s, c_t$.
+Sample sizes are randomly chosen from four overlapping levels.
+Let ``small'' sample sizes be randomly sampled numbers in $[20,50]$, ``medium'' encompass $[30,75]$, ``large'' $[50,150]$, and ``huge'' $[100,200]$.
+Hence, we randomly sample prey and predator gut count observations for each time period from one of the sample size levels, then cycle generate data for all models.
 This is repeated for each level of sample size.
-We simulate data for each of the $16$ scenarios above under both the standard maximum likelihood theory and under the EM algorithm.
-Only the details of the EM algorithm are presented here, as the other simulations are standard theory of point estimation.
+We simulate $500$ replicate datasets for each of the twelve scenarios above for both types of data, fully observed count data, $X_{jst}$, and for non-count data, when we observe only a binary response, $Z_{jst} = 1(X_{jst}>0)$.
+Each scenario is then fit with the true model that generated the data.
+A subset of the examples are provided here; the interested reader is referred to the supplementary materials for the complete set of simulation results.
+
+For all simulated data, the true parameter values for the rate at which prey species are encountered in the wild are fixed to be $\gamma_{st} = \pi, \, \forall s,t$.
+The values of $\lambda_{st}$ are set with respect to each data generating model.
+For the model $c_{st} = c$, where predator preferences don't vary by either time or species, we consider $\lambda_{st} = 2\pi, \forall s,t$.
+Under the model $c_s$, the ratio of rates vary by species only, so we put $\lambda_{1t} = \sqrt{2}$ and $\lambda_{2t} = \pi$.
+Hence, $c_1 = \sqrt{2}/\pi \approx 0.45$ and $c_2 = 1$.
+For the last model, $c_t$, the ratio of rates vary by time $t$.
+Here, we put $\lambda_{st} = t$ for $t \in \{1, \ldots, 5 \}$.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.5]{nonem}
+  \caption{Density plots of all $500$ estimates of fitting the true model to the data generated from models $c_s,c$ are shown with sample sizes small and medium, repsepectively.}
+  \label{fig:nonem}
+\end{figure}
+
+% more words about how these numbers were estimated; estimtes from across all simulations
+% simulations show nearly unbiased
+% drop decimal places; we don't have precision to state further. 2
+
+Figure~\ref{fig:nonem} shows density plots of the estimates of $c_s,c$ when fitting the true model to the fully observed count data generated under models $c_s$ and $c$.
+The plots provide evaluations of parameter bias under each scenario.
+In the first display, the parameters $c_1 \approx 0.45$ and $c_2 = 1$ are on average, across all $500$ simulations, estimated as $\hat{c}_1 = 0.45$ and $\hat{c}_2 = 1.00$, with standard errors of $\text{se}(\hat{c}_1) = 0.03$ and $\text{se}(\hat{c}_2) = 0.06$.
+The second display provides a similar conclusion.
+Averaging across all $500$ simulations, the parameter $c=2$ is estimated as $\hat{c} = 2.00$.
+
+Though only a subset of the results are shown, when count data are fully observed, simulations show our model to have very little bias.
+This is further seen in figure~\ref{fig:bp}, where box plots of the parameter estimates, centered at true parameter values, of the correct model fit to data generated from both $c_s$ and $c_t$ show empirically almost no bias.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.5]{bp}
+  \caption{Shown are all $500$ estimates, centered at the true parameter values, from fitting the true model to the data generated from models $c_s,c$ with sample sizes small and medium, repsepectively.}
+  \label{fig:bp}
+\end{figure}
+
+When we generate data with unobserved count data, the story is slightly different.
+As noted above under certain circumstatnces our EM model accurately estimates the parameters of interest, and at different times can greatly over-estimate the parameters.
+To investigate this issue further, we consider the same scenarios mentioned above, but now we reduce all of our count data down to binary observations.
+For each scenario's generated data, we fit our EM algorithm as if we knew a priori the true underlying model that generated the observed data.
+
+Figure~\ref{fig:em} contains density plots of fitting the true EM model, for all $500$ replications, of the data generating models $c_s,c_t$ with small and huge sample sizes, respectively.
+When data are generated under the model $c_s$ and the true model fit to the data, we find even for the small sample size that point estimates are reasonably accurate.
+When parameter values are of sufficient size to make zeros in the simulated data less common, the estimates from fitting the correct model to the generated data are occassionally over-estimated.
+This effect is easily seen in Figure~\ref{fig:em} for larger values of $c_t$ despite the increased sample size, but is also seen, less dramatically, in the density plot for the $c_s$ generated data.
+
+% The plots of $\gamma_{st}$ under the EM algorithm are not given as we do not consider missing data in the estimation of these parameters.
+
+% more of which model are we fitting, which model is generating the data
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.5]{em}
+  \caption{Density plots of all $500$ estimates of fitting the true model to the data generated from models $c_s,c_t$ are shown with sample sizes small and huge, repsepectively.}
+  \label{fig:em}
+\end{figure}
+
+The over-estimation of parameters, a symptom of the loss of information due to the unobserved count data, can also be seen with box plots of the $500$ point estimates, centered at their respective true parameter values.
+Figure~\ref{fig:em_bp} contains box plots of the same scenarios in Figure~\ref{fig:em}, but with the small and huge sample sizes.
+It is clear that as the parameters values increase and zeros in the observed data $Z_{jst}$ become less prevalent, over-estimation occurs more frequently.
+Because of the tendency for the EM algorithm to over-estimate some parameter values, bias for non-count data is worse than it is for the count data.
+
+%Table~\ref{tab:bias} displays the max absolute bias across all indices $s,t$ of $\lambda_{st}$ for each simulation scenario.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.5]{em_bp}
+  \caption{Shown are all $500$ estimates, centered at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$ are shown with sample sizes small and huge, repsepectively.}
+  \label{fig:em_bp}
+\end{figure}
+
+% \begin{table}
+%   \centering
+%   \begin{tabular}{ll|rrrr}
+%     & & small & medium & large & huge \\ 
+%     \hline
+%     count & $c_t$ & 0.040 & 0.013 & 0.020 & 0.011 \\ 
+%     & $c_s$ & 0.028 & 0.015 & 0.010 & 0.005 \\ 
+%     & $c$ & 0.030 & 0.016 & 0.017 & 0.012 \\ 
+%     non-count & $c_t$ & 4.151 & 3.155 & 0.941 & 1.423 \\ 
+%     & $c_s$ & 0.138 & 0.096 & 0.044 & 0.032 \\ 
+%     & $c$ & 2.924 & 2.078 & 0.669 & 0.547 \\ 
+%   \end{tabular}
+%   \caption{Max absolute bias of all $\lambda_{st}$ for each simulation scenario.}
+%   \label{tab:bias}
+% \end{table}
 

### 2014-10-13 spiders.Rcheck/00_pkg_src/spiders/R/contrasts.R

@@ -1,0 +1,120 @@
+##' @title linear contrast of c_st
+##'
+##' @description estimates linear contrasts of the elements of c, c_s, c_t, c_st
+##'
+##' @return A list with class '"htest"' containing the following components:
+##' 
+##' statistic: the value of the t-statistic.
+##' 
+##' parameter: the degrees of freedom for the t-statistic.
+##' 
+##' p.value: the p-value for the test.
+##' 
+##' conf.int: a confidence interval for the mean appropriate to the specified
+##' alternative hypothesis.
+##' 
+##' estimate: the estimated mean or difference in means depending on whether
+##' it was a one-sample test or a two-sample test.
+##' 
+##' null.value: the specified hypothesized value of the mean or mean difference
+##' depending on whether it was a one-sample test or a two-sample test.
+##' 
+##' alternative: a character string describing the alternative hypothesis.
+##' 
+##' method: a character string indicating what type of t-test was performed.
+##' 
+##' data.name: a character string giving the names of the data.
+##' 
+##' @details The input vector b performs the linear transformation
+##' t(b) \%*\% matrix(c_st), so that c_st becomes a column vector by indexing
+##' t first and then s.  Hence there is no requirement of a linear
+##' contrast, a linear transformation such that
+##' t(b) \%*\% matrix(1, nrow=length(b)) != 0 is allowed.
+##'
+##' @param x a prefPref object as fit by the eponymous function
+##' @param b a vector to linearly transform c_st
+##' @param mu a number to test the linear contrast against in the null
+##' @param alternative string to specify alternative hypothesis
+##' @param conf.level confidence level of the interval
+##' 
+##' @examples
+##' # set parameters
+##' Predators <- Traps <- 100
+##' PreySpecies <- 2
+##' Times <- 5
+##' g <- matrix(sqrt(2), nrow=Times, ncol=PreySpecies)     # gamma
+##' l <- matrix(seq(0.4,1.8,length.out=5)*sqrt(2), nrow=Times, ncol=PreySpecies) # ct
+##'
+##' # fit model
+##' \dontrun{
+##' fdata <- simPref(PreySpecies, Times, Predators, Traps, l, g, EM=FALSE)
+##' pref <- predPref(fdata$eaten, fdata$caught, hypotheses=c('ct', 'cst'))
[…]

### 2014-10-13 spiders.Rcheck/00_pkg_src/spiders/R/estimate_1.R

@@ -1,0 +1,70 @@
+##' @title estimate hypothesis c_st = 1
+##'
+##' @description estimates parameters from hypothesis lambda = gamma
+##'
+##' @details There are S*T free parameters under this hypothesis.
+##'
+##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param J vector of predators caught in each time period
+##' @param I vector of number of days all traps were left out in a given time period
+##' @param EM boolean; whether or not EM algorithm is used
+##' @param em_maxiter integer specifying max number of EM iterations
+##' @param BALANCED boolean; whether or not data are BALANCED
+est1 <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
+
+    JI <- J + I
+    if (EM) {
+        em_iter <- 1
+        init <- est1(Xdst, Ydst, J, I, FALSE, em_maxiter, BALANCED)
+        gammaHat <- gammaHat_old <- init$gamma
+
+        ## iterate EM
+        while (EM) {
+            
+            ## some numbers
+            lambda <- gammaHat
+            elambda <- exp(lambda)
+            EX <- lambda*elambda / (elambda - 1)
+            ZEX <- Xdst*EX
+
+            ## update parameters
+            gammaHat <- (ZEX + Ydst) / JI
+
+            ## check convergence of EM
+            if ( converged(gammaHat, gammaHat_old) ) break
+            
+            ## if not converged, store updated estimates
+            gammaHat_old <- gammaHat
+
+            ## limit iterations
+            em_iter <- em_iter+1
+            if ( em_iter > em_maxiter )
+                stop(sprintf('est1: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
+        }
+
+        ## calc standard error with est params
+        egamma <- exp(gammaHat)
+        Info <- diag(unlist(Ydst/gammaHat^2 + J*egamma/(egamma - 1)^2))
+        
+        ## calc loglik with est params
+        loglik <- llEM(Xdst, Ydst, gammaHat, gammaHat, J, I)
+        
+        list('lambda' = as.matrix(gammaHat), 'gamma' = as.matrix(gammaHat),
[…]

### 2014-10-13 spiders.Rcheck/00_pkg_src/spiders/R/estimate_C.R

@@ -1,0 +1,158 @@
+##' @title estimate hypothesis c_st = c
+##'
+##' @description estimate parameters from hypothesis lambda = c*gamma
+##'
+##' @details There are S*T + 1 free parameters under this hypothesis.
+##' 
+##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param J vector of predators caught in each time period
+##' @param I vector of number of days all traps were left out in a given time period
+##' @param EM boolean; whether or not EM algorithm is used
+##' @param em_maxiter integer specifying max number of EM iterations
+##' @param BALANCED boolean; whether or not data are BALANCED
+estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
+
+    ## some numbers
+    S <- ncol(Xdst); s <- seq_len(S)
+    T <- nrow(Xdst); t <- seq_len(T)
+    ST <- S*T
+
+    if (EM) {
+        ## initialize some values
+        em_iter <- 1
+        cHat <- cHat_old <- runif(1)        
+
+        init <- est1(Xdst, Ydst, J, I, FALSE, em_maxiter, BALANCED)
+        gammaHat <- gammaHat_old <- init$gamma 
+        lambda <- elambda <- init$lambda
+
+        ## iterate EM
+        while ( TRUE ) {
+            
+            ## expected value of Xjst
+            lambda <- cHat*gammaHat
+            elambda <- exp(-lambda)
+            EX <- lambda / (1-elambda)
+
+            ## convenience
+            ZEX <- Xdst*EX
+
+            ## iterate only once simultaneous eqs
+            gammaHat <- (ZEX + Ydst) / (cHat*J + I)
+            cHat <- sumST(ZEX) / sumT(J*sumSp(gammaHat))
+
+            ## check convergence of EM
+            if ( converged(cHat, cHat_old) &&
+                converged(gammaHat, gammaHat_old) ) break
+            
+            ## if not converged, store updated estimates
+            cHat_old <- cHat
+            gammaHat_old <- gammaHat
+            ## print(sprintf('EM iteration %d found values: c = %f', em_iter, cHat))
+            em_iter <- em_iter+1
+            
+            ## limit iterations
[…]

### 2014-10-13 spiders.Rcheck/00_pkg_src/spiders/R/estimate_Cs.R

@@ -1,0 +1,136 @@
+##' @title estimate hypothesis c_st = c_s
+##'
+##' @description estimates parameters from hypothesis lambda_s = c_s * gamma_s
+##'
+##' @details There are S*T + S free parameters under this hypothesis.
+##' 
+##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param J vector of predators caught in each time period
+##' @param I vector of number of days all traps were left out in a given time period
+##' @param EM boolean; whether or not EM algorithm is used
+##' @param em_maxiter integer specifying max number of EM iterations
+##' @param BALANCED boolean; whether or not data are BALANCED
+estCs <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
+
+    ## some numbers
+    S <- ncol(Xdst); s <- seq_len(S)
+    T <- nrow(Xdst); t <- seq_len(T)
+    ST <- S*T
+
+    if (EM) {
+        ## initialize some values
+        em_iter <- 1
+        cHat <- cHat_old <- runif(S)
+        init <- est1(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED)
+        gammaHat <- gammaHat_old <- init$gamma 
+
+        ## iterate EM
+        while ( TRUE ) {
+            
+            ## expected value of Xjst
+            lambda <- sapply(s, function(j) cHat[j]*gammaHat[,j]) # col-wise `*`
+            elambda <- exp(-lambda)
+            EX <- lambda/(1-elambda)
+
+            ## convenience
+            ZEX <- Xdst*EX
+
+            ## iterate only once simultaneous eqs
+            ZEXY <- ZEX + Ydst
+            gammaHat <- sapply(s, function(j) ZEXY[,j] / (cHat[j]*J + I)) # col-wise `/`
+            cHat <- sumT(ZEX) / sumT(J*gammaHat)
+            
+            ## check convergence of EM
+            if ( converged(cHat, cHat_old) &&
+                converged(gammaHat, gammaHat_old) ) break
+            
+            ## if not converged, store updated estimates
+            cHat_old <- cHat
+            gammaHat_old <- gammaHat
[…]

### 2014-10-13 spiders.Rcheck/00_pkg_src/spiders/R/estimate_Cst.R

@@ -1,0 +1,137 @@
+##' @title estimate hypothesis c_st
+##'
+##' @description estimates parameters from hypothesis lambda = c_st * gamma
+##'
+##' @details There are 2*S*T free parameters under this hypothesis
+##' 
+##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param J vector of predators caught in each time period
+##' @param I vector of number of days all traps were left out in a given time period    
+##' @param EM boolean; whether or not EM algorithm is used
+##' @param em_maxiter integer specifying max number of EM iterations
+##' @param BALANCED boolean; whether or not data are BALANCED
+estCst <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
+
+    ## some numbers
+    S <- ncol(Ydst); s <- seq_len(S)
+    T <- nrow(Ydst); t <- seq_len(T)
+    ST <- S*T; st <- seq_len(ST)
+
+    if (EM) {
+        ## initialize some values
+        em_iter <- 1
+        cHat <- cHat_old <- matrix(runif(ST), nrow=T)
+        init <- est1(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED)
+        gammaHat <- gammaHat_old <- init$gamma 
+        
+        ## iterate EM
+        while ( TRUE ) {
+            ## update lambda
+            lambda <- cHat*gammaHat
+            elambda <- exp(-lambda)
+            EX <- lambda/(1-elambda)
+
+            ## convenience
+            ZEX <- Xdst*EX
+
+            ## iterate only once simultaneous eqs
+            gammaHat <-  (ZEX + Ydst)/(J*cHat + I)                 
+            cHat <- ZEX/(J*gammaHat)
+
+            ## check convergence
+            if ( converged(cHat, cHat_old) &&
+                converged(gammaHat, gammaHat_old) ) break
+
+            ## if not converged, store updated estimates
+            cHat_old <- cHat
+            gammaHat_old <- gammaHat
+            ## print(sprintf('EM iteration %d found values: c = %f', em_iter, cHat))
+            em_iter <- em_iter+1
+            
+            ## limit iterations
[…]

### 2014-10-13 spiders.Rcheck/00_pkg_src/spiders/R/estimate_gen.R

@@ -1,0 +1,91 @@
+##' @title estimate a reparameterization of the  hypothesis c_st
+##'
+##' @description estimates parameters from hypothesis lambda != gamma, where lambda
+##' is indepdent of gamma
+##'
+##' @details There are 2*S*T free parameters under this hypothesis.
+##' 
+##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param J vector of predators caught in each time period
+##' @param I vector of number of days all traps were left out in a given time period    
+##' @param EM boolean; whether or not EM algorithm is used
+##' @param em_maxiter integer specifying max number of EM iterations
+##' @param BALANCED boolean; whether or not data are BALANCED
+estGen <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
+
+    ## some numbers
+    S <- ncol(Ydst); s <- seq_len(S)
+    T <- nrow(Ydst); t <- seq_len(T)
+    ST <- S*T; st <- seq_len(ST)
+
+    if (EM) {
+        
+        ## estimate gamma
+        gammaHat <- Ydst/I
+
+        ## initialize some things
+        em_iter <- 1
+        lambdaHat <- lambdaHat_old <- gammaHat
+
+        ## iterate EM
+        while ( TRUE ) {
+            ## update lambda
+            elambda <- exp(-lambdaHat)
+            EX <- lambdaHat/(1-elambda)
+            lambdaHat <- Xdst*EX/J
+
+            ## check convergence
+            if ( converged(lambdaHat, lambdaHat_old) ) break
+            lambdaHat_old <- lambdaHat
+            em_iter <- em_iter+1
+            
+            ## limit iterations
+            if ( em_iter > em_maxiter )
+                stop(sprintf('estGen: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
+        }
+        
+        ## since estimated 0s don't follow from calculations above
+        lambdaHat[which(Xdst == 0, arr.ind=T)] <- 0
+
+        ## calc standard error with est params
+        ## SE <- seEM(lambdaHat, gammaHat, NULL, Xdst, Ydst, J, I)
[…]

### 2014-10-13 spiders.Rcheck/00_pkg_src/spiders/R/hypotheses.R

@@ -1,0 +1,77 @@
+##' @title calculate hypotheses
+##'
+##' @description calculates hypotheses, given a user specifed null and alternative
+##'
+##' @param hyp a 2-tuple specifying the null and alternative hypotheses, respectively
+##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
+##' @param J vector of predators caught in each time period
+##' @param I vector of number of days all traps were left out in a given time period
+##' @param balanced boolean specifying balanced data or not
+##' @param EM boolean specifying if EM algorithm should be used
+##' @param em_maxiter maximum number of iterations allowed for EM algorithm
+calcHypotheses <- function(hyp, Xdst, Ydst, J, I, balanced, EM, em_maxiter) {
+
+    ## create map of estimators
+    fns <- list()
+    fns[['1']] <- est1; fns[['c']] <- estC
+    fns[['Cs']] <- estCs; fns[['Ct']] <- estCt
+    fns[['Cst']] <- estCst; fns[['gen']] <- estGen
+
+    ## calculate null and alternative hypotheses
+    null <- fns[[hyp[1]]](Xdst, Ydst, J, I, EM, em_maxiter, balanced)
+    alt <- fns[[hyp[2]]](Xdst, Ydst, J, I, EM, em_maxiter, balanced)        
+
+    ## calculate degrees of freedom
+    S <- ncol(Xdst); T <- nrow(Xdst); ST <- S*T
+    nullDF <- length(null[['c']])
+    altDF <- ifelse(is.null(alt[['c']]), ST, length(alt[['c']]))
+    df <- altDF - nullDF
+    
+    list('llH0' = null[['ll']], 'llH1' = alt[['ll']],
+         'null' = null, 'alt' = alt, 'df' = df)    
+}
+
+##' function to check user specified hypotheses
+##'
+##' @param hyp a 2-tuple specifying the null and alternative hypotheses, respectively
+checkHypotheses <- function(hyp) {
+
+    hyp <- tolower((as.character(hyp)))
+    
+    ## initialize output
+    H <- rep(0, 2)
+    
+    ## H0
[…]

### 2014-10-13 spiders.Rcheck/00_pkg_src/spiders/R/predPref.R

@@ -1,0 +1,110 @@
+##' @title parameter estimation
+##'
+##' @description Estimates parameters of predator preferences model and calculates LRT.
+##' Eaten and caught dataframes are indexed with rows across time points
+##' and columns of prey species.
+##'
+##' @return A list of class 'predPref' with the following elements:
+##'
+##' null: parameters as estimated under the specified null hypothesis.
+##' 
+##' alt: parameters as estimated under the specified alternative hypothesis.
+##' 
+##' loglikH0: the null hypothesis log-likelihood, with constants not accounted for.
+##' 
+##' loglikH1: the alternative hypothesis log-likelihood, with constants not accounted for.
+##' 
+##' J: a column vector of dimension T containing the number of predators in
+##' each time period.
+##'
+##' I: a column vector of dimension T containing the number of traps in each time period.
+##'
+##' LRT: the likelihood ratio test statistics.
+##'
+##' hypotheses: a 2-tuple of the user specified hypotheses.
+##'
+##' data.name: a character string giving the names of the data.
+##'
+##' @param eaten a dataframes of eatings preferences; TxS
+##' @param caught a dataframes of caught prey species; TxS
+##' @param hypotheses a 2-tuple specifying the null and alternative hypotheses, respectively
+##' @param alpha LRT level of significance
+##' @param em_maxiter maximum number of iterations allowed for EM algorithm
+##'
+##' @seealso \code{\link{simPref}}
+##'
+##' @examples
+##' # set parameters
+##' Predators <- Traps <- 100
+##' PreySpecies <- 2
+##' Times <- 5
+##' g <- matrix(sqrt(2), nrow=Times, ncol=PreySpecies)     # gamma
+##' l <- matrix(seq(0.4,1.8,length.out=5)*sqrt(2), nrow=Times, ncol=PreySpecies) # ct
+##'
+##' # fit model
+##' \dontrun{
+##' fdata <- simPref(PreySpecies, Times, Predators, Traps, l, g, EM=FALSE)
+##' predPref(fdata$eaten, fdata$caught, hypotheses=c('ct', 'cst'))
+##' }
+##' 
+##' @export
+predPref <- function(eaten, caught, hypotheses = c('c', 'Ct'), alpha=0.05, em_maxiter=1000) {
+
[…]

### 2014-10-13 spiders.Rcheck/00_pkg_src/spiders/R/simPref.R

@@ -1,0 +1,69 @@
+##' @title simulate data
+##'
+##' @description  simulate data for predator preferences model
+##'
+##' @details Both lambda and gamma must be specified as a matrix with rows indexing
+##' time and columns indexing the number of species.
+##'
+##' @return A list consisting of two dataframes, eaten and caught, made
+##' specifically for the function \code{predPref}.
+##'
+##' @param S number of prey species
+##' @param T number of time periods
+##' @param J scalar or vector (of length T) number of predators caught at each time
+##' @param I scalar or vector (of length T) effective number of traps at each time
+##' @param lambda matrix of rates at which predator eats prey species; TxS
+##' @param gamma matrix of rates at which prey species is seen in habitat; TxS
+##' @param EM boolean specifying test of EM algorithm
+##'
+##' @seealso \code{\link{predPref}}, \code{\link{testPref}}
+##' @export
+simPref <- function(S, T, J, I, lambda, gamma, EM=F) {
+
+    ## checks
+    if (!isTRUE(all.equal(J, round(J)))) J <- round(J)
+    if (!isTRUE(all.equal(I, round(I)))) I <- round(I)
+    
+    ## some numbers
+    ns <- seq_len(S)                    # index prey species
+    nt <- seq_len(T)                    # index times
+    lJ <- length(J)
+    lI <- length(I)
+    if (lJ != 1 && lJ != T) stop('J not a scalar or vector of appropriate size.')
+    if (lI != 1 && lI != T) stop('J not a scalar or vector of appropriate size.')
+    if (lJ == 1) {
+        J <- rep(J[1], T)
+    }
+    if (lI == 1) {
+        I <- rep(I[1], T)
+    }
+    
+    ## initialize data frame
+    eaten <- as.data.frame(matrix(NA, nrow=sum(J), ncol=S+1))
+    caught <- as.data.frame(matrix(NA, nrow=sum(I), ncol=S+1))
+
+    ## fill times
+    colnames(eaten) <- colnames(caught) <- c('time', paste('preySpecies', ns, sep=''))
+    eaten[,1] <- unlist(lapply(nt, function(x) rep(paste('time', x, sep=''), J[x])))
+    caught[,1] <- unlist(lapply(nt, function(x) rep(paste('time', x, sep=''), I[x])))
+
+    ## fill data frame
+    times <- unique(eaten$time)
+    for ( i in ns ) {
[…]

### 2014-10-13 spiders.Rcheck/00_pkg_src/spiders/R/test.R

@@ -1,0 +1,253 @@
+##' @title simulate and test predPref
+##'
+##' @description simulate data and test function \code{predPref} on each simulated dataset
+##'
+##' @return A list of two elements, one for each hypothesis. Each element contains
+##' parameter estimates as calculated by the respective hypothesis, and the number of
+##' iterations used to fit said model.
+##'
+##' @param J number of predators caught at each time
+##' @param I effective number of traps at each time
+##' @param lambda matrix of rates at which predator eats prey species; TxS
+##' @param gamma matrix of rates at which prey species is seen in habitat; TxS
+##' @param M number of simulated datasets
+##' @param hyp a 2-tuple specifying the null and alternative hypotheses, respectively
+##' @param EM boolean specifying test of EM algorithm
+##' @param em_maxiter maximum number of iterations allowed for EM algorithm
+##'
+##' @seealso \code{\link{simPref}}
+##'
+##' @examples
+##'
+##' # set parameters
+##' Predators <- Traps <- 100
+##' PreySpecies <- 2
+##' Times <- 5
+##' g <- matrix(sqrt(2), nrow=Times, ncol=PreySpecies)     # gamma
+##' l <- matrix(seq(0.4,1.8,length.out=5)*sqrt(2), nrow=Times, ncol=PreySpecies) # ct
+##'
+##' # test functions
+##' \dontrun{
+##' testPref(Predators, Traps, l, g, M=10, hyp=c('c', 'cst'), EM=F)
+##' }
+##' 
+##' @export
+testPref <- function(J, I, lambda, gamma, M=100, hyp=c('C', 'Cst'), EM = FALSE, em_maxiter = 100) {
+
+    ## initialize output structure
+    out <- vector('list', 2)
+    names(out) <- c('null', 'alt')
+    need_set_output <- TRUE             # need initialize storage within output?
+
+    ## some numbers 
+    S <- ncol(lambda)
+    T <- nrow(lambda)
+    
+    for ( m in seq_len(M) ) {
+        
+        ## simulate data and fit model
+        fdata <- simPref(S, T, J, I, lambda, gamma, EM=EM)
+        prefs <- predPref(fdata$eaten, fdata$caught, hypotheses = hyp, em_maxiter = em_maxiter)
+
+        ## initialize storage within output structure
+        if ( need_set_output ) {
+            ST <- length(unlist(prefs$null$gamma))
+            out$null[['gamma']] <- out$alt[['gamma']] <- matrix(0, M, ST)
[…]

### 2014-10-13 spiders.Rcheck/00_pkg_src/spiders/R/util.R

@@ -1,0 +1,48 @@
+##' sum specified columns by time
+##'
+##' @param data a dataframe
+##' @param vars column variables in data to sum over
+##' @param by extra variables to sum by
+getTimeCounts <- function(data, vars, by) {
+    if (missing(by)) by <- 'time' else by <- c('time', by)
+    ## ddply(data, by, function(dfr, idx) colSums(as.matrix(dfr[,idx])), vars)
+    ddply(data, by, colwise(sum, vars))
+}
+
+##' count number of spiders or traps in each unit of time
+##'
+##' @param data a dataframe
+getUnitCounts <- function(data) {
+    ddply(data, .(time), summarize, total=length(time))$total
+}
+
+##' checks convergence of the parameters for the estimation functions
+##'
+##' @param theta an object, convertible to a matrix, of current parameter values
+##' @param theta_old an object, convertible to a matrix, of old parameter values
+##' @param eps tolerance to determine convergence
+converged <- function(theta, theta_old, eps=1e-5) {
+    isTRUE(all.equal(theta, theta_old, tolerance=eps, check.names=F, check.attr=F))
+}
+
+## sum over species to get a vector of values for each time period
+sumSp <- function(mat) {
+    matrix(rowSums(mat), nrow=nrow(mat))
+}
+
+## sum over times to get a vector of values for each species
+sumT <- function(mat) {
+    matrix(colSums(mat), ncol=ncol(mat))
+}
+
+sumST <- function(mat) {
+    sum(mat)
+}
+
+## colors stolen from http://geography.uoregon.edu/datagraphics/color_scales.htm
+cols <- c('orange1' = "#FFBF80", 'orange2' = "#FF8000",
+          'yellow1' = "#FFFF99", 'yellow2' = "#FFFF33",
+          'green1' = "#B2FF8C", 'green2' = "#33FF00",
+          'blue1' = "#A6EDFF", 'blue2' = "#1AB2FF",
+          'purple1' = "#CCBFFF", 'puple2' = "#664CFF",
+          'red1' = "#FF99BF", 'red2' = "#E61A33")

### 2014-10-13 spiders.Rcheck/tests/testthat/test_predPref.R

@@ -1,0 +1,155 @@
+context('Output predPref.')
+
+## set up some data
+Predators <- c(200, 210, 220, 230, 240)
+Traps <- Predators
+PreySpecies <- 2
+Times <- 5
+TxS <- c(Times, PreySpecies)
+g <- matrix(sqrt(2), nrow=Times, ncol=PreySpecies)
+
+
+## noEM
+test_that('noEM, 1 vs c, null', {
+    l <- matrix(sqrt(2), nrow=Times, ncol=PreySpecies)
+    fdata <- simPref(PreySpecies, Times, Predators, Traps, l, g, EM=F)
+    pref <- predPref(fdata$eaten, fdata$caught, hypotheses=c('1', 'c'), em_maxiter = 50000)
+    expect_is(pref, 'predPref')             # inherits class
+    expect_equal(length(pref$null$c), 0)    # hypotheses correctly chosen
+    expect_equal(length(pref$alt$c), 1)
+    expect_equal(dim(pref$null$gamma), TxS) # dimensions of estimates
+    expect_equal(dim(pref$alt$gamma), TxS)
+    expect_more_than(pref$LRT$p.value, 0.001)     # correct conclusion
+    expect_that(all(diag(pref$null$var)>0), is_true()) # variances (null) are positive
+    expect_that(all(diag(pref$alt$var)>0), is_true()) # variances (alt) are positive
+})                                      
+
+test_that('noEM, c vs c^t, null', {
+    l <- matrix(2*sqrt(2), nrow=Times, ncol=PreySpecies)
+    fdata <- simPref(PreySpecies, Times, Predators, Traps, l, g, EM=F)
+    pref <- predPref(fdata$eaten, fdata$caught, hypotheses=c('c', 'Ct'), em_maxiter = 50000)
+    expect_is(pref, 'predPref')             # inherits class
+    expect_equal(length(pref$null$c), 1)    # hypotheses correctly chosen
+    expect_equal(length(pref$alt$c), Times)
+    expect_equal(dim(pref$null$gamma), TxS) # dimensions of estimates
+    expect_equal(dim(pref$alt$gamma), TxS)
+    expect_more_than(pref$LRT$p.value, 0.001)     # correct conclusion
+    expect_that(all(diag(pref$null$var)>0), is_true()) # variances (null) are positive
+    expect_that(all(diag(pref$alt$var)>0), is_true()) # variances (alt) are positive
+})
+
[…]

### 2014-10-21 paper/data.tex
first words about a read dataset; Simon's methods updates

@@
 
-Interest in the predator genus Schizocosa and the prey orders? Diptera, Collembola, Ensifera, for time periods October $2011$ to March $2013$.  
+We use the dataset from {\color{red} our Biologist colleagues}, where interest lies in the Wolf spider, genus Schizocosa, and the prey orders Diptera, Collembola, for each month from October $2011$ to March $2013$.
+Following our hierarchy of hypotheses, we find that the most parameter rich model, $\lambda_{st} = c_{st} \gamma_{st}$ where each month and each prey species is given its own parameter, fits these data best.
+This estimates $72$ parameters in total, $32$ of which estimate $c_{st}$.
+
+We demonstrate our method's ability to test linear constraints on $c_{st}$, by setting up a linear contrast that evaluates whether or not the model $c_t$ appropriately fits the data.
+This linear contrast approach is the same as testing the null hypothesis $H_0: \lambda_{st} = c_t \gamma_{st}, \forall t$ against the alternative hypothesis $H_1: \lambda_{st} = c_{st} \gamma_{st}$, but with a different statistical test than the likelihood ratio test.
+
+To do this we use the thirty-six estimates of $c_{st}$ fit from our EM algorithm, together with the linear contrast, $a^t = (1, \ldots, 1, -1, \ldots, -1)^t$.
+This linear contrast asks if a model with only one parameter for both prey orders within each time period fits as well as a model with two parameters for both prey orders in each time period.
+Hence, the null hypothesis for this linear contrast is $H_0: a^tc_{st} = 0$ against the alternative hypothesis $H_1: a^tc_{st} \ne 0$.
+The test gives $p-value < 2.2\times 10^{-16}$, indicating that the more parameter rich model is strongly favored, just as did the likelihood ratio test.
+
 %%% Local Variables: 

### 2014-10-21 paper/methods.tex
first words about a read dataset; Simon's methods updates

@@
 
-Working with biologists who study spider foraging, we have found that not all predators allow for easily counted prey species in their guts.
+% Working with biologists who study spider foraging, we have found that not all predators allow for easily counted prey species in their guts.
 As an alternative strategy, we can rely on the DNA sequencing of a sample from the predators' guts.
@@
 
-We denote the binary response that the predator did in fact eat at least one prey species $s$ in time period $t$ by $Z_{jst} = 1(X_{jst} > 0)$.
-Now the observed data are independent and identically distributed Bernoulli observations with $p_{st} = 1-\exp\{-\lambda_{st}\}$.
-Despite not observing $X_{jst}$, the EM algorithm is able to find maximum likelihood estimates of the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$ using the complete data log-likelihood, $l_{comp}(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log f_{X,Y,Z}(x_{jst},y_{ist},z_{jst}|\boldsymbol{\lambda}, \boldsymbol{\gamma})$.
+Many times it may not be possible to count the number of individuals of each prey species that are in a predators gut.
+Instead, it may only be possible to detect whether or not a predator consumed the prey species during a given period.
+This is the case in the study of ?
+spiders which relied on DNA sequencing to identify the prey species in a spider's gut contents.
+Despite this, we can still make inference about the predators preferences for the different prey species by using the EM algorithm to compute maximum likelihood estimates.
 
-With the distribution of $Z_{jst}$ in hand, we calculate the complete data likelihood by first noting that the conditional distribution of the unobserved data $X_{jst}$ given $Z_{jst}, Y_{ist}, \boldsymbol{\lambda}, \boldsymbol{\gamma}$ is a truncated Poisson distribution
+We denote the binary random variable indicating that the $j^{th}$ predator did in fact eat at least one individual of prey species $s$ in time period $t$ by $Z_{jst} = 1(X_{jst} > 0)$.
+The variables are independent Bernoulli observations with success probability $p_{st} = P(Z_{jst}=1)= 1-\exp\{-\lambda_{st}\}$.
+Despite not observing $X_{jst}$, we can compute maximum likelihood estimates of the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$ through the EM algorithm using the complete data log-likelihood
+\[
+l_{comp}(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log f_{X,Y,Z}(\boldsymbol x, \boldsymbol y, \boldsymbol z|\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \sum_{s=1}^{S} \sum_{t=1}^T [\sum_{j=1}^{J_t} \log f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda}) + \sum_{i=1}^{I_t}\log f_Y(y_{jst}|\boldsymbol{\gamma}).
+\]
 
+The density of $Y_{jst}$ is exactly as given above and so we focus on deriving the joint density of $X_{jst}$ and $Z_{jst}$.
+With the distribution of $Z_{jst}$ given above, we can compute $f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda})$ by noting that $X_{jst}=0$ with probability 1 if $Z_{jst}=0$ and that $X_{jst}|Z_{jst}=0$ has a truncated Poisson distribution with density
 \begin{equation*}
-  f_{X|Y,Z,\boldsymbol{\lambda},\boldsymbol{\gamma}}(x_{jst}) =
-  \frac{\exp{\{-\lambda_{st}\}} \lambda_{st}^{x_{jst}}}{(1 - \exp{\{-\lambda_{st}\}}) x_{jst}!}1(x_{jst} > 0) \quad \text{ where } \quad \E_{X|Y,Z}X_{jst} = \frac{\lambda_{st} \exp{\{\lambda_{st} \}}}{\exp{\{ \lambda_{st} \}} - 1}.
+  f_{X|Y,Z,\boldsymbol{\lambda},\boldsymbol{\gamma}}(x_{jst}|z_{jst}) =
+  \frac{\exp{\{-\lambda_{st}\}} \lambda_{st}^{x_{jst}}}{(1 - \exp{\{-\lambda_{st}\}}) x_{jst}!}1(x_{jst} > 0) \quad \text{ where } 
 \end{equation*}
-
-\noindent From this conditional distribution, we get the joint distribution of $X_{jst}, Z_{jst}$
-
+and expected value
+\[
+\E_{X|Y,Z}X_{jst} = \frac{\lambda_{st} \exp{\{\lambda_{st} \}}}{\exp{\{ \lambda_{st} \}} - 1}.
+\]
+\noindent The joint density of $X_{jst}, Z_{jst}$ is then 
 \begin{equation*}
@@
     \end{array}
-  \right.
+  \right..
 \end{equation*}
@@
 
-The EM algorithm works in two steps \citep{Dempster:1977,McLachlan:2007}.
-To drop dependence on the unobserved data, the E-step consists of taking the expectation of $l_{comp}$ with respect to the conditional distribution of $X$ given the observed data and a current estimate of the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$, namely $f_{X|Y,Z,\boldsymbol{\lambda}}(x_{jst})$.
-The current estimate of the parameters are used within the calculation of $\E_{X|Y,Z,\boldsymbol{\lambda}}X_{jst}$ and not within the distribution functions of the underlying model.
-The M-step then maximizes $Q = \E l_{comp}$ with respect to the parameters in the model.
-The E-step and M-step are then iteratively updated until a convergence criterion is met.
+The EM algorithm works by iterating two steps, the E-step and M-step, until the optimum is reached \citep{Dempster:1977,McLachlan:2007}.
+Let $k$ index the iterations in the EM algorithm so that $\boldsymbol \lambda^{(k)}$ and $\boldsymbol \gamma^{(k)}$ denote the estimates computed on the $k^{th}$ M-step.
+The E-step consists of computing the expectation of $l_{comp}$ with respect to the conditional distribution of $X$ given the current estimates of the parameters
+\[
+Q^{(k)}(\boldsymbol{\lambda},\boldsymbol{\gamma}) = \E_{X|Y,Z,\boldsymbol{\lambda}^{(k)}} l_{comp}
+\]
+in order to remove the unobserved data. The M-step then involves maximizing $Q = \E l_{comp}$ with respect to the parameters in the model to obtain updated estimates of the parameters,
+\[
+(\boldsymbol{\lambda}^{(k+1)}, \boldsymbol{\gamma}^{(k+1)}) = \argmax_{(\boldsymbol{\lambda}, \boldsymbol{\gamma})} Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma}).
+\]
+These steps are then alternated until a convergence criterion monitoring subsequent differences in the parameter estimates/likelihood values is met.
 
-Denoting iterations and current estimates of parameters of the EM algorithm with a superscript $(k)$, the two EM steps are E-step: $Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \E_{X|Y,Z,\boldsymbol{\lambda}^{(k)}} l_{comp}$, and M-step: $(\boldsymbol{\lambda}^{(k+1)}, \boldsymbol{\gamma}^{(k+1)}) = \argmax_{(\boldsymbol{\lambda}, \boldsymbol{\gamma})} Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})$.
-The calculation of $Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})$ is not difficult and is given in equation~\ref{eq:estep}.
-
+The calculation of $Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})$ is not difficult and is given by:
 \begin{align}
@@
   Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})
-  & = \E \log{f_{X,Y,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})} + \log{f_{Y|\boldsymbol{\gamma}}(y_{ist})} \\
+  & = \E \log{f_{X,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})} + \log{f_{Y|\boldsymbol{\gamma}}(y_{ist})} \\
   & = \sum_{s=1}^S \sum_{t=1}^T \sum_{j=1}^{J_t} \E \log{f_{X,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})}
@@
 \end{align}
-
-In this case, no analytic solution to the M-step exists so we choose to maximize $Q$ with coordinate descent \citep{Luo:1992}.
-In fact, as we only need find parameters that increase the value of $Q$, we forgo fully iterating to find the maximum and instead perform just one step uphill within each EM iteration.
+\noindent No analytic solution to the M-step exists, however, so we chose to maximize $Q$ with coordinate descent \citep{Luo:1992}.
+In fact, as we only need find parameters that increase the value of $Q$ on each iteration, we forgo fully iterating to find the maximum and instead perform just one step uphill within each EM iteration.
 Since $Q^{(k)}$ is concave and smooth in the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$, we are able to use the convergence of parameter estimates, $||(\boldsymbol{\lambda}^{(k)}, \boldsymbol{\gamma}^{(k)}) - (\boldsymbol{\lambda}^{(k+1)}, \boldsymbol{\gamma}^{(k+1)})||_{\infty} < \tau $, for some $\tau>0$, as our stopping criterion.
 
-This generalized EM algorithm accurately estimates the parameters when values of $\lambda_{st}$ are relatively small, such that zeros are prevalent in the data $Z_{jst}$.
+As we show in our simulation study, this generalized EM algorithm accurately estimates the parameters when values of $\lambda_{st}$ are relatively small, such that zeros are prevalent in the data $Z_{jst}$.
 In this case, not too much information is lost since estimation of $\E Z_{jst}$ can be estimated well by the proportion of observed zeros.
@@
 As the proportion of ones in the observed data increase, we expect $\lambda_{st}$ to grow exponentially large.
-When no zeros are present in the data, where only ones are observed, the maximum likelihood estimate can be made arbitrarily large by sending the parameter off to infinity.
+When no zeros are present in the data, so that where only ones are observed, the likelihood can be made arbitrarily large by sending the parameter off to infinity.
 

### 2014-10-21 R/contrasts.R

@@ -31,7 +31,7 @@
 ##' contrast, a linear transformation such that
 ##' t(b) \%*\% matrix(1, nrow=length(b)) != 0 is allowed.
 ##'
-##' @param x a prefPref object as fit by the eponymous function
+##' @param x a predPref object as fit by the eponymous function
 ##' @param b a vector to linearly transform c_st
 ##' @param mu a number to test the linear contrast against in the null
 ##' @param alternative string to specify alternative hypothesis

### 2014-10-21 R/predpref.R

@@ -31,7 +31,7 @@
 ##' @param alpha LRT level of significance
 ##' @param em_maxiter maximum number of iterations allowed for EM algorithm
 ##'
-##' @seealso \code{\link{simPref}}
+##' @seealso \code{\link{simPref}} \code{\link{summary.predPref}}
 ##'
 ##' @examples
 ##' # set parameters

### 2014-10-21 R/test.R

@@ -15,7 +15,7 @@
 ##' @param EM boolean specifying test of EM algorithm
 ##' @param em_maxiter maximum number of iterations allowed for EM algorithm
 ##'
-##' @seealso \code{\link{simPref}}
+##' @seealso \code{\link{simPref}} \code{\link{plotTestPref}} \code{\link{mean.testPref}}
 ##'
 ##' @examples
 ##'
@@ -30,7 +30,7 @@
 ##' \dontrun{
 ##' testPref(Predators, Traps, l, g, M=10, hyp=c('c', 'cst'), EM=F)
 ##' }
-##' 
+##'
 ##' @export
 testPref <- function(J, I, lambda, gamma, M=100, hyp=c('C', 'Cst'), EM = FALSE, em_maxiter = 100) {
 
@@ -116,8 +116,11 @@ testPref <- function(J, I, lambda, gamma, M=100, hyp=c('C', 'Cst'), EM = FALSE, 
 ##' @param hypothesis specify which hypothesis to plot
 ##' @param lambda a matrix of true values of the parameter lambda; TxS
 ##' @param gamma a matrix of true values of the parameter gamma; TxS
+##'
+##' @seealso \code{\link{simPref}} \code{\link{testPref}} \code{\link{mean.testPref}}
+##' 
 ##' @export
-plotTestPref <- function(x, hypothesis = c('null', 'alt', 'both'), lambda = NULL, gamma = NULL) {
+plotTestPref <- function(x, hypothesis = c('null', 'alt'), lambda = NULL, gamma = NULL) {
     
     ## some numbers
     hypothesis <- match.arg(hypothesis)
@@ -150,7 +153,6 @@ plotTestPref <- function(x, hypothesis = c('null', 'alt', 'both'), lambda = NULL
 
     ## plot data
     require(lattice)
-    require(gridExtra)
 
     if ( params ) {
         two <- densityplot(~gamma, data=nullGamma, groups=index,
@@ -174,11 +176,13 @@ plotTestPref <- function(x, hypothesis = c('null', 'alt', 'both'), lambda = NULL
         if ( params ) {
             trueC <- unique(apply(lambda/gamma, 1, unique))
             one <- densityplot(~c, data=nullC, groups=index,
+                               main='Null Hypothesis',
                                panel = function(...) {
                                    panel.densityplot(...)
                                    panel.rug(x = trueC, lwd=3)})
         } else {
[…]

### 2014-10-22 R/estimateC.R
create option to store random seed in testPref

@@ -44,7 +44,9 @@ estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
 
             ## check convergence of EM
             if ( converged(cHat, cHat_old) &&
-                converged(gammaHat, gammaHat_old) ) break
+                converged(gammaHat, gammaHat_old) ) {
+                break
+            }
             
             ## if not converged, store updated estimates
             cHat_old <- cHat
@@ -53,8 +55,10 @@ estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
             em_iter <- em_iter+1
             
             ## limit iterations
-            if ( em_iter > em_maxiter )
+            if ( em_iter > em_maxiter ) {
                 stop(sprintf('estC: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
+            }
+            
         }
         
         ## calc standard error with est params
@@ -110,8 +114,12 @@ estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
 
             ## not sure this is the right spot for these checks
             ## ensure J & I have dimension T or 1
-            if ( length(J) != T ) stop("J indexed oddly says est0.")
-            if ( length(I) != T ) stop("I indexed oddly says est0.")
+            if ( length(J) != T ) {
+                stop("J indexed oddly says est0.")
+            }
+            if ( length(I) != T ) {
+                stop("I indexed oddly says est0.")
+            }
 
             ## initialize some values
             cHat <- cHat_old <- runif(1)
@@ -126,15 +134,18 @@ estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
 
                 ## check convergence
                 if ( converged(gammaHat, gammaHat_old) &&
-                    converged(cHat, cHat_old) ) break
+                    converged(cHat, cHat_old) ) {
+                    break
+                }
 
                 ## if not converged, update estimates for next iteration
                 gammaHat_old <- gammaHat
                 cHat_old <- cHat
                 iter <- iter+1
                 
-                if ( iter > maxiter )
+                if ( iter > maxiter ) {
                     stop(sprintf('estC: %d not sufficient iterations for simultaneous equations.', maxiter))
+                }
             }
             
             ## calc standard error with est params

### 2014-10-22 R/estimateCt.R

@@ -1,8 +1,8 @@
-##' @title estimate hypothesis c_st = c_s
+##' @title estimate hypothesis c_st = c_t
 ##'
-##' @description estimates parameters from hypothesis lambda_s = c_s * gamma_s
+##' @description estimates parameters from hypothesis lambda_t = c_t * gamma_t
 ##'
-##' @details There are S*T + S free parameters under this hypothesis.
+##' @details There are S*T + T free parameters under this hypothesis.
 ##' 
 ##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
 ##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
@@ -11,7 +11,7 @@
 ##' @param EM boolean; whether or not EM algorithm is used
 ##' @param em_maxiter integer specifying max number of EM iterations
 ##' @param BALANCED boolean; whether or not data are BALANCED
-estCs <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
+estCt <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
 
     ## some numbers
     S <- ncol(Xdst); s <- seq_len(S)
@@ -21,7 +21,7 @@ estCs <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
     if (EM) {
         ## initialize some values
         em_iter <- 1
-        cHat <- cHat_old <- runif(S)
+        cHat <- cHat_old <- runif(length(J))
         init <- est1(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED)
         gammaHat <- gammaHat_old <- init$gamma 
 
@@ -29,7 +29,7 @@ estCs <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
         while ( TRUE ) {
             
             ## expected value of Xjst
-            lambda <- sapply(s, function(j) cHat[j]*gammaHat[,j]) # col-wise `*`
+            lambda <- cHat*gammaHat
             elambda <- exp(-lambda)
             EX <- lambda/(1-elambda)
 
@@ -37,13 +37,14 @@ estCs <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
             ZEX <- Xdst*EX
 
             ## iterate only once simultaneous eqs
-            ZEXY <- ZEX + Ydst
[…]

### 2014-10-22 R/estimateGen.R

@@ -36,13 +36,16 @@ estGen <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
             lambdaHat <- Xdst*EX/J
 
             ## check convergence
-            if ( converged(lambdaHat, lambdaHat_old) ) break
+            if ( converged(lambdaHat, lambdaHat_old) ) {
+                break
+            }
             lambdaHat_old <- lambdaHat
             em_iter <- em_iter+1
             
             ## limit iterations
-            if ( em_iter > em_maxiter )
+            if ( em_iter > em_maxiter ) {
                 stop(sprintf('estGen: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
+            }
         }
         
         ## since estimated 0s don't follow from calculations above
@@ -67,8 +70,12 @@ estGen <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
         ## some numbers
         ## not sure this is the right spot for these checks
         T <- nrow(Xdst)
-        if ( length(J) != T ) stop("J indexed oddly says estGen")
-        if ( length(I) != T ) stop("I indexed oddly says estGen")
+        if ( length(J) != T ) {
+            stop("J indexed oddly says estGen")
+        }
+        if ( length(I) != T ) {
+            stop("I indexed oddly says estGen")
+        }
 
         gammaHat <- Ydst / I                # row-wise division
         lambdaHat <- Xdst / J

### 2014-10-22 R/hypotheses.R

@@ -71,7 +71,9 @@ checkHypotheses <- function(hyp) {
     }
 
     ## both
-    if ( H[1] == H[2] ) stop('Null and Alternative hypotheses must be different.')
+    if ( H[1] == H[2] ) {
+        stop('Null and Alternative hypotheses must be different.')
+    }
 
     H
 }

### 2014-10-22 R/simPref.R

@@ -21,16 +21,24 @@
 simPref <- function(S, T, J, I, lambda, gamma, EM=F) {
 
     ## checks
-    if (!isTRUE(all.equal(J, round(J)))) J <- round(J)
-    if (!isTRUE(all.equal(I, round(I)))) I <- round(I)
+    if ( !isTRUE(all.equal(J, round(J))) ) {
+        J <- round(J)
+    }
+    if ( !isTRUE(all.equal(I, round(I))) ) {
+        I <- round(I)
+    }
     
     ## some numbers
     ns <- seq_len(S)                    # index prey species
     nt <- seq_len(T)                    # index times
     lJ <- length(J)
     lI <- length(I)
-    if (lJ != 1 && lJ != T) stop('J not a scalar or vector of appropriate size.')
-    if (lI != 1 && lI != T) stop('J not a scalar or vector of appropriate size.')
+    if ( lJ != 1 && lJ != T ) {
+        stop('J not a scalar or vector of appropriate size.')
+    }
+    if ( lI != 1 && lI != T ) {
+        stop('J not a scalar or vector of appropriate size.')
+    }
     if (lJ == 1) {
         J <- rep(J[1], T)
     }

### 2014-10-22 R/util.R

@@ -4,7 +4,11 @@
 ##' @param vars column variables in data to sum over
 ##' @param by extra variables to sum by
 getTimeCounts <- function(data, vars, by) {
-    if (missing(by)) by <- 'time' else by <- c('time', by)
+    if ( missing(by) ) {
+        by <- 'time'
+    } else {
+        by <- c('time', by)
+    }
     ## ddply(data, by, function(dfr, idx) colSums(as.matrix(dfr[,idx])), vars)
     ddply(data, by, colwise(sum, vars))
 }

### 2014-10-27 paper/data.tex
towards a reworked real data section

--- 2014-10-13 paper/data.tex
@@
 
-Interest in the predator genus Schizocosa and the prey orders? Diptera, Collembola, Ensifera, for time periods October $2011$ to March $2013$.  
+We analyze a dataset that was collected to investigate the eating preferences of the wolf spider, genus \textit{Schizocosa}, towards the prey orders \textit{Diptera}, \textit{Collembola}.
+Predators, caught in traps, were DNA sequenced to determine whether or not any \textit{Diptera} or \textit{Collembola} existed in the predators' guts for each time period.The left panel of figure~\ref{fig:gutP} plots the percentage of spiders that had the prey in their guts across time.
+Prey were similarly collected, in traps distributed across the field site of interest {\color{red} where?}.
+These data were collected for each month from October $2011$ to March $2013$.
+On average, about $34$ and $69$ spiders and prey, respectively, were caught in each time period.
+The range of the sample sizes, across all $18$ months, is $11$ to $181$ for caught spiders, and is from $24$ to $76$ for caught prey.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.5]{time}
+  \caption{moar words}
+  \label{fig:gutP}
+\end{figure}
+
+ Following our hierarchy of hypotheses, we find that the most parameter rich model, $\lambda_{st} = c_{st} \gamma_{st}$ fits these data best.
+This model estimates $72$ parameters in total; since, in this case, there are two prey of interest and $18$ time periods, it takes $36$ parameters to estimate each $c_{st}$ and $\gamma_{st}$.
+The right panel of figure~\ref{fig:gutP} plots the points estimates of $c_{st}$ for both prey and across all time periods.
+
+With the model $c_{st}$ in mind, we can now test any number of linear contrasts.
+Here, we test $c_{1t} = c_{2t}$, for $t \in \{1, \ldots, 18\}$.
+Using a level of significance of $0.05$, and after making a Bonferroni multiple comparisons adjustment, we find that the two prey are similarly prefered in October, November, and December of $2011$ and for March and July of $2012$.
+
 %%% Local Variables: 

### 2014-10-27 paper/discussion.tex
towards a reworked real data section

@@
 
-The model developed here advances the statistics literature available to determine predators' preferences by testing simultaneously across an array of multiple prey species and time.
+The model developed here allows practicioners to determine predators' preferences by testing simultaneously across an array of multiple prey species and time.
 This is acheived via a simple, but statistically powerful, likelihood ratio test.

### 2014-10-27 paper/simulations.tex
towards a reworked real data section

@@
 Of the hierarchy of hypotheses, we generate data under three models: $c, c_s, c_t$.
-Sample sizes are randomly chosen from four overlapping levels.
+Sample sizes for both prey species and predator gut count observations are randomly chosen from four overlapping levels.
 Let ``small'' sample sizes be randomly sampled numbers in $[20,50]$, ``medium'' encompass $[30,75]$, ``large'' $[50,150]$, and ``huge'' $[100,200]$.
-Hence, we randomly sample prey and predator gut count observations for each time period from one of the sample size levels, then cycle generate data for all models.
+Hence, we randomly sample for each time period from one of the sample size levels, then generate data for all models.
 This is repeated for each level of sample size.
@@
 Each scenario is then fit with the true model that generated the data.
+All simulations of non-count data use $\tau = 10^{-5}$ as the convergence tolerance.
 A subset of the examples are provided here; the interested reader is referred to the supplementary materials for the complete set of simulation results.
@@
 
-Figure~\ref{fig:nonem} shows density plots of the estimates of $c_s,c$ when fitting the true model to the fully observed count data generated under models $c_s$ and $c$.
+Figure~\ref{fig:nonem} shows density plots of the estimates of $c_s, c$ when fitting the true model to the fully observed count data generated under models $c_s$ and $c$.
 The plots provide evaluations of parameter bias under each scenario.
@@
 Averaging across all $500$ simulations, the parameter $c=2$ is estimated as $\hat{c} = 2.00$.
-
-Though only a subset of the results are shown, when count data are fully observed, simulations show our model to have very little bias.
 This is further seen in figure~\ref{fig:bp}, where box plots of the parameter estimates, centered at true parameter values, of the correct model fit to data generated from both $c_s$ and $c_t$ show empirically almost no bias.
@@
 
-When we generate data with unobserved count data, the story is slightly different.
+When we generate data with unobserved counts, the story is slightly different.
 As noted above under certain circumstatnces our EM model accurately estimates the parameters of interest, and at different times can greatly over-estimate the parameters.
@@
 
-Figure~\ref{fig:em} contains density plots of fitting the true EM model, for all $500$ replications, of the data generating models $c_s,c_t$ with small and huge sample sizes, respectively.
-When data are generated under the model $c_s$ and the true model fit to the data, we find even for the small sample size that point estimates are reasonably accurate.
+Figure~\ref{fig:em} contains density plots of fitting the true model, for all $500$ replications, of the data generating models $c_s,c_t$ with small and huge sample sizes, respectively.
+When data are generated under the model $c_s$ and the true model fit to the non-count data, we find even for the small sample size that point estimates are reasonable.
 When parameter values are of sufficient size to make zeros in the simulated data less common, the estimates from fitting the correct model to the generated data are occassionally over-estimated.
-This effect is easily seen in Figure~\ref{fig:em} for larger values of $c_t$ despite the increased sample size, but is also seen, less dramatically, in the density plot for the $c_s$ generated data.
+This effect is easily seen in figure~\ref{fig:em} for larger values of $c_t$ despite the increased sample size, but is also seen, less dramatically, in the density plot for the $c_s$ generated data.
+
+The seemingly arbitrary clustering of estimates near four seen in figure~\ref{fig:em} is due to two competing factors.
+As the proportion of ones in the $Z_{jst}$ observations increases, the observed log-likelihood becomes flatter.
+Thus, any finite convergence tolerance will cause the algorithm to terminate eventually.
+By decreasing $\tau$, and increasing, as appropriate, the maximum iterations allowed within the EM algorithm, one can gradually increase the magnitude of a maximum likelihood estimate.
 
@@
   \includegraphics[scale=0.5]{em}
-  \caption{Density plots of all $500$ estimates of fitting the true model to the data generated from models $c_s,c_t$ are shown with sample sizes small and huge, repsepectively.}
+  \caption{Density plots of all $500$ estimates of fitting the true model to the data generated from models $c_s,c_t$, when counts are not observed, are shown with sample sizes small and huge, repsepectively.}
   \label{fig:em}
@@
 
-The over-estimation of parameters, a symptom of the loss of information due to the unobserved count data, can also be seen with box plots of the $500$ point estimates, centered at their respective true parameter values.
+The over-estimation of parameters, a symptom of the loss of information due to the unobserved counts, can also be seen with box plots of the $500$ point estimates, centered at their respective true parameter values.
 Figure~\ref{fig:em_bp} contains box plots of the same scenarios in Figure~\ref{fig:em}, but with the small and huge sample sizes.
 It is clear that as the parameters values increase and zeros in the observed data $Z_{jst}$ become less prevalent, over-estimation occurs more frequently.
-Because of the tendency for the EM algorithm to over-estimate some parameter values, bias for non-count data is worse than it is for the count data.
+This over-estimation, when it occurs, will generally increase the magnitude of the bias.
 
@@
   \includegraphics[scale=0.5]{em_bp}
-  \caption{Shown are all $500$ estimates, centered at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$ are shown with sample sizes small and huge, repsepectively.}
+  \caption{Shown are all $500$ estimates, centered at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$, when counts are not observed, with sample sizes small and huge, repsepectively.}
   \label{fig:em_bp}

### 2014-10-27 R/test.R

@@ -138,9 +138,8 @@ testPref <- function(J, I, lambda, gamma, M=100, hyp=c('C', 'Cst'), EM = FALSE, 
 ##'
 ##' @description plot the output from \code{testPref}
 ##'
-##' @details Function relies on packages \code{lattice} and \code{gridExtra}.  If
-##' true values of lambda and gamma are given, then tick marks will be placed
-##' at those values along the x-axis.
+##' @details If true values of lambda and gamma are given, then tick
+##' (rugs) marks will be placed at those values along the x-axis.
 ##'
 ##' @param x a testPref object as returned by the eponymous function
 ##' @param hypothesis specify which hypothesis to plot
@@ -183,7 +182,6 @@ plotTestPref <- function(x, hypothesis = c('null', 'alt'), lambda = NULL, gamma 
 
     ## plot data
     require(lattice)
-
     if ( params ) {
         two <- densityplot(~gamma, data=nullGamma, groups=index,
                            panel = function(...) {

### 2014-10-27 spiders.Rcheck/00_pkg_src/spiders/R/contrasts.R

@@ -1,120 +1,0 @@
-##' @title linear contrast of c_st
-##'
-##' @description estimates linear contrasts of the elements of c, c_s, c_t, c_st
-##'
-##' @return A list with class '"htest"' containing the following components:
-##' 
-##' statistic: the value of the t-statistic.
-##' 
-##' parameter: the degrees of freedom for the t-statistic.
-##' 
-##' p.value: the p-value for the test.
-##' 
-##' conf.int: a confidence interval for the mean appropriate to the specified
-##' alternative hypothesis.
-##' 
-##' estimate: the estimated mean or difference in means depending on whether
-##' it was a one-sample test or a two-sample test.
-##' 
-##' null.value: the specified hypothesized value of the mean or mean difference
-##' depending on whether it was a one-sample test or a two-sample test.
-##' 
-##' alternative: a character string describing the alternative hypothesis.
-##' 
-##' method: a character string indicating what type of t-test was performed.
-##' 
-##' data.name: a character string giving the names of the data.
-##' 
-##' @details The input vector b performs the linear transformation
-##' t(b) \%*\% matrix(c_st), so that c_st becomes a column vector by indexing
-##' t first and then s.  Hence there is no requirement of a linear
-##' contrast, a linear transformation such that
-##' t(b) \%*\% matrix(1, nrow=length(b)) != 0 is allowed.
-##'
-##' @param x a prefPref object as fit by the eponymous function
-##' @param b a vector to linearly transform c_st
-##' @param mu a number to test the linear contrast against in the null
-##' @param alternative string to specify alternative hypothesis
-##' @param conf.level confidence level of the interval
-##' 
-##' @examples
-##' # set parameters
-##' Predators <- Traps <- 100
-##' PreySpecies <- 2
-##' Times <- 5
-##' g <- matrix(sqrt(2), nrow=Times, ncol=PreySpecies)     # gamma
-##' l <- matrix(seq(0.4,1.8,length.out=5)*sqrt(2), nrow=Times, ncol=PreySpecies) # ct
-##'
-##' # fit model
-##' \dontrun{
-##' fdata <- simPref(PreySpecies, Times, Predators, Traps, l, g, EM=FALSE)
-##' pref <- predPref(fdata$eaten, fdata$caught, hypotheses=c('ct', 'cst'))
[…]

### 2014-10-27 spiders.Rcheck/00_pkg_src/spiders/R/estimate_1.R

@@ -1,70 +1,0 @@
-##' @title estimate hypothesis c_st = 1
-##'
-##' @description estimates parameters from hypothesis lambda = gamma
-##'
-##' @details There are S*T free parameters under this hypothesis.
-##'
-##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
-##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
-##' @param J vector of predators caught in each time period
-##' @param I vector of number of days all traps were left out in a given time period
-##' @param EM boolean; whether or not EM algorithm is used
-##' @param em_maxiter integer specifying max number of EM iterations
-##' @param BALANCED boolean; whether or not data are BALANCED
-est1 <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
-
-    JI <- J + I
-    if (EM) {
-        em_iter <- 1
-        init <- est1(Xdst, Ydst, J, I, FALSE, em_maxiter, BALANCED)
-        gammaHat <- gammaHat_old <- init$gamma
-
-        ## iterate EM
-        while (EM) {
-            
-            ## some numbers
-            lambda <- gammaHat
-            elambda <- exp(lambda)
-            EX <- lambda*elambda / (elambda - 1)
-            ZEX <- Xdst*EX
-
-            ## update parameters
-            gammaHat <- (ZEX + Ydst) / JI
-
-            ## check convergence of EM
-            if ( converged(gammaHat, gammaHat_old) ) break
-            
-            ## if not converged, store updated estimates
-            gammaHat_old <- gammaHat
-
-            ## limit iterations
-            em_iter <- em_iter+1
-            if ( em_iter > em_maxiter )
-                stop(sprintf('est1: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
-        }
-
-        ## calc standard error with est params
-        egamma <- exp(gammaHat)
-        Info <- diag(unlist(Ydst/gammaHat^2 + J*egamma/(egamma - 1)^2))
-        
-        ## calc loglik with est params
-        loglik <- llEM(Xdst, Ydst, gammaHat, gammaHat, J, I)
-        
-        list('lambda' = as.matrix(gammaHat), 'gamma' = as.matrix(gammaHat),
[…]

### 2014-10-27 spiders.Rcheck/00_pkg_src/spiders/R/estimate_gen.R

@@ -1,91 +1,0 @@
-##' @title estimate a reparameterization of the  hypothesis c_st
-##'
-##' @description estimates parameters from hypothesis lambda != gamma, where lambda
-##' is indepdent of gamma
-##'
-##' @details There are 2*S*T free parameters under this hypothesis.
-##' 
-##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
-##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time, and cols indexed by prey species, TxS
-##' @param J vector of predators caught in each time period
-##' @param I vector of number of days all traps were left out in a given time period    
-##' @param EM boolean; whether or not EM algorithm is used
-##' @param em_maxiter integer specifying max number of EM iterations
-##' @param BALANCED boolean; whether or not data are BALANCED
-estGen <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
-
-    ## some numbers
-    S <- ncol(Ydst); s <- seq_len(S)
-    T <- nrow(Ydst); t <- seq_len(T)
-    ST <- S*T; st <- seq_len(ST)
-
-    if (EM) {
-        
-        ## estimate gamma
-        gammaHat <- Ydst/I
-
-        ## initialize some things
-        em_iter <- 1
-        lambdaHat <- lambdaHat_old <- gammaHat
-
-        ## iterate EM
-        while ( TRUE ) {
-            ## update lambda
-            elambda <- exp(-lambdaHat)
-            EX <- lambdaHat/(1-elambda)
-            lambdaHat <- Xdst*EX/J
-
-            ## check convergence
-            if ( converged(lambdaHat, lambdaHat_old) ) break
-            lambdaHat_old <- lambdaHat
-            em_iter <- em_iter+1
-            
-            ## limit iterations
-            if ( em_iter > em_maxiter )
-                stop(sprintf('estGen: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
-        }
-        
-        ## since estimated 0s don't follow from calculations above
-        lambdaHat[which(Xdst == 0, arr.ind=T)] <- 0
-
-        ## calc standard error with est params
-        ## SE <- seEM(lambdaHat, gammaHat, NULL, Xdst, Ydst, J, I)
[…]

### 2014-10-27 spiders.Rcheck/00_pkg_src/spiders/R/likelihood.R

@@ -1,63 +1,0 @@
-##' @title observed count log-likelihood of predator preferances model
-##'
-##' @description log-likelihood of fully observed count data predator preferances model
-##'
-##' @param Xdst matrix of sums of number of eaten prey species s during occurrence t; rows indexed by time and cols indexed by prey species, TxS
-##' @param Ydst matrix sum of number of caught prey species s during occurrence t; rows indexed by time and cols indexed by prey species, TxS
-##' @param lambda matrix of parameters representing rates predator ate prey species s in time period t; TxS
-##' @param gamma matrix of parameters representing rates traps caught prey species s in time period t; TxS
-##' @param J vector of predators caught in each time period
-##' @param I vector of number of days all traps were left out in a given time period
-##' @param c scalar in null hypotheses
-ll <- function(Xdst, Ydst, lambda, gamma, J, I, c=NULL) {
-    if ( is.null(c) ) {
-        out <- -sumT(J*sumSp(lambda)) + sumT(log(J)*sumSp(Xdst)) + sumST(Xdst*log(lambda))
-    } else {
-        lc <- length(c)
-        if ( lc == nrow(Xdst) || lc == 1 ) {
-            out <- -sumT(c*J*sumSp(gamma)) + sumT(log(c*J)*sumSp(Xdst)) + sumST(Xdst*log(gamma)) 
-        } else if ( lc == ncol(Xdst) ) {
-            s <- seq_len(ncol(Xdst))
-            cGamma <- sapply(s, function(j) gamma[,j]*c[j])
-            out <- -sumT(J*sumSp(cGamma)) + sumT(log(J)*sumSp(Xdst)) + sumST(Xdst*log(cGamma))
-        } else {
-            out <- -sumT(J*sumSp(c*gamma)) + sumT(log(J)*sumSp(Xdst)) + sumST(Xdst*log(c*gamma))
-        }
-    }
-    out - sumT(I*sumSp(gamma)) + sumST(Ydst*log(gamma))
-}
-
-##' @title non-observed count log-likelihood of predators preferances
-##'
-##' @description  log-likelihood for non-observed count data; model with EM
-##'
-##' @param Zdst matrix of sums of indicators whether or not predator ate prey species s during occurrence t; TxS
-##' @param Ydst matrix sum of number of caught prey species s during occurrence t; TxS
[…]

### 2014-10-27 spiders.Rcheck/00_pkg_src/spiders/R/predPref.R

@@ -1,110 +1,0 @@
-##' @title parameter estimation
-##'
-##' @description Estimates parameters of predator preferences model and calculates LRT.
-##' Eaten and caught dataframes are indexed with rows across time points
-##' and columns of prey species.
-##'
-##' @return A list of class 'predPref' with the following elements:
-##'
-##' null: parameters as estimated under the specified null hypothesis.
-##' 
-##' alt: parameters as estimated under the specified alternative hypothesis.
-##' 
-##' loglikH0: the null hypothesis log-likelihood, with constants not accounted for.
-##' 
-##' loglikH1: the alternative hypothesis log-likelihood, with constants not accounted for.
-##' 
-##' J: a column vector of dimension T containing the number of predators in
-##' each time period.
-##'
-##' I: a column vector of dimension T containing the number of traps in each time period.
-##'
-##' LRT: the likelihood ratio test statistics.
-##'
-##' hypotheses: a 2-tuple of the user specified hypotheses.
-##'
-##' data.name: a character string giving the names of the data.
-##'
-##' @param eaten a dataframes of eatings preferences; TxS
-##' @param caught a dataframes of caught prey species; TxS
-##' @param hypotheses a 2-tuple specifying the null and alternative hypotheses, respectively
-##' @param alpha LRT level of significance
-##' @param em_maxiter maximum number of iterations allowed for EM algorithm
-##'
-##' @seealso \code{\link{simPref}}
-##'
-##' @examples
-##' # set parameters
-##' Predators <- Traps <- 100
-##' PreySpecies <- 2
-##' Times <- 5
-##' g <- matrix(sqrt(2), nrow=Times, ncol=PreySpecies)     # gamma
-##' l <- matrix(seq(0.4,1.8,length.out=5)*sqrt(2), nrow=Times, ncol=PreySpecies) # ct
-##'
-##' # fit model
-##' \dontrun{
-##' fdata <- simPref(PreySpecies, Times, Predators, Traps, l, g, EM=FALSE)
-##' predPref(fdata$eaten, fdata$caught, hypotheses=c('ct', 'cst'))
-##' }
-##' 
-##' @export
-predPref <- function(eaten, caught, hypotheses = c('c', 'Ct'), alpha=0.05, em_maxiter=1000) {
-
[…]

### 2014-10-27 spiders.Rcheck/00_pkg_src/spiders/R/summary.predPref.R

@@ -1,65 +1,0 @@
-##' @title predPref summary
-##'
-##' @description  summary method for predPref objects as returned by the
-##' function \code{predPref}
-##'
-##' @param object predPref object as returned from predPref()
-##' @param ... additional arguments
-##' @param sig.level significance level used in hypothesis test
-##' @export
-summary.predPref <- function(object, ..., sig.level=0.05) {
-    out <- vector('list', 7)
-    names(out) <- c('loglikH0', 'loglikH1', 'p.value',
-                    'estimates', 'df', 'Lambda', 'hypotheses')
-    if ( object$LRT$p.value > sig.level ) {
-        out[['estimates']] <- object$null
-    } else {
-        out[['estimates']] <- object$alt
-    }
-    out[['p.value']] <- object$LRT$p.value
-    out[['df']] <- object$LRT$df
-    out[['Lambda']] <- object$LRT$Lambda
-    out[['loglikH0']] <- object$loglikH0
-    out[['loglikH1']] <- object$loglikH1
-    out[['hypotheses']] <- object$hypotheses
-    class(out) <- 'summary.predPref'
-    out
-}
-
-print.summary.predPref <- function(x) {
-    cat("\nPredator Preferences Model:\n\n")
-    cat("Hypotheses:\n\tH0 =", x$hypotheses[1], "\n\tH1 =", x$hypotheses[2], "\n")
-    cat("Likelihood Ratio Test:\n\t-2*(llH0-llH1) =", x$Lambda, "on", x$df, "degrees of freedom\n")
-    cat("\tp-value =", x$p.value, "\n")
-    cat("Parameter Estimates:\n")
-
-    ## some numbers
-    S <- ncol(x$estimates$gamma); s <- seq_len(S)
-    T <- nrow(x$estimates$gamma); t <- seq_len(T)
-    indices <- unlist(lapply(s,
-                             function(y) sapply(t,
-                                                function(x) paste(x, '_', y, sep=''))))
-    if ( !is.null(x$estimates$c) ) {
-        est <- cbind(c(as.vector(x$estimates$c),
-                       as.vector(x$estimates$gamma)),
-                     as.vector(sqrt(diag(x$estimates$var))))
-        rn <- c(paste('c_', seq_along(x$estimates$c), sep=''),
-                paste('gamma_', indices, sep=''))
-    } else {
-        if ( converged(x$estimates$lambda, x$estimates$gamma) ) {
-            est <- cbind(as.vector(x$estimates$gamma),
[…]

### 2014-10-27 spiders.Rcheck/00_pkg_src/spiders/R/util.R

@@ -1,48 +1,0 @@
-##' sum specified columns by time
-##'
-##' @param data a dataframe
-##' @param vars column variables in data to sum over
-##' @param by extra variables to sum by
-getTimeCounts <- function(data, vars, by) {
-    if (missing(by)) by <- 'time' else by <- c('time', by)
-    ## ddply(data, by, function(dfr, idx) colSums(as.matrix(dfr[,idx])), vars)
-    ddply(data, by, colwise(sum, vars))
-}
-
-##' count number of spiders or traps in each unit of time
-##'
-##' @param data a dataframe
-getUnitCounts <- function(data) {
-    ddply(data, .(time), summarize, total=length(time))$total
-}
-
-##' checks convergence of the parameters for the estimation functions
-##'
-##' @param theta an object, convertible to a matrix, of current parameter values
-##' @param theta_old an object, convertible to a matrix, of old parameter values
-##' @param eps tolerance to determine convergence
-converged <- function(theta, theta_old, eps=1e-5) {
-    isTRUE(all.equal(theta, theta_old, tolerance=eps, check.names=F, check.attr=F))
-}
-
-## sum over species to get a vector of values for each time period
-sumSp <- function(mat) {
-    matrix(rowSums(mat), nrow=nrow(mat))
-}
-
-## sum over times to get a vector of values for each species
-sumT <- function(mat) {
-    matrix(colSums(mat), ncol=ncol(mat))
-}
-
-sumST <- function(mat) {
-    sum(mat)
-}
-
-## colors stolen from http://geography.uoregon.edu/datagraphics/color_scales.htm
-cols <- c('orange1' = "#FFBF80", 'orange2' = "#FF8000",
-          'yellow1' = "#FFFF99", 'yellow2' = "#FFFF33",
-          'green1' = "#B2FF8C", 'green2' = "#33FF00",
-          'blue1' = "#A6EDFF", 'blue2' = "#1AB2FF",
-          'purple1' = "#CCBFFF", 'puple2' = "#664CFF",
-          'red1' = "#FF99BF", 'red2' = "#E61A33")

### 2014-10-27 spiders.Rcheck/00_pkg_src/spiders/tests/testthat/test_predPref.R

@@ -1,155 +1,0 @@
-context('Output predPref.')
-
-## set up some data
-Predators <- c(200, 210, 220, 230, 240)
-Traps <- Predators
-PreySpecies <- 2
-Times <- 5
-TxS <- c(Times, PreySpecies)
-g <- matrix(sqrt(2), nrow=Times, ncol=PreySpecies)
-
-
-## noEM
-test_that('noEM, 1 vs c, null', {
-    l <- matrix(sqrt(2), nrow=Times, ncol=PreySpecies)
-    fdata <- simPref(PreySpecies, Times, Predators, Traps, l, g, EM=F)
-    pref <- predPref(fdata$eaten, fdata$caught, hypotheses=c('1', 'c'), em_maxiter = 50000)
-    expect_is(pref, 'predPref')             # inherits class
-    expect_equal(length(pref$null$c), 0)    # hypotheses correctly chosen
-    expect_equal(length(pref$alt$c), 1)
-    expect_equal(dim(pref$null$gamma), TxS) # dimensions of estimates
-    expect_equal(dim(pref$alt$gamma), TxS)
-    expect_more_than(pref$LRT$p.value, 0.001)     # correct conclusion
-    expect_that(all(diag(pref$null$var)>0), is_true()) # variances (null) are positive
-    expect_that(all(diag(pref$alt$var)>0), is_true()) # variances (alt) are positive
-})                                      
-
-test_that('noEM, c vs c^t, null', {
-    l <- matrix(2*sqrt(2), nrow=Times, ncol=PreySpecies)
-    fdata <- simPref(PreySpecies, Times, Predators, Traps, l, g, EM=F)
-    pref <- predPref(fdata$eaten, fdata$caught, hypotheses=c('c', 'Ct'), em_maxiter = 50000)
-    expect_is(pref, 'predPref')             # inherits class
-    expect_equal(length(pref$null$c), 1)    # hypotheses correctly chosen
-    expect_equal(length(pref$alt$c), Times)
-    expect_equal(dim(pref$null$gamma), TxS) # dimensions of estimates
-    expect_equal(dim(pref$alt$gamma), TxS)
-    expect_more_than(pref$LRT$p.value, 0.001)     # correct conclusion
-    expect_that(all(diag(pref$null$var)>0), is_true()) # variances (null) are positive
-    expect_that(all(diag(pref$alt$var)>0), is_true()) # variances (alt) are positive
-})
-
[…]

### 2014-11-01 paper/data.tex
real data section changes, moar plots, moar words

--- 2014-10-13 paper/data.tex
@@
 
-Interest in the predator genus Schizocosa and the prey orders? Diptera, Collembola, Ensifera, for time periods October $2011$ to March $2013$.  
+We analyze a dataset that was collected to investigate the eating preferences of the wolf spider, genus \textit{Schizocosa}, towards the prey orders \textit{Diptera} and \textit{Collembola}.
+Predators, collected in traps, had their gut-contents analyzed to determine whether or not the spiders ate any \textit{Diptera} or \textit{Collembola} within each time period.
+The left panel of figure~\ref{fig:data} plots the percentage of spiders that had the prey in their guts across time.
+Prey were similarly collected in traps distributed across the Berea College Forest in Madison County, Kentucky, USA.
+These data were collected for each month from October $2011$ to March $2013$.
+On average, $69$ spiders, and $111$ and $297$ \textit{Diptera} and \textit{Collembola}, respectively, were caught in each time period.
+The range of the sample sizes, across all $18$ months is, $11$ to $181$ for caught spiders, from $7$ to $322$ for trapped \textit{Diptera}, and from $101$ to $755$ for trapped \textit{Collembola}.
+The right panel of figure~\ref{fig:data} plots the total number of each order that was caught during each time period.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.5]{data}
+  \caption{For both \textit{Diptera} and \textit{Collembola}, the plots show the proportion of spiders with these orders found in their guts, and the number of these orders trapped in each time period.}
+  \label{fig:data}
+\end{figure}
+
+These data provide an example of our hierarchy of hypotheses.
+First, we test the model $c_{st} = c$ against $c_{st} = c_s$, to determine whether or not the wolf spider has different preferences for the two orders \textit{Diptera}  and \textit{Collembola}.
+With, one degree of freedom, this likelihood ratio test indicates, $p-value < 0.0001$,  that two parameters, one for each order, fits these data better than one parameter for both.
+Similarly, we test whether or not there is a significant effect across time by testing the model $c_{st} = c$ against $c_{st} = c_t$.
+Here, the likelihood ratio test, with $17$ degrees of freedom, implies that the wolf spiders of the Berea College Forest eat these prey orders at different rates across the months of the year, $p-value < 0.0001$.
+In fact, we find that the most parameter rich model, $\lambda_{st} = c_{st} \gamma_{st}$ fits these data best.
+This model estimates $72$ parameters in total; since, in this case, there are two prey of interest and $18$ time periods, it takes $36$ parameters to estimate each $c_{st}$ and $\gamma_{st}$.
+Figure~\ref{fig:cst} plots the point estimates and confidence intervals of $c_{st}$, for both prey across all time periods.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.5]{cst}
+  \caption{One plot for each prey order displays the point estimates and confidence intervals as estimated from the model $c_{st}$.}
+  \label{fig:cst}
+\end{figure}
+
+With point estimates of $c_{st}$ under the model $\lambda_{st} = c_{st} \gamma_{st}$, we can test any number of linear contrasts.
+For instance, consider $c_{1t} = c_{2t}$, for $t \in \{1, \ldots, 18\}$.
+Using a level of significance of $0.05$, and after making a Bonferroni multiple comparisons adjustment, the data do not indicate that the two prey are differently preferred in October, November, and December of $2011$ and for March and July of $2012$.
+
 %%% Local Variables: 

### 2014-11-01 paper/simulations.tex
real data section changes, moar plots, moar words

@@
 
-The seemingly arbitrary clustering of estimates near four seen in figure~\ref{fig:em} is due to two competing factors.
+The seemingly arbitrary clustering of estimates near four seen in figure~\ref{fig:em} is caused by both the shape of the log-likelihood and the convergence tolerance, $\tau$.
 As the proportion of ones in the $Z_{jst}$ observations increases, the observed log-likelihood becomes flatter.
@@
 By decreasing $\tau$, and increasing, as appropriate, the maximum iterations allowed within the EM algorithm, one can gradually increase the magnitude of a maximum likelihood estimate.
+For the tolerance we used in these simulations, $\tau = 10^{-5}$, four just happens to be where the algorithm stops.
 

### 2014-11-06 paper/data.tex
better tracking / setting of seeds and sample sizes

@@
 We analyze a dataset that was collected to investigate the eating preferences of the wolf spider, genus \textit{Schizocosa}, towards the prey orders \textit{Diptera} and \textit{Collembola}.
-Predators, collected in traps, had their gut-contents analyzed to determine whether or not the spiders ate any \textit{Diptera} or \textit{Collembola} within each time period.
+Predators were collected in traps within the deciduous Berea College Forest in Madison County, Kentucky, USA and had their gut-contents analyzed to determine whether or not the spiders ate any \textit{Diptera} or \textit{Collembola} within each time period.
 The left panel of figure~\ref{fig:data} plots the percentage of spiders that had the prey in their guts across time.
-Prey were similarly collected in traps distributed across the Berea College Forest in Madison County, Kentucky, USA.
+Prey were also collected in traps from the same area.
 These data were collected for each month from October $2011$ to March $2013$.
@@
 These data provide an example of our hierarchy of hypotheses.
-First, we test the model $c_{st} = c$ against $c_{st} = c_s$, to determine whether or not the wolf spider has different preferences for the two orders \textit{Diptera}  and \textit{Collembola}.
-With, one degree of freedom, this likelihood ratio test indicates, $p-value < 0.0001$,  that two parameters, one for each order, fits these data better than one parameter for both.
-Similarly, we test whether or not there is a significant effect across time by testing the model $c_{st} = c$ against $c_{st} = c_t$.
+First, tested model $c_{st} = c$ against $c_{st} = c_s$, to determine whether or not the wolf spider has different preferences for the two orders \textit{Diptera}  and \textit{Collembola}.
+With, one degree of freedom, this likelihood ratio test indicated, $p-value < 0.0001$,  that two parameters, one for each order, fits these data better than one parameter for both.
+Similarly, we tested whether or not there was a significant effect across time by testing model $c_{st} = c$ against $c_{st} = c_t$.
 Here, the likelihood ratio test, with $17$ degrees of freedom, implies that the wolf spiders of the Berea College Forest eat these prey orders at different rates across the months of the year, $p-value < 0.0001$.
-In fact, we find that the most parameter rich model, $\lambda_{st} = c_{st} \gamma_{st}$ fits these data best.
-This model estimates $72$ parameters in total; since, in this case, there are two prey of interest and $18$ time periods, it takes $36$ parameters to estimate each $c_{st}$ and $\gamma_{st}$.
+In fact, we find that the most parameter rich model, $\lambda_{st} = c_{st} \gamma_{st}$ fits these data better than is expected by chance, $p-value < 0.0001$.
+Model $c_{st}$ estimates $72$ parameters in total; since, in this case, there are two prey of interest and $18$ time periods, it takes $36$ parameters to estimate each $c_{st}$ and $\gamma_{st}$.
 Figure~\ref{fig:cst} plots the point estimates and confidence intervals of $c_{st}$, for both prey across all time periods.
@@
 With point estimates of $c_{st}$ under the model $\lambda_{st} = c_{st} \gamma_{st}$, we can test any number of linear contrasts.
-For instance, consider $c_{1t} = c_{2t}$, for $t \in \{1, \ldots, 18\}$.
-Using a level of significance of $0.05$, and after making a Bonferroni multiple comparisons adjustment, the data do not indicate that the two prey are differently preferred in October, November, and December of $2011$ and for March and July of $2012$.
+For instance, $c_{1t} = c_{2t}$, for $t \in \{1, \ldots, 18\}$.
+Using a level of significance of $0.05$, and after making a Bonferroni multiple comparisons adjustment, the data can not say that the two prey are differently preferred in October, November, and December of $2011$ and for March and July of $2012$.
 

### 2014-11-06 paper/simulations.tex
better tracking / setting of seeds and sample sizes

--- 2014-10-27 paper/simulations.tex
@@
 Of the hierarchy of hypotheses, we generate data under three models: $c, c_s, c_t$.
-Sample sizes for both prey species and predator gut count observations are randomly chosen from four overlapping levels.
-Let ``small'' sample sizes be randomly sampled numbers in $[20,50]$, ``medium'' encompass $[30,75]$, ``large'' $[50,150]$, and ``huge'' $[100,200]$.
-Hence, we randomly sample for each time period from one of the sample size levels, then generate data for all models.
+Sample sizes for both prey species and predator gut count observations are randomly chosen from four overlapping levels: ``small'' sample sizes are randomly sampled numbers in $[20,50]$, ``medium'' $[30,75]$, ``large'' $[50,150]$, and ``huge'' $[100,200]$.
 This is repeated for each level of sample size.
@@
 All simulations of non-count data use $\tau = 10^{-5}$ as the convergence tolerance.
-A subset of the examples are provided here; the interested reader is referred to the supplementary materials for the complete set of simulation results.
+A subset of the examples are provided here; the interested reader is referred to the supplementary materials for the complete simulation results.
 
@@
 The values of $\lambda_{st}$ are set with respect to each data generating model.
-For the model $c_{st} = c$, where predator preferences don't vary by either time or species, we consider $\lambda_{st} = 2\pi, \forall s,t$.
-Under the model $c_s$, the ratio of rates vary by species only, so we put $\lambda_{1t} = \sqrt{2}$ and $\lambda_{2t} = \pi$.
+For model $c_{st} = c$, where predator preferences don't vary by either time or species, we put $\lambda_{st} = 2\pi, \forall s,t$.
+Under model $c_s$, the ratio of rates vary by species only, so we put $\lambda_{1t} = \sqrt{2}$ and $\lambda_{2t} = \pi$.
 Hence, $c_1 = \sqrt{2}/\pi \approx 0.45$ and $c_2 = 1$.
@@
   \includegraphics[scale=0.5]{nonem}
-  \caption{Density plots of all $500$ estimates of fitting the true model to the data generated from models $c_s,c$ are shown with sample sizes small and medium, repsepectively.}
+  \caption{Density plots of all $500$ estimates of fitting the true model to the data generated from models $c_s,c$ are shown with sample sizes small and medium, respectively.}
   \label{fig:nonem}
@@
 Figure~\ref{fig:nonem} shows density plots of the estimates of $c_s, c$ when fitting the true model to the fully observed count data generated under models $c_s$ and $c$.
-The plots provide evaluations of parameter bias under each scenario.
+The plots provide evaluations of parameter estimates under each scenario.
 In the first display, the parameters $c_1 \approx 0.45$ and $c_2 = 1$ are on average, across all $500$ simulations, estimated as $\hat{c}_1 = 0.45$ and $\hat{c}_2 = 1.00$, with standard errors of $\text{se}(\hat{c}_1) = 0.03$ and $\text{se}(\hat{c}_2) = 0.06$.
-The second display provides a similar conclusion.
+The second display provides results for model $c_{st} = c$.
 Averaging across all $500$ simulations, the parameter $c=2$ is estimated as $\hat{c} = 2.00$.
@@
   \includegraphics[scale=0.5]{bp}
-  \caption{Shown are all $500$ estimates, centered at the true parameter values, from fitting the true model to the data generated from models $c_s,c$ with sample sizes small and medium, repsepectively.}
+  \caption{Shown are all $500$ estimates, centered at the true parameter values, from fitting the true model to the data generated from models $c_s,c$ with sample sizes small and medium, respectively.}
   \label{fig:bp}
@@
 
-When we generate data with unobserved counts, the story is slightly different.
-As noted above under certain circumstatnces our EM model accurately estimates the parameters of interest, and at different times can greatly over-estimate the parameters.
+We next generated data with unobserved counts.
+As noted above under certain circumstances our unobserved counts model accurately estimates the parameters of interest, and at different times can greatly over-estimate the parameters.
 To investigate this issue further, we consider the same scenarios mentioned above, but now we reduce all of our count data down to binary observations.
-For each scenario's generated data, we fit our EM algorithm as if we knew a priori the true underlying model that generated the observed data.
+For each scenario, we fit our unobserved counts model as if we knew the true underlying model that generated the observed data.
 
-Figure~\ref{fig:em} contains density plots of fitting the true model, for all $500$ replications, of the data generating models $c_s,c_t$ with small and huge sample sizes, respectively.
-When data are generated under the model $c_s$ and the true model fit to the non-count data, we find even for the small sample size that point estimates are reasonable.
-When parameter values are of sufficient size to make zeros in the simulated data less common, the estimates from fitting the correct model to the generated data are occassionally over-estimated.
-This effect is easily seen in figure~\ref{fig:em} for larger values of $c_t$ despite the increased sample size, but is also seen, less dramatically, in the density plot for the $c_s$ generated data.
+Figure~\ref{fig:em} contains density plots for all $500$ replications of the data generating models $c_s,c_t$ with small and huge sample sizes, respectively.
+When data are generated under the model $c_s$ and the true model fit to the non-count data, we find even for the small sample size that point estimates are nearly unbiased.
+When parameter values are of sufficient size to make zeros in the simulated data less common, the estimates from fitting the correct model to the generated data are occasionally over-estimated.
+This effect is easily seen in the bottom panel of figure~\ref{fig:em} for larger values of $c_t$ despite the increased sample size, but is also seen, less dramatically, in the density plot for the $c_s$ generated data.
 
-The seemingly arbitrary clustering of estimates near four seen in figure~\ref{fig:em} is due to two competing factors.
+{\color{red} Need to figure this out on computer then reconsider this paragraph.} The cluster of estimates for $c_5$ between $3.5$ and $4.0$ in the bottom panel of figure~\ref{fig:em} comes from datasets in which $X_{js5}>0$ so that $Z_{js5}>0$ for all $j$ and $s$.
+The probability that this happens is ?
+and we observed this in ?
+of the 500 data sets.
+As mentioned above, the estimate of $c_5$ is infinite in this case.
+However, the EM algorithm will always provide a finite estimate for all parameters when it terminates.
+In this case, we set $\tau=10^{-5}$ and it just happened that this caused the algorithm to terminate with $\hat c_5$ between 3.5 and 4.0 in all cases.
+To confirm that this is due to the arbitrary choice of $\tau$, we repeated the algorithm with smaller values of $\tau$ for several data sets.
+As expected, $\hat c_5$ increased without bound as we refit the model with smaller and smaller values of $\tau$.
+
+% The seemingly arbitrary clustering of estimates near four seen in figure~\ref{fig:em} is caused by both the shape of the log-likelihood and the convergence tolerance, $\tau$.
 As the proportion of ones in the $Z_{jst}$ observations increases, the observed log-likelihood becomes flatter.
@@
   \includegraphics[scale=0.5]{em}
-  \caption{Density plots of all $500$ estimates of fitting the true model to the data generated from models $c_s,c_t$, when counts are not observed, are shown with sample sizes small and huge, repsepectively.}
+  \caption{Density plots of all $500$ estimates of fitting the true model to the data generated from models $c_s,c_t$, when counts are not observed, are shown with sample sizes small and huge, respectively.}
   \label{fig:em}
@@
 
-The over-estimation of parameters, a symptom of the loss of information due to the unobserved counts, can also be seen with box plots of the $500$ point estimates, centered at their respective true parameter values.
+{\color{red} Also address this paragraph after figuring out the computer issue.}  The over-estimation of parameters, a symptom of the loss of information due to the unobserved counts, can also be seen with box plots of the $500$ point estimates, centered at their respective true parameter values.
 Figure~\ref{fig:em_bp} contains box plots of the same scenarios in Figure~\ref{fig:em}, but with the small and huge sample sizes.
@@
   \includegraphics[scale=0.5]{em_bp}
-  \caption{Shown are all $500$ estimates, centered at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$, when counts are not observed, with sample sizes small and huge, repsepectively.}
+  \caption{Shown are all $500$ estimates, centered at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$, when counts are not observed, with sample sizes small and huge, respectively.}
   \label{fig:em_bp}

### 2014-11-06 R/estimateC.R

@@ -21,11 +21,10 @@ estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
     if (EM) {
         ## initialize some values
         em_iter <- 1
-        cHat <- cHat_old <- runif(1)        
-
         init <- est1(Xdst, Ydst, J, I, FALSE, em_maxiter, BALANCED)
         gammaHat <- gammaHat_old <- init$gamma 
         lambda <- elambda <- init$lambda
+        cHat <- cHat_old <- as.numeric(sumST(Xdst) / sumT(J*sumSp(gammaHat)))
 
         ## iterate EM
         while ( TRUE ) {
@@ -122,8 +121,9 @@ estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
             }
 
             ## initialize some values
-            cHat <- cHat_old <- runif(1)
-            gammaHat <- gammaHat_old <- XYdst / (J*cHat + I)
+            ## cHat <- cHat_old <- 0.5 # runif(1)
+            ## gammaHat <- gammaHat_old <- XYdst / (J*cHat + I)
+            gammaHat <- gammaHat_old <- XYdst / (J + I)
 
             ## iteratively update; relies on concavity of log-lik
             while ( TRUE ) {

### 2014-11-06 R/estimateCs.R

@@ -21,9 +21,10 @@ estCs <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
     if (EM) {
         ## initialize some values
         em_iter <- 1
-        cHat <- cHat_old <- runif(S)
         init <- est1(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED)
-        gammaHat <- gammaHat_old <- init$gamma 
+        gammaHat <- gammaHat_old <- init$gamma
+        cHat <- cHat_old <- sumT(Xdst) / sumT(J*gammaHat) # rep(0.5, S) # runif(S)
+
 
         ## iterate EM
         while ( TRUE ) {
@@ -94,8 +95,9 @@ estCs <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
         if ( length(I) != T ) stop("I indexed oddly says est0.")
 
         ## initialize some values
-        cHat <- cHat_old <- runif(S)
-        gammaHat <- gammaHat_old <- sapply(s, function(j) XYdst[,j] / (J*cHat[j] + I))
+        ## cHat <- cHat_old <- rep(0.5, S) # runif(S)
+        ## gammaHat <- gammaHat_old <- sapply(s, function(j) XYdst[,j] / (J*cHat[j] + I))
+        gammaHat <- gammaHat_old <- sapply(s, function(j) XYdst[,j] / (J + I))
 
         ## iteratively update; relies on concavity of log-lik
         while ( TRUE ) {

### 2014-11-06 R/estimateCst.R

@@ -21,9 +21,10 @@ estCst <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
     if (EM) {
         ## initialize some values
         em_iter <- 1
-        cHat <- cHat_old <- matrix(runif(ST), nrow=T)
+        ## cHat <- cHat_old <- matrix(0.5, ncol=S, nrow=T) # matrix(runif(ST), nrow=T)
         init <- est1(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED)
-        gammaHat <- gammaHat_old <- init$gamma 
+        gammaHat <- gammaHat_old <- init$gamma
+        cHat <- cHat_old <- Xdst/(J*gammaHat)
         
         ## iterate EM
         while ( TRUE ) {
@@ -93,8 +94,9 @@ estCst <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
         if ( length(I) != T ) stop("I indexed oddly says estGen")
 
         ## initialize some values
-        cHat <- cHat_old <- matrix(runif(ST), nrow=T)
-        gammaHat <- gammaHat_old <- XYdst / (J*cHat + I)
+        ## cHat <- cHat_old <- matrix(0.5, ncol=S, nrow=T) # matrix(runif(ST), nrow=T)
+        ## gammaHat <- gammaHat_old <- XYdst / (J*cHat + I)
+        gammaHat <- gammaHat_old <- XYdst / (J + I)
 
         ## iteratively update; relies on concavity of log-lik
         while ( TRUE ) {

### 2014-11-06 R/predpref.R

@@ -59,8 +59,12 @@ predPref <- function(eaten, caught, hypotheses = c('c', 'Ct'), alpha=0.05, em_ma
     yNames <- colnames(caught)
 
     ## traps left out for different numbers of days?
-    if ( !('adj' %in% xNames) ) eaten$adj <- 1
-    if ( !('adj' %in% yNames) ) caught$adj <- 1
+    if ( !('adj' %in% xNames) ) {
+        eaten$adj <- 1
+    }
+    if ( !('adj' %in% yNames) ) {
+        caught$adj <- 1
+    }
 
     ## prey names
     extraVars <- c('time', 'adj')
@@ -70,7 +74,9 @@ predPref <- function(eaten, caught, hypotheses = c('c', 'Ct'), alpha=0.05, em_ma
     X <- eaten[,preyNames]
     Y <- caught[,preyNames]
 
-    if ( any(X < 0) || any(Y < 0) ) stop("Count data can not be less than 0.")
+    if ( any(X < 0) || any(Y < 0) ) {
+        stop("Count data can not be less than 0.")
+    }
 
     ## predators (J), traps (I)
     J <- getTimeCounts(eaten, 'adj')[,2]
@@ -87,7 +93,14 @@ predPref <- function(eaten, caught, hypotheses = c('c', 'Ct'), alpha=0.05, em_ma
     BAL <- length(unique(J)) == 1 && length(unique(I)) == 1
 
     ## errors with time points
-    if ( nrow(Xdst) != nrow(Ydst) ) stop("Differing number of time points in eaten/caught data.")
+    if ( nrow(Xdst) != nrow(Ydst) ) {
+        stop("Differing number of time points in eaten/caught data.")
+    }
+
+    ## check hypotheses
+    if ( is.null(hypotheses) ) {
+        stop('Need to speficy hypotheses.')
+    }
 
     calcs <- calcHypotheses(hyp = hypotheses,
                       Xdst = Xdst, Ydst = Ydst, J=J, I=I,

### 2014-11-06 R/test.R

@@ -33,29 +33,26 @@
 ##' }
 ##'
 ##' @export
-testPref <- function(J, I, lambda, gamma, M=100, hyp=c('C', 'Cst'), EM = FALSE, em_maxiter = 100, storeSeed = FALSE) {
+testPref <- function(J, I, lambda, gamma, M=100, hyp=c('C', 'Cst'), EM = FALSE, em_maxiter=100, storeSeed=FALSE) {
 
     ## initialize output structure
     out <- vector('list', 2)
     names(out) <- c('null', 'alt')
     needSetOutput <- TRUE             # need initialize storage within output?
-    if ( storeSeed ) {
-        rseed <- vector('list', M)
-        invisible(runif(1))             # ensure random seed exists
-    }
 
     ## some numbers 
     S <- ncol(lambda)
     T <- nrow(lambda)
+    if ( storeSeed ) {
+        rseed <- sample(.Machine$integer.max/2, M)
+    } else {
+        rseed <- NULL
+    }
     
     for ( m in seq_len(M) ) {
-
-        if ( storeSeed ) {
-            rseed[[m]] <- .Random.seed
-        }
         
         ## simulate data and fit model
-        fdata <- simPref(S, T, J, I, lambda, gamma, EM=EM)
+        fdata <- simPref(S, T, J, I, lambda, gamma, EM=EM, seed=rseed[m])
         prefs <- predPref(fdata$eaten, fdata$caught, hypotheses = hyp, em_maxiter = em_maxiter)
 
         ## initialize storage within output structure
@@ -128,9 +125,9 @@ testPref <- function(J, I, lambda, gamma, M=100, hyp=c('C', 'Cst'), EM = FALSE, 
     class(out) <- 'testPref'
     attr(out, 'ST') <- S*T
     if ( storeSeed ) {
-        attr(out, "seed") <- rseed        
+        attr(out, "seeds") <- rseed
     }
-    
+    attr(out, 'sample_size') <- list(J = J, I = I)
     out
 }
 

### 2014-11-15 paper/simulations.tex
all simulations performed with BatchExperiments; moar words about infinite bias; better idea of when exactly infinite bias will occur

@@
 \label{sec:sim}
-
-% use language like: data generating model and fitted model, not hypotheses
 
@@
 A subset of the examples are provided here; the interested reader is referred to the supplementary materials for the complete simulation results.
+For the simulations we used the \texttt{R} \cite{Core-Team:2014} package \texttt{BatchExperiments} by \cite{Bischl:2014}.
 
@@
 
-% more words about how these numbers were estimated; estimtes from across all simulations
-% simulations show nearly unbiased
-% drop decimal places; we don't have precision to state further. 2
-
 Figure~\ref{fig:nonem} shows density plots of the estimates of $c_s, c$ when fitting the true model to the fully observed count data generated under models $c_s$ and $c$.
@@
 Averaging across all $500$ simulations, the parameter $c=2$ is estimated as $\hat{c} = 2.00$.
-This is further seen in figure~\ref{fig:bp}, where box plots of the parameter estimates, centered at true parameter values, of the correct model fit to data generated from both $c_s$ and $c_t$ show empirically almost no bias.
+This is further seen in figure~\ref{fig:bp}, where box plots of the parameter estimates, centered at true parameter values, of the correct model fit to data generated from both $c_s$ and $c_t$ show empirically very little bias.
 
@@
   \includegraphics[scale=0.5]{bp}
-  \caption{Shown are all $500$ estimates, centered at the true parameter values, from fitting the true model to the data generated from models $c_s,c$ with sample sizes small and medium, respectively.}
+  \caption{Shown are all $500$ estimates, centered at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$ with sample sizes small and medium, respectively.}
   \label{fig:bp}
@@
 We next generated data with unobserved counts.
-As noted above under certain circumstances our unobserved counts model accurately estimates the parameters of interest, and at different times can greatly over-estimate the parameters.
-To investigate this issue further, we consider the same scenarios mentioned above, but now we reduce all of our count data down to binary observations.
-For each scenario, we fit our unobserved counts model as if we knew the true underlying model that generated the observed data.
+As noted above under certain circumstances our unobserved counts model accurately estimates the parameters of interest, and at other times can infinitely over-estimate parameters.
+To investigate this issue further, we consider the same scenarios mentioned above, but now we reduce all of the count data down to binary observations.
+For each scenario, we fit the unobserved counts model as if we knew the true underlying model that generated the observed data.
 
 Figure~\ref{fig:em} contains density plots for all $500$ replications of the data generating models $c_s,c_t$ with small and huge sample sizes, respectively.
-When data are generated under the model $c_s$ and the true model fit to the non-count data, we find even for the small sample size that point estimates are nearly unbiased.
+When data are generated under the model $c_s$ and the true model fit to the non-count data, we find even for the small sample size that point estimates are only very slightly biased.
 When parameter values are of sufficient size to make zeros in the simulated data less common, the estimates from fitting the correct model to the generated data are occasionally over-estimated.
@@
 
-{\color{red} Need to figure this out on computer then reconsider this paragraph.} The cluster of estimates for $c_5$ between $3.5$ and $4.0$ in the bottom panel of figure~\ref{fig:em} comes from datasets in which $X_{js5}>0$ so that $Z_{js5}>0$ for all $j$ and $s$.
-The probability that this happens is ?
-and we observed this in ?
-of the 500 data sets.
+The cluster of estimates for $c_5$ between $3.5$ and $4.0$ in the bottom panel of figure~\ref{fig:em} comes from datasets in which $X_{js5}>0$ so that $Z_{js5}=1$ for all $j,s$.
+For the data shown in figure~\ref{fig:em}, this happened $73$ times out of the $500$ replicated datasets.
 As mentioned above, the estimate of $c_5$ is infinite in this case.
 However, the EM algorithm will always provide a finite estimate for all parameters when it terminates.
-In this case, we set $\tau=10^{-5}$ and it just happened that this caused the algorithm to terminate with $\hat c_5$ between 3.5 and 4.0 in all cases.
-To confirm that this is due to the arbitrary choice of $\tau$, we repeated the algorithm with smaller values of $\tau$ for several data sets.
-As expected, $\hat c_5$ increased without bound as we refit the model with smaller and smaller values of $\tau$.
-
-% The seemingly arbitrary clustering of estimates near four seen in figure~\ref{fig:em} is caused by both the shape of the log-likelihood and the convergence tolerance, $\tau$.
-As the proportion of ones in the $Z_{jst}$ observations increases, the observed log-likelihood becomes flatter.
-Thus, any finite convergence tolerance will cause the algorithm to terminate eventually.
-By decreasing $\tau$, and increasing, as appropriate, the maximum iterations allowed within the EM algorithm, one can gradually increase the magnitude of a maximum likelihood estimate.
-For the tolerance we used in these simulations, $\tau = 10^{-5}$, four just happens to be where the algorithm stops.
-
-% The plots of $\gamma_{st}$ under the EM algorithm are not given as we do not consider missing data in the estimation of these parameters.
-
-% more of which model are we fitting, which model is generating the data
+In this case, we set $\tau=10^{-5}$ and it just happened that this caused the algorithm to terminate with $\hat c_5$ between $3.5$ and $4.0$ in all cases.
+To confirm that this is due to the arbitrary choice of $\tau$, we repeated the algorithm with smaller values of $\tau$ for several datasets.
+As expected, $\hat c_5$ increased without bound as we refit the model with increasingly small values of $\tau$.
 
@@
 
-{\color{red} Also address this paragraph after figuring out the computer issue.}  The over-estimation of parameters, a symptom of the loss of information due to the unobserved counts, can also be seen with box plots of the $500$ point estimates, centered at their respective true parameter values.
-Figure~\ref{fig:em_bp} contains box plots of the same scenarios in Figure~\ref{fig:em}, but with the small and huge sample sizes.
-It is clear that as the parameters values increase and zeros in the observed data $Z_{jst}$ become less prevalent, over-estimation occurs more frequently.
-This over-estimation, when it occurs, will generally increase the magnitude of the bias.
-
-%Table~\ref{tab:bias} displays the max absolute bias across all indices $s,t$ of $\lambda_{st}$ for each simulation scenario.
+The over-estimation of parameters, a symptom of the loss of information due to the unobserved counts, can also be seen with box plots of the $500$ point estimates centered at their respective true parameter values.
+Figure~\ref{fig:em_bp} contains box plots of the same scenarios in Figure~\ref{fig:em}.
+For the $73$ cases in which $Z_{js5} = 1$ for all $j,s$ under model $c_t$ with the huge sample size, the bias is infinite since parameter estimates will, theoretically, be infinite.
+The finite bias shown in these plots is due to the finite estimates provided by the termination of the EM algorithm.
+Thus, conditional on a mixture of $0$s and $1$s in the data the correspoding estimators are unbiased, but when no $0$s exist in the data the theoretical bias is infinite.
 
@@
 
-% \begin{table}
-%   \centering
-%   \begin{tabular}{ll|rrrr}
-%     & & small & medium & large & huge \\ 
-%     \hline
-%     count & $c_t$ & 0.040 & 0.013 & 0.020 & 0.011 \\ 
-%     & $c_s$ & 0.028 & 0.015 & 0.010 & 0.005 \\ 
-%     & $c$ & 0.030 & 0.016 & 0.017 & 0.012 \\ 
-%     non-count & $c_t$ & 4.151 & 3.155 & 0.941 & 1.423 \\ 
-%     & $c_s$ & 0.138 & 0.096 & 0.044 & 0.032 \\ 
-%     & $c$ & 2.924 & 2.078 & 0.669 & 0.547 \\ 
-%   \end{tabular}
-%   \caption{Max absolute bias of all $\lambda_{st}$ for each simulation scenario.}
-%   \label{tab:bias}
-% \end{table}
-
 %%% Local Variables: 

### 2014-11-29 paper/data.tex
edit and typo check

@@
 
-We analyze a dataset that was collected to investigate the eating preferences of the wolf spider, genus \textit{Schizocosa}, towards the prey orders \textit{Diptera} and \textit{Collembola}.
+We analyzed a dataset that was collected to investigate the eating preferences of the wolf spider, genus \textit{Schizocosa}, towards the prey orders \textit{Diptera} and \textit{Collembola}.
 Predators were collected in traps within the deciduous Berea College Forest in Madison County, Kentucky, USA and had their gut-contents analyzed to determine whether or not the spiders ate any \textit{Diptera} or \textit{Collembola} within each time period.
@@
 These data provide an example of our hierarchy of hypotheses.
-First, tested model $c_{st} = c$ against $c_{st} = c_s$, to determine whether or not the wolf spider has different preferences for the two orders \textit{Diptera}  and \textit{Collembola}.
+First, we tested model $c_{st} = c$ against $c_{st} = c_s$, to determine whether or not the wolf spider has different preferences for the two orders \textit{Diptera}  and \textit{Collembola}.
 With, one degree of freedom, this likelihood ratio test indicated, $p-value < 0.0001$,  that two parameters, one for each order, fits these data better than one parameter for both.
@@
 Model $c_{st}$ estimates $72$ parameters in total; since, in this case, there are two prey of interest and $18$ time periods, it takes $36$ parameters to estimate each $c_{st}$ and $\gamma_{st}$.
-Figure~\ref{fig:cst} plots the point estimates and confidence intervals of $c_{st}$, for both prey across all time periods.
+Figure~\ref{fig:cst} plots the point estimates and $95\%$ confidence intervals of $c_{st}$, for both prey across all time periods.
 
@@
   \includegraphics[scale=0.5]{cst}
-  \caption{One plot for each prey order displays the point estimates and confidence intervals as estimated from the model $c_{st}$.}
+  \caption{One plot for each prey order displays the point estimates and $95\%$ confidence intervals as estimated from the model $c_{st}$.}
   \label{fig:cst}

### 2014-11-29 paper/discussion.tex
edit and typo check

@@
 
-The model developed here allows practicioners to determine predators' preferences by testing simultaneously across an array of multiple prey species and time.
-This is acheived via a simple, but statistically powerful, likelihood ratio test.
+The model developed here allows practitioners to determine predators' preferences by testing simultaneously across an array of multiple prey species and time.
+This is achieved via a simple, but statistically powerful, likelihood ratio test.
 Further testing of the ratio of rates for which predators eat to encounter prey species allows researchers to make specific conclusions about predators' preferences.
-For instance, rates acros time can be estimated to make statements about seasonal effects on a predator's eating habits, or relative rates across species groups allows for statements about the preferences for different species.
+For instance, rates across time can be estimated to make statements about seasonal effects on a predator's eating habits, or relative rates across species groups allows for statements about the relative preferences for different species.
 
 When counts of predators' gut contents are not fully observed, and instead only a binary response indicating the existence of the prey species in the gut is observed, we are able to treat the counts as missing data.
-By modelling all of the observed data, both the binary responses and the number of prey species caught, and the missing count data, we are able to use the EM algorithm to extract as much information from the data as possible.
+By modeling all of the observed data, both the binary responses and the number of prey species caught, and the missing count data, we are able to use the EM algorithm to extract as much information from the data as possible.
 Though this is nice in theory, in practice the success of this modification to our original model is limited by the magnitude of the unknown parameters $\lambda_{st}$.
@@
 Further developments of our model could be beneficial.
-Taking into account other environmental variables that might effect a predators' eating habits, such as rain or temperature, say, might be advantageous.
+Taking into account other environmental variables that might effect a predator's eating habits, such as rain or temperature, say, might be advantageous.
 

### 2014-11-29 paper/introduction.tex
edit and typo check

--- 2014-08-27 paper/introduction.tex
@@
-\section*{Introduction}
+\section{Introduction}
 \label{sec:intro}
+
+Modeling a predator's food selection was given much attention for a short period of time; see \citet{Strauss:1979} and the references within.
+Many, if not all, of the indices developed, though intuitive, focused on a snapshot in time and on only one prey species, and rarely obtained practical asymptotic properties.
+This restricts the interested practitioner to the most computationally manageable of techniques, and has completely ignored simultaneous testing across an array of both species and time.
+
+The model presented here maintains tractable asymptotic properties while being general enough to take into account multiple species and time points.
+By modeling both time and any number of prey species, we are able to see a more detailed analysis of the predator's eating preferences.
+The simplicity of the model allows us to consider predators for which exact tallies of the number of each prey species eaten within any given time period is not observed.
+Instead, we rely on the researcher being able to DNA sequence the contents of the predator's gut and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.
+Despite the data's lost information, under certain situations our model is able to accurately estimate the parameters of interest.
+
 %%% Local Variables: 

### 2014-11-29 paper/methods.tex
create option to store random seed in testPref
edit and typo check

@@
 
-The use of Poisson distributions make the following implicit assumptions: $1)$ traps independently catch the prey species of interest, $2)$ predators eat prey species independently, $3)$ predators eat indepedent of each other.
+The use of Poisson distributions make the following implicit assumptions: $1)$ traps independently catch the prey species of interest, $2)$ predators eat prey species independently, $3)$ predators eat independent of each other.
 
-We denote the number of predators and the numer of prey species caught, in each time period $t$, by $J_t$ and $I_t$, respectively.
+We denote the number of predators and the number of prey species caught, in each time period $t$, by $J_t$ and $I_t$, respectively.
 Let $X_{jst} \iid \mathcal{P}(\lambda_{st})$ represent the number of prey species $s$ that predator $j$ ate during occurrence $t$, where $j \in \{1, \ldots, J_t\}$.
@@
 \begin{enumerate}
-\item $c_{st} = 1, s = 1, \ldots, S; t = 1, \ldots, T$
-\item $c_{st} = c, s = 1, \ldots, S; t = 1, \ldots, T$
-\item $c_{st} = c_s, s = 1, \ldots, S$
-\item $c_{st} = c_t, t = 1, \ldots, T$
-\item $c_{st} = c_{st}, s = 1, \ldots, S; t = 1, \ldots, T$
+\item $c_{st} = 1$
+\item $c_{st} = c$
+\item $c_{st} = c_s$
+\item $c_{st} = c_t$
+\item $c_{st} = c_{st}$
 \end{enumerate}
@@
 The first hypothesis states that predators and traps sample all prey species at the same rate.
-One imagines this is the case if the predator simply eats that which comes within its reach, thus suggesting a diet indifferene.
+One imagines this is the case if the predator simply eats that which comes within its reach, thus suggesting a diet indifference.
 The second says that predators sample prey proportionally across all time periods.
@@
 \subsection{Fully Observed Count Data}
+\label{sec:count}
 
@@
 \subsection{Unobserved Counts}
+\label{sec:noncount}
 
@@
 
-Many times it may not be possible to count the number of individuals of each prey species that are in a predators gut.
+Many times it may not be possible to count the number of individuals of each prey species that are in a predator's gut.
 Instead, it may only be possible to detect whether or not a predator consumed the prey species during a given period.
-This is the case in the study of ?
-spiders which relied on DNA sequencing to identify the prey species in a spider's gut contents.
-Despite this, we can still make inference about the predators preferences for the different prey species by using the EM algorithm to compute maximum likelihood estimates.
+In this case we can still make inference about the predators' preferences for the different prey species by using the EM algorithm to compute maximum likelihood estimates.
 
@@
 \[
-l_{comp}(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log f_{X,Y,Z}(\boldsymbol x, \boldsymbol y, \boldsymbol z|\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \sum_{s=1}^{S} \sum_{t=1}^T [\sum_{j=1}^{J_t} \log f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda}) + \sum_{i=1}^{I_t}\log f_Y(y_{jst}|\boldsymbol{\gamma}).
+l_{comp}(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log f_{X,Y,Z}(\boldsymbol x, \boldsymbol y, \boldsymbol z|\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \sum_{s=1}^{S} \sum_{t=1}^T \left[ \sum_{j=1}^{J_t} \log f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda}) + \sum_{i=1}^{I_t}\log f_Y(y_{jst}|\boldsymbol{\gamma}) \right].
 \]
 
-The density of $Y_{jst}$ is exactly as given above and so we focus on deriving the joint density of $X_{jst}$ and $Z_{jst}$.
-With the distribution of $Z_{jst}$ given above, we can compute $f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda})$ by noting that $X_{jst}=0$ with probability 1 if $Z_{jst}=0$ and that $X_{jst}|Z_{jst}=0$ has a truncated Poisson distribution with density
-\begin{equation*}
+The density of $Y_{jst}$ is exactly as in section~\ref{sec:count} and so we focus on deriving the joint density of $X_{jst}$ and $Z_{jst}$.
+With the distribution of $Z_{jst}$ given above, we can compute $f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda})$ by noting that $X_{jst}=0$ with probability 1 if $Z_{jst}=0$, and that $[X_{jst}|Z_{jst}=0]$ has a truncated Poisson distribution with density
+\[
   f_{X|Y,Z,\boldsymbol{\lambda},\boldsymbol{\gamma}}(x_{jst}|z_{jst}) =
-  \frac{\exp{\{-\lambda_{st}\}} \lambda_{st}^{x_{jst}}}{(1 - \exp{\{-\lambda_{st}\}}) x_{jst}!}1(x_{jst} > 0) \quad \text{ where } 
-\end{equation*}
+  \frac{\exp{\{-\lambda_{st}\}} \lambda_{st}^{x_{jst}}}{(1 - \exp{\{-\lambda_{st}\}}) x_{jst}!}1(x_{jst} > 0)
+\]
 and expected value
@@
     \begin{array}{lr}
-      \exp{\{ -\lambda_{st} \}}, & x_{jst}=0 \mbox{ and } Z_{jst} = 0 \\
-      \frac{\exp{\{-\lambda_{st} \}} \lambda_{st}^{x_{jst}}}{x_{jst}!}, & x_{jst} > 0 \mbox{ and } Z_{jst} = 1\\
+      \exp{\{ -\lambda_{st} \}}, & x_{jst}=0 \mbox{ and } z_{jst} = 0 \\
+      \frac{\exp{\{-\lambda_{st} \}} \lambda_{st}^{x_{jst}}}{x_{jst}!}, & x_{jst} > 0 \mbox{ and } z_{jst} = 1 \\
       0 & \mbox{otherwise}
@@
 \]
-These steps are then alternated until a convergence criterion monitoring subsequent differences in the parameter estimates/likelihood values is met.
+These steps are then alternated until a convergence criterion monitoring subsequent differences in the parameter estimates/likelihood is met.
 
@@
   & \propto \sum_{s,t,j} \left( - \lambda_{st} 
-    + z_{jst} \E X_{jst} \log{\lambda_{st}}  \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right) \\
-  & \propto \sum_{s,t} \left( -J_t \lambda_{st} + z_{\cdot st} \E (X_{jst}|\lambda_{st}^{(k)},\gamma_{st}^{(k)}) \log{\lambda_{st}} \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right).
+    + z_{jst} \log{\lambda_{st}} \E X_{jst} \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right) \\
+  & \propto \sum_{s,t} \left( -J_t \lambda_{st} + z_{\cdot st} \log{\lambda_{st}} \E (X_{jst}|\lambda_{st}^{(k)},\gamma_{st}^{(k)}) \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right).
 \end{split}
@@
 On the other hand, if the predator consistently eats a given prey species, few to no zeros will show up in the observed data and $\E Z_{jst}$ is estimated to be nearly $1$.
-The loss of information is best seen by attempting to solve for $\lambda_{st}$ in the equation $\E Z_{jst} = 1 = 1 - \exp\{-\lambda_{st}\}$.
-As the proportion of ones in the observed data increase, we expect $\lambda_{st}$ to grow exponentially large.
+The loss of information is best seen by attempting to solve for $\lambda_{st}$ in the equation $1 = \E Z_{jst} 1 - \exp\{-\lambda_{st}\}$.
+As the proportion of ones in the observed data increases, we expect $\lambda_{st}$ to grow exponentially large.
 When no zeros are present in the data, so that where only ones are observed, the likelihood can be made arbitrarily large by sending the parameter off to infinity.
@@
 
-The likelihood ratio test (LRT) statistic is
+The likelihood ratio test statistic is
 
@@
 
-\subsection{Testing $c_{st}$}
+\subsection{Linear Transformations of $c_{st}$}
 
@@
 For instance, let $a$ be a vector of the same dimension of $\hat{\mathbf{c}}$.
-Then $a^t\hat{\mathbf{c}}$ is asymptotically distrbuted as $\mathcal{N}(a^t\mathbf{c}, a^t\Sigma a)$, where $\Sigma$ is the covariance matrix of the asymptotic distrbution of $\hat{\mathbf{c}}$.
+Then $a^t\hat{\mathbf{c}}$ is asymptotically distributed as $\mathcal{N}(a^t\mathbf{c}, a^t\Sigma a)$, where $\Sigma$ is the covariance matrix of the asymptotic distribution of $\hat{\mathbf{c}}$.
 
@@
 Confidence intervals of any size are similarly, readily obtained.
+Of course, one could also use a $t$ distribution as a small sample size correction.
 

### 2014-11-29 paper/simulations.tex
edit and typo check

@@
 However, the EM algorithm will always provide a finite estimate for all parameters when it terminates.
-In this case, we set $\tau=10^{-5}$ and it just happened that this caused the algorithm to terminate with $\hat c_5$ between $3.5$ and $4.0$ in all cases.
+In this case, we set $\tau=10^{-5}$ and it just happened that this caused the algorithm to terminate with $\hat c_5$ between $3.5$ and $4.0$.
 To confirm that this is due to the arbitrary choice of $\tau$, we repeated the algorithm with smaller values of $\tau$ for several datasets.
@@
 The finite bias shown in these plots is due to the finite estimates provided by the termination of the EM algorithm.
-Thus, conditional on a mixture of $0$s and $1$s in the data the correspoding estimators are unbiased, but when no $0$s exist in the data the theoretical bias is infinite.
+Thus, conditional on a mixture of $0$s and $1$s in the data the corresponding estimators are unbiased, but when no $0$s exist in the data the theoretical bias is infinite.
 

### 2014-11-29 tests/testthat/test_predPref.R

@@ -7,6 +7,7 @@ PreySpecies <- 2
 Times <- 5
 TxS <- c(Times, PreySpecies)
 g <- matrix(sqrt(2), nrow=Times, ncol=PreySpecies)
+set.seed(1)
 
 
 ## noEM

### 2014-12-02 paper/introduction.tex
add references to intro, and connect past indices with our model

@@
 
-Modeling a predator's food selection was given much attention for a short period of time; see \citet{Strauss:1979} and the references within.
-Many, if not all, of the indices developed, though intuitive, focused on a snapshot in time and on only one prey species, and rarely obtained practical asymptotic properties.
-This restricts the interested practitioner to the most computationally manageable of techniques, and has completely ignored simultaneous testing across an array of both species and time.
+Modeling a predator's food preferences, or selectivity, is a relatively old topic \citep{Ivlev:1964,Jacobs:1974,Chesson:1978,Strauss:1979,Vanderploeg:1979,Chesson:1983}, and yet continues to generate numerous publications.
+Many of the indices developed, though intuitive, focused on a snapshot in time, few allow more than one prey species, and practical asymptotic statistical properties are even more rare.
+This restricts the interested practitioner to the most computationally manageable of techniques, and has completely ignored simultaneous testing across an array of both time and species.
+
+A comprehensive overview by \citet{Lechowicz:1982} details the benefits and faults of the most common methods at that time.
+It was demonstrated there, that a majority of the indices give comparable results, save Strauss' linear index $L$, despite the fact that most of the methods differ by range and linearity of response.
+Lechowicz generally finds that indices are either biologically interpretable, but statistically intractable, or are of little practical use.
+By the end of the paper, the selectivity index $E^*$ by \citet{Vanderploeg:1979} is recommended as the ``single best,'' albeit imperfect, index.
+This index is recommended on the basis of random feeding denoted by the index value $0$, a possible range of $[-1,1]$ (though $E^*=1$ is nigh impossible), and because the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the availability of all other prey.
+The downside to this index is its lack of reasonable statistical properties.
+
+A few years later, \citet{Chesson:1983} developed a stochastic waiting time model that assumes the searching time necessary to find a prey of interest follows an exponential distribution.
+This model, due to the memory-less property of the exponential distribution, inherently assumes that when a plausible prey species is encountered, but not eaten, that the time to find the next prey is again exponentially distributed.
+Thus, this model naturally considers the availability of other prey species by means of the predator's search time.
+We feel this model while intuitive in theory does not accord well with the data collection of the other indices.
+Therefore, we propose, loosely speaking, to invert Chesson's model and quantify the expected number of prey encounters by the predator of interest.
+To do this, our model uses the Poisson distribution to measure the rate at which encounters of a given prey species happens and the rate at which a predator eats said prey.
+In the process, however, we give up the somewhat arbitrary preference for an index to have $0$ denote random eating and the intuitive range of $[-1,1]$.
 

### 2014-12-04 paper/introduction.tex
update intro: GLMs from 90s, clarify last paragraph, mention R package

--- 2014-11-29 paper/introduction.tex
@@
 
-Modeling a predator's food selection was given much attention for a short period of time; see \citet{Strauss:1979} and the references within.
-Many, if not all, of the indices developed, though intuitive, focused on a snapshot in time and on only one prey species, and rarely obtained practical asymptotic properties.
-This restricts the interested practitioner to the most computationally manageable of techniques, and has completely ignored simultaneous testing across an array of both species and time.
+The indices most commonly used to estimate a predator's food preferences, or selectivity, are relatively old \citep{Ivlev:1964,Jacobs:1974,Chesson:1978,Strauss:1979,Vanderploeg:1979,Chesson:1983}, and yet continue to generate numerous publications; a quick search returns hundreds of publications that cite these fundamental papers, many of which are quite recent \citep{Clements:2014,Hansen:2014,Hellstrom:2014,Lyngdoh:2014,Madduppa:2014}.
+These indices, though intuitive, lack statisical justification, focus on a snapshot in time, few allow more than one prey species to be considered, and practical asymptotic statistical properties are even more rare \citep{Lechowicz:1982}.
+We propose an intiutive statistical model to determine and statistically test predators' prey preferences across an array of time and multiple prey species.
 
-The model presented here maintains tractable asymptotic properties while being general enough to take into account multiple species and time points.
-By modeling both time and any number of prey species, we are able to see a more detailed analysis of the predator's eating preferences.
-The simplicity of the model allows us to consider predators for which exact tallies of the number of each prey species eaten within any given time period is not observed.
+A comprehensive overview by \citet{Lechowicz:1982}, which was later summarized by \citet{Manly:1992}, details the benefits and faults of the most popular indices.
+According to these reviews, a majority of the indices give comparable results, save Strauss' linear index $L$, despite the fact that most of the methods differ by range and linearity of response.
+While \citet{Lechowicz:1982} recommends just one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best,'' albeit imperfect, index, \citet{Manly:1992} instead takes the approach of suggesting against the subset of indices which don't ``estimate any biologically meaningful value.''  $E^*$, an element of the \citet{Manly:1992} suggested indices, is recommended on the basis of random feeding denoted by the index value $0$, a possible range of $[-1,1]$ (though $E^*=1$ is nigh impossible), and because the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the availability of all other prey.
+The downside to this index is its lack of reasonable statistical properties \citep{Lechowicz:1982}, thus making standard errors and hypothesis testing somewhat of a mystery.
+This is, in fact, a common fault amongst most of the indices.
+
+To encourage more formal statistical inference, and simultaneously generalize predators' selectivity to animal resource selection, \citet{Manly:1992} proposed the use of generalized linear models (GLM).
+The well established literature on GLMs allowed for hypothesis testing to replace the indices.
+The model we present here, while restricted to predators' eating preferences, is a comprimise between these two extremes, indices and GLMs.
+Our model offers formal hypothesis testing and inference similar to the GLMs of \citet{Manly:1992}, but also provides meaningfull single number summaries of the predator's dietary preferenes.
+To do this, we estimate the rate at which a predator consumes the prey of interest instead of estimating the proportion of consumed to available prey.
+An outcome of our model, is that we give up the somewhat arbitrary preference for an index to have $0$ denote random eating and the range of $[-1,1]$.
+
+Our model enables formal hypothesis testing and statistical inference, while being general enough to perform statistical tests across multiple species and time points.
+This provides researchers a more detailed analysis of the predator's eating preferences.
+Further, because our model is based on the Poisson distribution, members of the well studied exponential family, we are able to estimate the parameters of interest even when exact tallies of the number of each prey species eaten within any given time period is not observed.
 Instead, we rely on the researcher being able to DNA sequence the contents of the predator's gut and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.
-Despite the data's lost information, under certain situations our model is able to accurately estimate the parameters of interest.
+Alongside our model, we offer an \texttt{R} \cite{Core-Team:2014} package named \texttt{spiders} that fits all of our methods.
 

### 2014-12-10 paper/discussion.tex
more intro edits

@@
 
-The model developed here allows practitioners to determine predators' preferences by testing simultaneously across an array of multiple prey species and time.
+The model developed here allows practitioners to determine predators' preferences by testing simultaneously across an array of multiple prey species and time points.
 This is achieved via a simple, but statistically powerful, likelihood ratio test.

### 2014-12-10 paper/introduction.tex
more intro edits

--- 2014-11-29 paper/introduction.tex
@@
 
-Modeling a predator's food selection was given much attention for a short period of time; see \citet{Strauss:1979} and the references within.
-Many, if not all, of the indices developed, though intuitive, focused on a snapshot in time and on only one prey species, and rarely obtained practical asymptotic properties.
-This restricts the interested practitioner to the most computationally manageable of techniques, and has completely ignored simultaneous testing across an array of both species and time.
+The indices most commonly used to estimate a predator's food preferences, or selectivity, are relatively old \citep{Ivlev:1964,Jacobs:1974,Chesson:1978,Strauss:1979,Vanderploeg:1979,Chesson:1983}, and yet many applied papers continue to use them; a quick search of just $2014$ returns hundreds of publications that cite these fundamental papers, a few being \citet{Clements:2014,Hansen:2014,Hellstrom:2014,Lyngdoh:2014,Madduppa:2014}.
+These indices, though intuitive, lack the statistical rigor of a model, focus on a snapshot in time, and rarely allow more than one prey species to be considered \citep{Lechowicz:1982}.
+We propose an intuitive statistical model to determine and statistically test differences in a predators' prey preferences across an array of time points and between multiple prey species.
 
-The model presented here maintains tractable asymptotic properties while being general enough to take into account multiple species and time points.
-By modeling both time and any number of prey species, we are able to see a more detailed analysis of the predator's eating preferences.
-The simplicity of the model allows us to consider predators for which exact tallies of the number of each prey species eaten within any given time period is not observed.
+A comprehensive overview by \citet{Lechowicz:1982}, which was later summarized by \citet{Manly:1992}, details the benefits and faults of the most popular indices.
+According to these reviews, a majority of the indices give comparable results, save Strauss' linear index $L$, despite the fact that most of the methods differ by range and linearity of response.
+While \citet{Lechowicz:1982} recommends just one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best,'' albeit imperfect, index, \citet{Manly:1992} instead take the approach of excluding the subset of indices which don't ``estimate any biologically meaningful value.''  \citet{Lechowicz:1982} recommends the index $E^*$, an element of the \citet{Manly:1992} suggested indices, because random feeding the index value $0$ denotes random feeding, the index has a range restricted to $[-1,1]$ (though $E^*=1$ is nigh impossible), and because the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the number of available prey types (assumed known).
+The downside to this index is its lack of reasonable statistical properties \citep{Lechowicz:1982}, thus making the computation of standard errors, and hypothesis testing difficult.
+This is, in fact, a common fault amongst most of the indices.
+
+To encourage more formal statistical inference, and simultaneously generalize predators' selectivity to animal resource selection, \citet{Manly:1992} proposed the use of generalized linear models (GLM).
+The well established literature on GLMs allows for hypothesis testing to replace the indices, by estimating the proportion of eaten prey species to that which is available while using environmental variables as predictors.
+The model we present here, while restricted to predators' eating preferences, is a compromise between these two extremes, indices and GLMs.
+Our model offers formal hypothesis testing and inference similar to the GLMs of \citet{Manly:1992}, but also provides meaningful single number summaries of the predator's dietary preferences.
+To do this, we estimate the rate at which a predator consumes the prey of interest instead of estimating the proportion of consumed to available prey.
+An outcome of our model, is that we give up the somewhat arbitrary preference for an index to have the range $[-1,1]$, and random feeding, now denoted by $1$, is formally testable across time points and across prey species.
+
+Our model enables formal hypothesis testing and statistical inference, while being general enough to perform statistical tests across multiple species and time points.
+This provides researchers a more detailed analysis of the predator's eating preferences.
+Further, because our model is based on underlying Poisson distributions, members of the well studied exponential family, we are able to estimate the parameters of interest even when exact tallies of the number of each prey species eaten within any given time period is not observed.
 Instead, we rely on the researcher being able to DNA sequence the contents of the predator's gut and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.
-Despite the data's lost information, under certain situations our model is able to accurately estimate the parameters of interest.
+
+This paper is organized as follows.
+Section~\ref{sec:methods} describes our statistical model, for both fully observed count data, and for the non-observed count data for which we use the Expectation-Maximization (EM) algorithm, and the statistical tests used to make statements about the population parameters of interest.
+In section~\ref{sec:sim}, we offer a simulation study that demonstrates the accuracy of our methods.
+Section~\ref{sec:data} provides a real dataset, which investigates the eating preferences of the wolf spider, genus \textit{Schizocosa}, found in the Berea College Forest in Madison County, Kentucky, USA, to demonstrate how interested practitioners could apply our methods.
+A brief discussion concludes the paper in section~\ref{sec:discussion}.
+Alongside our model, we offer an \texttt{R} \citep{Core-Team:2014} package named \texttt{spiders} that fits all the methods discussed.
 

### 2015-01-09 paper/acknowledgments.tex
first bio edits, realized need Adobe products

\section{Acknowledgments}
\label{sec:acknowledge}

The information reported in this paper (No. xx-xx-xxx) is part of a project of the Kentucky Agricultural Experiment Station and is published with the approval of the Director.  Support for this research was provided by the University of Kentucky Agricultural Experiment Station State Project KY008055 and the National Science Foundation Graduate Research Fellowship Program.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2015-01-09 paper/data.tex
first bio edits, realized need Adobe products

@@
 
-We analyzed a dataset that was collected to investigate the eating preferences of the wolf spider, genus \textit{Schizocosa}, towards the prey orders \textit{Diptera} and \textit{Collembola}.
-Predators were collected in traps within the deciduous Berea College Forest in Madison County, Kentucky, USA and had their gut-contents analyzed to determine whether or not the spiders ate any \textit{Diptera} or \textit{Collembola} within each time period.
+We analyzed a dataset that was collected to investigate the eating preferences of the wolf spider, genus \textit{Schizocosa}, towards the prey orders Diptera and Collembola.
+Predators were collected in traps within the deciduous Berea College Forest in Madison County, Kentucky, USA and had their gut-contents analyzed to determine whether or not the spiders ate any Diptera or Collembola within each time period.
 The left panel of figure~\ref{fig:data} plots the percentage of spiders that had the prey in their guts across time.
@@
 These data were collected for each month from October $2011$ to March $2013$.
-On average, $69$ spiders, and $111$ and $297$ \textit{Diptera} and \textit{Collembola}, respectively, were caught in each time period.
-The range of the sample sizes, across all $18$ months is, $11$ to $181$ for caught spiders, from $7$ to $322$ for trapped \textit{Diptera}, and from $101$ to $755$ for trapped \textit{Collembola}.
+On average, $69$ spiders, and $111$ and $297$ Diptera and Collembola, respectively, were caught in each time period.
+The range of the sample sizes, across all $18$ months is, $11$ to $181$ for caught spiders, from $7$ to $322$ for trapped Diptera, and from $101$ to $755$ for trapped Collembola.
 The right panel of figure~\ref{fig:data} plots the total number of each order that was caught during each time period.
@@
   \includegraphics[scale=0.5]{data}
-  \caption{For both \textit{Diptera} and \textit{Collembola}, the plots show the proportion of spiders with these orders found in their guts, and the number of these orders trapped in each time period.}
+  \caption{For both Diptera and Collembola, the plots show the proportion of spiders with these orders found in their guts, and the number of these orders trapped in each time period.}
   \label{fig:data}
@@
 These data provide an example of our hierarchy of hypotheses.
-First, we tested model $c_{st} = c$ against $c_{st} = c_s$, to determine whether or not the wolf spider has different preferences for the two orders \textit{Diptera}  and \textit{Collembola}.
+First, we tested model $c_{st} = c$ against $c_{st} = c_s$, to determine whether or not the wolf spider has different preferences for the two orders Diptera  and Collembola.
 With, one degree of freedom, this likelihood ratio test indicated, $p-value < 0.0001$,  that two parameters, one for each order, fits these data better than one parameter for both.

### 2015-01-09 paper/discussion.tex
first bio edits, realized need Adobe products

--- 2014-11-29 paper/discussion.tex
@@
 
-The model developed here allows practitioners to determine predators' preferences by testing simultaneously across an array of multiple prey species and time.
+The model developed here allows predators' preferences by testing simultaneously across an array of multiple prey species and time points.
 This is achieved via a simple, but statistically powerful, likelihood ratio test.
@@
 Further developments of our model could be beneficial.
-Taking into account other environmental variables that might effect a predator's eating habits, such as rain or temperature, say, might be advantageous.
-
-An \texttt{R} package, named \texttt{spiders}, is available on CRAN at \url{http://cran.r-project.org/web/packages/spiders/index.html} and fits all the methods discussed above.
+Taking into account other environmental variables that might effect a predator's eating habits, such as rain or temperature might be advantageous.
 

### 2015-01-09 paper/introduction.tex
first bio edits, realized need Adobe products

@@
 According to these reviews, a majority of the indices give comparable results, save Strauss' linear index $L$, despite the fact that most of the methods differ by range and linearity of response.
-While \citet{Lechowicz:1982} recommends just one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best,'' albeit imperfect, index, \citet{Manly:1992} instead take the approach of excluding the subset of indices which don't ``estimate any biologically meaningful value.''  \citet{Lechowicz:1982} recommends the index $E^*$, an element of the \citet{Manly:1992} suggested indices, because random feeding the index value $0$ denotes random feeding, the index has a range restricted to $[-1,1]$ (though $E^*=1$ is nigh impossible), and because the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the number of available prey types (assumed known).
+While \citet{Lechowicz:1982} recommends one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best,'' albeit imperfect, index, \citet{Manly:1992} instead take the approach of suggesting the indices which  estimate biologically meaningful values.
+\citet{Lechowicz:1982} recommends the index $E^*$, an element of the \citet{Manly:1992} suggested indices, because the index value $0$ denotes random feeding, the index has a range restricted to $[-1,1]$ (though $E^*=1$ is nigh impossible), and because the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the number of available prey types (assumed known).
 The downside to this index is its lack of reasonable statistical properties \citep{Lechowicz:1982}, thus making the computation of standard errors, and hypothesis testing difficult.
@@
 The well established literature on GLMs allows for hypothesis testing to replace the indices, by estimating the proportion of eaten prey species to that which is available while using environmental variables as predictors.
-The model we present here, while restricted to predators' eating preferences, is a compromise between these two extremes, indices and GLMs.
+The model we present here, while restricted to predators' preferences, is a compromise between these two extremes, indices and GLMs.
 Our model offers formal hypothesis testing and inference similar to the GLMs of \citet{Manly:1992}, but also provides meaningful single number summaries of the predator's dietary preferences.
@@
 In section~\ref{sec:sim}, we offer a simulation study that demonstrates the accuracy of our methods.
-Section~\ref{sec:data} provides a real dataset, which investigates the eating preferences of the wolf spider, genus \textit{Schizocosa}, found in the Berea College Forest in Madison County, Kentucky, USA, to demonstrate how interested practitioners could apply our methods.
+Section~\ref{sec:data} provides a real dataset, which investigates the eating preferences of the wolf spider, found in the Berea College Forest in Madison County, Kentucky, USA, to demonstrate how interested practitioners could apply our methods.
 A brief discussion concludes the paper in section~\ref{sec:discussion}.

### 2015-01-09 paper/main.tex
unlucky roll last time due to my testing
edit and typo check
more intro edits
first bio edits, realized need Adobe products

@@
-\documentclass[12pt]{article}
+\documentclass[12pt,titlepage]{article}
 \input{TexOpts.tex}
@@
 
-\title{Modeling Predator Preferences}
-\author{Edward A. Roualdes and Simon Bonner \\ University of Kentucky}
+\title{Modelling Predator Preferences using Expectation-Maximization Algorithm and Molecular Gut-Content Analysis}
+\author[1]{Edward A. Roualdes}
+\author[1]{Simon Bonner}
+\author[2]{Thomas D. Whitney}
+\author[3]{James D. Harwood}
+\affil[1]{Department of Statistics, University of Kentucky, Lexington, KY 40536, USA}
+\affil[2]{Warnell School of Forestry and Natural Resources, University of Georgia, Athens, GA 30602, USA}
+\affil[3]{Department of Entomology, University of Kentucky, Lexington, KY 40546, USA}
+
 \date{\today}
@@
 Further, we apply the Expectation-Maximization algorithm to cases where exact counts of the number of prey species eaten in a particular timer period is not observed.
-A simulation is provided to demonstrate the accuracy of our methods.
+We conduct a simulation study to demonstrate the accuracy of our method, and a real dataset, collected on the wolf spider, illustrates how to apply our methods.
+\\
+\textit{Keywords: moar words}
 \end{abstract}
@@
 \input{discussion.tex}
+\input{acknowledgments.tex}
+\input{accessibility.tex}
 

### 2015-01-09 paper/methods.tex
more intro edits
first bio edits, realized need Adobe products

@@
 Formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda}$ and $\boldsymbol{\gamma}$ offer insights to the relative rates at which predators eat particular prey species.
-
-% \begin{figure}
-%   \centering
-%   \begin{tabular}{rrcc}
-%     & & \multicolumn{2}{c}{Time} \\
-%     & & constant & varies \\
-%     \cline{3-4}
-%     \multirow{2}{*}{Species} & \multicolumn{1}{r|}{constant} & \multicolumn{1}{c}{$c$} & \multicolumn{1}{|c|}{$c_t$} \\ 
-%     \cline{3-4}
-%     & \multicolumn{1}{r|}{varies} & \multicolumn{1}{c}{$c_s$} & \multicolumn{1}{|c|}{$c_{st}$} \\
-%     \cline{3-4}
-%   \end{tabular}
-%   \caption{The four hypotheses considered are shown by their symbolic representations, highlighting which indices are allowed to vary.  This is essentially the range of the mapping $\xi$.  }
-%   \label{tab:hyp}
-% \end{figure}
 
@@
 
-% Working with biologists who study spider foraging, we have found that not all predators allow for easily counted prey species in their guts.
-As an alternative strategy, we can rely on the DNA sequencing of a sample from the predators' guts.
-If such sequencing returns a positive response, say a $1$ if a particular predator ate prey species $s$, and $0$ otherwise, we can, albeit with these incomplete data, model predators' eating preferences with the above framework using the EM algorithm.
-
 Many times it may not be possible to count the number of individuals of each prey species that are in a predator's gut.
 Instead, it may only be possible to detect whether or not a predator consumed the prey species during a given period.
-In this case we can still make inference about the predators' preferences for the different prey species by using the EM algorithm to compute maximum likelihood estimates.
+In this case we can still make inference about the predators' preferences for the different prey species by using the Expectation-Maximization (EM) algorithm to compute maximum likelihood estimates.
 
@@
 \end{equation*}
-
-% introduce EM and its iterative steps earlier
-% just a couple of general sentences; explain the two steps and what they do
-% rubin and little (~1980); mclchalan and Krishnan book
-% split up E and M step, with \cdot ^(k) notation earlier
-% mention convergence criterion
 

### 2015-01-09 paper/simulations.tex
first bio edits, realized need Adobe products

@@
 Of the hierarchy of hypotheses, we generate data under three models: $c, c_s, c_t$.
-Sample sizes for both prey species and predator gut count observations are randomly chosen from four overlapping levels: ``small'' sample sizes are randomly sampled numbers in $[20,50]$, ``medium'' $[30,75]$, ``large'' $[50,150]$, and ``huge'' $[100,200]$.
+Sample sizes for both prey species and predator gut count observations are randomly chosen from four overlapping levels: ``smallest'' sample sizes are randomly sampled numbers in $[20,50]$, ``small'' $[30,75]$, ``large'' $[50,150]$, and ``largest'' $[100,200]$.
 This is repeated for each level of sample size.
@@
   \includegraphics[scale=0.5]{nonem}
-  \caption{Density plots of all $500$ estimates of fitting the true model to the data generated from models $c_s,c$ are shown with sample sizes small and medium, respectively.}
+  \caption{Density plots of all $500$ estimates of fitting the true model to the data generated from models $c_s,c$ are shown with sample sizes smallest and small, respectively.}
   \label{fig:nonem}
@@
   \includegraphics[scale=0.5]{bp}
-  \caption{Shown are all $500$ estimates, centered at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$ with sample sizes small and medium, respectively.}
+  \caption{Shown are all $500$ estimates, centered at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$ with sample sizes smallest and small, respectively.}
   \label{fig:bp}
@@
 
-Figure~\ref{fig:em} contains density plots for all $500$ replications of the data generating models $c_s,c_t$ with small and huge sample sizes, respectively.
+Figure~\ref{fig:em} contains density plots for all $500$ replications of the data generating models $c_s,c_t$ with the smallest and the largest sample sizes, respectively.
 When data are generated under the model $c_s$ and the true model fit to the non-count data, we find even for the small sample size that point estimates are only very slightly biased.
@@
   \includegraphics[scale=0.5]{em}
-  \caption{Density plots of all $500$ estimates of fitting the true model to the data generated from models $c_s,c_t$, when counts are not observed, are shown with sample sizes small and huge, respectively.}
+  \caption{Density plots of all $500$ estimates of fitting the true model to the data generated from models $c_s,c_t$, when counts are not observed, are shown with sample sizes smallest and largest, respectively.}
   \label{fig:em}
@@
 Figure~\ref{fig:em_bp} contains box plots of the same scenarios in Figure~\ref{fig:em}.
-For the $73$ cases in which $Z_{js5} = 1$ for all $j,s$ under model $c_t$ with the huge sample size, the bias is infinite since parameter estimates will, theoretically, be infinite.
+For the $73$ cases in which $Z_{js5} = 1$ for all $j,s$ under model $c_t$ with the largest sample size, the bias is infinite since parameter estimates will, theoretically, be infinite.
 The finite bias shown in these plots is due to the finite estimates provided by the termination of the EM algorithm.
@@
   \includegraphics[scale=0.5]{em_bp}
-  \caption{Shown are all $500$ estimates, centered at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$, when counts are not observed, with sample sizes small and huge, respectively.}
+  \caption{Shown are all $500$ estimates, centered at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$, when counts are not observed, with sample sizes smallest and largest, respectively.}
   \label{fig:em_bp}

### 2015-01-12 paper/data.tex
first bio comments taken. need more references...

@@
-\section{Real Data}
+\section{Application}
 \label{sec:data}
 
-We analyzed a dataset that was collected to investigate the eating preferences of the wolf spider, genus \textit{Schizocosa}, towards the prey orders Diptera and Collembola.
-Predators were collected in traps within the deciduous Berea College Forest in Madison County, Kentucky, USA and had their gut-contents analyzed to determine whether or not the spiders ate any Diptera or Collembola within each time period.
-The left panel of figure~\ref{fig:data} plots the percentage of spiders that had the prey in their guts across time.
-Prey were also collected in traps from the same area.
-These data were collected for each month from October $2011$ to March $2013$.
-On average, $69$ spiders, and $111$ and $297$ Diptera and Collembola, respectively, were caught in each time period.
-The range of the sample sizes, across all $18$ months is, $11$ to $181$ for caught spiders, from $7$ to $322$ for trapped Diptera, and from $101$ to $755$ for trapped Collembola.
-The right panel of figure~\ref{fig:data} plots the total number of each order that was caught during each time period.
+We analyzed a dataset that was collected to investigate the feeding preferences of two species of wolf spider, Schizocosa ocreata and Schizocosa stridulans (Araneae: Lycosidae).
+Every $6-12$ days, $10$ to $40$ spiders were hand-collected between October $2011$ and April $2013$ within Berea College Forest in Madison County, Kentucky, USA.
+Spiders were removed from the litter using an aspirator, placed in separate $1.5$ mL microcentrifuge tubes filled with $95$\% {\color{red}EtOH}, and preserved at $-20\celsius$ until DNA extraction.
+In parallel, we also surveyed availability of forest floor prey using pitfall traps $(n = 32)$.
+For the analysis, both species of Schizocosa were pooled and the number of spiders and prey were analysed by month.
+On average, $69$ spiders, $111$ Diptera, and $297$ Collembola were caught in each time period.
+The range of the sample sizes across all $18$ months was $11$ to $181$ for caught spiders, $7$ to $322$ for trapped Diptera, and $101$ to $755$ for trapped Collembola.
+Figure~\ref{fig:trapped} plots the total number of each order that was caught during each time period.
+
+\begin{table}
+  \label{tab:s1}
+  \centering
+  \begin{tabular}{llll}
+    \hline
+    \textbf{Target group} & \textbf{Primer names and sequences} $5'-3'$ & \textbf{Size (bp)} & \textbf{Source}\\
+    \hline
+    Collembola & Col3F: GGACGATYTTRTTRGTTCGT & 228 & Sint et al. 2012 \\
+    & Col-gen-A246: TTTCACCTCTAACGTCGCAG & & \\
+    Diptera & DIPS16: CACTTGCTTCTTAAATrGACAAATT & 198 & Eitzinger et al. 2013 \\
+    & DIPA17: TTyATGTGAACAGTTTCAGTyCA &  & \\
+  \end{tabular}
+  \caption{Targeted prey orders, primer names and sequences, size of amplicon, and source of design for the detection of prey taxa within the guts of Schizocosa spiders.
+Both primer sets were used in singleplex PCR assays.}
+\end{table}
+
+To determine whether spiders consumed dipterans and/or collemblans, we conducted a molecular analysis of their gut-contents.
+First, DNA from spiders was extracted using Qiagen DNEasy\circledR Tissue Extraction Kit {\color{red}(Qiagen Inc., Chatsworth, California, USA)} following the animal tissue protocol outlined by the manufacturer, with minor modifications.
+Whole bodies of the spiders were first crushed to release prey DNA from within their alimentary canal for extraction.
+The $200 \mu$L extractions were stored at $-20\celsius$ until PCR.
+Second, order-specific primers from the literature were used to detect the DNA of Collembola and Diptera within the guts of the spiders.
+Primer pairs designed by {\color{red} Sint et al.
+(2012)}, targeting the $18$S rDNA gene, were used to detect Collembola predation table~\ref{tab:s1}.
+A PCR cycling protocol for $12.5 \mu$L reactions containing $1x$ Takara buffer {\color{red}(Takara Bio Inc., Shiga, Japan)}, $0.2$ mM dNTPs, $0.2 \mu$M of each primer, $0.625$ U Takara Ex TaqTM and $1.5 \mu$L of template DNA, using BioRad PTC$-200$ and C$1000$ thermal cyclers {\color{red}(Bio-Rad Laboratories, Hercules, California, USA)}, was optimized as follows: $95\celsius$ for $1$ minute, followed by $35$ cycles of $94\celsius$ for $30$ seconds, $61.2\celsius$ for $90$ seconds, and $72\celsius$ for $60$ seconds.
+Primer pairs designed by {\color{red}Eitzinger et al.
+(2013)}, targeting the $18$S rDNA gene, were used to detect Diptera predation table~\ref{tab:s1}.
+PCR cycling protocol for $12.5 \mu$L reactions with Takara reagents (as above) and $2 \mu$L of template DNA was optimized as follows: $95\celsius$ for $1$ minute, followed by $40$ cycles of $94$ for $45$ seconds, $60\celsius$ for $45$ seconds, and $72\celsius$ for $45$ seconds.
+Both primer pairs were tested for cross-reactivity against a range of prey and predator species from the field site and in all cases, no amplification of DNA was observed, confirming suitable specificity of the primers for this study.
+Lastly, electrophoresis of $10 \mu$L of each PCR product was later conducted to determine success of DNA amplification using $2\%$ Seakem agarose {\color{red}(Lonza, Rockland, Maine, USA)} stained with $1x$ GelRed\texttrademark nucleic acid stain {\color{red}(Biotium, Hayward, California, USA)}.
+This procedure allowed us to determine a presence or an absence of Diptera and Collembola DNA within each spider.
 
@@
   \centering
-  \includegraphics[scale=0.5]{data}
-  \caption{For both Diptera and Collembola, the plots show the proportion of spiders with these orders found in their guts, and the number of these orders trapped in each time period.}
-  \label{fig:data}
+  \includegraphics[scale=0.35]{prey_trapped}
+  \caption{For both Collembola (pink/dashed) and Diptera (blue/solid), the plot shows the number of the prey trapped in each time period.}
+  \label{fig:trapped}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.35]{prey_props}
+  \caption{For both Collembola (pink/dashed) and Diptera (blue/solid), the plot shows the prey proportions in the sampled wolf spiders' guts in each time period.}
+  \label{fig:props}
 \end{figure}
@@
 Model $c_{st}$ estimates $72$ parameters in total; since, in this case, there are two prey of interest and $18$ time periods, it takes $36$ parameters to estimate each $c_{st}$ and $\gamma_{st}$.
-Figure~\ref{fig:cst} plots the point estimates and $95\%$ confidence intervals of $c_{st}$, for both prey across all time periods.
+Figures~\ref{fig:est_coll},~\ref{fig:est_dipt} plot the point estimates and $95\%$ confidence intervals of $c_{st}$, for both prey across all time periods.
 
@@
   \centering
-  \includegraphics[scale=0.5]{cst}
-  \caption{One plot for each prey order displays the point estimates and $95\%$ confidence intervals as estimated from the model $c_{st}$.}
-  \label{fig:cst}
+  \includegraphics[scale=0.35]{est_coll}
+  \caption{Point estimates (blue/solid) and $95\%$ confidence intervals (pink/dashed) as estimated from the model $c_{st}$ for Collembola.}
+  \label{fig:est_coll}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.35]{est_dipt}
+  \caption{Point estimates (blue/solid) and $95\%$ confidence intervals (pink/dashed) as estimated from the model $c_{st}$ for Diptera.}
+  \label{fig:est_dipt}
 \end{figure}

### 2015-01-12 paper/discussion.tex
first bio comments taken. need more references...

@@
 
-The model developed here allows predators' preferences by testing simultaneously across an array of multiple prey species and time points.
+The model developed here allows for the determination of a predator's preferences by testing simultaneously across an array of multiple prey species and time points.
 This is achieved via a simple, but statistically powerful, likelihood ratio test.
@@
 Further developments of our model could be beneficial.
-Taking into account other environmental variables that might effect a predator's eating habits, such as rain or temperature might be advantageous.
+Taking into account other environmental variables that might effect a predator's eating habits, such as rain or temperature might also be advantageous.
 

### 2015-01-12 paper/introduction.tex
first bio comments taken. need more references...

@@
 
-The indices most commonly used to estimate a predator's food preferences, or selectivity, are relatively old \citep{Ivlev:1964,Jacobs:1974,Chesson:1978,Strauss:1979,Vanderploeg:1979,Chesson:1983}, and yet many applied papers continue to use them; a quick search of just $2014$ returns hundreds of publications that cite these fundamental papers, a few being \citet{Clements:2014,Hansen:2014,Hellstrom:2014,Lyngdoh:2014,Madduppa:2014}.
-These indices, though intuitive, lack the statistical rigor of a model, focus on a snapshot in time, and rarely allow more than one prey species to be considered \citep{Lechowicz:1982}.
+The indices most commonly used to estimate a predator's food preferences, or selectivity, are relatively old \citep{Ivlev:1964,Jacobs:1974,Chesson:1978,Strauss:1979,Vanderploeg:1979,Chesson:1983}, and yet many applied papers continue to use them; a quick search of papers published in $2014$ returns hundreds of publications that cite these fundamental papers, a few being \citet{Clements:2014,Hansen:2014,Hellstrom:2014,Lyngdoh:2014,Madduppa:2014}.
+These indices, though intuitive, lack the statistical rigour of a model, focus on a snapshot in time, and rarely allow more than one prey species to be considered \citep{Lechowicz:1982}.
 We propose an intuitive statistical model to determine and statistically test differences in a predators' prey preferences across an array of time points and between multiple prey species.
@@
 According to these reviews, a majority of the indices give comparable results, save Strauss' linear index $L$, despite the fact that most of the methods differ by range and linearity of response.
-While \citet{Lechowicz:1982} recommends one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best,'' albeit imperfect, index, \citet{Manly:1992} instead take the approach of suggesting the indices which  estimate biologically meaningful values.
-\citet{Lechowicz:1982} recommends the index $E^*$, an element of the \citet{Manly:1992} suggested indices, because the index value $0$ denotes random feeding, the index has a range restricted to $[-1,1]$ (though $E^*=1$ is nigh impossible), and because the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the number of available prey types (assumed known).
+While \citet{Lechowicz:1982} recommends one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best,'' albeit imperfect, index, \citet{Manly:1992} instead take the approach of excluding the subset of indices which do not ``estimate any biologically meaningful value.''  \citet{Lechowicz:1982} recommends the index $E^*$, an element of the \citet{Manly:1992} suggested indices, because the index value $0$ denotes random feeding, the index has a range restricted to $[-1,1]$ (though $E^*=1$ is nigh impossible), and because the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the number of available prey types (assumed known).
 The downside to this index is its lack of reasonable statistical properties \citep{Lechowicz:1982}, thus making the computation of standard errors, and hypothesis testing difficult.
@@
 
-To encourage more formal statistical inference, and simultaneously generalize predators' selectivity to animal resource selection, \citet{Manly:1992} proposed the use of generalized linear models (GLM).
+To encourage more formal statistical inference, and simultaneously generalise predators' selectivity to animal resource selection, \citet{Manly:1992} proposed the use of generalised linear models (GLM).
 The well established literature on GLMs allows for hypothesis testing to replace the indices, by estimating the proportion of eaten prey species to that which is available while using environmental variables as predictors.
@@
 Our model enables formal hypothesis testing and statistical inference, while being general enough to perform statistical tests across multiple species and time points.
-This provides researchers a more detailed analysis of the predator's eating preferences.
+This provides researchers a more detailed analysis of the predator's feeding preferences.
 Further, because our model is based on underlying Poisson distributions, members of the well studied exponential family, we are able to estimate the parameters of interest even when exact tallies of the number of each prey species eaten within any given time period is not observed.
-Instead, we rely on the researcher being able to DNA sequence the contents of the predator's gut and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.
+Instead, we rely on the researcher being able to detect prey DNA within the predator's gut \citep{Schmidt:2014,Raso:2014,Madduppa:2014} and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.
 
@@
 In section~\ref{sec:sim}, we offer a simulation study that demonstrates the accuracy of our methods.
-Section~\ref{sec:data} provides a real dataset, which investigates the eating preferences of the wolf spider, found in the Berea College Forest in Madison County, Kentucky, USA, to demonstrate how interested practitioners could apply our methods.
+Section~\ref{sec:data} provides a real dataset, which investigates the eating preferences of wolf spiders (Araneae: Lycosidae), found in the Berea College Forest in Madison County, Kentucky, USA, to demonstrate how those interested in assessing trophic interactions with gut-content analyses could apply our methods.
 A brief discussion concludes the paper in section~\ref{sec:discussion}.

### 2015-01-12 paper/main.tex
first bio comments taken. need more references...

--- 2014-10-13 paper/main.tex
@@
 \begin{document}
+\subsection*{Title:}  Formal Modelling of Predator Preferences using Molecular Gut-Content Analysis
 
-\title{Modeling Predator Preferences}
-\author{Edward A. Roualdes and Simon Bonner \\ University of Kentucky}
-\date{\today}
-\maketitle
+\subsection*{Author Information:}
+Edward A. Roualdes$^{1,2}$, Simon Bonner$^1$, Thomas D. Whitney$^3$, and James D. Harwood$^4$
 
-\begin{abstract}
-The literature on modeling a predator's prey selection includes many intuitive indices, few of which have both reasonable statistical justification and tractable asymptotic properties.
-Here, we provide a simple model that meets both of these criteria, while extending previous work to include an array of multiple species and time points.
-Further, we apply the Expectation-Maximization algorithm to cases where exact counts of the number of prey species eaten in a particular timer period is not observed.
-A simulation is provided to demonstrate the accuracy of our methods.
-\end{abstract}
+\begin{tabular}{ll}
+  1 & Department of Statistics\\
+  & University of Kentucky\\
+  & Rm.~311 Multidisciplinary Science Building\\
+  & 725 Rose Street\\
+  & Lexington, KY 40536-0082\\
+  & USA\\
+  \\
+  2 & Email: edward.roualdes@uky.edu\\
+  & Phone: 530-570-2674\\
+  \\
+  3 & Warnell School of Forestry and Natural Resources \\
+  & University of Georgia \\
+  & Athens, GA 30602\\
+  & USA \\
+  \\
+  4 & Department of Entomology\\
+  & University of Kentucky \\
+  & Lexington, KY 40546\\
+  & USA \\
+\end{tabular}
+
+\subsection*{Short Title:}
+Modelling Predator Preferances
+
+\subsection*{Word Count:}
+This manuscript contains $3253$ words including figure captions and headers, but excluding title page and equations.
+
+\subsection*{Keywords}
+electivity; Expectation-Maximization;
+
+\newpage
+\section*{Summary}
+\label{sec:abstract}
+
+\begin{enumerate}
+\item The literature on modelling the prey selection a predator includes many intuitive indices, few of which have both reasonable statistical justification and tractable asymptotic properties.
+\item Here, we provide a simple model that meets both of these criteria, while extending previous work to include an array of multiple species and time points.
+\item Further, we apply the Expectation-Maximisation algorithm to cases where exact counts of the number of prey species eaten in a particular timer period is not observed.
+\item We conduct a simulation study to demonstrate the accuracy of our method, and a real dataset, collected on wolf spiders using molecular gut-content analysis, illustrates how to apply our methods.
+\end{enumerate}
 

### 2015-01-12 paper/methods.tex
first bio comments taken. need more references...

@@
 
-The use of Poisson distributions make the following implicit assumptions: $1)$ traps independently catch the prey species of interest, $2)$ predators eat prey species independently, $3)$ predators eat independent of each other.
+The use of Poisson distributions make the following implicit assumptions: $1)$ traps independently catch the prey species of interest, $2)$ predators eat independent of each other.
 
@@
 The first hypothesis states that predators and traps sample all prey species at the same rate.
-One imagines this is the case if the predator simply eats that which comes within its reach, thus suggesting a diet indifference.
-The second says that predators sample prey proportionally across all time periods.
-The third hypothesis says that predators sample different prey species at different rates, but each rate is steady across time.
+One imagines this is the case if the predator simply eats prey which comes within its reach, thus suggesting no selection for a particular prey item.
+The second states that predators sample prey proportionally across all time periods.
+The third hypothesis states that predators sample different prey species at different rates, but each rate is steady across time.
 This implies that the predator expresses preferences for one prey species over another, but is unresponsive to changes due to time.
 Conversely, the forth hypothesis implies that each prey species is sampled similarly within each time period, while the rates across time are allowed to change.
-The fifth assumes a predator's selection varies by both time and prey species.
-This would make sense if environmental variables, say weather, or prey availability, and taste were affecting predators' selection strategies.
+The fifth hypothesis assumes a predator's selection varies by both time and prey species.
+This would make sense if environmental and biological variables, such as weather, prey availability, and/or palatability were affecting predators' selection strategies.
 
@@
 
-  \caption{Hierarchy of hypotheses.}
+  \caption{This hierarchy of hypotheses suggests the order in which the discussed models should be tested.
+One begins with the models at the top and sequentially, following the arrows, tests the hypotheses using the formal test described in section~\ref{sec:test}.}
   \label{fig:hier}
@@
 The likelihood function that allows for estimation of these parameters is as follows.
-Since we assume $X_{jst}$ is independent of $Y_{ist}$ we can simply multiply the respective Poisson probability density functions, and then form products over all $s,t$ to get the likelihood.
+Since we assume $X_{jst}$ is independent of $Y_{ist}$ we can simply multiply the respective Poisson probability density functions, and then form products over all $s,t$ to obtain the likelihood.
 
@@
 
-Many times it may not be possible to count the number of individuals of each prey species that are in a predator's gut.
-Instead, it may only be possible to detect whether or not a predator consumed the prey species during a given period.
+In many applications, such as DNA-based gut-content anslysis, it is not possible to count the number of individuals of each prey species that are in a predator's gut.
+Instead, it is only possible to detect whether or not a predator consumed the prey species during a given time period, based on the rate at which prey DNA decays in the predator gut \citep{Greenstone:2013}.
 In this case we can still make inference about the predators' preferences for the different prey species by using the Expectation-Maximization (EM) algorithm to compute maximum likelihood estimates.
@@
 \noindent No analytic solution to the M-step exists, however, so we chose to maximize $Q$ with coordinate descent \citep{Luo:1992}.
-In fact, as we only need find parameters that increase the value of $Q$ on each iteration, we forgo fully iterating to find the maximum and instead perform just one step uphill within each EM iteration.
+In fact, as we only need to find parameters that increase the value of $Q$ on each iteration, we forgo fully iterating to find the maximum and instead perform just one step uphill within each EM iteration.
 Since $Q^{(k)}$ is concave and smooth in the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$, we are able to use the convergence of parameter estimates, $||(\boldsymbol{\lambda}^{(k)}, \boldsymbol{\gamma}^{(k)}) - (\boldsymbol{\lambda}^{(k+1)}, \boldsymbol{\gamma}^{(k+1)})||_{\infty} < \tau $, for some $\tau>0$, as our stopping criterion.
@@
 In this case, not too much information is lost since estimation of $\E Z_{jst}$ can be estimated well by the proportion of observed zeros.
-On the other hand, if the predator consistently eats a given prey species, few to no zeros will show up in the observed data and $\E Z_{jst}$ is estimated to be nearly $1$.
-The loss of information is best seen by attempting to solve for $\lambda_{st}$ in the equation $1 = \E Z_{jst} 1 - \exp\{-\lambda_{st}\}$.
+In contrast, if the predator consistently eats a given prey species, few to no zeros will show up in the observed data and $\E Z_{jst}$ is estimated to be nearly $1$.
+The loss of information is best seen by attempting to solve for $\lambda_{st}$ in the equation $1 = \E Z_{jst} = 1 - \exp\{-\lambda_{st}\}$.
 As the proportion of ones in the observed data increases, we expect $\lambda_{st}$ to grow exponentially large.
@@
 \subsection{Testing}
-
+\label{sec:test}
 The likelihood ratio test statistic is

### 2015-01-12 paper/simulations.tex
first bio comments taken. need more references...

@@
-\section{Simulations}
+\section{Simulation Study}
 \label{sec:sim}
@@
 Of the hierarchy of hypotheses, we generate data under three models: $c, c_s, c_t$.
-Sample sizes for both prey species and predator gut count observations are randomly chosen from four overlapping levels: ``smallest'' sample sizes are randomly sampled numbers in $[20,50]$, ``small'' $[30,75]$, ``large'' $[50,150]$, and ``largest'' $[100,200]$.
+Sample sizes for both prey species and predator gut count observations are randomly chosen from four overlapping levels: ``small'' sample sizes are randomly sampled numbers in $[20,50]$, ``medium'' $[30,75]$, ``large'' $[50,150]$, and ``larger'' $[100,200]$.
 This is repeated for each level of sample size.
 We simulate $500$ replicate datasets for each of the twelve scenarios above for both types of data, fully observed count data, $X_{jst}$, and for non-count data, when we observe only a binary response, $Z_{jst} = 1(X_{jst}>0)$.
-Each scenario is then fit with the true model that generated the data.
+Each scenario is then fitted with the true model that generated the data.
 All simulations of non-count data use $\tau = 10^{-5}$ as the convergence tolerance.
@@
 For the simulations we used the \texttt{R} \cite{Core-Team:2014} package \texttt{BatchExperiments} by \cite{Bischl:2014}.
+
+\begin{figure}
+  \textbf{Small Sample Size}\par
+  \centering
+  \includegraphics[scale=0.75]{small}
+  \caption{The density plot of all $500$ estimates of fitting the true model to the data generated from models $c_s$ with the small sample sizes is shown.
+Each element of $c_s$ is color coded for clarity, and ticks on the x-axis show the true parameter values.}
+  \label{fig:small}
+\end{figure}
 
@@
 \begin{figure}
+  \textbf{Medium Sample Size}\par
   \centering
-  \includegraphics[scale=0.5]{nonem}
-  \caption{Density plots of all $500$ estimates of fitting the true model to the data generated from models $c_s,c$ are shown with sample sizes smallest and small, respectively.}
-  \label{fig:nonem}
+  \includegraphics[scale=0.75]{medium}
+  \caption{The density plot of all $500$ estimates of fitting the true model to the data generated from model $c$ with the small sample sizes is shown.
+A tick on the x-axis shows the true parameter value.}
+  \label{fig:medium}
 \end{figure}
 
-Figure~\ref{fig:nonem} shows density plots of the estimates of $c_s, c$ when fitting the true model to the fully observed count data generated under models $c_s$ and $c$.
+Figure~\ref{fig:small} shows the density plot of the estimates of $c_s$ when fitting the true model to the fully observed count data generated under models $c_s$, while figure~\ref{fig:medium} shows the same for the estimates of $c$.
 The plots provide evaluations of parameter estimates under each scenario.
-In the first display, the parameters $c_1 \approx 0.45$ and $c_2 = 1$ are on average, across all $500$ simulations, estimated as $\hat{c}_1 = 0.45$ and $\hat{c}_2 = 1.00$, with standard errors of $\text{se}(\hat{c}_1) = 0.03$ and $\text{se}(\hat{c}_2) = 0.06$.
-The second display provides results for model $c_{st} = c$.
+For model $c_s$ in figure~\ref{fig:small}, the parameters $c_1 \approx 0.45$ and $c_2 = 1$ are on average, across all $500$ simulations, estimated as $\hat{c}_1 = 0.45$ and $\hat{c}_2 = 1.00$, with standard errors of $\text{se}(\hat{c}_1) = 0.03$ and $\text{se}(\hat{c}_2) = 0.06$.
+Figure~\ref{fig:medium} provides results for model $c_{st} = c$.
 Averaging across all $500$ simulations, the parameter $c=2$ is estimated as $\hat{c} = 2.00$.
@@
 
-Figure~\ref{fig:em} contains density plots for all $500$ replications of the data generating models $c_s,c_t$ with the smallest and the largest sample sizes, respectively.
+Figures \ref{fig:smallEM} and \ref{fig:largerEM} contain density plots for all $500$ replications of the data generating models $c_s,c_t$ with the small and the larger sample sizes, respectively.
 When data are generated under the model $c_s$ and the true model fit to the non-count data, we find even for the small sample size that point estimates are only very slightly biased.
 When parameter values are of sufficient size to make zeros in the simulated data less common, the estimates from fitting the correct model to the generated data are occasionally over-estimated.
-This effect is easily seen in the bottom panel of figure~\ref{fig:em} for larger values of $c_t$ despite the increased sample size, but is also seen, less dramatically, in the density plot for the $c_s$ generated data.
+This effect is easily seen in figure~\ref{fig:largerEM} for the two greatest values of $c_t$ despite the increased sample size, but is also seen, less dramatically, in the density plot for the $c_s$ generated data.
 
-The cluster of estimates for $c_5$ between $3.5$ and $4.0$ in the bottom panel of figure~\ref{fig:em} comes from datasets in which $X_{js5}>0$ so that $Z_{js5}=1$ for all $j,s$.
-For the data shown in figure~\ref{fig:em}, this happened $73$ times out of the $500$ replicated datasets.
+\begin{figure}
+  \textbf{Small Sample Size, Non-Count Data}\par
+  \centering
+  \includegraphics[scale=0.75]{smallEM}
+  \caption{The density plot of all $500$ estimates of fitting the true model to the data generated from model $c_s$, when counts are not observed, is shown with the small sample size.
+Each element of $c_s$ is color coded for clarity, and ticks on the x-axis show the true parameter values.}
+  \label{fig:smallEM}
+\end{figure}
+
+The cluster of estimates for $c_5$ between $3.5$ and $4.0$ in figure~\ref{fig:largerEM} comes from datasets in which $X_{js5}>0$ so that $Z_{js5}=1$ for all $j,s$.
+For the data shown in figure~\ref{fig:largerEM}, this happened $73$ times out of the $500$ replicated datasets.
 As mentioned above, the estimate of $c_5$ is infinite in this case.
 However, the EM algorithm will always provide a finite estimate for all parameters when it terminates.
-In this case, we set $\tau=10^{-5}$ and it just happened that this caused the algorithm to terminate with $\hat c_5$ between $3.5$ and $4.0$.
+In this case, we set $\tau=10^{-5}$ and this caused the algorithm to terminate with $\hat c_5$ between $3.5$ and $4.0$.
 To confirm that this is due to the arbitrary choice of $\tau$, we repeated the algorithm with smaller values of $\tau$ for several datasets.
@@
 \begin{figure}
+  \textbf{Medium Sample Size, Non-Count Data}\par
   \centering
-  \includegraphics[scale=0.5]{em}
-  \caption{Density plots of all $500$ estimates of fitting the true model to the data generated from models $c_s,c_t$, when counts are not observed, are shown with sample sizes smallest and largest, respectively.}
-  \label{fig:em}
+  \includegraphics[scale=0.75]{largerEM}
+  \caption{The density plot of all $500$ estimates of fitting the true model to the data generated from model $c_t$, when counts are not observed, is shown with the larger sample size.
+Each element of $c_t$ is color coded for clarity, and ticks on the x-axis show the true parameter values.}
+  \label{fig:largerEM}
 \end{figure}
@@
 The over-estimation of parameters, a symptom of the loss of information due to the unobserved counts, can also be seen with box plots of the $500$ point estimates centered at their respective true parameter values.
-Figure~\ref{fig:em_bp} contains box plots of the same scenarios in Figure~\ref{fig:em}.
-For the $73$ cases in which $Z_{js5} = 1$ for all $j,s$ under model $c_t$ with the largest sample size, the bias is infinite since parameter estimates will, theoretically, be infinite.
+Figure~\ref{fig:em_bp} contains box plots of the same scenarios in figures \ref{fig:smallEM} and \ref{fig:largerEM}.
+For the $73$ cases in which $Z_{js5} = 1$ for all $j,s$ under model $c_t$ with the larger sample size, the bias is infinite since parameter estimates will, theoretically, be infinite.
 The finite bias shown in these plots is due to the finite estimates provided by the termination of the EM algorithm.
@@
   \includegraphics[scale=0.5]{em_bp}
-  \caption{Shown are all $500$ estimates, centered at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$, when counts are not observed, with sample sizes smallest and largest, respectively.}
+  \caption{Shown are all $500$ estimates, centered at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$, when counts are not observed, with sample sizes small and larger, respectively.}
   \label{fig:em_bp}

### 2015-01-13 paper/acknowledgments.tex
last of bio's comments

@@
-\section{Acknowledgments}
+\section{Acknowledgements}
 \label{sec:acknowledge}
 
-The information reported in this paper (No.
-xx-xx-xxx) is part of a project of the Kentucky Agricultural Experiment Station and is published with the approval of the Director.
+The information reported in this paper (No.\ $15-08-008$) is part of a project of the Kentucky Agricultural Experiment Station and is published with the approval of the Director.
 Support for this research was provided by the University of Kentucky Agricultural Experiment Station State Project KY008055 and the National Science Foundation Graduate Research Fellowship Program.

### 2015-01-13 paper/data.tex
last of bio's comments

@@
 Every $6-12$ days, $10$ to $40$ spiders were hand-collected between October $2011$ and April $2013$ within Berea College Forest in Madison County, Kentucky, USA.
-Spiders were removed from the litter using an aspirator, placed in separate $1.5$ mL microcentrifuge tubes filled with $95$\% {\color{red}EtOH}, and preserved at $-20\celsius$ until DNA extraction.
+Spiders were removed from the litter using an aspirator, placed in separate $1.5$ mL microcentrifuge tubes filled with $95$\% EtOH, and preserved at $-20\celsius$ until DNA extraction.
 In parallel, we also surveyed availability of forest floor prey using pitfall traps $(n = 32)$.
-For the analysis, both species of Schizocosa were pooled and the number of spiders and prey were analysed by month.
+For the analysis, both species of Schizocosa were pooled and the number of spiders and prey were analyzed by month.
 On average, $69$ spiders, $111$ Diptera, and $297$ Collembola were caught in each time period.
@@
     \hline
-    Collembola & Col3F: GGACGATYTTRTTRGTTCGT & 228 & Sint et al. 2012 \\
+    Collembola & Col3F: GGACGATYTTRTTRGTTCGT & 228 & \citet{Sint:2012} \\
     & Col-gen-A246: TTTCACCTCTAACGTCGCAG & & \\
-    Diptera & DIPS16: CACTTGCTTCTTAAATrGACAAATT & 198 & Eitzinger et al. 2013 \\
+    Diptera & DIPS16: CACTTGCTTCTTAAATrGACAAATT & 198 & \citet{Eitzinger:2014} \\
     & DIPA17: TTyATGTGAACAGTTTCAGTyCA &  & \\
@@
 To determine whether spiders consumed dipterans and/or collemblans, we conducted a molecular analysis of their gut-contents.
-First, DNA from spiders was extracted using Qiagen DNEasy\circledR Tissue Extraction Kit {\color{red}(Qiagen Inc., Chatsworth, California, USA)} following the animal tissue protocol outlined by the manufacturer, with minor modifications.
+First, DNA from spiders was extracted using Qiagen DNEasy\circledR Tissue Extraction Kit (Qiagen Inc., Chatsworth, California, USA) following the animal tissue protocol outlined by the manufacturer, with minor modifications.
 Whole bodies of the spiders were first crushed to release prey DNA from within their alimentary canal for extraction.
@@
 Second, order-specific primers from the literature were used to detect the DNA of Collembola and Diptera within the guts of the spiders.
-Primer pairs designed by {\color{red} Sint et al.
-(2012)}, targeting the $18$S rDNA gene, were used to detect Collembola predation table~\ref{tab:s1}.
-A PCR cycling protocol for $12.5 \mu$L reactions containing $1x$ Takara buffer {\color{red}(Takara Bio Inc., Shiga, Japan)}, $0.2$ mM dNTPs, $0.2 \mu$M of each primer, $0.625$ U Takara Ex TaqTM and $1.5 \mu$L of template DNA, using BioRad PTC$-200$ and C$1000$ thermal cyclers {\color{red}(Bio-Rad Laboratories, Hercules, California, USA)}, was optimized as follows: $95\celsius$ for $1$ minute, followed by $35$ cycles of $94\celsius$ for $30$ seconds, $61.2\celsius$ for $90$ seconds, and $72\celsius$ for $60$ seconds.
-Primer pairs designed by {\color{red}Eitzinger et al.
-(2013)}, targeting the $18$S rDNA gene, were used to detect Diptera predation table~\ref{tab:s1}.
+Primer pairs designed by \citet{Sint:2012}, targeting the $18$S rDNA gene, were used to detect Collembola predation table~\ref{tab:s1}.
+A PCR cycling protocol for $12.5 \mu$L reactions containing $1x$ Takara buffer (Takara Bio Inc., Shiga, Japan), $0.2$ mM dNTPs, $0.2 \mu$M of each primer, $0.625$ U Takara Ex TaqTM and $1.5 \mu$L of template DNA, using BioRad PTC$-200$ and C$1000$ thermal cyclers (Bio-Rad Laboratories, Hercules, California, USA), was optimized as follows: $95\celsius$ for $1$ minute, followed by $35$ cycles of $94\celsius$ for $30$ seconds, $61.2\celsius$ for $90$ seconds, and $72\celsius$ for $60$ seconds.
+Primer pairs designed by \citet{Eitzinger:2014}, targeting the $18$S rDNA gene, were used to detect Diptera predation table~\ref{tab:s1}.
 PCR cycling protocol for $12.5 \mu$L reactions with Takara reagents (as above) and $2 \mu$L of template DNA was optimized as follows: $95\celsius$ for $1$ minute, followed by $40$ cycles of $94$ for $45$ seconds, $60\celsius$ for $45$ seconds, and $72\celsius$ for $45$ seconds.
 Both primer pairs were tested for cross-reactivity against a range of prey and predator species from the field site and in all cases, no amplification of DNA was observed, confirming suitable specificity of the primers for this study.
-Lastly, electrophoresis of $10 \mu$L of each PCR product was later conducted to determine success of DNA amplification using $2\%$ Seakem agarose {\color{red}(Lonza, Rockland, Maine, USA)} stained with $1x$ GelRed\texttrademark nucleic acid stain {\color{red}(Biotium, Hayward, California, USA)}.
+Lastly, electrophoresis of $10 \mu$L of each PCR product was later conducted to determine success of DNA amplification using $2\%$ Seakem agarose (Lonza, Rockland, Maine, USA) stained with $1x$ GelRed\texttrademark nucleic acid stain (Biotium, Hayward, California, USA).
 This procedure allowed us to determine a presence or an absence of Diptera and Collembola DNA within each spider.

### 2015-01-13 paper/main.tex
last of bio's comments

@@
 \subsection*{Author Information:}
-Edward A. Roualdes$^{1,2}$, Simon Bonner$^1$, Thomas D. Whitney$^3$, and James D. Harwood$^4$
+Edward A. Roualdes$^{1,2}$, Simon Bonner$^1$, Thomas D. Whitney$^{3,*}$, and James D. Harwood$^3$
 
@@
   \\
-  3 & Warnell School of Forestry and Natural Resources \\
+  3 & Department of Entomology\\
+  & University of Kentucky \\
+  & Lexington, KY 40546\\
+  & USA \\
+  \\
+  * & Current address: \\
+  & Warnell School of Forestry and Natural Resources \\
   & University of Georgia \\
@@
   \\
-  4 & Department of Entomology\\
-  & University of Kentucky \\
-  & Lexington, KY 40546\\
-  & USA \\
+
 \end{tabular}
@@
 \subsection*{Short Title:}
-Modelling Predator Preferances
+Modelling Predator Preferences
 
@@
 \item Further, we apply the Expectation-Maximisation algorithm to cases where exact counts of the number of prey species eaten in a particular timer period is not observed.
-\item We conduct a simulation study to demonstrate the accuracy of our method, and a real dataset, collected on wolf spiders using molecular gut-content analysis, illustrates how to apply our methods.
+\item We conduct a simulation study to demonstrate the accuracy of our method, and a real dataset, collected on wolf spiders using molecular gut-content analysis, illustrates the utility of the approach for field analysis of predation.
 \end{enumerate}

### 2015-01-13 paper/methods.tex
last of bio's comments

@@
 
-In many applications, such as DNA-based gut-content anslysis, it is not possible to count the number of individuals of each prey species that are in a predator's gut.
+In many applications, such as DNA-based gut-content analysis, it is not possible to count the number of individuals of each prey species that are in a predator's gut.
 Instead, it is only possible to detect whether or not a predator consumed the prey species during a given time period, based on the rate at which prey DNA decays in the predator gut \citep{Greenstone:2013}.

### 2015-01-16 paper/main.tex
last of Simon's comments

@@
 \subsection*{Word Count:}
-This manuscript contains $3253$ words including figure captions and headers, but excluding title page and equations.
+This manuscript contains $3709$ words including figure captions and headers, but excluding title page and equations.
 
 \subsection*{Keywords}
-electivity; Expectation-Maximization;
+electivity; expectation-maximization; predator-prey interactions; generalist predators; food web analysis
 
@@
 \begin{enumerate}
-\item The literature on modelling the prey selection a predator includes many intuitive indices, few of which have both reasonable statistical justification and tractable asymptotic properties.
-\item Here, we provide a simple model that meets both of these criteria, while extending previous work to include an array of multiple species and time points.
-\item Further, we apply the Expectation-Maximisation algorithm to cases where exact counts of the number of prey species eaten in a particular timer period is not observed.
-\item We conduct a simulation study to demonstrate the accuracy of our method, and a real dataset, collected on wolf spiders using molecular gut-content analysis, illustrates the utility of the approach for field analysis of predation.
+\item The literature on modelling a predator's prey selection describes many intuitive indices, few of which have both reasonable statistical justification and tractable asymptotic properties.
+\item Here, we provide a simple model that meets both of these criteria, while extending previous work to include an array of data from multiple species and time points.
+\item Further, we apply the expectation-maximisation algorithm to compute estimates if exact counts of the number of prey species eaten in a particular time period are not observed.
+\item We conduct a simulation study to demonstrate the accuracy of our method, and illustrate the utility of the approach for field analysis of predation using a real dataset, collected on wolf spiders using molecular gut-content analysis.
 \end{enumerate}
 
-\input{introduction.tex}
-\input{methods.tex}
-\input{simulations.tex}
-\input{data.tex}
-\input{discussion.tex}
-\input{acknowledgments.tex}
-\input{accessibility.tex}
+\section{Introduction}
+\label{sec:intro}
+
+The indices most commonly used to describe a predator's food preferences, or selectivity, are relatively old \citep{Ivlev:1964,Jacobs:1974,Chesson:1978,Strauss:1979,Vanderploeg:1979,Chesson:1983}, and yet many applied papers continue to use them.
+A quick search of papers published in $2014$ returns hundreds of publications that cite these fundamental papers, a few being \citet{Clements:2014,Hansen:2014,Hellstrom:2014,Lyngdoh:2014,Madduppa:2014}.
+These indices, though intuitive, lack the statistical rigour of a full model, focus on a snapshot in time, and rarely allow more than one prey species to be considered \citep{Lechowicz:1982}.
+We propose an intuitive statistical model to estimate and statistically test differences in a predators' prey preferences across an array of time points and between multiple prey species.
+
+A comprehensive overview by \citet{Lechowicz:1982}, later summarised by \citet{Manly:1992}, details the benefits and faults of the most popular indices.
+According to these reviews, a majority of the indices give comparable results, save Strauss's linear index $L$, despite the fact that most of the methods differ by range and linearity of response.
+While Lechowicz recommends one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best'' \citep{Lechowicz:1982}, albeit imperfect, index, Manly et al.\ instead take the approach of excluding the subset of indices which do not ``estimate any biologically meaningful value'' \citep{Manly:1992}.
+\citet{Lechowicz:1982} recommends the index $E^*$, an element of the \citet{Manly:1992} suggested indices, because the index value $0$ denotes random feeding, the index has a range restricted to $[-1,1]$ (though $E^*=1$ is nigh impossible), and the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the number of available prey types (assumed known).
+The downside to this index is its lack of reasonable statistical properties \citep{Lechowicz:1982}, thus making the computation of standard errors, and hypothesis testing difficult.
+This is, in fact, a common fault amongst most of the indices.
+
+To encourage more formal statistical inference, and simultaneously generalise predators' selectivity to animal resource selection, \citet{Manly:1992} proposed the use of generalised linear models (GLM).
+The well established literature on GLMs allows for hypothesis testing to replace the indices, by estimating the proportion of eaten relative to the population of prey species, while using environmental variables as predictors.
+The model we present here, while restricted to predators' preferences, is a compromise between these two extremes, indices and GLMs.
+Our model offers formal hypothesis testing and inference similar to the GLMs of \citet{Manly:1992}, but also provides meaningful single number summaries of the predator's dietary preferences.
+To do this, we estimate the rate at which a predator consumes the prey of interest instead of estimating the proportion of consumed to available prey.
+Outcomes of our model, are that we give up the somewhat arbitrary preference for an index to have the range $[-1,1]$, and random feeding, now denoted by $1$, is formally testable across time points and across prey species.
+
+Our model enables formal hypothesis testing and statistical inference, while being general enough to perform statistical tests across multiple species and time points.
+This provides researchers a more detailed analysis of the predator's feeding preferences.
+Further, because our model is based on underlying Poisson distributions, members of the well studied exponential family, we are able to estimate the parameters of interest even when exact counts of each prey species eaten within any given time period are not observed.
+Instead, we rely on the researcher being able to detect prey DNA within the predator's gut \citep{Schmidt:2014,Raso:2014,Madduppa:2014} and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.
+
+This paper is organised as follows.
+Section~\ref{sec:methods} describes our statistical model, for both fully observed count data, and for the non-observed count data for which we use the expectation-maximisation (EM) algorithm, and the statistical tests used to make statements about the population parameters of interest.
+In section~\ref{sec:sim}, we offer a simulation study that demonstrates the accuracy of our methods.
+Section~\ref{sec:data} provides a real data set, which investigates the eating preferences of wolf spiders (Araneae: Lycosidae), found in the Berea College Forest in Madison County, Kentucky, USA, to demonstrate how those interested in assessing trophic interactions with gut-content analyses could apply our methods.
+A brief discussion concludes the paper in section~\ref{sec:discussion}.
+Alongside our model, we offer an \texttt{R} \citep{Core-Team:2014} package named \texttt{spiders} that implements the methods discussed.
+
+\section{Methods}
+\label{sec:methods}
+
+We assume data are collected in the following manner.
+Traps are dispersed, for $T$ time periods, throughout the habitat of the predator and prey of interest.
+Predators and prey species are collected in the traps and counted at each time period.
+We denote the number of predators and the number of prey species caught in each time period $t \in \{1, \ldots, T\}$ by $J_t$ and $I_t$, respectively.
+Prey species will be indexed by $s \in \{1, \ldots, S \}$.
+Let $X_{jst}$ represent the number of prey species $s$ that predator $j$ ate during occurrence $t$, where $j \in \{1, \ldots, J_t\}$.
+Let $Y_{ist}$ represent the number of prey species $s$ found in trap $i$ during occurrence $t$, $i \in \{1, \ldots, I_t\}$.
+
+The number of prey species $s$ that predator $j$ ate during occurrence $t$ is assumed to follow a Poisson distribution with rate parameter $\lambda_{st}$, $X_{jst} \iid \mathcal{P}(\lambda_{st})$.
+The parameter $\lambda_{st}$ represents the rate at which the predator ate prey species $s$ during time period $t$.
+The number of prey species $s$ found in trap $i$ during occurrence $t$ is assumed to follow a Poisson distibution with rate parameter $\gamma_{st}$.
+By modelling $\lambda_{st}$ and $\gamma_{st}$ we are able to test claims about a predator's eating preferences.
+Formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda} = (\lambda_{11}, \ldots, \lambda_{ST})^t$ and $\boldsymbol{\gamma} = (\gamma_{11}, \ldots, \gamma_{ST})^t$ offer insights to the relative rates at which predators eat particular prey species.
+
+The use of Poisson distributions make the following implicit assumptions: $1)$ traps independently catch the prey species of interest, $2)$ predators eat independently of each other.
+
+We consider five variations on the relative magnitude of $c_{st} = \lambda_{st}/\gamma_{st}$.
+These five hypotheses each allow $c_{st}$ to vary by time, prey species, both, or neither.
+Because the five hypotheses are nested, a natural testing order is suggested in Figure~\ref{fig:hier}.
+
+\begin{enumerate}
+\item $c_{st} = 1$
+\item $c_{st} = c$
+\item $c_{st} = c_s$
+\item $c_{st} = c_t$
+\item $c_{st} = c_{st}$
+\end{enumerate}
+
+The first hypothesis states that the relative rate of sampling for the predator and the traps is the same for all species on all occassions.
+One imagines this is the case if they prey move randomly and the predator simply eats prey which comes within its reach, thus suggesting no selection for a particular prey item.
+The second states that predators sample prey proportionally across all time periods.
+The third hypothesis states that predators sample different prey species at different rates, but each rate is steady across time.
+This implies that the predator expresses preferences for one prey species over another, but is unresponsive to changes due to time.
+Conversely, the forth hypothesis implies that each prey species is sampled similarly within each time period, while the rates across time are allowed to change.
+The fifth hypothesis assumes a predator's selection varies by both time and prey species.
+This would make sense if environmental and biological variables, such as weather, prey availability, and/or palatability were affecting predators' selection strategies.
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2.8cm, semithick]
+
+    \node[state] (A)                    {$1$};
+    \node[state] (B) [below of=A]       {$c$};
+    \node[state] (C) [below left of=B]  {$c_s$};
+    \node[state] (D) [below right of=B] {$c_t$};
+    \node[state] (E) [below left of=D]  {$c_{st}$};
+
+    \path       (B) edge node {} (A)
+                (C) edge node {} (B)
+                (D) edge node {} (B)
+                (E) edge node {} (C)
+                (E) edge node {} (D);
+  \end{tikzpicture}
+
+  \caption{This hierarchy of hypotheses suggests the order in which the discussed models should be tested.
+One begins with the most complex models at the bottom and sequentially, following the arrows, tests simpler hypotheses using the formal test described in section~\ref{sec:test} until a final model is established.}
+  \label{fig:hier}
+\end{figure}
+
+\subsection{Fully Observed Count Data}
+\label{sec:count}
+
+The likelihood function that allows for estimation of these parameters is as follows.
+Since we assume $X_{jst}$ is independent of $Y_{ist}$ we can simply multiply the respective Poisson probability density functions, and then form products over all $s,t$ to obtain the likelihood.
+
+\begin{equation}
+  \label{eq:likelihood}
+  L(x_{jst}, y_{ist} |\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \prod_{t = 1}^{T} \prod_{s=1}^S \left\{ \prod_{j=1}^{J_t} f_X(x_{jst}|\boldsymbol{\lambda}) \prod_{i=1}^{I_t} f_Y(y_{ist} | \boldsymbol{\gamma}) \right\}.
+\end{equation}
+
+\noindent Writing all five hypotheses as $\lambda_{st} = c_{st}\gamma_{st}$, we can, in the simplest cases, find analytic solutions for the maximum likelihood estimates (MLEs) of $c_{st}$, $\gamma_{st}$, and by invariance $\lambda_{st}$, when the data are balanced $J_t = J$, $I_t = I$, and $c_{st} = c$.
+Namely, these solutions are
+
+\begin{equation*}
+  \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_t + I_t}, \quad \text{ and } \quad \hat{c} = \frac{I \sum_{s,t} X_{\cdot st}}{J \sum_{s,t} Y_{\cdot st}}, \quad \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{I \left( \frac{\sum_{st} X_{\cdot st}}{\sum_{st} Y_{\cdot st}} + 1 \right)}
+\end{equation*}
+
+\noindent respectively, where $X_{\cdot st} = \sum_{j=1}^{J_t}X_{jst}$ and $Y_{\cdot st} = \sum_{i=1}^{I_t} Y_{ist}$.
+
+In all other cases, analytic solutions are not readily available and instead we rely on the fact that the log-likelihood $l(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log{L}$ is concave.
+To compute MLEs, we maximise the log-likelihood, using coordinate descent \citep{Luo:1992}, by iteratively solving partial derivatives of $l$, with respect to $c_{st}$ and $\gamma_{st}$, set equal to zero.
+The partial derivatives of the log-likelihood with respect to the parameters of interest, in the models $c, c_t, c_s,$ and $c_{st}$, set equal to zero and solved are as follows.
+
+\begin{equation*}
+  \hat{c} = \frac{\sum_{s,t} X_{\cdot st}}{\sum_t J_t \sum_s \gamma_{st}}, \, \hat{c}_t =  \frac{\sum_s X_{\cdot st}}{J_t \sum_s \gamma_{st}}, \, \hat{c}_s = \frac{\sum_{t}X_{\cdot st}}{\sum_t J_t \gamma_{st}}, \text{ or} \, \hat{c}_{st} = \frac{X_{\cdot st}}{J_t\gamma_{st}}, \text{ and } \, \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_t c_{st} + I_t}.
+\end{equation*}
+
+\noindent The equation shown for $\gamma_{st}$ is the general solution for which $c_{st}$ could be replaced by $c, c_t,$ or $c_s$.
+
+\subsection{Unobserved Counts}
+\label{sec:noncount}
+
+In many applications, such as DNA-based gut-content analysis, it is not possible to count the number of individuals of each prey species that are in a predator's gut.
+Instead, it is only possible to detect whether or not a predator consumed the prey species during a given time period, based on the rate at which prey DNA decays in the predator gut \citep{Greenstone:2013}.
+In this case we can still make inference about the predators' preferences for the different prey species by using the expectation-maximisation (EM) algorithm to compute MLEs.
+
+We denote the binary random variable indicating whether the $j^{th}$ predator did in fact eat at least one individual of prey species $s$ in time period $t$ by $Z_{jst} = 1(X_{jst} > 0)$.
+Given the Poisson assumptions above, these variables are independent Bernoulli observations with success probability $p_{st} = P(Z_{jst}=1)= 1-\exp\{-\lambda_{st}\}$.
+Despite not observing $X_{jst}$, we can compute maximum likelihood estimates of the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$ through the EM algorithm using the complete data log-likelihood
+\[
+l_{comp}(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log f_{X,Y,Z}(\boldsymbol x, \boldsymbol y, \boldsymbol z|\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \sum_{s=1}^{S} \sum_{t=1}^T \left[ \sum_{j=1}^{J_t} \log f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda}) + \sum_{i=1}^{I_t}\log f_Y(y_{jst}|\boldsymbol{\gamma}) \right].
+\]
+
+The density of $Y_{jst}$ is exactly as in section~\ref{sec:count} and so we focus on deriving the joint density of $X_{jst}$ and $Z_{jst}$.
+With the distribution of $Z_{jst}$ given above, we can compute $f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda})$ by noting that $X_{jst}=0$ with probability 1 if $Z_{jst}=0$, and that $[X_{jst}|Z_{jst}=0]$ has a truncated Poisson distribution with density
+\[
+  f_{X|Y,Z,\boldsymbol{\lambda},\boldsymbol{\gamma}}(x_{jst}|z_{jst}) =
+  \frac{\exp{\{-\lambda_{st}\}} \lambda_{st}^{x_{jst}}}{(1 - \exp{\{-\lambda_{st}\}}) x_{jst}!}1(x_{jst} > 0)
+\]
+and expected value
+\[
+\E_{X|Y,Z}X_{jst} = \frac{\lambda_{st} \exp{\{\lambda_{st} \}}}{\exp{\{ \lambda_{st} \}} - 1}.
+\]
+\noindent The joint density of $X_{jst}, Z_{jst}$ is then 
+\begin{equation*}
+    f_{X,Z|\boldsymbol{\lambda}}(x_{jst},z_{jst}) = \left\{
+    \begin{array}{lr}
+      \exp{\{ -\lambda_{st} \}}, & x_{jst}=0 \mbox{ and } z_{jst} = 0 \\
+      \frac{\exp{\{-\lambda_{st} \}} \lambda_{st}^{x_{jst}}}{x_{jst}!}, & x_{jst} > 0 \mbox{ and } z_{jst} = 1 \\
+      0 & \mbox{otherwise}
+    \end{array}
+  \right..
+\end{equation*}
+
+The EM algorithm works by iterating two steps, the E-step and M-step, until the optimum is reached \citep{Dempster:1977,McLachlan:2007}.
+Let $k$ index the iterations in the EM algorithm so that $\boldsymbol \lambda^{(k)}$ and $\boldsymbol \gamma^{(k)}$ denote the estimates computed on the $k^{th}$ M-step.
+The E-step consists of computing the expectation of $l_{comp}$ with respect to the conditional distribution of $X$ given the current estimates of the parameters
+\[
+Q^{(k)}(\boldsymbol{\lambda},\boldsymbol{\gamma}) = \E_{X|Y,Z,\boldsymbol{\lambda}^{(k)}} l_{comp}
+\]
+in order to remove the unobserved data.
+The M-step then involves maximising $Q^{(k)}(\boldsymbol{\lambda},\boldsymbol{\gamma})$ with respect to the parameters in the model to obtain updated estimates of the parameters,
+\[
+(\boldsymbol{\lambda}^{(k+1)}, \boldsymbol{\gamma}^{(k+1)}) = \argmax_{(\boldsymbol{\lambda}, \boldsymbol{\gamma})} Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma}).
+\]
+These steps are alternated until a convergence criterion monitoring subsequent differences in the parameter estimates/likelihood is met.
+
+The calculation of $Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})$ is not difficult and is given by:
+\begin{align}
+  \label{eq:estep}
+  \begin{split}
+  Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})
+  & = \E \log{f_{X,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})} + \log{f_{Y|\boldsymbol{\gamma}}(y_{ist})} \\
+  & = \sum_{s=1}^S \sum_{t=1}^T \sum_{j=1}^{J_t} \E \log{f_{X,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})}
+  + \sum_{s=1}^S \sum_{t=1}^T \sum_{i=1}^{I_t} \log{f_{Y|\boldsymbol{\gamma}}(y)} \\
+  & \propto \sum_{s,t,j} \left( - \lambda_{st} 
+    + z_{jst} \log{\lambda_{st}} \E X_{jst} \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right) \\
+  & \propto \sum_{s,t} \left( -J_t \lambda_{st} + z_{\cdot st} \log{\lambda_{st}} \E (X_{jst}|\lambda_{st}^{(k)},\gamma_{st}^{(k)}) \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right).
+\end{split}
+\end{align}
+\noindent No analytic solution to the M-step exists, however, so we again chose to maximise $Q$ with coordinate descent \citep{Luo:1992}.
+In fact, as we only need to find parameters that increase the value of $Q$ on each iteration, we forgo fully iterating the coordinate descent algorithm to find the maximum and instead perform just one step uphill within each EM iteration \citep{Givens:2012}.
+Since $Q^{(k)}$ is concave and smooth in the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$, we are able to use the convergence of parameter estimates, $||(\boldsymbol{\lambda}^{(k)}, \boldsymbol{\gamma}^{(k)}) - (\boldsymbol{\lambda}^{(k+1)}, \boldsymbol{\gamma}^{(k+1)})||_{\infty} < \tau $, for some $\tau>0$, as our stopping criterion.
+
+As we show in our simulation study, this generalised EM algorithm accurately estimates the parameters when values of $\lambda_{st}$ are relatively small, such that zeros are prevalent in the data $Z_{jst}$.
+In contrast, if the predator consistently eats a given prey species, few to no zeros will show up in the observed data and $\E Z_{jst}$ is estimated to be nearly $1$.
+The loss of information is best seen by attempting to solve for $\lambda_{st}$ in the equation $1 = \E Z_{jst} = 1 - \exp\{-\lambda_{st}\}$.
+As the proportion of ones in the observed data increases, we expect $\lambda_{st}$ to grow exponentially large.
+When no zeros are present in the data, so that only ones are observed, the likelihood can be made arbitrarily large by sending the parameter off to infinity.
+
+\subsection{Testing}
+\label{sec:test}
+The likelihood ratio test statistic is
+
+\begin{equation*}
+  \label{eq:LRT}
+    \Lambda(X,Y) := -2 \log{ \frac{ \sup_{\theta_0} L(\theta_0|X,Y)}{ \sup_{\theta_1} L(\theta_1|X,Y)} },
+\end{equation*}
+
+\noindent where $\theta_0, \theta_1$ represent the parameters estimated under the null and alternative hypotheses, respectively.
+It is well known that the asymptotic distribution of $\Lambda$ is a $\chi_{\rho}^2$ distribution with $\rho$ degrees of freedom \citep{Wilks:1938}.
+The degrees of freedom $\rho$ equal the number of free parameters available in the stated hypotheses under question.
+If we put the null hypothesis to be $H_0: \lambda_t = c_t \gamma_t$, for all $t$ and contrast this against $H_1: \lambda_{st} = c_{st}\gamma_{st}$ then there are $\rho = 2(S \cdot T) - S \cdot T - T = S \cdot T - T$ degrees of freedom.
+When the observations $X_{jst}$ are not observed, we use $L_{obs}(Z,Y)$ as the likelihood in the calculation of $\Lambda$.
+The level of significance, $\alpha$, is used to reject the null hypothesis in favour of the alternative hypothesis if $\mathbb{P}(\chi^2_{\rho} > \Lambda) < \alpha$.
+
+\subsection{Linear Transformations of $c_{st}$}
+
+After determining which model best fits the data, more detail may be extracted through specific hypothesis test of the elements of $c_{st}$, or in vector notation as $\mathbf{c} \in \mathbb{R}^{S\cdot T}$.
+Let the elements of $\hat{\mathbf{c}}$ be the maximum likelihood estimates, $\hat{c}_{st}$, as found via the framework above.
+Since $\hat{\mathbf{c}}$ is asymptotically normally distributed, any linear combination of the elements is also asymptotically normally distributed.
+For instance, let $a$ be a vector of the same dimension of $\hat{\mathbf{c}}$.
+Then $a^t\hat{\mathbf{c}}$ is asymptotically distributed as $\mathcal{N}(a^t\mathbf{c}, a^t\Sigma a)$, where $\Sigma$ is the covariance matrix of the asymptotic distribution of $\hat{\mathbf{c}}$.
+Tests of the form $H_0: a^t\mathbf{c} = \mu$ against any alternative of interest are then approximate $Z$-tests.
+Confidence intervals of any size are similarly, readily obtained.
+Suppose, for example, that the hypothesis $c_s$ is determined to best fit the data with $s$ ranging $s = 1, 2, 3$.
+We can test to see whether or not the first two species are equally preferred under the null hypothesis $c_{1} = c_{2}$.
+This hypothesis is alternatively written in vector notation as $a^t\mathbf{c} = 0$, where $a = (1, -1, 0)^t$.
+
+\section{Simulation Study}
+\label{sec:sim}
+
+Our simulations assume two prey species and five time points, throughout.
+Of the hierarchy of hypotheses, we generate data under three models: $c, c_s, c_t$.
+Sample sizes for both prey species and predator gut count observations are randomly chosen from four overlapping levels: ``small'' sample sizes are randomly sampled numbers in $[20,50]$, ``medium'' $[30,75]$, ``large'' $[50,150]$, and ``larger'' $[100,200]$.
+This is repeated for each level of sample size.
+We simulate $500$ replicate data sets for each of the twelve scenarios above for both types of data, fully observed count data, $X_{jst}$, and for non-count data, when we observe only a binary response, $Z_{jst} = 1(X_{jst}>0)$.
+Each scenario is then fitted with the true model that generated the data.
+All simulations of non-count data use $\tau = 10^{-5}$ as the convergence tolerance.
+A subset of the examples are provided here; the interested reader is referred to the supplementary materials for the complete simulation results.
+For the simulations we used the \texttt{R} \cite{Core-Team:2014} package \texttt{BatchExperiments} by \cite{Bischl:2014}.
+
+\begin{figure}
+  \textbf{Small Sample Size}\par
+  \centering
+  \includegraphics[scale=0.75]{small}
+  \caption{The density plot of all $500$ estimates of fitting the true model to the data generated from models $c_s$ with the small sample size is shown.
+Each element of $c_s$ is colour coded for clarity, and ticks on the x-axis show the true parameter values.}
+  \label{fig:small}
+\end{figure}
+
+For all simulated data, the true parameter values for the rate at which prey species are encountered in the wild are fixed to be $\gamma_{st} = \pi \approx 3.14, \, \forall s,t$.
+The values of $\lambda_{st}$ are set with respect to each data generating model.
+For model $c_{st} = c$, where predator preferences don't vary by either time or species, we put $\lambda_{st} = 2\pi, \forall s,t$.
+Under model $c_s$, the ratio of rates vary by species only, so we put $\lambda_{1t} = \sqrt{2}$ and $\lambda_{2t} = \pi$.
+Hence, $c_1 = \sqrt{2}/\pi \approx 0.45$ and $c_2 = 1$.
+For the last model, $c_t$, the ratio of rates vary by time $t$.
+Here, we put $\lambda_{st} = t$ for $t \in \{1, \ldots, 5 \}$.
+
+\begin{figure}
+  \textbf{Medium Sample Size}\par
+  \centering
+  \includegraphics[scale=0.75]{medium}
+  \caption{The density plot of all $500$ estimates of fitting the true model to the data generated from model $c$ with the medium sample size is shown.
+A tick on the x-axis shows the true parameter value.}
+  \label{fig:medium}
+\end{figure}
+
+We consider results when the correct model is fit to the simulated data.
+Figure~\ref{fig:small} shows the density plot of the estimates of $c_s$ when fitting the true model to the fully observed count data generated under models $c_s$, while figure~\ref{fig:medium} shows the same for the estimates of $c$ when data is generated under model $c$.
+The plots provide evaluations of parameter estimates under each scenario.
+For model $c_s$ in figure~\ref{fig:small}, the parameters $c_1 \approx 0.45$ and $c_2 = 1$ are on average, across all $500$ simulations, estimated as $\hat{c}_1 = 0.45$ and $\hat{c}_2 = 1.00$, with sample standard deviations of $\text{SD}(\hat{c}_1) = 0.03$ and $\text{SD}(\hat{c}_2) = 0.06$.
+Figure~\ref{fig:medium} provides results for model $c_{st} = c$.
+Averaging across all $500$ simulations, the parameter $c=2$ is estimated as $\hat{c} = 2.00$ with sample standard deviation $\text{SD}(0.06)$.
+This is further seen in figure~\ref{fig:bp}, where box plots of the parameter estimates, centred at true parameter values, of the correct model fit to data generated from both $c_s$ and $c_t$ show empirically very little bias.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.5]{bp}
+  \caption{Shown are all $500$ estimates, centred at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$ with sample sizes small and medium, respectively.}
+  \label{fig:bp}
+\end{figure}
+
+We next generated data with unobserved counts.
+As noted above under certain circumstances our unobserved counts model accurately estimates the parameters of interest, and at other times can infinitely over-estimate parameters.
+To investigate this issue further, we consider the same scenarios mentioned above, but reduce all of the count data down to binary observations.
+For each scenario, we fit the unobserved counts model as if we knew the true underlying model that generated the observed data.
+
+Figures \ref{fig:smallEM} and \ref{fig:largerEM} contain density plots of the estimates of $c_s, c_t$ for all $500$ replications of the data generating models $c_s,c_t$ with the small and the larger sample sizes, respectively.
+When data are generated under the model $c_s$ and the true model is fit to the non-count data, we find even for the small sample size that point estimates are only very slightly biased.
+When parameter values are of sufficient size to make zeros in the simulated data less common, the estimates from fitting the correct model to the generated data are occasionally over-estimated.
+This effect is easily seen in figure~\ref{fig:largerEM} for the two greatest values of $c_t$ despite the increased sample size, but is also seen, less dramatically, in the density plot for the $c_s$ generated data.
+
+\begin{figure}
+  \textbf{Small Sample Size, Non-Count Data}\par
+  \centering
+  \includegraphics[scale=0.75]{smallEM}
+  \caption{The density plot of all $500$ estimates of fitting the true model to the data generated from model $c_s$, when counts are not observed, is shown with the small sample size.
+Each element of $c_s$ is colour coded for clarity, and ticks on the x-axis show the true parameter values.}
+  \label{fig:smallEM}
+\end{figure}
+
+The cluster of estimates for $c_5$ between $3.5$ and $4.0$ in figure~\ref{fig:largerEM} comes from data sets in which $Z_{js5}=1$ for all $j,s$.
+For the data shown in figure~\ref{fig:largerEM}, this happened $73$ times out of the $500$ replicated data sets.
+As mentioned above, the estimate of $c_5$ is infinite in this case.
+However, the EM algorithm will always provide a finite estimate for all parameters when it terminates.
+In this case, we set $\tau=10^{-5}$ and this caused the algorithm to terminate with $\hat c_5$ between $3.5$ and $4.0$.
+To confirm that this is due to the arbitrary choice of $\tau$, we repeated the algorithm with smaller values of $\tau$ for several data sets.
+As expected, $\hat c_5$ increased without bound as we refit the model with increasingly small values of $\tau$.
+
+\begin{figure}
+  \textbf{Larger Sample Size, Non-Count Data}\par
+  \centering
+  \includegraphics[scale=0.75]{largerEM}
+  \caption{The density plot of all $500$ estimates of fitting the true model to the data generated from model $c_t$, when counts are not observed, is shown with the larger sample size.
+Each element of $c_t$ is colour coded for clarity, and ticks on the x-axis show the true parameter values.}
+  \label{fig:largerEM}
+\end{figure}
+
+The over-estimation of parameters, a symptom of the loss of information due to the unobserved counts, can also be seen with box plots of the $500$ point estimates centred at their respective true parameter values.
+Figure~\ref{fig:em_bp} contains box plots of the same scenarios in figures \ref{fig:smallEM} and \ref{fig:largerEM}.
+For the $73$ cases in which $Z_{js5} = 1$ for all $j,s$ under model $c_t$ with the larger sample size, the bias is infinite since parameter estimates will, theoretically, be infinite.
+The finite bias shown in these plots is due to the finite estimates provided by the termination of the EM algorithm.
+Thus, conditional on a mixture of $0$s and $1$s in the data the corresponding estimators appear to be unbiased, but when no $0$s exist in the data the theoretical bias is infinite.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.5]{em_bp}
+  \caption{Shown are all $500$ estimates, centred at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$, when counts are not observed, with sample sizes small and larger, respectively.}
+  \label{fig:em_bp}
+\end{figure}
+
+\section{Application}
+\label{sec:data}
+
+To illustrate these methods, we analysed a dataset that was collected to investigate the feeding preferences of two species of wolf spider, Schizocosa ocreata and Schizocosa stridulans (Araneae: Lycosidae).
+Every $6-12$ days, $10$ to $40$ spiders were hand-collected between October $2011$ and April $2013$ within Berea College Forest in Madison County, Kentucky, USA.
+Spiders were removed from the litter using an aspirator, placed in separate $1.5$ mL microcentrifuge tubes filled with $95$\% EtOH, and preserved at $-20\celsius$ until DNA extraction.
+In parallel, we also surveyed availability of forest floor prey using pitfall traps $(n = 32)$.
+For the analysis, both species of Schizocosa were pooled and the number of spiders and prey were analysed by month.
+On average, $69$ spiders, $111$ Diptera, and $297$ Collembola were caught in each time period.
+The range of the sample sizes across all $18$ months was $11$ to $181$ for caught spiders, $7$ to $322$ for trapped Diptera, and $101$ to $755$ for trapped Collembola.
+Figure~\ref{fig:trapped} plots the total number of each order that was caught during each time period.
+
+\begin{table}
+  \label{tab:s1}
+  \centering
+  \begin{tabular}{llll}
+    \hline
+    \textbf{Target group} & \textbf{Primer names and sequences} $5'-3'$ & \textbf{Size (bp)} & \textbf{Source}\\
+    \hline
+    Collembola & Col3F: GGACGATYTTRTTRGTTCGT & 228 & \citet{Sint:2012} \\
+    & Col-gen-A246: TTTCACCTCTAACGTCGCAG & & \\
+    Diptera & DIPS16: CACTTGCTTCTTAAATrGACAAATT & 198 & \citet{Eitzinger:2014} \\
+    & DIPA17: TTyATGTGAACAGTTTCAGTyCA &  & \\
+  \end{tabular}
+  \caption{Targeted prey orders, primer names and sequences, size of amplicon, and source of design for the detection of prey taxa within the guts of Schizocosa spiders.
+Both primer sets were used in singleplex PCR assays.}
+\end{table}
+
+To determine whether spiders had consumed dipterans and/or collemblans, we conducted a molecular analysis of their gut-contents.
+First, DNA from spiders was extracted using Qiagen DNEasy\circledR Tissue Extraction Kit (Qiagen Inc., Chatsworth, California, USA) following the animal tissue protocol outlined by the manufacturer, with minor modifications.
+Whole bodies of the spiders were first crushed to release prey DNA from within their alimentary canal for extraction.
+The $200 \mu$L extractions were stored at $-20\celsius$ until PCR.
+Second, order-specific primers from the literature were used to detect the DNA of Collembola and Diptera within the guts of the spiders.
+Primer pairs designed by \citet{Sint:2012}, targeting the $18$S rDNA gene, were used to detect Collembola predation table~\ref{tab:s1}.
+A PCR cycling protocol for $12.5 \mu$L reactions containing $1x$ Takara buffer (Takara Bio Inc., Shiga, Japan), $0.2$ mM dNTPs, $0.2 \mu$M of each primer, $0.625$ U Takara Ex TaqTM and $1.5 \mu$L of template DNA, using BioRad PTC$-200$ and C$1000$ thermal cyclers (Bio-Rad Laboratories, Hercules, California, USA), was optimised as follows: $95\celsius$ for $1$ minute, followed by $35$ cycles of $94\celsius$ for $30$ seconds, $61.2\celsius$ for $90$ seconds, and $72\celsius$ for $60$ seconds.
+Primer pairs designed by \citet{Eitzinger:2014}, targeting the $18$S rDNA gene, were used to detect Diptera predation table~\ref{tab:s1}.
+PCR cycling protocol for $12.5 \mu$L reactions with Takara reagents (as above) and $2 \mu$L of template DNA was optimised as follows: $95\celsius$ for $1$ minute, followed by $40$ cycles of $94$ for $45$ seconds, $60\celsius$ for $45$ seconds, and $72\celsius$ for $45$ seconds.
+Both primer pairs were tested for cross-reactivity against a range of prey and predator species from the field site and in all cases, no amplification of DNA was observed, confirming suitable specificity of the primers for this study.
+Lastly, electrophoresis of $10 \mu$L of each PCR product was later conducted to determine success of DNA amplification using $2\%$ Seakem agarose (Lonza, Rockland, Maine, USA) stained with $1x$ GelRed\texttrademark nucleic acid stain (Biotium, Hayward, California, USA).
+This procedure allowed us to determine a presence or an absence of Diptera and Collembola DNA within each spider.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.35]{prey_trapped}
+  \caption{For both Collembola (pink/dashed) and Diptera (blue/solid), the plot shows the number of the prey trapped in each time period.}
+  \label{fig:trapped}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.35]{prey_props}
+  \caption{For both Collembola (pink/dashed) and Diptera (blue/solid), the plot shows the prey proportions in the sampled wolf spiders' guts in each time period.}
+  \label{fig:props}
+\end{figure}
+
+These data provide an example of our hierarchy of hypotheses.
+First, we tested model $c_{st} = c$ against $c_{st} = c_s$, to determine whether or not the wolf spider has different preferences for the two orders Diptera  and Collembola.
+With, one degree of freedom, this likelihood ratio test indicated, $p-value < 0.0001$,  that two parameters, one for each order, fits these data better than one parameter for both.
+Similarly, we tested whether or not there was a significant effect across time by testing model $c_{st} = c$ against $c_{st} = c_t$.
+Here, the likelihood ratio test, with $17$ degrees of freedom, implies that the wolf spiders of the Berea College Forest eat these prey orders at different rates across the months of the year, $p-value < 0.0001$.
+In fact, we find that the most parameter rich model, $\lambda_{st} = c_{st} \gamma_{st}$ fits these data better than is expected by chance, when compared to either model $c_t$ $(p-value < 0.0001)$ or model $c_s$ $(p-value < 0.0001)$.
+Model $c_{st}$ estimates $72$ parameters in total; since, in this case, there are two prey of interest and $18$ time periods, it takes $36$ parameters to estimate each $c_{st}$ and $\gamma_{st}$.
+Figures~\ref{fig:est_coll},~\ref{fig:est_dipt} plot the point estimates and $95\%$ confidence intervals of $c_{st}$, for both prey across all time periods.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.35]{est_coll}
+  \caption{Point estimates (blue/solid) and $95\%$ confidence intervals (pink/dashed) as estimated from the model $c_{st}$ for Collembola.}
+  \label{fig:est_coll}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.35]{est_dipt}
+  \caption{Point estimates (blue/solid) and $95\%$ confidence intervals (pink/dashed) as estimated from the model $c_{st}$ for Diptera.}
+  \label{fig:est_dipt}
+\end{figure}
+
+With point estimates of $c_{st}$ under the model $\lambda_{st} = c_{st} \gamma_{st}$, we can test any number of linear contrasts.
+For instance, the hypotheses $c_{1t} = c_{2t}$, for $t \in \{1, \ldots, 18\}$ state that wolf spiders equally prefer the orders Diptera and Collembola at each of the $18$ time points.
+Using a level of significance of $0.05$, and after making a Bonferroni multiple comparisons adjustment, the data can not say that the two prey are differently preferred in October, November, and December of $2011$ and for March and July of $2012$.
+
+\section{Discussion}
+\label{sec:discussion}
+
+The model developed here allows for the determination of a predator's eating preferences by testing simultaneously across an array of multiple prey species and time points.
+This is achieved via a simple, but statistically powerful, likelihood ratio test.
+Further testing of the ratio of rates for which predators eat to encounter prey species allows researchers to make specific conclusions about predators' preferences.
+For instance, rates across time can be estimated to make statements about seasonal effects on a predator's eating habits, while relative rates across species groups allows for statements about the relative preferences for different species.
+
+When counts of predators' gut contents are not fully observed, and instead only a binary response indicating the existence of the prey species in the gut is observed, we are able to treat the counts as missing data.
+By modelling all of the observed data, both the binary responses and the number of prey species caught, and the missing count data, we are able to use the EM algorithm to extract as much information from the data as possible.
+
+Further developments of our model could take into account other environmental variables that might affect a predator's eating habits, such as rain or temperature.
+
+\section{Acknowledgements}
+\label{sec:acknowledge}
+
+The information reported in this paper (No.\ $15-08-008$) is part of a project of the Kentucky Agricultural Experiment Station and is published with the approval of the Director.
+Support for this research was provided by the University of Kentucky Agricultural Experiment Station State Project KY008055 and the National Science Foundation Graduate Research Fellowship Program.
+
+\section{Data Accessibility}
+\label{sec:accessibility}
+
+An \texttt{R} package, named \texttt{spiders}, is available on CRAN at \url{http://cran.r-project.org/web/packages/spiders/index.html} and fits all the methods discussed above.
 

### 2015-01-20 paper/main.tex
working our way to the last edits

@@
 \item Further, we apply the expectation-maximisation algorithm to compute estimates if exact counts of the number of prey species eaten in a particular time period are not observed.
-\item We conduct a simulation study to demonstrate the accuracy of our method, and illustrate the utility of the approach for field analysis of predation using a real dataset, collected on wolf spiders using molecular gut-content analysis.
+\item We conduct a simulation study to demonstrate the accuracy of our method, and illustrate the utility of the approach for field analysis of predation using a real data set, collected on wolf spiders using molecular gut-content analysis.
 \end{enumerate}
@@
 To encourage more formal statistical inference, and simultaneously generalise predators' selectivity to animal resource selection, \citet{Manly:1992} proposed the use of generalised linear models (GLM).
-The well established literature on GLMs allows for hypothesis testing to replace the indices, by estimating the proportion of eaten relative to the population of prey species, while using environmental variables as predictors.
+The well established literature on GLMs allows for hypothesis testing to replace the indices, by estimating the proportion of eaten prey species, relative to the abundance of that prey species, while using environmental variables as predictors.
 The model we present here, while restricted to predators' preferences, is a compromise between these two extremes, indices and GLMs.
@@
 Section~\ref{sec:methods} describes our statistical model, for both fully observed count data, and for the non-observed count data for which we use the expectation-maximisation (EM) algorithm, and the statistical tests used to make statements about the population parameters of interest.
-In section~\ref{sec:sim}, we offer a simulation study that demonstrates the accuracy of our methods.
+In Section~\ref{sec:sim}, we offer a simulation study that demonstrates the accuracy of our methods.
 Section~\ref{sec:data} provides a real data set, which investigates the eating preferences of wolf spiders (Araneae: Lycosidae), found in the Berea College Forest in Madison County, Kentucky, USA, to demonstrate how those interested in assessing trophic interactions with gut-content analyses could apply our methods.
-A brief discussion concludes the paper in section~\ref{sec:discussion}.
+A brief discussion concludes the paper in Section~\ref{sec:discussion}.
 Alongside our model, we offer an \texttt{R} \citep{Core-Team:2014} package named \texttt{spiders} that implements the methods discussed.
@@
 
-We assume data are collected in the following manner.
-Traps are dispersed, for $T$ time periods, throughout the habitat of the predator and prey of interest.
-Predators and prey species are collected in the traps and counted at each time period.
+We assume that samples of both the predators and prey are captured from the study area on $T$ occasions.
+Depending on the species involved and the design of the study the predators and prey may be sampled in the same way or using different methods.
+In the spider experiment described in Section~\ref{sec:data}, for example, prey are captured using pitfall traps that are dispersed throughout the study area whereas the predators (spiders) were captured by hand.
+Predators and prey species are collected and counted at each time period.
 We denote the number of predators and the number of prey species caught in each time period $t \in \{1, \ldots, T\}$ by $J_t$ and $I_t$, respectively.
 Prey species will be indexed by $s \in \{1, \ldots, S \}$.
-Let $X_{jst}$ represent the number of prey species $s$ that predator $j$ ate during occurrence $t$, where $j \in \{1, \ldots, J_t\}$.
-Let $Y_{ist}$ represent the number of prey species $s$ found in trap $i$ during occurrence $t$, $i \in \{1, \ldots, I_t\}$.
+Let $X_{jst}$ represent the number of prey species $s$ that predator $j$ ate during period $t$, where $j \in \{1, \ldots, J_t\}$.
+Let $Y_{ist}$ represent the number of prey species $s$ found in trap $i$ during period $t$, $i \in \{1, \ldots, I_t\}$.
 
-The number of prey species $s$ that predator $j$ ate during occurrence $t$ is assumed to follow a Poisson distribution with rate parameter $\lambda_{st}$, $X_{jst} \iid \mathcal{P}(\lambda_{st})$.
+The number of prey species $s$ that predator $j$ ate during period $t$ is assumed to follow a Poisson distribution with rate parameter $\lambda_{st}$, $X_{jst} \iid \mathcal{P}(\lambda_{st})$.
 The parameter $\lambda_{st}$ represents the rate at which the predator ate prey species $s$ during time period $t$.
-The number of prey species $s$ found in trap $i$ during occurrence $t$ is assumed to follow a Poisson distibution with rate parameter $\gamma_{st}$.
-By modelling $\lambda_{st}$ and $\gamma_{st}$ we are able to test claims about a predator's eating preferences.
-Formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda} = (\lambda_{11}, \ldots, \lambda_{ST})^t$ and $\boldsymbol{\gamma} = (\gamma_{11}, \ldots, \gamma_{ST})^t$ offer insights to the relative rates at which predators eat particular prey species.
-
+The number of prey species $s$ found in trap $i$ during period $t$ is assumed to follow a Poisson distribution with rate parameter $\gamma_{st}$.
 The use of Poisson distributions make the following implicit assumptions: $1)$ traps independently catch the prey species of interest, $2)$ predators eat independently of each other.
 
+By modelling $\lambda_{st}, \gamma_{st}$ we are able to test claims about a predator's eating preferences.
+Formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda} = (\lambda_{11}, \ldots, \lambda_{ST})^t$ and $\boldsymbol{\gamma} = (\gamma_{11}, \ldots, \gamma_{ST})^t$ offer insights to the relative rates at which predators eat particular prey species.
 We consider five variations on the relative magnitude of $c_{st} = \lambda_{st}/\gamma_{st}$.
@@
 
-The first hypothesis states that the relative rate of sampling for the predator and the traps is the same for all species on all occassions.
-One imagines this is the case if they prey move randomly and the predator simply eats prey which comes within its reach, thus suggesting no selection for a particular prey item.
+\noindent The first hypothesis states that the relative rate of sampling for the predator and the traps is the same for all species on all occasions.
+One imagines this is the case if the prey move randomly and the predator simply eats prey which comes within its reach, thus suggesting no selection for a particular prey item.
 The second states that predators sample prey proportionally across all time periods.
@@
 This implies that the predator expresses preferences for one prey species over another, but is unresponsive to changes due to time.
-Conversely, the forth hypothesis implies that each prey species is sampled similarly within each time period, while the rates across time are allowed to change.
+Conversely, the fourth hypothesis implies that each prey species is sampled similarly within each time period, while the rates across time are allowed to change.
 The fifth hypothesis assumes a predator's selection varies by both time and prey species.
@@
   \caption{This hierarchy of hypotheses suggests the order in which the discussed models should be tested.
-One begins with the most complex models at the bottom and sequentially, following the arrows, tests simpler hypotheses using the formal test described in section~\ref{sec:test} until a final model is established.}
+One begins with the most complex models at the bottom and sequentially, following the arrows, tests simpler hypotheses using the formal test described in Section~\ref{sec:test} until a final model is established.}
   \label{fig:hier}
@@
 The likelihood function that allows for estimation of these parameters is as follows.
-Since we assume $X_{jst}$ is independent of $Y_{ist}$ we can simply multiply the respective Poisson probability density functions, and then form products over all $s,t$ to obtain the likelihood.
+Since we assume $X_{jst}$ is independent of $Y_{ist}$ we can simply multiply the respective Poisson probability density functions, and then form products over all $s,t$ to obtain the likelihood, $L( \boldsymbol{\lambda}, \boldsymbol{\gamma} | x_{jst}, y_{ist})$.
 
@@
   \label{eq:likelihood}
-  L(x_{jst}, y_{ist} |\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \prod_{t = 1}^{T} \prod_{s=1}^S \left\{ \prod_{j=1}^{J_t} f_X(x_{jst}|\boldsymbol{\lambda}) \prod_{i=1}^{I_t} f_Y(y_{ist} | \boldsymbol{\gamma}) \right\}.
+ L( \boldsymbol{\lambda}, \boldsymbol{\gamma} | x_{jst}, y_{ist}) = \prod_{t = 1}^{T} \prod_{s=1}^S \left\{ \prod_{j=1}^{J_t} f_X(x_{jst}|\boldsymbol{\lambda}) \prod_{i=1}^{I_t} f_Y(y_{ist} | \boldsymbol{\gamma}) \right\}.
 \end{equation}
 
-\noindent Writing all five hypotheses as $\lambda_{st} = c_{st}\gamma_{st}$, we can, in the simplest cases, find analytic solutions for the maximum likelihood estimates (MLEs) of $c_{st}$, $\gamma_{st}$, and by invariance $\lambda_{st}$, when the data are balanced $J_t = J$, $I_t = I$, and $c_{st} = c$.
-Namely, these solutions are
+\noindent Writing all five hypotheses as $\lambda_{st} = c_{st}\gamma_{st}$, we can, in the simplest cases, find analytic solutions for the maximum likelihood estimates (MLEs) of $c_{st}$, $\gamma_{st}$, and by invariance $\lambda_{st}$.
+Such solutions exist when $c_{st} = 1$ so that only $\gamma_{st}$ need be estimated, and when the data are balanced: $J_t = J$, $I_t = I$ and $c_{st} = c$.
+Respectively, these solutions are
 
@@
 
-\noindent respectively, where $X_{\cdot st} = \sum_{j=1}^{J_t}X_{jst}$ and $Y_{\cdot st} = \sum_{i=1}^{I_t} Y_{ist}$.
+\noindent where $X_{\cdot st} = \sum_{j=1}^{J_t}X_{jst}$ and $Y_{\cdot st} = \sum_{i=1}^{I_t} Y_{ist}$.
 
-In all other cases, analytic solutions are not readily available and instead we rely on the fact that the log-likelihood $l(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log{L}$ is concave.
-To compute MLEs, we maximise the log-likelihood, using coordinate descent \citep{Luo:1992}, by iteratively solving partial derivatives of $l$, with respect to $c_{st}$ and $\gamma_{st}$, set equal to zero.
-The partial derivatives of the log-likelihood with respect to the parameters of interest, in the models $c, c_t, c_s,$ and $c_{st}$, set equal to zero and solved are as follows.
+In all other cases, analytic solutions are not readily available and instead we rely on the fact that the log-likelihood $l(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log{L}$ is concave to maximise the likelihood numerically.
+To compute MLEs, we maximise the log-likelihood, using coordinate descent \citep{Luo:1992}, by iteratively solving the likelihood equation (i.e.\, the equations obtained by setting the partial derivatives of the likelihood with respect to the parameters equal to zero).
+The partial derivatives of the log-likelihood with respect to the parameters of interest, in the models $c, c_t$, and $c_s$, set equal to zero and solved are as follows.
 
 \begin{equation*}
-  \hat{c} = \frac{\sum_{s,t} X_{\cdot st}}{\sum_t J_t \sum_s \gamma_{st}}, \, \hat{c}_t =  \frac{\sum_s X_{\cdot st}}{J_t \sum_s \gamma_{st}}, \, \hat{c}_s = \frac{\sum_{t}X_{\cdot st}}{\sum_t J_t \gamma_{st}}, \text{ or} \, \hat{c}_{st} = \frac{X_{\cdot st}}{J_t\gamma_{st}}, \text{ and } \, \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_t c_{st} + I_t}.
+  \hat{c} = \frac{\sum_{s,t} X_{\cdot st}}{\sum_t J_t \sum_s \gamma_{st}}, \quad \hat{c}_t =  \frac{\sum_s X_{\cdot st}}{J_t \sum_s \gamma_{st}}, \quad \hat{c}_s = \frac{\sum_{t}X_{\cdot st}}{\sum_t J_t \gamma_{st}}
 \end{equation*}
 
-\noindent The equation shown for $\gamma_{st}$ is the general solution for which $c_{st}$ could be replaced by $c, c_t,$ or $c_s$.
+\noindent The equation for updating $\hat \gamma_{st}$ is
+
+\[ \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_t c_{st} + I_t} \]
+
+\noindent where $c_{st}$ would be replaced by $c, c_t$, or $c_s$ depending on the chosen model.
 
@@
 \[
-l_{comp}(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log f_{X,Y,Z}(\boldsymbol x, \boldsymbol y, \boldsymbol z|\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \sum_{s=1}^{S} \sum_{t=1}^T \left[ \sum_{j=1}^{J_t} \log f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda}) + \sum_{i=1}^{I_t}\log f_Y(y_{jst}|\boldsymbol{\gamma}) \right].
+l_{comp}(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log f_{X,Y,Z}(\boldsymbol x, \boldsymbol y, \boldsymbol z|\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \sum_{s,t}^{S,T} \left[ \sum_{j=1}^{J_t} \log f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda}) + \sum_{i=1}^{I_t}\log f_Y(y_{jst}|\boldsymbol{\gamma}) \right].
 \]
 
-The density of $Y_{jst}$ is exactly as in section~\ref{sec:count} and so we focus on deriving the joint density of $X_{jst}$ and $Z_{jst}$.
+The density of $Y_{jst}$ is exactly as in Section~\ref{sec:count} and so we focus on deriving the joint density of $X_{jst}$ and $Z_{jst}$.
 With the distribution of $Z_{jst}$ given above, we can compute $f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda})$ by noting that $X_{jst}=0$ with probability 1 if $Z_{jst}=0$, and that $[X_{jst}|Z_{jst}=0]$ has a truncated Poisson distribution with density
@@
 If we put the null hypothesis to be $H_0: \lambda_t = c_t \gamma_t$, for all $t$ and contrast this against $H_1: \lambda_{st} = c_{st}\gamma_{st}$ then there are $\rho = 2(S \cdot T) - S \cdot T - T = S \cdot T - T$ degrees of freedom.
-When the observations $X_{jst}$ are not observed, we use $L_{obs}(Z,Y)$ as the likelihood in the calculation of $\Lambda$.
-The level of significance, $\alpha$, is used to reject the null hypothesis in favour of the alternative hypothesis if $\mathbb{P}(\chi^2_{\rho} > \Lambda) < \alpha$.
+When the observations $X_{jst}$ are not observed, we use $L_{obs}(\boldsymbol{\lambda}, \boldsymbol{\gamma}|Z,Y)$ as the likelihood in the calculation of $\Lambda$
+
+\[ L_{obs}(\boldsymbol{\lambda}, \boldsymbol{\gamma}|Z,Y) = \prod_{s,t}^{S,T} \left\{ \prod_{j=1}^{J_t}f_Z(z_{jst}|\boldsymbol{\lambda}) \prod_{i=1}^{I_t}f_Y(y_{ist}|\boldsymbol{\gamma}) \right\}. \]
+
+\noindent The level of significance $\alpha$ is used to reject the null hypothesis in favour of the alternative hypothesis if $\mathbb{P}(\chi^2_{\rho} > \Lambda) < \alpha$.
 
@@
 A subset of the examples are provided here; the interested reader is referred to the supplementary materials for the complete simulation results.
-For the simulations we used the \texttt{R} \cite{Core-Team:2014} package \texttt{BatchExperiments} by \cite{Bischl:2014}.
 
@@
 We consider results when the correct model is fit to the simulated data.
-Figure~\ref{fig:small} shows the density plot of the estimates of $c_s$ when fitting the true model to the fully observed count data generated under models $c_s$, while figure~\ref{fig:medium} shows the same for the estimates of $c$ when data is generated under model $c$.
+Figure~\ref{fig:small} shows the density plot of the estimates of $c_s$ when fitting the true model to the fully observed count data generated under models $c_s$, while Figure~\ref{fig:medium} shows the same for the estimates of $c$ when data is generated under model $c$.
 The plots provide evaluations of parameter estimates under each scenario.
-For model $c_s$ in figure~\ref{fig:small}, the parameters $c_1 \approx 0.45$ and $c_2 = 1$ are on average, across all $500$ simulations, estimated as $\hat{c}_1 = 0.45$ and $\hat{c}_2 = 1.00$, with sample standard deviations of $\text{SD}(\hat{c}_1) = 0.03$ and $\text{SD}(\hat{c}_2) = 0.06$.
+For model $c_s$ in Figure~\ref{fig:small}, the parameters $c_1 \approx 0.45$ and $c_2 = 1$ are on average, across all $500$ simulations, estimated as $\hat{c}_1 = 0.45$ and $\hat{c}_2 = 1.00$, with sample standard deviations of $\text{SD}(\hat{c}_1) = 0.03$ and $\text{SD}(\hat{c}_2) = 0.06$.
 Figure~\ref{fig:medium} provides results for model $c_{st} = c$.
-Averaging across all $500$ simulations, the parameter $c=2$ is estimated as $\hat{c} = 2.00$ with sample standard deviation $\text{SD}(0.06)$.
-This is further seen in figure~\ref{fig:bp}, where box plots of the parameter estimates, centred at true parameter values, of the correct model fit to data generated from both $c_s$ and $c_t$ show empirically very little bias.
+Averaging across all $500$ simulations, the parameter $c=2$ is estimated as $\hat{c} = 2.00$ with sample standard deviation $\text{SD}(\hat{c})=0.06$.
+This is further seen in Figure~\ref{fig:bp}, where box plots of the parameter estimates, centred at true parameter values, of the correct model fit to data generated from both $c_s$ and $c_t$ show empirically very little bias.
 
@@
   \includegraphics[scale=0.5]{bp}
-  \caption{Shown are all $500$ estimates, centred at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$ with sample sizes small and medium, respectively.}
+  \caption{Shown are all $500$ estimates, cent-red at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$ with sample sizes small and medium, respectively.}
   \label{fig:bp}
@@
 When parameter values are of sufficient size to make zeros in the simulated data less common, the estimates from fitting the correct model to the generated data are occasionally over-estimated.
-This effect is easily seen in figure~\ref{fig:largerEM} for the two greatest values of $c_t$ despite the increased sample size, but is also seen, less dramatically, in the density plot for the $c_s$ generated data.
+This effect is easily seen in Figure~\ref{fig:largerEM} for the two greatest values of $c_t$ despite the increased sample size, but is also seen, less dramatically, in the density plot for the $c_s$ generated data.
 
@@
 
-The cluster of estimates for $c_5$ between $3.5$ and $4.0$ in figure~\ref{fig:largerEM} comes from data sets in which $Z_{js5}=1$ for all $j,s$.
-For the data shown in figure~\ref{fig:largerEM}, this happened $73$ times out of the $500$ replicated data sets.
+The cluster of estimates for $c_5$ between $3.5$ and $4.0$ in Figure~\ref{fig:largerEM} comes from data sets in which $Z_{js5}=1$ for all $j,s$.
+For the data shown in Figure~\ref{fig:largerEM}, this happened $73$ times out of the $500$ replicated data sets.
 As mentioned above, the estimate of $c_5$ is infinite in this case.
@@
 The over-estimation of parameters, a symptom of the loss of information due to the unobserved counts, can also be seen with box plots of the $500$ point estimates centred at their respective true parameter values.
-Figure~\ref{fig:em_bp} contains box plots of the same scenarios in figures \ref{fig:smallEM} and \ref{fig:largerEM}.
+Figure~\ref{fig:em_bp} contains box plots of the same scenarios in Figures \ref{fig:smallEM} and \ref{fig:largerEM}.
 For the $73$ cases in which $Z_{js5} = 1$ for all $j,s$ under model $c_t$ with the larger sample size, the bias is infinite since parameter estimates will, theoretically, be infinite.
@@
 
-To illustrate these methods, we analysed a dataset that was collected to investigate the feeding preferences of two species of wolf spider, Schizocosa ocreata and Schizocosa stridulans (Araneae: Lycosidae).
+To illustrate these methods, we analysed a data set that was collected to investigate the feeding preferences of two species of wolf spider, Schizocosa ocreata and Schizocosa stridulans (Araneae: Lycosidae).
 Every $6-12$ days, $10$ to $40$ spiders were hand-collected between October $2011$ and April $2013$ within Berea College Forest in Madison County, Kentucky, USA.
-Spiders were removed from the litter using an aspirator, placed in separate $1.5$ mL microcentrifuge tubes filled with $95$\% EtOH, and preserved at $-20\celsius$ until DNA extraction.
+Spiders were removed from the leaf litter using an aspirator, placed in separate $1.5$ mL microcentrifuge tubes filled with $95$\% EtOH, and preserved at $-20\celsius$ until DNA extraction.
 In parallel, we also surveyed availability of forest floor prey using pitfall traps $(n = 32)$.
@@
 Second, order-specific primers from the literature were used to detect the DNA of Collembola and Diptera within the guts of the spiders.
-Primer pairs designed by \citet{Sint:2012}, targeting the $18$S rDNA gene, were used to detect Collembola predation table~\ref{tab:s1}.
+Primer pairs designed by \citet{Sint:2012}, targeting the $18$S rDNA gene, were used to detect Collembola predation Table~\ref{tab:s1}.
 A PCR cycling protocol for $12.5 \mu$L reactions containing $1x$ Takara buffer (Takara Bio Inc., Shiga, Japan), $0.2$ mM dNTPs, $0.2 \mu$M of each primer, $0.625$ U Takara Ex TaqTM and $1.5 \mu$L of template DNA, using BioRad PTC$-200$ and C$1000$ thermal cyclers (Bio-Rad Laboratories, Hercules, California, USA), was optimised as follows: $95\celsius$ for $1$ minute, followed by $35$ cycles of $94\celsius$ for $30$ seconds, $61.2\celsius$ for $90$ seconds, and $72\celsius$ for $60$ seconds.
-Primer pairs designed by \citet{Eitzinger:2014}, targeting the $18$S rDNA gene, were used to detect Diptera predation table~\ref{tab:s1}.
+Primer pairs designed by \citet{Eitzinger:2014}, targeting the $18$S rDNA gene, were used to detect Diptera predation Table~\ref{tab:s1}.
 PCR cycling protocol for $12.5 \mu$L reactions with Takara reagents (as above) and $2 \mu$L of template DNA was optimised as follows: $95\celsius$ for $1$ minute, followed by $40$ cycles of $94$ for $45$ seconds, $60\celsius$ for $45$ seconds, and $72\celsius$ for $45$ seconds.
@@
 These data provide an example of our hierarchy of hypotheses.
-First, we tested model $c_{st} = c$ against $c_{st} = c_s$, to determine whether or not the wolf spider has different preferences for the two orders Diptera  and Collembola.
-With, one degree of freedom, this likelihood ratio test indicated, $p-value < 0.0001$,  that two parameters, one for each order, fits these data better than one parameter for both.
-Similarly, we tested whether or not there was a significant effect across time by testing model $c_{st} = c$ against $c_{st} = c_t$.
-Here, the likelihood ratio test, with $17$ degrees of freedom, implies that the wolf spiders of the Berea College Forest eat these prey orders at different rates across the months of the year, $p-value < 0.0001$.
-In fact, we find that the most parameter rich model, $\lambda_{st} = c_{st} \gamma_{st}$ fits these data better than is expected by chance, when compared to either model $c_t$ $(p-value < 0.0001)$ or model $c_s$ $(p-value < 0.0001)$.
+From the bottom of the graph in Figure~\ref{fig:hier}, we tested the most parameter rich model $c_{st} = \lambda_{st}/\gamma_{st}$ against models $c_{s}, c_{t}$.
+In both cases, the more parameter rich model fits these data better than is expected by chance; $H_0: c_s$ versus $H_1: c_{st}$ gives p-value $< 0.0001$ and $H_0: c_t$ versus $H_1: c_{st}$ gives p-value $< 0.0001$.
 Model $c_{st}$ estimates $72$ parameters in total; since, in this case, there are two prey of interest and $18$ time periods, it takes $36$ parameters to estimate each $c_{st}$ and $\gamma_{st}$.
@@
 
-With point estimates of $c_{st}$ under the model $\lambda_{st} = c_{st} \gamma_{st}$, we can test any number of linear contrasts.
+With point estimates of $c_{st}$ under the model $c_{st} = \lambda_{st}/\gamma_{st}$, we can test any number of linear contrasts.
 For instance, the hypotheses $c_{1t} = c_{2t}$, for $t \in \{1, \ldots, 18\}$ state that wolf spiders equally prefer the orders Diptera and Collembola at each of the $18$ time points.
@@
 An \texttt{R} package, named \texttt{spiders}, is available on CRAN at \url{http://cran.r-project.org/web/packages/spiders/index.html} and fits all the methods discussed above.
+\nocite{Wickham:2011a,Wickham:2011,Bischl:2014}
 

### 2015-01-20 paper/main.tex
add c_{st} updating equation

@@
 The likelihood function that allows for estimation of these parameters is as follows.
-Since we assume $X_{jst}$ is independent of $Y_{ist}$ we can simply multiply the respective Poisson probability density functions, and then form products over all $s,t$ to obtain the likelihood, $L( \boldsymbol{\lambda}, \boldsymbol{\gamma} | x_{jst}, y_{ist})$.
+Since we assume $X_{jst}$ is independent of $Y_{ist}$ we can simply multiply the respective Poisson probability density functions, and then form products over all $s,t$ to obtain the likelihood
 
@@
 \begin{equation*}
-  \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_t + I_t}, \quad \text{ and } \quad \hat{c} = \frac{I \sum_{s,t} X_{\cdot st}}{J \sum_{s,t} Y_{\cdot st}}, \quad \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{I \left( \frac{\sum_{st} X_{\cdot st}}{\sum_{st} Y_{\cdot st}} + 1 \right)}
+  \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_t + I_t}, \quad \text{ and } \quad \hat{c} = \frac{I \sum_{s,t} X_{\cdot st}}{J \sum_{s,t} Y_{\cdot st}}, \quad \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{I \left( \frac{\sum_{st} X_{\cdot st}}{\sum_{st} Y_{\cdot st}} + 1 \right)},
 \end{equation*}
@@
 In all other cases, analytic solutions are not readily available and instead we rely on the fact that the log-likelihood $l(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log{L}$ is concave to maximise the likelihood numerically.
-To compute MLEs, we maximise the log-likelihood, using coordinate descent \citep{Luo:1992}, by iteratively solving the likelihood equation (i.e.\, the equations obtained by setting the partial derivatives of the likelihood with respect to the parameters equal to zero).
-The partial derivatives of the log-likelihood with respect to the parameters of interest, in the models $c, c_t$, and $c_s$, set equal to zero and solved are as follows.
+To compute MLEs, we maximise the log-likelihood, using coordinate descent \citep{Luo:1992}, by iteratively solving the likelihood equations (i.e.\, the equations obtained by setting the partial derivatives of the likelihood with respect to the parameters equal to zero).
+This yields the following equations for updating $c$ in the models $c$, $c_t$, and $c_s$ respectively
 
 \begin{equation*}
-  \hat{c} = \frac{\sum_{s,t} X_{\cdot st}}{\sum_t J_t \sum_s \gamma_{st}}, \quad \hat{c}_t =  \frac{\sum_s X_{\cdot st}}{J_t \sum_s \gamma_{st}}, \quad \hat{c}_s = \frac{\sum_{t}X_{\cdot st}}{\sum_t J_t \gamma_{st}}
+  \hat{c} = \frac{\sum_{s,t} X_{\cdot st}}{\sum_t J_t \sum_s \gamma_{st}}, \quad \hat{c}_t =  \frac{\sum_s X_{\cdot st}}{J_t \sum_s \gamma_{st}}, \quad \hat{c}_s = \frac{\sum_{t}X_{\cdot st}}{\sum_t J_t \gamma_{st}}, \quad \text{ and } \quad \hat{c}_{st} = \frac{X_{\cdot st}}{J_t \gamma_{st}}.
 \end{equation*}
@@
 
-\[ \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_t c_{st} + I_t} \]
+\[ \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_t c_{st} + I_t}, \]
 

### 2015-01-20 paper/main.tex
quite possibly the last edits

@@
 \noindent Writing all five hypotheses as $\lambda_{st} = c_{st}\gamma_{st}$, we can, in the simplest cases, find analytic solutions for the maximum likelihood estimates (MLEs) of $c_{st}$, $\gamma_{st}$, and by invariance $\lambda_{st}$.
-Such solutions exist when $c_{st} = 1$ so that only $\gamma_{st}$ need be estimated, and when the data are balanced: $J_t = J$, $I_t = I$ and $c_{st} = c$.
+Such solutions exist for the simplest model, $c_{st}=1$, for which only $\gamma_{st}$ need be estimated, and for the second simplest model, $c_{st}=c$, provided that the data are balanced so that $J_t=J$ and $I_t=I$.
 Respectively, these solutions are

### 2015-04-03 paper/main.tex
html vignette, starting to address reviewer's comments

@@
-\documentclass[12pt]{article}
+\documentclass[smallextended]{svjour3}
 \input{TexOpts.tex}
+\begin{document}
+\title{Formal Modelling of Predator Preferences using Molecular Gut-Content Analysis}
+\author{Edward A. Roualdes \and Simon Bonner \and Thomas D. Whitney \and James D. Harwood}
+\institute{E.A. Roualdes \and S. Bonner
+        \at Department of Statistics, University of Kentucky, Rm.~311 Multidisciplinary Science Building, 725 Rose Street, Lexington, KY 40536-0082, \\\email{edward.roualdes@uky.edu} \and
+        T.D. Whitney 
+        \at Department of Entomology, University of Kentucky, Lexington, KY 40546, USA \emph{Present address: Warnell School of Forestry and Natural Resources, University of Georgia, Athens, GA 30602, USA}
+        \and
+        J.D. Harwood
+        \at Department of Entomology, University of Kentucky, Lexington, KY 40546, USA
+}
 
-\begin{document}
-\subsection*{Title:}  Formal Modelling of Predator Preferences using Molecular Gut-Content Analysis
+% \subsection*{Keywords}
+% 
+\maketitle
 
-\subsection*{Author Information:}
-Edward A. Roualdes$^{1,2}$, Simon Bonner$^1$, Thomas D. Whitney$^{3,*}$, and James D. Harwood$^3$
+\begin{abstract}
+  The literature on modelling a predator's prey selection describes many intuitive indices, few of which have both reasonable statistical justification and tractable asymptotic properties.
+Here, we provide a simple model that meets both of these criteria, while extending previous work to include an array of data from multiple species and time points.
+Further, we apply the expectation-maximisation algorithm to compute estimates if exact counts of the number of prey species eaten in a particular time period are not observed.
+We conduct a simulation study to demonstrate the accuracy of our method, and illustrate the utility of the approach for field analysis of predation using a real data set, collected on wolf spiders using molecular gut-content analysis.
+  \keywords{electivity \and expectation-maximization \and predator-prey interactions \and generalist predators \and food web analysis}
+\end{abstract}
 
-\begin{tabular}{ll}
-  1 & Department of Statistics\\
-  & University of Kentucky\\
-  & Rm.~311 Multidisciplinary Science Building\\
-  & 725 Rose Street\\
-  & Lexington, KY 40536-0082\\
-  & USA\\
-  \\
-  2 & Email: edward.roualdes@uky.edu\\
-  & Phone: 530-570-2674\\
-  \\
-  3 & Department of Entomology\\
-  & University of Kentucky \\
-  & Lexington, KY 40546\\
-  & USA \\
-  \\
-  * & Current address: \\
-  & Warnell School of Forestry and Natural Resources \\
-  & University of Georgia \\
-  & Athens, GA 30602\\
-  & USA \\
-  \\
-
-\end{tabular}
-
-\subsection*{Short Title:}
-Modelling Predator Preferences
-
-\subsection*{Word Count:}
-This manuscript contains $3709$ words including figure captions and headers, but excluding title page and equations.
-
-\subsection*{Keywords}
-electivity; expectation-maximization; predator-prey interactions; generalist predators; food web analysis
-
-\newpage
-\section*{Summary}
-\label{sec:abstract}
-
-\begin{enumerate}
-\item The literature on modelling a predator's prey selection describes many intuitive indices, few of which have both reasonable statistical justification and tractable asymptotic properties.
-\item Here, we provide a simple model that meets both of these criteria, while extending previous work to include an array of data from multiple species and time points.
-\item Further, we apply the expectation-maximisation algorithm to compute estimates if exact counts of the number of prey species eaten in a particular time period are not observed.
-\item We conduct a simulation study to demonstrate the accuracy of our method, and illustrate the utility of the approach for field analysis of predation using a real data set, collected on wolf spiders using molecular gut-content analysis.
-\end{enumerate}
+% tweet: Formal modelling of a predator's prey selection across multiple species and time points using molecular gut-content analysis.
 
 \section{Introduction}
-\label{sec:intro}
+\label{intro}
 
@@
 These indices, though intuitive, lack the statistical rigour of a full model, focus on a snapshot in time, and rarely allow more than one prey species to be considered \citep{Lechowicz:1982}.
+Other authors \citep{Davey:2013,King:2010} are using the Monte Carlo based method of \citet{Agusti:2003}, but this also neglects to take into account multiple species across multiple time points.
 We propose an intuitive statistical model to estimate and statistically test differences in a predators' prey preferences across an array of time points and between multiple prey species.
 
-A comprehensive overview by \citet{Lechowicz:1982}, later summarised by \citet{Manly:1992}, details the benefits and faults of the most popular indices.
+A comprehensive overview by \citet{Lechowicz:1982}, later summarised by \citet{Manly:2002}, details some of the benefits and faults of the most popular indices.
 According to these reviews, a majority of the indices give comparable results, save Strauss's linear index $L$, despite the fact that most of the methods differ by range and linearity of response.
-While Lechowicz recommends one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best'' \citep{Lechowicz:1982}, albeit imperfect, index, Manly et al.\ instead take the approach of excluding the subset of indices which do not ``estimate any biologically meaningful value'' \citep{Manly:1992}.
-\citet{Lechowicz:1982} recommends the index $E^*$, an element of the \citet{Manly:1992} suggested indices, because the index value $0$ denotes random feeding, the index has a range restricted to $[-1,1]$ (though $E^*=1$ is nigh impossible), and the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the number of available prey types (assumed known).
-The downside to this index is its lack of reasonable statistical properties \citep{Lechowicz:1982}, thus making the computation of standard errors, and hypothesis testing difficult.
+While Lechowicz recommends one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best'' \citep{Lechowicz:1982}, albeit imperfect, index, Manly et al.\ instead take the approach of excluding the subset of indices which do not ``estimate any biologically meaningful value'' \citep{Manly:2002}.
+\citet{Lechowicz:1982} recommends the index $E^*$, an element of the \citet{Manly:2002} suggested indices, because the index value $0$ denotes random feeding, the index has a range restricted to $[-1,1]$ (though $E^*=1$ is nigh impossible), and the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the number of available prey types (assumed known).
+The downside to this index is its lack of reasonable statistical properties \citep{Lechowicz:1982}, thus making the computation of standard errors and thus hypothesis testing difficult.
 This is, in fact, a common fault amongst most of the indices.
 
-To encourage more formal statistical inference, and simultaneously generalise predators' selectivity to animal resource selection, \citet{Manly:1992} proposed the use of generalised linear models (GLM).
-The well established literature on GLMs allows for hypothesis testing to replace the indices, by estimating the proportion of eaten prey species, relative to the abundance of that prey species, while using environmental variables as predictors.
-The model we present here, while restricted to predators' preferences, is a compromise between these two extremes, indices and GLMs.
-Our model offers formal hypothesis testing and inference similar to the GLMs of \citet{Manly:1992}, but also provides meaningful single number summaries of the predator's dietary preferences.
-To do this, we estimate the rate at which a predator consumes the prey of interest instead of estimating the proportion of consumed to available prey.
+\citet{Manly:2002} recognized the need for more formal statistical inference and proposed the use of generalised linear models (GLM).
+The well established literature on GLMs allows for formal hypothesis testing.
+Much of this work however focues on general resource selection and doesn't directly address the problems of food selection.
+Our model, while focusing on predators' food selection not general resource selection, offers formal hypothesis testing and inference similar to the GLMs of \citet{Manly:2002}, but also provides meaningful single number summaries of the predator's dietary preferences.
+Thus, we are proposing here a model that 1) maintains the intuitiveness of the indices summarised by \citet{Lechowicz:1982, Manly:2002}, 2) offers the formal statistical justification of GLMs, 3) models multiple prey species across multiple time points, and 4) estimates parameters of interest even when exact counts of prey species eaten are not fully observed.
+
+% To do this, we estimate the rate at which a predator consumes the prey of interest instead of estimating the proportion of consumed to available prey.
 Outcomes of our model, are that we give up the somewhat arbitrary preference for an index to have the range $[-1,1]$, and random feeding, now denoted by $1$, is formally testable across time points and across prey species.
 
-Our model enables formal hypothesis testing and statistical inference, while being general enough to perform statistical tests across multiple species and time points.
-This provides researchers a more detailed analysis of the predator's feeding preferences.
-Further, because our model is based on underlying Poisson distributions, members of the well studied exponential family, we are able to estimate the parameters of interest even when exact counts of each prey species eaten within any given time period are not observed.
+Our model is based on two Poisson distributions with parameters describing the rates at which the predator consumes prey species and the rate at which the predator encounters prey species.
+Because the Poisson distirbution is a member of the well studied exponential family, we are able to estimate the parameters of interest even when exact counts of each prey species eaten within any given time period are not observed.
 Instead, we rely on the researcher being able to detect prey DNA within the predator's gut \citep{Schmidt:2014,Raso:2014,Madduppa:2014} and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.
@@
 These five hypotheses each allow $c_{st}$ to vary by time, prey species, both, or neither.
-Because the five hypotheses are nested, a natural testing order is suggested in Figure~\ref{fig:hier}.
 
@@
   \centering
-  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2.8cm, semithick]
-
-    \node[state] (A)                    {$1$};
-    \node[state] (B) [below of=A]       {$c$};
-    \node[state] (C) [below left of=B]  {$c_s$};
-    \node[state] (D) [below right of=B] {$c_t$};
-    \node[state] (E) [below left of=D]  {$c_{st}$};
-
-    \path       (B) edge node {} (A)
-                (C) edge node {} (B)
-                (D) edge node {} (B)
-                (E) edge node {} (C)
-                (E) edge node {} (D);
-  \end{tikzpicture}
-
+  \includegraphics{hyps.pdf}
   \caption{This hierarchy of hypotheses suggests the order in which the discussed models should be tested.
@@
 \end{figure}
+
+Because the five hypotheses are nested, a natural testing order is suggested in Figure~\ref{fig:hier}.
+For instance, suppose interest lies in the predator's dietary preferences with respect to prey species $s_1,s_2$ across $3$ time periods.
+If it is believed that hypothesis $4.$ best describes the predator's eating, i.e.\ that the prey species are equally preferred, such that $s_1,s_2$ can be thought of as one prey, but that this preference changes across time, then
+
+\begin{align*}
+  H_0&: \lambda_t = c_t \gamma_t, \forall t \\  
+  H_1&: \lambda_{st} = c_{st}\gamma_{st}
+\end{align*}
+
+would test, statistically, if the data provide evidence for such a belief beyond the model that treats each prey species $s_1$ and $s_2$ individually across the $3$ time points.  Similarly,
+
+\begin{align*}
+  H_0&: \lambda_s = c_s \gamma_s, \forall s \\  
+  H_1&: \lambda_{st} = c_{st}\gamma_{st}
+\end{align*}
+
+would test if $P$'s eating varies by prey species, but not by time, against the most parameter rich model $c_{st}$.
+The above two tests should be performed before any of the simpler methods are tested.
 
@@
 
-The model developed here allows for the determination of a predator's eating preferences by testing simultaneously across an array of multiple prey species and time points.
-This is achieved via a simple, but statistically powerful, likelihood ratio test.
-Further testing of the ratio of rates for which predators eat to encounter prey species allows researchers to make specific conclusions about predators' preferences.
-For instance, rates across time can be estimated to make statements about seasonal effects on a predator's eating habits, while relative rates across species groups allows for statements about the relative preferences for different species.
+All the methods mentioned here, whether they be others' methods or our model, are susceptible to a number of issues that must be handled as best as possible by the investigator.
+The most obvious are sampling issues of prey and predator species of interest.
+It is almost impossible to know how well one's data represents the habitat and the availability of the prey of interest.
+Variables not recorded when collecting data could have influenced the relative abundance of species that were caught.
+And yet this is just the beginning.
+Other practical challenges of analysing the data once in the lab provide immense opportunity for data to be comprimised.
+We therefore, in this paper, focus solely on the methodologies themselves, assuming the best of possible scenarios brought your data to our model.
 
-When counts of predators' gut contents are not fully observed, and instead only a binary response indicating the existence of the prey species in the gut is observed, we are able to treat the counts as missing data.
-By modelling all of the observed data, both the binary responses and the number of prey species caught, and the missing count data, we are able to use the EM algorithm to extract as much information from the data as possible.
+The earliest attempts to estimate predators' dietary preferences, summarised by \citet{Lechowicz:1982} and \citet{Manly:2002},  focused on producing one number summaries that rarely considered more than one prey species of interest, did not take into account changes across time, and had minimal statistical justification.
+Instead, the indices developed were justified by arguing in favor of each index's unique, and claimed ``optimal,'' properties.
+We feel these indices missed the point; an index is only as good as it is understood to estimate and model our world of observable data.
+
+A more recent and notable method to esimate predator's preferences uses a Monte Carlo, or bootstrap to be more accurate, approach \citep{Agusti:2003}.
+This method deserves credit for its lack of distributional assumptions, and yet it still misses the benefits of formally modelling multiple prey species and multiple time points.
+Further, it should be noted that this Monte Carlo approach though creative and widely applicable, easily confuses a key statistical idea: re-sampling does not in any way increase one's sample size or the information contained there within.
+
+The model presented here is a formal statistical model, which enables hypothesis testing of multiple prey species across an array of time points.
+When appropriately fit to data our methods offer greater statistical power than any index or distribution-free Monte Carlo method could.
+We were able to justify our method with robust simulations that provide insight into the statistical properties of our model.
+Moreover, we offer the means to estimate parameters of interest even when the predator's gut contents are not fully observed.
+
+Throughout this paper we assume the observed data were generated from Poisson distributions.
+This enabled us to use the expectation-maximisation algorithm to fit our model when only incomplete data is observed.
+Of course other distributional assumptions could be made in the case of fully observed count data, but it is not clear if this is true in the case of the unobserved count data.
+For instance, consider simply replacing the Poisson distribution with a distribution with more than one parameter.
+Since, in the case of non-count data, only $0$s and $1$s are observed, an underlying distribution with more than one parameter would be over-identified.
+The only other option is to specify a distribution that better fits the observed data than the Poisson distribution, but here one needs to be careful about underyling processes that are being modelled.
+
+Need words about users of these methods.
 
@@
 
-\section{Acknowledgements}
-\label{sec:acknowledge}
-
+\begin{acknowledgements}
 The information reported in this paper (No.\ $15-08-008$) is part of a project of the Kentucky Agricultural Experiment Station and is published with the approval of the Director.
 Support for this research was provided by the University of Kentucky Agricultural Experiment Station State Project KY008055 and the National Science Foundation Graduate Research Fellowship Program.
+\end{acknowledgements}
+% \section{Data Accessibility}
+% \label{sec:accessibility}
 
-\section{Data Accessibility}
-\label{sec:accessibility}
-
-An \texttt{R} package, named \texttt{spiders}, is available on CRAN at \url{http://cran.r-project.org/web/packages/spiders/index.html} and fits all the methods discussed above.
+% An \texttt{R} package, named \texttt{spiders}, is available on CRAN at \url{http://cran.r-project.org/web/packages/spiders/index.html} and fits all the methods discussed above.
 \nocite{Wickham:2011a,Wickham:2011,Bischl:2014}
 
-\bibliographystyle{plainnat}
+\bibliographystyle{spbasic}
 \bibliography{refs}
 \end{document}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2015-04-03 R/summary.predPref.R
summary method fixed

@@ -68,4 +68,5 @@ print.summary.predPref <- function(x) {
     colnames(est) <- c('estimate', 'Std. Err.')
     rownames(est) <- rn
     printCoefmat(est)
+    invisible(x)
 }

### 2015-04-07 paper/main.tex
respond to reviewer's comments, Edward's take 1

@@
 These indices, though intuitive, lack the statistical rigour of a full model, focus on a snapshot in time, and rarely allow more than one prey species to be considered \citep{Lechowicz:1982}.
-Other authors \citep{Davey:2013,King:2010} are using the Monte Carlo based method of \citet{Agusti:2003}, but this also neglects to take into account multiple species across multiple time points.
+Other authors \citep{Davey:2013,King:2010} are using the Monte Carlo based method of \citet{Agusti:2003}, but this also neglects to take into account multiple prey species across multiple time points.
 We propose an intuitive statistical model to estimate and statistically test differences in a predators' prey preferences across an array of time points and between multiple prey species.
@@
 Much of this work however focues on general resource selection and doesn't directly address the problems of food selection.
-Our model, while focusing on predators' food selection not general resource selection, offers formal hypothesis testing and inference similar to the GLMs of \citet{Manly:2002}, but also provides meaningful single number summaries of the predator's dietary preferences.
-Thus, we are proposing here a model that 1) maintains the intuitiveness of the indices summarised by \citet{Lechowicz:1982, Manly:2002}, 2) offers the formal statistical justification of GLMs, 3) models multiple prey species across multiple time points, and 4) estimates parameters of interest even when exact counts of prey species eaten are not fully observed.
+Our model, while focusing on predators' food selection, not general resource selection, offers formal hypothesis testing and inference similar to the GLMs of \citet{Manly:2002}, but also provides meaningful single number summaries of the predator's dietary preferences.
+Thus, we are proposing here a model that 1) maintains the intuitiveness of the indices summarised by \citet{Lechowicz:1982, Manly:2002}, 2) offers formal statistical justification similar to GLMs, 3) models multiple prey species across multiple time points, and 4) estimates parameters of interest even when exact counts of prey species eaten are not fully observed.
 
@@
 Because the Poisson distirbution is a member of the well studied exponential family, we are able to estimate the parameters of interest even when exact counts of each prey species eaten within any given time period are not observed.
-Instead, we rely on the researcher being able to detect prey DNA within the predator's gut \citep{Schmidt:2014,Raso:2014,Madduppa:2014} and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.
+When exact counts are unobserved, we instead rely on the researcher being able to detect prey DNA within the predator's gut \citep{Schmidt:2014,Raso:2014,Madduppa:2014} and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.
 
 This paper is organised as follows.
-Section~\ref{sec:methods} describes our statistical model, for both fully observed count data, and for the non-observed count data for which we use the expectation-maximisation (EM) algorithm, and the statistical tests used to make statements about the population parameters of interest.
+Section~\ref{sec:methods} describes our statistical model, for both fully observed count data and for the non-observed count data for which we use the expectation-maximisation (EM) algorithm to estimate parameters of interest, and the statistical tests used to make statements about the parameters of interest.
 In Section~\ref{sec:sim}, we offer a simulation study that demonstrates the accuracy of our methods.
@@
 
-To illustrate these methods, we analysed a data set that was collected to investigate the feeding preferences of two species of wolf spider, Schizocosa ocreata and Schizocosa stridulans (Araneae: Lycosidae).
+To illustrate these methods, we analysed a data set collected to investigate the feeding preferences of two species of wolf spider, Schizocosa ocreata and Schizocosa stridulans (Araneae: Lycosidae).
 Every $6-12$ days, $10$ to $40$ spiders were hand-collected between October $2011$ and April $2013$ within Berea College Forest in Madison County, Kentucky, USA.
@@
 Using a level of significance of $0.05$, and after making a Bonferroni multiple comparisons adjustment, the data can not say that the two prey are differently preferred in October, November, and December of $2011$ and for March and July of $2012$.
+{\color{red} What does this tell us/someone?
+Just a quick sentence.}
 
@@
 Other practical challenges of analysing the data once in the lab provide immense opportunity for data to be comprimised.
-We therefore, in this paper, focus solely on the methodologies themselves, assuming the best of possible scenarios brought your data to our model.
+We therefore, in this paper, focused solely on the methodologies themselves, assuming the best of possible scenarios brought your data to our model.
 
-The earliest attempts to estimate predators' dietary preferences, summarised by \citet{Lechowicz:1982} and \citet{Manly:2002},  focused on producing one number summaries that rarely considered more than one prey species of interest, did not take into account changes across time, and had minimal statistical justification.
+The earliest estimators of predators' dietary preferences, summarised by \citet{Lechowicz:1982} and \citet{Manly:2002},  produced one number summaries that rarely considered more than one prey species of interest, did not take into account changes across time, and had minimal statistical justification.
 Instead, the indices developed were justified by arguing in favor of each index's unique, and claimed ``optimal,'' properties.
@@
 The model presented here is a formal statistical model, which enables hypothesis testing of multiple prey species across an array of time points.
-When appropriately fit to data our methods offer greater statistical power than any index or distribution-free Monte Carlo method could.
-We were able to justify our method with robust simulations that provide insight into the statistical properties of our model.
-Moreover, we offer the means to estimate parameters of interest even when the predator's gut contents are not fully observed.
+When appropriately fit to data our methods, via a likelihood ratio test, offer greater statistical power than any index or distribution-free Monte Carlo methods do.
+Robust simulations provided insight into the statistical properties of our model.
+Moreover, we can estimate parameters of interest even when the predator's gut contents are not fully observed.
 
@@
 This enabled us to use the expectation-maximisation algorithm to fit our model when only incomplete data is observed.
-Of course other distributional assumptions could be made in the case of fully observed count data, but it is not clear if this is true in the case of the unobserved count data.
-For instance, consider simply replacing the Poisson distribution with a distribution with more than one parameter.
-Since, in the case of non-count data, only $0$s and $1$s are observed, an underlying distribution with more than one parameter would be over-identified.
-The only other option is to specify a distribution that better fits the observed data than the Poisson distribution, but here one needs to be careful about underyling processes that are being modelled.
+Of course, other distributional assumptions could be made in the case of fully observed count data, but it is not clear if this is true in the case of the unobserved count data.
+For instance, consider simply replacing the Poisson distribution with a distribution that has more than one parameter.
+Since, in the case of non-count data only $0$s and $1$s are observed, a distribution with more than one parameter would be over-identified.
+The only other option is to specify a distribution that better fits the observed data than the Poisson distribution, but here one needs to be careful about the underyling processes that are being modelled.
 
-Need words about users of these methods.
+{\color{red} So what does our model offer ecologists? Just a quick paragraph.}
 

### 2015-05-11 paper/main.tex
take 1, Agusti paper

@@
 \title{Formal Modelling of Predator Preferences using Molecular Gut-Content Analysis}
-\author{Edward A. Roualdes \and Simon Bonner \and Thomas D. Whitney \and James D. Harwood}
-\institute{E.A. Roualdes \and S. Bonner
+\author{Edward A. Roualdes \and Simon J. Bonner \and Thomas D. Whitney \and James D. Harwood}
+\institute{E.A. Roualdes \and S. J. Bonner
         \at Department of Statistics, University of Kentucky, Rm.~311 Multidisciplinary Science Building, 725 Rose Street, Lexington, KY 40536-0082, \\\email{edward.roualdes@uky.edu} \and
@@
 These indices, though intuitive, lack the statistical rigour of a full model, focus on a snapshot in time, and rarely allow more than one prey species to be considered \citep{Lechowicz:1982}.
-Other authors \citep{Davey:2013,King:2010} are using the Monte Carlo based method of \citet{Agusti:2003}, but this also neglects to take into account multiple prey species across multiple time points.
-We propose an intuitive statistical model to estimate and statistically test differences in a predators' prey preferences across an array of time points and between multiple prey species.
+Other authors are using the Monte Carlo based method of \citet{Agusti:2003} \citep[see][]{Davey:2013,King:2010}, but this is also unable to take into account multiple prey species across multiple time points.
+We propose an intuitive statistical model to estimate and statistically test differences in a predator's prey preferences across an array of time points and between multiple prey species.
 
@@
 The well established literature on GLMs allows for formal hypothesis testing.
-Much of this work however focues on general resource selection and doesn't directly address the problems of food selection.
-Our model, while focusing on predators' food selection, not general resource selection, offers formal hypothesis testing and inference similar to the GLMs of \citet{Manly:2002}, but also provides meaningful single number summaries of the predator's dietary preferences.
-Thus, we are proposing here a model that 1) maintains the intuitiveness of the indices summarised by \citet{Lechowicz:1982, Manly:2002}, 2) offers formal statistical justification similar to GLMs, 3) models multiple prey species across multiple time points, and 4) estimates parameters of interest even when exact counts of prey species eaten are not fully observed.
+Much of this work however focuses on general resource selection and doesn't directly address the problems of food selection.
+Our model focuses specifically on the problem of predators' food selection, offers formal inference and hypothesis testing similar to that of GLMs \citet{Manly:2002}, and provides meaningful single number summaries of the predators' food preferences.
+The model we are proposing 1) maintains the intuitiveness of the indices summarised by \citet{Lechowicz:1982, Manly:2002}, 2) offers formal statistical justification similar to GLMs, 3) accommodates data collected from multiple prey species across multiple time points, and 4) provides parameters estimates even when exact counts of prey species eaten are not fully observed.
 
@@
 
-Our model is based on two Poisson distributions with parameters describing the rates at which the predator consumes prey species and the rate at which the predator encounters prey species.
-Because the Poisson distirbution is a member of the well studied exponential family, we are able to estimate the parameters of interest even when exact counts of each prey species eaten within any given time period are not observed.
+Our model assumes that numbers of each prey species captured and the numbers consumed by an individual predator follow independent Poisson distributions with separate rates that might possibly vary between species and over time.
+We are able to estimate the parameters of interest even when exact counts of each prey species eaten within any given time period are not observed.
 When exact counts are unobserved, we instead rely on the researcher being able to detect prey DNA within the predator's gut \citep{Schmidt:2014,Raso:2014,Madduppa:2014} and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.
@@
 Because the five hypotheses are nested, a natural testing order is suggested in Figure~\ref{fig:hier}.
-For instance, suppose interest lies in the predator's dietary preferences with respect to prey species $s_1,s_2$ across $3$ time periods.
-If it is believed that hypothesis $4.$ best describes the predator's eating, i.e.\ that the prey species are equally preferred, such that $s_1,s_2$ can be thought of as one prey, but that this preference changes across time, then
+For instance, suppose interest lies in the predator's dietary preferences with respect to prey species $s_1$ and $s_2$ across three time periods.
+The hypothesis that the predator has the same preference for the two prey species can be tested with
 
@@
   H_0&: \lambda_t = c_t \gamma_t, \forall t \\  
-  H_1&: \lambda_{st} = c_{st}\gamma_{st}
+  H_1&: \lambda_{st} = c_{st}\gamma_{st}.
 \end{align*}
 
-would test, statistically, if the data provide evidence for such a belief beyond the model that treats each prey species $s_1$ and $s_2$ individually across the $3$ time points.  Similarly,
+The alternative here allows for the predators preferences to vary independently for the two species and across the three time points. Similarly, the hypotheses
 
@@
 
-would test if $P$'s eating varies by prey species, but not by time, against the most parameter rich model $c_{st}$.
+would test if the predator's eating varies by prey species, but not by time, against the most parameter rich model $c_{st}$.
 The above two tests should be performed before any of the simpler methods are tested.
@@
 
-All the methods mentioned here, whether they be others' methods or our model, are susceptible to a number of issues that must be handled as best as possible by the investigator.
-The most obvious are sampling issues of prey and predator species of interest.
-It is almost impossible to know how well one's data represents the habitat and the availability of the prey of interest.
-Variables not recorded when collecting data could have influenced the relative abundance of species that were caught.
-And yet this is just the beginning.
-Other practical challenges of analysing the data once in the lab provide immense opportunity for data to be comprimised.
-We therefore, in this paper, focused solely on the methodologies themselves, assuming the best of possible scenarios brought your data to our model.
+The earliest estimators of predators' dietary preferences, summarised by \citet{Lechowicz:1982} and \citet{Manly:2002}, produced one number summaries that rarely considered more than one prey species, did not take into account changes across time, and had minimal statistical justification.
+Instead, the indices developed were justified by arguing in favor of each index's unique, and claimed ``optimal,'' properties.
+Further, these indices could not handle unobserved count data.
 
-The earliest estimators of predators' dietary preferences, summarised by \citet{Lechowicz:1982} and \citet{Manly:2002},  produced one number summaries that rarely considered more than one prey species of interest, did not take into account changes across time, and had minimal statistical justification.
-Instead, the indices developed were justified by arguing in favor of each index's unique, and claimed ``optimal,'' properties.
-We feel these indices missed the point; an index is only as good as it is understood to estimate and model our world of observable data.
+A more recent method, by \citet{Agusti:2003}, esimates predator's preferences with unobserved count data using a Monte Carlo approach, the bootstrap to be more accurate.
+This method deserves credit for its attempt to do away with distributional assumptions, however we find that the lack of a distribution causes some unresolved problems.
+\citet{Agusti:2003} suggest resampling the data to estimate proprotions of prey species present in the mesocosms\footnote{Mesocosm was defined by \citet{Agusti:2003} and is there assumed representative of the predator's habitat.} and in predators' guts.
+A seemingly intuitive null hypothesis to test random feeding states that pairs of proportions, relative to a given prey species, are equal.
+For instance, if the proportion of prey species $j$ in the mesocosm is denoted $\pi_j$, and $p_j$ denotes the proportion of prey species $j$ in the predator's gut, then their null hypothesis is written $\pi_j = p_j, \, \forall j$.
+Since the $\pi_j$s depend on the unknown number of prey species consumed, this null hypothesis is not easily tested.
+Even if one assumed known the distribution of prey species consumed, this assumed distribution would depend on unkown parameters, thus only complicating any derived null hypothesis further.
+Exactly what \citet{Agusti:2003} are testing in their paper is unclear.
 
-A more recent and notable method to esimate predator's preferences uses a Monte Carlo, or bootstrap to be more accurate, approach \citep{Agusti:2003}.
-This method deserves credit for its lack of distributional assumptions, and yet it still misses the benefits of formally modelling multiple prey species and multiple time points.
-Further, it should be noted that this Monte Carlo approach though creative and widely applicable, easily confuses a key statistical idea: re-sampling does not in any way increase one's sample size or the information contained there within.
-
-The model presented here is a formal statistical model, which enables hypothesis testing of multiple prey species across an array of time points.
-When appropriately fit to data our methods, via a likelihood ratio test, offer greater statistical power than any index or distribution-free Monte Carlo methods do.
+The model presented here is a formal statistical model, which enables clear and intuitive hypothesis testing of multiple prey species across an array of time points.
+Our methods, via a likelihood ratio test, offer statistically powerful conclusions about the hypotheses of interest.
 Robust simulations provided insight into the statistical properties of our model.
@@
 
-Throughout this paper we assume the observed data were generated from Poisson distributions.
-This enabled us to use the expectation-maximisation algorithm to fit our model when only incomplete data is observed.
-Of course, other distributional assumptions could be made in the case of fully observed count data, but it is not clear if this is true in the case of the unobserved count data.
-For instance, consider simply replacing the Poisson distribution with a distribution that has more than one parameter.
-Since, in the case of non-count data only $0$s and $1$s are observed, a distribution with more than one parameter would be over-identified.
-The only other option is to specify a distribution that better fits the observed data than the Poisson distribution, but here one needs to be careful about the underyling processes that are being modelled.
+Throughout this paper we assume the observed data were generated from underlying Poisson distributions.
+Though this assumption is not easily tested, and other factors affecting our model may distort our methods, e.g.\ sampling issues, environmental variables, differential decay rates, we do not believe a better assumption is readily available to handle all the cases we considered.
+The Poisson distribution provides reasonable biological interpretation to the available data, data obtained by ecologists interested in prey electivity.
+One could hypothesize a more complex distribution that similarly has good biological interpretation, say a distribution with more than one parameter.
+This may perform better than our model when full count data are observed.
+But when only unobserved count data is available, any distribution with more than one parameter is unidentifiable.
+Further, because the Poisson distribution {\color{red} SAY SOMETHIGN}, the EM algorithm is theoretically guaranteed to converge to the parameters of interest {\color{red} CITATION?
+Hoeting and Givens}.
+Future development of our methods might consider more complex distributions to handle variables not presently considered in the case of fully observed count data.
 
-{\color{red} So what does our model offer ecologists? Just a quick paragraph.}
+{\color{red} So what does our model offer ecologists? Multiple prey species and time points, contrasts, unobserved count data.  Just a quick paragraph. It should also mention our package.}
 
-Further developments of our model could take into account other environmental variables that might affect a predator's eating habits, such as rain or temperature.
+% An \texttt{R} package, named \texttt{spiders}, is available on CRAN at \url{http://cran.r-project.org/web/packages/spiders/index.html} and fits all the methods discussed above.
+\nocite{Wickham:2011a,Wickham:2011,Bischl:2014}
 
@@
 \end{acknowledgements}
-% \section{Data Accessibility}
-% \label{sec:accessibility}
-
-% An \texttt{R} package, named \texttt{spiders}, is available on CRAN at \url{http://cran.r-project.org/web/packages/spiders/index.html} and fits all the methods discussed above.
-\nocite{Wickham:2011a,Wickham:2011,Bischl:2014}
 

### 2015-05-11 paper/main.tex
more changes in Discussion

@@
 
-Throughout this paper we assume the observed data were generated from underlying Poisson distributions.
+Throughout this paper we assume the complete data were generated from underlying Poisson distributions.
 Though this assumption is not easily tested, and other factors affecting our model may distort our methods, e.g.\ sampling issues, environmental variables, differential decay rates, we do not believe a better assumption is readily available to handle all the cases we considered.
@@
 One could hypothesize a more complex distribution that similarly has good biological interpretation, say a distribution with more than one parameter.
-This may perform better than our model when full count data are observed.
+Such a distribution could theoretically perform better than our model when full count data are observed and the sampled data were not generated from a Poisson distribution -- if the data were generated from a Poisson distribution then our likelihood ratio test ensures we are using the most powerful test available.
 But when only unobserved count data is available, any distribution with more than one parameter is unidentifiable.
-Further, because the Poisson distribution {\color{red} SAY SOMETHIGN}, the EM algorithm is theoretically guaranteed to converge to the parameters of interest {\color{red} CITATION?
-Hoeting and Givens}.
+The Poisson distribution on the other hand allowed us to use a relatively simple EM algorithm to estimate parameters of interest when only unobserved count data are available.
 Future development of our methods might consider more complex distributions to handle variables not presently considered in the case of fully observed count data.

### 2015-05-18 paper/main.tex
some concluding words for the discussion

@@
 Using a level of significance of $0.05$, and after making a Bonferroni multiple comparisons adjustment, the data can not say that the two prey are differently preferred in October, November, and December of $2011$ and for March and July of $2012$.
-{\color{red} What does this tell us/someone?
-Just a quick sentence.}
 
@@
 
-A more recent method, by \citet{Agusti:2003}, esimates predator's preferences with unobserved count data using a Monte Carlo approach, the bootstrap to be more accurate.
-This method deserves credit for its attempt to do away with distributional assumptions, however we find that the lack of a distribution causes some unresolved problems.
-\citet{Agusti:2003} suggest resampling the data to estimate proprotions of prey species present in the mesocosms\footnote{Mesocosm was defined by \citet{Agusti:2003} and is there assumed representative of the predator's habitat.} and in predators' guts.
-A seemingly intuitive null hypothesis to test random feeding states that pairs of proportions, relative to a given prey species, are equal.
-For instance, if the proportion of prey species $j$ in the mesocosm is denoted $\pi_j$, and $p_j$ denotes the proportion of prey species $j$ in the predator's gut, then their null hypothesis is written $\pi_j = p_j, \, \forall j$.
-Since the $\pi_j$s depend on the unknown number of prey species consumed, this null hypothesis is not easily tested.
-Even if one assumed known the distribution of prey species consumed, this assumed distribution would depend on unkown parameters, thus only complicating any derived null hypothesis further.
-Exactly what \citet{Agusti:2003} are testing in their paper is unclear.
-
-The model presented here is a formal statistical model, which enables clear and intuitive hypothesis testing of multiple prey species across an array of time points.
-Our methods, via a likelihood ratio test, offer statistically powerful conclusions about the hypotheses of interest.
-Robust simulations provided insight into the statistical properties of our model.
-Moreover, we can estimate parameters of interest even when the predator's gut contents are not fully observed.
+A more recent method for testing prey preferences was presenteby by Agusti et al (2003).
+This method attempts to test the hypothesis that individuals are eating prey at random using a test based on Monte Carlo simulation (more specificially the bootstrap).
+Unfortunately, we believe that the method described is flawed specifically because it does not attempt to model the distribution of the number of prey items eaten, as we have done.
+The null hypothesis tested by Agusti et all (2003) is that the proportion of prey species $j=1,\ldots,J$ in the mesocosm, denoted $\pi_j$, is equal to the proportion of prey species $j$ in the predator's guts, denoted by $p_j$, at all of the sampling times.
+This is tested by repeatedly simulating the gut contents of spiders in each sample under this hypothesis and then comparing summary statistics computed from the observed data and the simulated data to generate a $p$-value.
+The problem with this approach is that the relationship between $\pi_j$ and $p_j$ depends on the number of prey items eaten by each predator.
+If each predator had eaten exactly one prey item then the null hypothesis of Agusti et al (2003) would be equivalent to random sampling of prey.
+However, the relationship between $\pi_j$ and $p_j$ is more complicated when predators can have more than one prey item in their gut.
+Given that a predator has $m$ prey items in its gut, randomly sampled from the mesocosm, the marginal probability that at least one item from prey species $j$ is present is $1-(1-\pi_j)^m$.
+The marginal probability that an individual has as least one item from prey species $j$ in its gut is then $\sum_{m=1}^{\infty} (1-(1-\pi_j)^m)P(m)$, where $P(m)$ is the probability that a predator has a total of $m$ species in its gut.
+The presence of species in a predators gut contents also cannot be simulated without modeling the number of items in the gut for exactly the same reason.
+This shows the importance of the distribution of prey items in the gut in testing the hypothesis of random sampling and because of this it is unclear what hypothesis is being tested by Agusti et al (2003).
+Monte Carlo simulation methods similar to this could be implemented using the output from our model, but these would be computationally intensive in comparison to the likelihood ratio tests which can be computed almost immediately.
 
@@
 
-{\color{red} So what does our model offer ecologists? Multiple prey species and time points, contrasts, unobserved count data.  Just a quick paragraph. It should also mention our package.}
+Our methods offer ecologists a formal statistical framework to model and test predators prey preferences across an array of time points and prey species.
+Because we build from intuitive assumptions and well behaved statistical distributions, we were able to extend our methods to unobserved count data.
+For both cases, observed or unobserved count data, a computationally efficient likelihood ratio test provides the most powerful test one can hope for.
+No methods before have offered such abilities.
+These methods are available to researchers via the \texttt{R} package \texttt{spiders}.
 

### 2015-05-19 paper/main.tex
final touches discussion; v1.1 ready for CRAN

@@
 
-Our methods offer ecologists a formal statistical framework to model and test predators prey preferences across an array of time points and prey species.
-Because we build from intuitive assumptions and well behaved statistical distributions, we were able to extend our methods to unobserved count data.
-For both cases, observed or unobserved count data, a computationally efficient likelihood ratio test provides the most powerful test one can hope for.
-No methods before have offered such abilities.
-These methods are available to researchers via the \texttt{R} package \texttt{spiders}.
+The methods we have presented offer ecologists a formal statistical framework to model and test predators feeding preferences across an array of time points and prey species.
+Building from intuitive assumptions, we have developed methodologies for analyzing counts of prey items consumed and have extended these methods for analyzing simple presence/absence data.
+We have provided computationally efficient methods for fitting these models using the EM algorithm and developed powerful tests for a hierarchy of hypotheses based on the likelihood ratio.
+We are not aware of previous methods that have offered such abilities.
+These methods are available to researchers via the R package spiders which is available at \url{http://cran.r-project.org/web/packages/spiders/index.html}.
 
-% An \texttt{R} package, named \texttt{spiders}, is available on CRAN at \url{http://cran.r-project.org/web/packages/spiders/index.html} and fits all the methods discussed above.
-\nocite{Wickham:2011a,Wickham:2011,Bischl:2014}
+\nocite{Wickham:2011a,Wickham:2011,Bischl:2014,Roualdes:2014}
 

### 2015-05-19 R/estimateC.R

@@ -73,12 +73,17 @@ estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
         Info[1,-1] <- unlist(-J/(expl-1) + l*tmp) # fill in off diags; upper tri only
         lowmat <- lower.tri(Info)
         Info[lowmat] <- t(Info)[lowmat]      # make symmetric from upper tri
+        tryCatch(var <- solve(Info),
+                 error=function(e) {
+                     print("Variances not calculated; system is singular.")
+                     var <- NULL
+                 })
 
         ## calc log-lik with est params
         loglik <- llEM(Xdst, Ydst, NA, gammaHat, J, I, as.numeric(cHat))
         
         list('c' = cHat, 'gamma' = as.matrix(gammaHat), 'em_iters' = em_iter,
-             'll' = loglik, 'var' = solve(Info))
+             'll' = loglik, 'var' = var)
     } else {
         if (BALANCED) {
 

### 2015-05-19 R/hypotheses.R

@@ -23,7 +23,9 @@ calcHypotheses <- function(hyp, Xdst, Ydst, J, I, balanced, EM, em_maxiter) {
     alt <- fns[[hyp[2]]](Xdst, Ydst, J, I, EM, em_maxiter, balanced)        
 
     ## calculate degrees of freedom
-    S <- ncol(Xdst); T <- nrow(Xdst); ST <- S*T
+    S <- ncol(Xdst)
+    T <- nrow(Xdst)
+    ST <- S*T
     nullDF <- length(null[['c']])
     altDF <- ifelse(is.null(alt[['c']]), ST, length(alt[['c']]))
     df <- altDF - nullDF

### 2015-05-19 R/contrasts.R
clarify testC, hopefully; update docs

@@ -33,13 +33,14 @@
 ##'
 ##' Of the two estimated hypotheses in the underlying call
 ##' to \code{\link{predPref}}, the linear transformation b is applied to the
-##' hypothesis that is determined by the choice of conf.level.  
+##' hypothesis that is determined by the choice of \code{sig.level}.
 ##'
 ##' @param x a predPref object as fit by the eponymous function
 ##' @param b a vector to linearly transform c_st
 ##' @param mu a number to test the linear contrast against in the null
 ##' @param alternative string to specify alternative hypothesis
 ##' @param conf.level confidence level of the interval
+##' @param sig.level determines null/alternative hypothesis value of c_st from predPref
 ##' 
 ##' @examples
 ##' # set parameters
@@ -49,15 +50,16 @@
 ##' g <- matrix(sqrt(2), nrow=Times, ncol=PreySpecies)     # gamma
 ##' l <- matrix(seq(0.4,1.8,length.out=5)*sqrt(2), nrow=Times, ncol=PreySpecies) # ct
 ##'
-##' # fit model
+##' # fit model and contrast
 ##' \dontrun{
-##' fdata <- simPref(PreySpecies, Times, Predators, Traps, l, g, EM=FALSE)
+##' set.seed(0)
+##' fdata <- simPref(PreySpecies, Times, Predators, Traps, l, g, EM=FALSE) # p-value=0.305
 ##' pref <- predPref(fdata$eaten, fdata$caught, hypotheses=c('ct', 'cst'))
-##' testC(pref, b = c(0,1, -1, 0, 0))
+##' testC(pref, b = c(0,1, -1, 0, 0)) # p-value > sig.level => ct is used, not cst
 ##' }
 ##' 
 ##' @export
-testC <- function(x, b, mu = 0, alternative = c("two.sided", "less", "greater"), conf.level = 0.95) {
+testC <- function(x, b, mu = 0, alternative = c("two.sided", "less", "greater"), conf.level = 0.95, sig.level=0.05) {
     
     ## some check on input; stolen from t.test
     if (!missing(mu) && (length(mu) != 1 || is.na(mu))) 
@@ -67,8 +69,7 @@ testC <- function(x, b, mu = 0, alternative = c("two.sided", "less", "greater"),
         stop("'conf.level' must be a single number between 0 and 1")
 
     ## get appropriate estimates
-    alpha <- 1-conf.level
[…]

### 2015-06-01 paper/main.tex
clarify testC, hopefully; update docs
last edits before resubmission

@@
 
-% \subsection*{Keywords}
-% 
 \maketitle
@@
 A comprehensive overview by \citet{Lechowicz:1982}, later summarised by \citet{Manly:2002}, details some of the benefits and faults of the most popular indices.
-According to these reviews, a majority of the indices give comparable results, save Strauss's linear index $L$, despite the fact that most of the methods differ by range and linearity of response.
+According to these reviews, a majority of the indices give comparable results, save the linear index $L$ by \citet{Strauss:1979}, despite the fact that most of the methods differ by range and linearity of response.
 While Lechowicz recommends one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best'' \citep{Lechowicz:1982}, albeit imperfect, index, Manly et al.\ instead take the approach of excluding the subset of indices which do not ``estimate any biologically meaningful value'' \citep{Manly:2002}.
@@
 The well established literature on GLMs allows for formal hypothesis testing.
-Much of this work however focuses on general resource selection and doesn't directly address the problems of food selection.
-Our model focuses specifically on the problem of predators' food selection, offers formal inference and hypothesis testing similar to that of GLMs \citet{Manly:2002}, and provides meaningful single number summaries of the predators' food preferences.
-The model we are proposing 1) maintains the intuitiveness of the indices summarised by \citet{Lechowicz:1982, Manly:2002}, 2) offers formal statistical justification similar to GLMs, 3) accommodates data collected from multiple prey species across multiple time points, and 4) provides parameters estimates even when exact counts of prey species eaten are not fully observed.
-
-% To do this, we estimate the rate at which a predator consumes the prey of interest instead of estimating the proportion of consumed to available prey.
-Outcomes of our model, are that we give up the somewhat arbitrary preference for an index to have the range $[-1,1]$, and random feeding, now denoted by $1$, is formally testable across time points and across prey species.
+Much of this work focuses on general resource selection and doesn't directly address the problems of food selection.
+By combining the food selectivity literature with formal statistical modelling, our new method 1) focuses specifically on the problem of predators' food selection, while maintaining the intuitiveness of the indices summarised by \citet{Lechowicz:1982} and \citet{Manly:2002}, 2) offers formal statistical justification, enabling proper inference and hypothesis testing similar to that of GLMs \citep{Manly:2002}, 3) accommodates data collected from multiple prey species across multiple time points, 4) provides parameter estimates even when exact counts of prey species eaten are not fully observed, and 5) provides meaningful single number summaries of the predators' food preferences.
 
@@
 
-To illustrate these methods, we analysed a data set collected to investigate the feeding preferences of two species of wolf spider, Schizocosa ocreata and Schizocosa stridulans (Araneae: Lycosidae).
+To illustrate these methods, we analysed a data set collected to investigate the feeding preferences of two species of wolf spider, \textit{Schizocosa ocreata} and \textit{Schizocosa stridulans} (Araneae: Lycosidae).
 Every $6-12$ days, $10$ to $40$ spiders were hand-collected between October $2011$ and April $2013$ within Berea College Forest in Madison County, Kentucky, USA.
@@
 In parallel, we also surveyed availability of forest floor prey using pitfall traps $(n = 32)$.
-For the analysis, both species of Schizocosa were pooled and the number of spiders and prey were analysed by month.
+For the analysis, both species of \textit{Schizocosa} were pooled and the number of spiders and prey were analysed by month.
 On average, $69$ spiders, $111$ Diptera, and $297$ Collembola were caught in each time period.
@@
   \end{tabular}
-  \caption{Targeted prey orders, primer names and sequences, size of amplicon, and source of design for the detection of prey taxa within the guts of Schizocosa spiders.
+  \caption{Targeted prey orders, primer names and sequences, size of amplicon, and source of design for the detection of prey taxa within the guts of \textit{Schizocosa} spiders.
 Both primer sets were used in singleplex PCR assays.}
@@
 Using a level of significance of $0.05$, and after making a Bonferroni multiple comparisons adjustment, the data can not say that the two prey are differently preferred in October, November, and December of $2011$ and for March and July of $2012$.
+In all other time periods, the data suggest that Diptera and Collembola are unequally preferred by wolf spiders.
+Contrasts such as these allow molecular ecologists to identify in greater detail and formally test changes in prey preferences across time points and multiple species.
 
@@
 The earliest estimators of predators' dietary preferences, summarised by \citet{Lechowicz:1982} and \citet{Manly:2002}, produced one number summaries that rarely considered more than one prey species, did not take into account changes across time, and had minimal statistical justification.
-Instead, the indices developed were justified by arguing in favor of each index's unique, and claimed ``optimal,'' properties.
+Instead, they were justified by arguing in favor of each index's unique, and claimed ``optimal,'' properties.
 Further, these indices could not handle unobserved count data.
 
-A more recent method for testing prey preferences was presenteby by Agusti et al (2003).
+A more recent method for testing prey preferences was developed by Agusti et al (2003).
 This method attempts to test the hypothesis that individuals are eating prey at random using a test based on Monte Carlo simulation (more specificially the bootstrap).
-Unfortunately, we believe that the method described is flawed specifically because it does not attempt to model the distribution of the number of prey items eaten, as we have done.
+Unfortunately, the method described does not attempt to model the distribution of the number of prey items eaten, as we have done, and therefore we provide a methodological advance on the approach described previously.
 The null hypothesis tested by Agusti et all (2003) is that the proportion of prey species $j=1,\ldots,J$ in the mesocosm, denoted $\pi_j$, is equal to the proportion of prey species $j$ in the predator's guts, denoted by $p_j$, at all of the sampling times.
@@
 The marginal probability that an individual has as least one item from prey species $j$ in its gut is then $\sum_{m=1}^{\infty} (1-(1-\pi_j)^m)P(m)$, where $P(m)$ is the probability that a predator has a total of $m$ species in its gut.
-The presence of species in a predators gut contents also cannot be simulated without modeling the number of items in the gut for exactly the same reason.
-This shows the importance of the distribution of prey items in the gut in testing the hypothesis of random sampling and because of this it is unclear what hypothesis is being tested by Agusti et al (2003).
+The presence of species in a predator's gut contents also cannot be simulated without modeling the number of items in the gut for exactly the same reason.
+This demonstrates the importance of knowing the distribution of prey items in the gut in testing the hypothesis of random sampling.
 Monte Carlo simulation methods similar to this could be implemented using the output from our model, but these would be computationally intensive in comparison to the likelihood ratio tests which can be computed almost immediately.
@@
 Throughout this paper we assume the complete data were generated from underlying Poisson distributions.
-Though this assumption is not easily tested, and other factors affecting our model may distort our methods, e.g.\ sampling issues, environmental variables, differential decay rates, we do not believe a better assumption is readily available to handle all the cases we considered.
+Though this assumption is not easily tested, and other factors affecting our model may distort our methods (e.g.\ sampling issues, environmental variables, differential decay rates), we do not believe a better assumption is readily available to handle all the cases we considered.
 The Poisson distribution provides reasonable biological interpretation to the available data, data obtained by ecologists interested in prey electivity.
@@
 Such a distribution could theoretically perform better than our model when full count data are observed and the sampled data were not generated from a Poisson distribution -- if the data were generated from a Poisson distribution then our likelihood ratio test ensures we are using the most powerful test available.
-But when only unobserved count data is available, any distribution with more than one parameter is unidentifiable.
+However, when only unobserved count data is available, any distribution with more than one parameter is unidentifiable.
 The Poisson distribution on the other hand allowed us to use a relatively simple EM algorithm to estimate parameters of interest when only unobserved count data are available.
@@
 
-The methods we have presented offer ecologists a formal statistical framework to model and test predators feeding preferences across an array of time points and prey species.
+The methods we have presented offer ecologists a formal statistical framework to model and test predator feeding preferences across an array of time points and prey species.
 Building from intuitive assumptions, we have developed methodologies for analyzing counts of prey items consumed and have extended these methods for analyzing simple presence/absence data.

### 2015-09-10 paper/main.tex
first draft revisions encouraged by EES reviewers

@@
 Our model assumes that numbers of each prey species captured and the numbers consumed by an individual predator follow independent Poisson distributions with separate rates that might possibly vary between species and over time.
-We are able to estimate the parameters of interest even when exact counts of each prey species eaten within any given time period are not observed.
+Further, we are able to estimate the rates of interest even when exact counts of each prey species eaten within any given time period are not observed.
 When exact counts are unobserved, we instead rely on the researcher being able to detect prey DNA within the predator's gut \citep{Schmidt:2014,Raso:2014,Madduppa:2014} and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.
@@
 
+% !116
 The likelihood function that allows for estimation of these parameters is as follows.
-Since we assume $X_{jst}$ is independent of $Y_{ist}$ we can simply multiply the respective Poisson probability density functions, and then form products over all $s,t$ to obtain the likelihood
+Since we assume $X_{jst}$ is independent of $Y_{ist}$ we can simply multiply the respective Poisson probability mass functions, and then form products over all $s,t$ to obtain the likelihood
 
@@
 \[
-l_{comp}(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log f_{X,Y,Z}(\boldsymbol x, \boldsymbol y, \boldsymbol z|\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \sum_{s,t}^{S,T} \left[ \sum_{j=1}^{J_t} \log f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda}) + \sum_{i=1}^{I_t}\log f_Y(y_{jst}|\boldsymbol{\gamma}) \right].
+l_{comp}(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log f_{X,Y,Z}(\boldsymbol x, \boldsymbol y, \boldsymbol z|\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \sum_{s,t}^{S,T} \left[ \sum_{j=1}^{J_t} \log f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda}) + \sum_{i=1}^{I_t}\log f_Y(y_{ist}|\boldsymbol{\gamma}) \right].
 \]
 
-The density of $Y_{jst}$ is exactly as in Section~\ref{sec:count} and so we focus on deriving the joint density of $X_{jst}$ and $Z_{jst}$.
-With the distribution of $Z_{jst}$ given above, we can compute $f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda})$ by noting that $X_{jst}=0$ with probability 1 if $Z_{jst}=0$, and that $[X_{jst}|Z_{jst}=0]$ has a truncated Poisson distribution with density
+% !143, 146
+The mass function of $Y_{jst}$ is exactly as in Section~\ref{sec:count} and so we focus on deriving the joint probability mass function of $X_{jst}$ and $Z_{jst}$.
+With the distribution of $Z_{jst}$ given above, we can compute $f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda})$ by noting that $X_{jst}=0$ with probability 1 if $Z_{jst}=0$, and that $[X_{jst}|Z_{jst}=0]$ has a truncated Poisson distribution with mass function
 \[
@@
 \]
-\noindent The joint density of $X_{jst}, Z_{jst}$ is then 
+% !147
+\noindent The joint probability mass function of $X_{jst}, Z_{jst}$ is then 
 \begin{equation*}
@@
   \includegraphics[scale=0.5]{bp}
-  \caption{Shown are all $500$ estimates, cent-red at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$ with sample sizes small and medium, respectively.}
+  \caption{Shown are all $500$ estimates, centred at the true parameter values, from fitting the true model to the data generated from models $c_s,c_t$ with sample sizes small and medium, respectively.}
   \label{fig:bp}
@@
 As expected, $\hat c_5$ increased without bound as we refit the model with increasingly small values of $\tau$.
+Thus, conditional on a mixture of $0$s and $1$s in the data the corresponding estimators appear to be unbiased, but when no $0$s exist in the data the theoretical bias is infinite.
+In practice however we imagine that no $0$s in the data seems an unlikely circumstance.
 
@@
 \end{figure}
-
-The over-estimation of parameters, a symptom of the loss of information due to the unobserved counts, can also be seen with box plots of the $500$ point estimates centred at their respective true parameter values.
-Figure~\ref{fig:em_bp} contains box plots of the same scenarios in Figures \ref{fig:smallEM} and \ref{fig:largerEM}.
-For the $73$ cases in which $Z_{js5} = 1$ for all $j,s$ under model $c_t$ with the larger sample size, the bias is infinite since parameter estimates will, theoretically, be infinite.
-The finite bias shown in these plots is due to the finite estimates provided by the termination of the EM algorithm.
-Thus, conditional on a mixture of $0$s and $1$s in the data the corresponding estimators appear to be unbiased, but when no $0$s exist in the data the theoretical bias is infinite.
 
@@
 
+% !344
 Throughout this paper we assume the complete data were generated from underlying Poisson distributions.
@@
 The Poisson distribution provides reasonable biological interpretation to the available data, data obtained by ecologists interested in prey electivity.
-One could hypothesize a more complex distribution that similarly has good biological interpretation, say a distribution with more than one parameter.
+The rate parameters of the Poisson distributions also avoid dependence on estimates of true abundance, a sensitive issue in and of itself.
+Alternatively, one could hypothesize a more complex distribution that similarly has good biological interpretation, say a distribution with more than one parameter.
 Such a distribution could theoretically perform better than our model when full count data are observed and the sampled data were not generated from a Poisson distribution -- if the data were generated from a Poisson distribution then our likelihood ratio test ensures we are using the most powerful test available.
@@
 
-The methods we have presented offer ecologists a formal statistical framework to model and test predator feeding preferences across an array of time points and prey species.
+Beyond the statistical assumptions, sampling issues should be given careful attention.
+In collecting prey of interest, we rely on human-made traps.
+Human-made traps are liable to collect prey at rates unequal to that of the predator of interest.
+Confounded with researchers' abilities to make sufficiently efficient traps are a host of environmental variables.
+Researchers should give consideration to these potentialities and attempt to determine to what degree their traps' efficiency is effected.
+Mathematically, these issues will directly effect the parameter $c_{st} = \lambda_{st}/\gamma_{st}$.
+When human-made traps are relatively more efficient at catching prey than are the predators, $c_{st}$ will be relatively smaller.
+Conversely, when the traps are relatively less efficient, $c_{st}$ will be relatively larger.
+Unfortunately, these concerns are difficult to control, statistically, practically, or otherwise, and are to some degree unavoidable.
+
+The methods we presented offer ecologists a formal statistical framework to model and test predator feeding preferences across an array of time points and prey species.
 Building from intuitive assumptions, we have developed methodologies for analyzing counts of prey items consumed and have extended these methods for analyzing simple presence/absence data.

### 2015-09-15 paper/main.tex
rando changes

--- 2015-06-01 paper/main.tex
@@
 \author{Edward A. Roualdes \and Simon J. Bonner \and Thomas D. Whitney \and James D. Harwood}
-\institute{E.A. Roualdes \and S. J. Bonner
+\institute{E. A. Roualdes \and S. J. Bonner
         \at Department of Statistics, University of Kentucky, Rm.~311 Multidisciplinary Science Building, 725 Rose Street, Lexington, KY 40536-0082, \\\email{edward.roualdes@uky.edu} \and
@@
 \end{abstract}
+
+\begin{acknowledgements}
+The information reported in this paper (No.\ $15-08-008$) is part of a project of the Kentucky Agricultural Experiment Station and is published with the approval of the Director.
+Support for this research was provided by the University of Kentucky Agricultural Experiment Station State Project KY008055 and the National Science Foundation Graduate Research Fellowship Program.
+\end{acknowledgements}
 
@@
 \nocite{Wickham:2011a,Wickham:2011,Bischl:2014,Roualdes:2014}
-
-\begin{acknowledgements}
-The information reported in this paper (No.\ $15-08-008$) is part of a project of the Kentucky Agricultural Experiment Station and is published with the approval of the Director.
-Support for this research was provided by the University of Kentucky Agricultural Experiment Station State Project KY008055 and the National Science Foundation Graduate Research Fellowship Program.
-\end{acknowledgements}
-
 \bibliographystyle{spbasic}

### 2015-09-15 paper/main.tex
Thomas changes in second to last paragraph.

--- 2015-09-10 paper/main.tex
@@
 
-Beyond the statistical assumptions, sampling issues should be given careful attention.
-In collecting prey of interest, we rely on human-made traps.
-Human-made traps are liable to collect prey at rates unequal to that of the predator of interest.
+Beyond the statistical assumptions, sampling of prey availability should be given careful attention.
+To survey prey of interest, we relied on pitfall trapping, which is a reasonable method to quantify potential prey encounters of sit-and-pursue predators, such as wolf spiders \citep{Uetz:1999}.
+In general, however, human-made traps are liable to collect prey at rates unequal to that of the predator of interest.
 Confounded with researchers' abilities to make sufficiently efficient traps are a host of environmental variables.
-Researchers should give consideration to these potentialities and attempt to determine to what degree their traps' efficiency is effected.
-Mathematically, these issues will directly effect the parameter $c_{st} = \lambda_{st}/\gamma_{st}$.
+Researchers should consider these potential sampling issues and attempt to determine, a priori, how well their traps will characterize prey availability according to the biology of their focal predator.
+Mathematically, these issues will directly affect the parameter $c_{st} = \lambda_{st}/\gamma_{st}$.
 When human-made traps are relatively more efficient at catching prey than are the predators, $c_{st}$ will be relatively smaller.
 Conversely, when the traps are relatively less efficient, $c_{st}$ will be relatively larger.
-Unfortunately, these concerns are difficult to control, statistically, practically, or otherwise, and are to some degree unavoidable.
+Unfortunately, these factors can be difficult to control, statistically, practically, or otherwise, and are to some degree unavoidable.
 

### 2015-09-17 paper/main.tex
add mark-recapture comment

@@
 The Poisson distribution provides reasonable biological interpretation to the available data, data obtained by ecologists interested in prey electivity.
-The rate parameters of the Poisson distributions also avoid dependence on estimates of true abundance, a sensitive issue in and of itself.
 Alternatively, one could hypothesize a more complex distribution that similarly has good biological interpretation, say a distribution with more than one parameter.
@@
 Unfortunately, these factors can be difficult to control, statistically, practically, or otherwise, and are to some degree unavoidable.
+Mark-recapture methods might be implemented to adjust for trap efficiency and estimate the true density of each prey species.
+However, recapture rates are likely to be low and this type of data is beyond what is regularly collected in these types of experiments.
 

### 2015-09-24 paper/main.tex
line 367 corrected

@@
 Researchers should consider these potential sampling issues and attempt to determine, a priori, how well their traps will characterize prey availability according to the biology of their focal predator.
-Mathematically, these issues will directly affect the parameter $c_{st} = \lambda_{st}/\gamma_{st}$.
-When human-made traps are relatively more efficient at catching prey than are the predators, $c_{st}$ will be relatively smaller.
-Conversely, when the traps are relatively less efficient, $c_{st}$ will be relatively larger.
-Unfortunately, these factors can be difficult to control, statistically, practically, or otherwise, and are to some degree unavoidable.
+If such issues are steady over time, then mathematically the parameter $c_{st} = \lambda_{st}/\gamma_{st}$ is directly affected.
+When human-made traps are relatively and consistently more efficient at catching prey than are the predators, $c_{st}$ will be relatively smaller.
+Conversely, when the traps are relatively and consistenly less efficient, $c_{st}$ will be relatively larger.
+Unfortunately, these factors can be difficult to control, practically, or otherwise, and are to some degree unavoidable.
 Mark-recapture methods might be implemented to adjust for trap efficiency and estimate the true density of each prey species.
 However, recapture rates are likely to be low and this type of data is beyond what is regularly collected in these types of experiments.
+
+If issues of trap efficiency or abundance are suspected to change across sampling times, we offer an alternative parameterization, details of which are found in Appendix~\ref{sec:efficiency}.
+The ratio $c_{st} = \lambda_{st} / \gamma_{st}$ implicitly treats $\gamma_{st}$ as an index proportional to true abundance.
+This assumption is valid if the traps maintain efficiency across all occassions, but otherwise is incorrect.
+Rewriting $c_{st}$ as $k_{st}d_t$, such that $\sum_{s=1}^S k_{st} = 1$ for all $t$, allows for trap efficiency to vary across time.
+The parameter $d_t$ captures changes in the absolute relationship between $\lambda_{st}$ and $\gamma_{st}$ which may be due to actual changes in abundance or to changes in trap efficiency.
+$k_{st}$ then appropriately measures the ratio of consumption to encounter rates, taking into account changes in abundance or trap efficiency.
 
@@
 
+\appendix
+\section{Varying Trap Efficiency}
+\label{sec:efficiency}
+
+When there is concern about trap efficiency varying across time,  we recommend the following reparameterization of $c_{st} = \lambda_{st}/\gamma_{st}$.  Let
+
+\[ k_{st}d_t = \frac{\lambda_{st}}{\gamma_{st}}, \]
+
+\noindent such that $\sum_{s=1}^S k_{st} = 1$, for all $t$.
+The parameter $d_t$ allows for changes in trap efficiency or in actual abundance over time.
+If such changes are believed to be present, this reparameterization allows for a more natural interpretation.
+The parameter $k_{st}$ models relative consumption to encounter rates, as expected.
+
+Since this reparameterization amounts to a linear transformation of the parameters, the asymptotic distribution for $\hat{d}_t$ can be evaluated directly.
+Because of the constraint $\sum_{s=1}^S k_{st} = 1$, we find that $d_t = \sum_{s=1}^S c_{st}$ or written in matrix notation  $\mathbf{d} = \mathbf{1}' \mathbf{c}$, where $\mathbf{d} \in \mathbb{R}^T$.
+The asymptotic distribution of $\mathbf{\hat{d}}$ follows immediately, $\mathbf{\hat{d}} \stackrel{\cdot}{\sim} \mathcal{N}(\mathbf{1}'\mathbf{d}, \mathbf{1}'\Sigma\mathbf{1})$, where $\Sigma$ is the covariance matrix of the asymptotic distribution of $\mathbf{\hat{c}}$.
+
+More importantly, we can find the asymptotic distribution of $\mathbf{\hat{k}}_t = (k_{1t}, \ldots, k_{St})' \in \mathbb{R}^S$.
+We make use of the delta method \citep{Serfling:2001}.
+The asymptotic variance relies on the derivatives, $d k_{rt} / d c_{st}$.
+Let $c_{\cdot t} = \sum_{s=1}^S c_{st}$.
+Because $k_{st} = c_{st}/ c_{\cdot t}$, we can find the necessary derivatives in two cases
+
+\[
+  \frac{d k_{rt}}{d c_{st}} =
+\begin{cases}
+  \frac{-c_{rt}}{c_{\cdot t}^2}, & r \ne s \\
+  \frac{c_{\cdot t} - c_{st}}{c_{\cdot t}^2}, & r = s.
+\end{cases}
+\]
+
+\noindent  Thus, the asymptotic distribution of $\mathbf{\hat{k}}_{t}$ is $\mathcal{N}(\frac{c_{st}}{c_{\cdot t}}, (\frac{d \mathbf{k}_t}{d \mathbf{c}_t}) \Sigma (\frac{d \mathbf{k}_t}{d \mathbf{c}_t})')$.
 \nocite{Wickham:2011a,Wickham:2011,Bischl:2014,Roualdes:2014}

### 2015-11-18 paper/main.tex
Simon's edits.

@@
 \author{Edward A. Roualdes \and Simon J. Bonner \and Thomas D. Whitney \and James D. Harwood}
-\institute{E. A. Roualdes \and S. J. Bonner
-        \at Department of Statistics, University of Kentucky, Rm.~311 Multidisciplinary Science Building, 725 Rose Street, Lexington, KY 40536-0082, \\\email{edward.roualdes@uky.edu} \and
+\institute{E. A. Roualdes 
+        \at Department of Statistics, University of Kentucky, Rm.~311 Multidisciplinary Science Building, 725 Rose Street, Lexington, KY 40536-0082, \\\email{edward.roualdes@uky.edu}
+        \and S. J. Bonner
+        \at Department of Statistical and Actuarial Sciences, University of Western Ontario, 1151 Richmond Street, London, Ontario, N6A 5B7, Canada
         T.D. Whitney 
@@
 According to these reviews, a majority of the indices give comparable results, save the linear index $L$ by \citet{Strauss:1979}, despite the fact that most of the methods differ by range and linearity of response.
-While Lechowicz recommends one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best'' \citep{Lechowicz:1982}, albeit imperfect, index, Manly et al.\ instead take the approach of excluding the subset of indices which do not ``estimate any biologically meaningful value'' \citep{Manly:2002}.
+While Lechowicz recommends one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best'' \citep{Lechowicz:1982}, albeit imperfect, index, \citet{Manly:2002} instead take the approach of excluding the subset of indices which do not ``estimate any biologically meaningful value'' \citep{Manly:2002}.
 \citet{Lechowicz:1982} recommends the index $E^*$, an element of the \citet{Manly:2002} suggested indices, because the index value $0$ denotes random feeding, the index has a range restricted to $[-1,1]$ (though $E^*=1$ is nigh impossible), and the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the number of available prey types (assumed known).
@@
 Further, we are able to estimate the rates of interest even when exact counts of each prey species eaten within any given time period are not observed.
-When exact counts are unobserved, we instead rely on the researcher being able to detect prey DNA within the predator's gut \citep{Schmidt:2014,Raso:2014,Madduppa:2014} and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.
+When exact counts are unobserved, we instead rely on the researcher being able to detect prey DNA within the predator's gut \citep{Schmidt:2014,Raso:2014,Madduppa:2014} and make a simple binary conclusion: this predator ate some of that prey species during this time period or did not.
 
@@
 Predators and prey species are collected and counted at each time period.
-We denote the number of predators and the number of prey species caught in each time period $t \in \{1, \ldots, T\}$ by $J_t$ and $I_t$, respectively.
+We denote the number of predators and the number of traps in each time period $t \in \{1, \ldots, T\}$ by $J_t$ and $I_t$, respectively.
 Prey species will be indexed by $s \in \{1, \ldots, S \}$.
@@
 \begin{enumerate}
-\item $c_{st} = 1$
 \item $c_{st} = c$
@@
 
-\noindent The first hypothesis states that the relative rate of sampling for the predator and the traps is the same for all species on all occasions.
+\noindent The first hypothesis states that the relative rates of sampling for the predator and the traps are the same for all species on all occasions.
 One imagines this is the case if the prey move randomly and the predator simply eats prey which comes within its reach, thus suggesting no selection for a particular prey item.
-The second states that predators sample prey proportionally across all time periods.
-The third hypothesis states that predators sample different prey species at different rates, but each rate is steady across time.
+The second hypothesis states that predators sample different prey species at different rates, but each rate is steady across time.
 This implies that the predator expresses preferences for one prey species over another, but is unresponsive to changes due to time.
-Conversely, the fourth hypothesis implies that each prey species is sampled similarly within each time period, while the rates across time are allowed to change.
-The fifth hypothesis assumes a predator's selection varies by both time and prey species.
+Conversely, the third hypothesis implies that each prey species is sampled similarly within each time period, while the rates across time are allowed to change.
+This would happen if the average amount of prey consumed by the predators changed over time while the relative rates of sampling remained the same, or if there were changes in the efficiency of the traps, or both.
+The final hypothesis assumes a predator's selection varies by both time and prey species.
 This would make sense if environmental and biological variables, such as weather, prey availability, and/or palatability were affecting predators' selection strategies.
@@
 The hypothesis that the predator has the same preference for the two prey species can be tested with
-
 \begin{align*}
@@
 \end{align*}
-
-The alternative here allows for the predators preferences to vary independently for the two species and across the three time points. Similarly, the hypotheses
-
+The alternative here allows for the predator's preferences to vary independently for the two species and across the three time points. Similarly, the hypotheses
 \begin{align*}
@@
 \end{align*}
-
 would test if the predator's eating varies by prey species, but not by time, against the most parameter rich model $c_{st}$.
-The above two tests should be performed before any of the simpler methods are tested.
+The above two tests should be performed before any of the simpler models are tested.
 
@@
 \noindent Writing all five hypotheses as $\lambda_{st} = c_{st}\gamma_{st}$, we can, in the simplest cases, find analytic solutions for the maximum likelihood estimates (MLEs) of $c_{st}$, $\gamma_{st}$, and by invariance $\lambda_{st}$.
-Such solutions exist for the simplest model, $c_{st}=1$, for which only $\gamma_{st}$ need be estimated, and for the second simplest model, $c_{st}=c$, provided that the data are balanced so that $J_t=J$ and $I_t=I$.
-Respectively, these solutions are
-
+Such solutions exist for the simplest model, $c_{st}=c$, provided that the data are balanced so that $J_t=J$ and $I_t=I$.
+The solutions are
 \begin{equation*}
-  \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_t + I_t}, \quad \text{ and } \quad \hat{c} = \frac{I \sum_{s,t} X_{\cdot st}}{J \sum_{s,t} Y_{\cdot st}}, \quad \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{I \left( \frac{\sum_{st} X_{\cdot st}}{\sum_{st} Y_{\cdot st}} + 1 \right)},
+  % \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{J_t + I_t}, \quad 
+  % \text{ and } \quad 
+  \hat{c} = \frac{I \sum_{s,t} X_{\cdot st}}{J \sum_{s,t} Y_{\cdot st}}
+  \quad
+  \text{ and }
+  \quad
+  \hat{\gamma}_{st} = \frac{X_{\cdot st} + Y_{\cdot st}}{I \left( \frac{\sum_{st} X_{\cdot st}}{\sum_{st} Y_{\cdot st}} + 1 \right)},
 \end{equation*}
-
-\noindent where $X_{\cdot st} = \sum_{j=1}^{J_t}X_{jst}$ and $Y_{\cdot st} = \sum_{i=1}^{I_t} Y_{ist}$.
+where $X_{\cdot st} = \sum_{j=1}^{J_t}X_{jst}$ and $Y_{\cdot st} = \sum_{i=1}^{I_t} Y_{ist}$.
 
@@
 \]
-
 % !143, 146
@@
 % !147
-\noindent The joint probability mass function of $X_{jst}, Z_{jst}$ is then 
+The joint probability mass function of $X_{jst}, Z_{jst}$ is then 
 \begin{equation*}
@@
 
-A more recent method for testing prey preferences was developed by Agusti et al (2003).
+A more recent method for testing prey preferences was developed by \citet{Agusti:2003}.
 This method attempts to test the hypothesis that individuals are eating prey at random using a test based on Monte Carlo simulation (more specificially the bootstrap).
 Unfortunately, the method described does not attempt to model the distribution of the number of prey items eaten, as we have done, and therefore we provide a methodological advance on the approach described previously.
-The null hypothesis tested by Agusti et all (2003) is that the proportion of prey species $j=1,\ldots,J$ in the mesocosm, denoted $\pi_j$, is equal to the proportion of prey species $j$ in the predator's guts, denoted by $p_j$, at all of the sampling times.
+The null hypothesis tested by \citet{Agusti:2003} is that the proportion of prey species $j=1,\ldots,J$ in the mesocosm, denoted $\pi_j$, is equal to the proportion of prey species $j$ in the predator's guts, denoted by $p_j$, at all of the sampling times.
 This is tested by repeatedly simulating the gut contents of spiders in each sample under this hypothesis and then comparing summary statistics computed from the observed data and the simulated data to generate a $p$-value.
 The problem with this approach is that the relationship between $\pi_j$ and $p_j$ depends on the number of prey items eaten by each predator.
-If each predator had eaten exactly one prey item then the null hypothesis of Agusti et al (2003) would be equivalent to random sampling of prey.
+If each predator had eaten exactly one prey item then the null hypothesis of \citet{Agusti:2003} would be equivalent to random sampling of prey.
 However, the relationship between $\pi_j$ and $p_j$ is more complicated when predators can have more than one prey item in their gut.
@@
 We are not aware of previous methods that have offered such abilities.
-These methods are available to researchers via the R package spiders which is available at \url{http://cran.r-project.org/web/packages/spiders/index.html}.
+These methods are available to researchers via the R package spiders which is available at \url{http://cran.r-project.org/web/packages/spiders/index.html} \citep{Roualdes:2014}.
 
@@
 \noindent  Thus, the asymptotic distribution of $\mathbf{\hat{k}}_{t}$ is $\mathcal{N}(\frac{c_{st}}{c_{\cdot t}}, (\frac{d \mathbf{k}_t}{d \mathbf{c}_t}) \Sigma (\frac{d \mathbf{k}_t}{d \mathbf{c}_t})')$.
-\nocite{Wickham:2011a,Wickham:2011,Bischl:2014,Roualdes:2014}
+%\nocite{Wickham:2011a,Wickham:2011,Bischl:2014,Roualdes:2014}
 \bibliographystyle{spbasic}

### 2015-12-09 paper/EEST/First_Submission/Response/EEST-D-15-00056_response.tex
Adding LaTeX version of response to reviewers to git.

\documentclass[12pt]{article}

\usepackage[margin=2.0cm]{geometry}

\begin{document}
\begin{center}
  \begin{Large}
    \textbf{Response to Reviews of EEST-15-00056:}\\
    \textsc{Formal Modelling of Predator Preferences using Molecular Gut-Content
      Analysis}
  \end{Large}
\end{center}

\section{Associate Editor:}

\begin{enumerate}
\item \textit{ ... called the probability mass function.}

Thank you for your attention to detail.  Changes throughout the manuscript have been made accordingly.

\end{enumerate}
\section{Reviewer \#1:}
\begin{enumerate}
\item  \textit{ ... commend the authors on an excellent contribution.}

  Thank you.

\end{enumerate}

\section{Reviewer \#2:}

\subsection{Major Comments}

\begin{enumerate}
\item \textit{ The method estimates the ratio between lambda and gamma, not lambda and gamma themselves. The method does not estimate true abundances of preys, and I am not sure why without estimating the true abundance, why the ratio can be estimated so accurately as shown in the Figures? Could you explain some essence (trick?) of the success?}

% In fact, we estimate gamma and the ratio of lambda and gamma, denoted by $c$, directly.  Estimates of lambda are found via these two quantities and the invariance property of maximum likelihood.  In the case of fully observed data, there is no trick.  We use maximum likelihood methods to estimate parameters of interest, as stated in line 121.  In the case of unobserved count data, this EM algorithm provides us a means to estimate the parameters of interest.  When unobserved counts are present we discuss times at which the EM algorithm will fail.  Though, this discussion itself has been reduced as per Minor Point 4.

This is an important point that we overlooked in our initial submission. We had considered $\gamma_{st}$ as a measure of absolute abundance of prey species $s$ on sampling occasion $t$, and you are absolutely correct that this is not the case. Instead, it represents the mean number of individuals from this prey species caught in the traps. To relate this to absolute abundance would require extra information on the efficiency of the traps that could be obtained from a mark-recapture study, for example, but this information is not available for our particular study and probably not available for many other studies either.

However, we can interpret $\gamma_{st}$ as a relative measure of abundance under some assumptions so that changes in $\gamma_{s_1t}/\gamma_{s_2t}$ represent changes in the relative abundance of species $s_1$ and $s_2$. This is the case if the capture efficiency of the traps is assumed to be constant for all species over all occasions. Under weaker assumptions we can interpret $\gamma_{st}$ as a relative measure of the availability of prey in the environment, rather than abundance, and $c_{st}$ as a relative measure of preference given abundance. We might assume that trap efficiency for a prey species depends on how much of the time individuals in that species are trappable. In this case, trappability might be interpreted as availability to the prey. In our application, prey species that are actively foraging in the leaf litter for longer periods may have higher trappability and also be more available to the spiders. In this case, we can interpret $\gamma_{s_1t}/\gamma_{s_2t}$ as a relative measure of the availability of prey and $c_{st}$ as a relative measure of preference given availability.

With respect to the figures, the apparent trick is simply a result of our incorrect interpretation of $\gamma_{st}$. We generated data for the simulations by simulating the number of individuals from each prey species caught in traps given $\gamma_{st}$. Our model is able to estimate the ratio of $\lambda_{st}$ and $\gamma_{st}$ accurately, but it was incorrect of us to interpret this as estimating absolute abundance accurately.

\item \textit{ The ratio is between the preference observed in gut-content and captured preys in the traps produced by human. The denominator is not the true abundance, but expresses how much that prey species are easily captured by that trapping system. What are ecological implications of the ratio between observed gut contents and the "efficiency" of human-produced traps? Of course, this could be unavoidable for population dynamics of animal species and for food-web studies, and I think that a few sentences about this point should be added as far as a journal is not specified to animal ecology.}

Following from our response to the last point, we have changed the text so it is clear that $\gamma_{st}$ is a measure of relative, not absolute, abundance. We have also added two paragraphs to the Discussion commenting on the issues of trap efficiency and the assumptions needed to interpret $\gamma_{st}$ as a relative measure of abundance.

The second paragraph we have added to the Discussion also introduces a new parameterization that helps to address these concerns. The new parametrization applies when the efficiency of traps cannot be assumed constant and explicitly separates changes in efficiency over time. Appendix A provides the mathematical details behind the new paramterization.

Recognizing $\gamma_{st}$ as a measure of relative abundance also led us to remove the first of our hypotheses about the way in which $c_{st}$ might vary, namely $c_{st} = 1$. This leaves only four hypotheses on the way $c_{st}$ might change. As in our previous response, the relationship between the samples collected from the predators and the traps depends on the traps' efficiency and may be affected by, among other things, how big the traps are.  Suppose that the experiment is repeated on two occasions with exactly the same properties that the traps double in size on the second occasion.  If $c_{st}=1$ on the first occasion then you would expect $c_{st}=.5$ on the second occasion, but the biological process has not changed.  This means that the hypothesis $c_{st} = 1$ does not have the interpretation in terms of absolute abundance that we previously thought.

\end{enumerate}

\subsection{Minor Points:}

\begin{enumerate}
\item  $Y_{jst} => Y_{ist}$ Change made.

\item  Means now presented instead of total counts.  This was not as simple as divide by the number of traps, since traps were left out for different numbers of days in different months.  Nonetheless, this was a good point.

\item  We doubled checked the calculations and stand by our results.  The correct intervals for November 2011 are (2.7, 7.1) for Diptera and (2.0, 3.9) for Collembola.  The overlap of the intervals is due to large variation in the point estimates, and hence we fail to see the difference between the two point estimates.

\item  ... such long descriptions are not needed.

A paragraph was removed, namely that which was the last paragraph in Section 3 Simulation Study.
\end{enumerate}
\end{document}

### 2015-12-09 paper/main.tex
Added the explicit statement of the distribution of $Y_{ist}$ like you had for $X_{jst}$.
Changed the references to five hypotheses to four hypotheses.

@@
 The parameter $\lambda_{st}$ represents the rate at which the predator ate prey species $s$ during time period $t$.
-The number of prey species $s$ found in trap $i$ during period $t$ is assumed to follow a Poisson distribution with rate parameter $\gamma_{st}$.
+The number of prey species $s$ found in trap $i$ during period $t$ is assumed to follow a Poisson distribution with rate parameter $\gamma_{st}$, $Y_{ist} \iid \mathcal{P}(\gamma_{st})$.
 The use of Poisson distributions make the following implicit assumptions: $1)$ traps independently catch the prey species of interest, $2)$ predators eat independently of each other.
@@
 Formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda} = (\lambda_{11}, \ldots, \lambda_{ST})^t$ and $\boldsymbol{\gamma} = (\gamma_{11}, \ldots, \gamma_{ST})^t$ offer insights to the relative rates at which predators eat particular prey species.
-We consider five variations on the relative magnitude of $c_{st} = \lambda_{st}/\gamma_{st}$.
-These five hypotheses each allow $c_{st}$ to vary by time, prey species, both, or neither.
+We consider four variations on the relative magnitude of $c_{st} = \lambda_{st}/\gamma_{st}$.
+These four hypotheses each allow $c_{st}$ to vary by time, prey species, both, or neither.
 
@@
 
-Because the five hypotheses are nested, a natural testing order is suggested in Figure~\ref{fig:hier}.
+Because the four hypotheses are nested, a natural testing order is suggested in Figure~\ref{fig:hier}.
 For instance, suppose interest lies in the predator's dietary preferences with respect to prey species $s_1$ and $s_2$ across three time periods.
@@
 
-\noindent Writing all five hypotheses as $\lambda_{st} = c_{st}\gamma_{st}$, we can, in the simplest cases, find analytic solutions for the maximum likelihood estimates (MLEs) of $c_{st}$, $\gamma_{st}$, and by invariance $\lambda_{st}$.
+\noindent Writing all four hypotheses as $\lambda_{st} = c_{st}\gamma_{st}$, we can, in the simplest cases, find analytic solutions for the maximum likelihood estimates (MLEs) of $c_{st}$, $\gamma_{st}$, and by invariance $\lambda_{st}$.
 Such solutions exist for the simplest model, $c_{st}=c$, provided that the data are balanced so that $J_t=J$ and $I_t=I$.

### 2015-12-09 paper/EEST/First_Submission/Response/EEST-D-15-00056_response.tex
Final edits to the response to reviewers.

@@
 \begin{enumerate}
-\item  $Y_{jst} => Y_{ist}$ Change made.
+\item  \textit{$Y_{jst} => Y_{ist}$}
+  Change made.
 
-\item  Means now presented instead of total counts.
+\item  \textit{Fig. 8 shows the total numbers of the preys in the 32 traps. I think the mean (divided by 32) is better because the ratio "c" uses gamma that are close to this mean.}
+
+  Means are now presented instead of total counts.
 This was not as simple as divide by the number of traps, since traps were left out for different numbers of days in different months.
@@
 
-\item  We doubled checked the calculations and stand by our results.
+\item  \textit{In Nov 2011, the MS says there is no significant difference between Collembola and Diptera, but the former looks 1.8-9 (Fig.
+10) while the latter looks 3.9-4.0 (Fig.
+11).
+There seems to be a big difference.
+Isn't it really significant?
+If not significant, could you explain why?}
+
+  We doubled checked the calculations and stand by our results.
 The correct intervals for November 2011 are (2.7, 7.1) for Diptera and (2.0, 3.9) for Collembola.
@@
 
-\item  ... such long descriptions are not needed.
+\item  \textit{... such long descriptions are not needed.}
 

### 2015-12-10 paper/EEST/First_Submission/Biographies/EEST-D-15-00056_biographies.tex
Added author biographies for EEST.

\documentclass[12pt]{article}

\setlength{\parindent}{0pt}
\setlength{\parskip}{12pt}

\begin{document}
\begin{center}
  \begin{Large}
    \textbf{Author Biographies for EEST-15-00056:}\\
    \textsc{Formal Modelling of Predator Preferences using Molecular Gut-Content
      Analysis}
  \end{Large}
\end{center}

\textbf{Simon Bonner} is an Assistant Professor jointly appointed between the Department of Statistical and Actuarial Sciences and the Department of Biology at the University of Western Ontario, London, Ontario, Canada (sbonner6@uwo.ca). His research focuses primarily on developing statistical methods for the analysis of ecological data with particular emphasis on Bayesian methods for analyzing data from mark-recapture studies of wild animals. This work was completed while Simon was an Assistant Professor in the Department of Statistics at the University of Kentucky.

\textbf{James Harwood} is an Associate Professor of Insect Ecology in the Department of Entomology at the University of Kentucky, Lexington, Kentucky, USA (james.harwood@uky.edu). His research seeks to understand mechanisms of foraging by generalist predators and identify their role in biological control through the integration of molecular techniques, behavioral studies and field experiments.

\end{document}

### 2015-12-13 paper/main.tex
some small changes before pulling

--- 2015-09-24 paper/main.tex
@@
 The Poisson distribution provides reasonable biological interpretation to the available data, data obtained by ecologists interested in prey electivity.
+The rate parameters of the Poisson distributions also avoid dependence on estimates of true abundance, a sensitive issue in and of itself.
 Alternatively, one could hypothesize a more complex distribution that similarly has good biological interpretation, say a distribution with more than one parameter.
@@
 Researchers should consider these potential sampling issues and attempt to determine, a priori, how well their traps will characterize prey availability according to the biology of their focal predator.
-If such issues are steady over time, then mathematically the parameter $c_{st} = \lambda_{st}/\gamma_{st}$ is directly affected.
-When human-made traps are relatively and consistently more efficient at catching prey than are the predators, $c_{st}$ will be relatively smaller.
-Conversely, when the traps are relatively and consistenly less efficient, $c_{st}$ will be relatively larger.
-Unfortunately, these factors can be difficult to control, practically, or otherwise, and are to some degree unavoidable.
-Mark-recapture methods might be implemented to adjust for trap efficiency and estimate the true density of each prey species.
-However, recapture rates are likely to be low and this type of data is beyond what is regularly collected in these types of experiments.
-
-If issues of trap efficiency or abundance are suspected to change across sampling times, we offer an alternative parameterization, details of which are found in Appendix~\ref{sec:efficiency}.
-The ratio $c_{st} = \lambda_{st} / \gamma_{st}$ implicitly treats $\gamma_{st}$ as an index proportional to true abundance.
-This assumption is valid if the traps maintain efficiency across all occassions, but otherwise is incorrect.
-Rewriting $c_{st}$ as $k_{st}d_t$, such that $\sum_{s=1}^S k_{st} = 1$ for all $t$, allows for trap efficiency to vary across time.
-The parameter $d_t$ captures changes in the absolute relationship between $\lambda_{st}$ and $\gamma_{st}$ which may be due to actual changes in abundance or to changes in trap efficiency.
-$k_{st}$ then appropriately measures the ratio of consumption to encounter rates, taking into account changes in abundance or trap efficiency.
+Mathematically, these issues will directly affect the parameter $c_{st} = \lambda_{st}/\gamma_{st}$.
+When human-made traps are relatively more efficient at catching prey than are the predators, $c_{st}$ will be relatively smaller.
+Conversely, when the traps are relatively less efficient, $c_{st}$ will be relatively larger.
+Unfortunately, these factors can be difficult to control, statistically, practically, or otherwise, and are to some degree unavoidable.
 
@@
 
-\appendix
-\section{Varying Trap Efficiency}
-\label{sec:efficiency}
-
-When there is concern about trap efficiency varying across time,  we recommend the following reparameterization of $c_{st} = \lambda_{st}/\gamma_{st}$.  Let
-
-\[ k_{st}d_t = \frac{\lambda_{st}}{\gamma_{st}}, \]
-
-\noindent such that $\sum_{s=1}^S k_{st} = 1$, for all $t$.
-The parameter $d_t$ allows for changes in trap efficiency or in actual abundance over time.
-If such changes are believed to be present, this reparameterization allows for a more natural interpretation.
-The parameter $k_{st}$ models relative consumption to encounter rates, as expected.
-
-Since this reparameterization amounts to a linear transformation of the parameters, the asymptotic distribution for $\hat{d}_t$ can be evaluated directly.
-Because of the constraint $\sum_{s=1}^S k_{st} = 1$, we find that $d_t = \sum_{s=1}^S c_{st}$ or written in matrix notation  $\mathbf{d} = \mathbf{1}' \mathbf{c}$, where $\mathbf{d} \in \mathbb{R}^T$.
-The asymptotic distribution of $\mathbf{\hat{d}}$ follows immediately, $\mathbf{\hat{d}} \stackrel{\cdot}{\sim} \mathcal{N}(\mathbf{1}'\mathbf{d}, \mathbf{1}'\Sigma\mathbf{1})$, where $\Sigma$ is the covariance matrix of the asymptotic distribution of $\mathbf{\hat{c}}$.
-
-More importantly, we can find the asymptotic distribution of $\mathbf{\hat{k}}_t = (k_{1t}, \ldots, k_{St})' \in \mathbb{R}^S$.
-We make use of the delta method \citep{Serfling:2001}.
-The asymptotic variance relies on the derivatives, $d k_{rt} / d c_{st}$.
-Let $c_{\cdot t} = \sum_{s=1}^S c_{st}$.
-Because $k_{st} = c_{st}/ c_{\cdot t}$, we can find the necessary derivatives in two cases
-
-\[
-  \frac{d k_{rt}}{d c_{st}} =
-\begin{cases}
-  \frac{-c_{rt}}{c_{\cdot t}^2}, & r \ne s \\
-  \frac{c_{\cdot t} - c_{st}}{c_{\cdot t}^2}, & r = s.
-\end{cases}
-\]
-
-\noindent  Thus, the asymptotic distribution of $\mathbf{\hat{k}}_{t}$ is $\mathcal{N}(\frac{c_{st}}{c_{\cdot t}}, (\frac{d \mathbf{k}_t}{d \mathbf{c}_t}) \Sigma (\frac{d \mathbf{k}_t}{d \mathbf{c}_t})')$.
 \nocite{Wickham:2011a,Wickham:2011,Bischl:2014,Roualdes:2014}

### 2015-12-13 paper/EEST/First_Submission/Biographies/EEST-D-15-00056_biographies.tex

@@
 
-\textbf{Simon Bonner} is an Assistant Professor jointly appointed between the Department of Statistical and Actuarial Sciences and the Department of Biology at the University of Western Ontario, London, Ontario, Canada (sbonner6@uwo.ca).
+\textbf{Edward Roualdes} is a swell guy who is more than capable of writing his own bio.
+
+\textbf{Simon Bonner} is an Assistant Professor of Environmetrics jointly appointed between the Department of Statistical and Actuarial Sciences and the Department of Biology at the University of Western Ontario, London, Ontario, Canada (sbonner6@uwo.ca).
 His research focuses primarily on developing statistical methods for the analysis of ecological data with particular emphasis on Bayesian methods for analyzing data from mark-recapture studies of wild animals.
 This work was completed while Simon was an Assistant Professor in the Department of Statistics at the University of Kentucky.
+
+\textbf{Thomas Whitney} is a PhD student in the Warnell School of Forestry and Natural Resources at the University of Georgia, Athens, GA (thomas.whitney@uga.edu).
+He is broadly interested in community ecology and uses molecular tools to characterize interactions between predators and prey, parasites and hosts, and herbivores and primary producers.
 

### 2015-12-14 paper/main.tex
update my bio

--- 2015-12-09 paper/main.tex
@@
 \institute{E. A. Roualdes 
-        \at Department of Statistics, University of Kentucky, Rm.~311 Multidisciplinary Science Building, 725 Rose Street, Lexington, KY 40536-0082, \\\email{edward.roualdes@uky.edu}
+        \at Department of Mathematics and Statistics, California State University, Chico, Chico, CA 95929, USA, \\\email{edward.roualdes@uky.edu}
         \and S. J. Bonner

### 2015-12-14 paper/main.tex
minor spelling mistakes

@@
 \institute{E. A. Roualdes 
-        \at Department of Mathematics and Statistics, California State University, Chico, Chico, CA 95929, USA, \\\email{edward.roualdes@uky.edu}
+        \at Department of Mathematics and Statistics, California State University, Chico, Chico, CA 95929, USA \\ \email{eroualdes@csuchico.edu}
         \and S. J. Bonner
         \at Department of Statistical and Actuarial Sciences, University of Western Ontario, 1151 Richmond Street, London, Ontario, N6A 5B7, Canada
-        T.D. Whitney 
+        \and T.D. Whitney 
         \at Department of Entomology, University of Kentucky, Lexington, KY 40546, USA \emph{Present address: Warnell School of Forestry and Natural Resources, University of Georgia, Athens, GA 30602, USA}
@@
 The number of prey species $s$ found in trap $i$ during period $t$ is assumed to follow a Poisson distribution with rate parameter $\gamma_{st}$, $Y_{ist} \iid \mathcal{P}(\gamma_{st})$.
-The use of Poisson distributions make the following implicit assumptions: $1)$ traps independently catch the prey species of interest, $2)$ predators eat independently of each other.
+The use of Poisson distributions makes the following implicit assumptions: $1)$ traps independently catch the prey species of interest, $2)$ predators eat independently of each other.
 
@@
 Instead, it is only possible to detect whether or not a predator consumed the prey species during a given time period, based on the rate at which prey DNA decays in the predator gut \citep{Greenstone:2013}.
-In this case we can still make inference about the predators' preferences for the different prey species by using the expectation-maximisation (EM) algorithm to compute MLEs.
+In this case we can still make inference about the predators' preferences for the different prey species by using an expectation-maximisation (EM) algorithm to compute MLEs.
 
@@
 Given the Poisson assumptions above, these variables are independent Bernoulli observations with success probability $p_{st} = P(Z_{jst}=1)= 1-\exp\{-\lambda_{st}\}$.
-Despite not observing $X_{jst}$, we can compute maximum likelihood estimates of the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$ through the EM algorithm using the complete data log-likelihood
+Despite not observing $X_{jst}$, we can compute maximum likelihood estimates of the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$ through an EM algorithm using the complete data log-likelihood
 \[
@@
 
-The EM algorithm works by iterating two steps, the E-step and M-step, until the optimum is reached \citep{Dempster:1977,McLachlan:2007}.
-Let $k$ index the iterations in the EM algorithm so that $\boldsymbol \lambda^{(k)}$ and $\boldsymbol \gamma^{(k)}$ denote the estimates computed on the $k^{th}$ M-step.
+EM algorithms work by iterating two steps, the E-step and M-step, until the optimum is reached \citep{Dempster:1977,McLachlan:2007}.
+Let $k$ index the iterations in this EM algorithm so that $\boldsymbol \lambda^{(k)}$ and $\boldsymbol \gamma^{(k)}$ denote the estimates computed on the $k^{th}$ M-step.
 The E-step consists of computing the expectation of $l_{comp}$ with respect to the conditional distribution of $X$ given the current estimates of the parameters
@@
 
-After determining which model best fits the data, more detail may be extracted through specific hypothesis test of the elements of $c_{st}$, or in vector notation as $\mathbf{c} \in \mathbb{R}^{S\cdot T}$.
+After determining which model best fits the data, more detail may be extracted through specific hypothesis tests of the elements of $c_{st}$, or in vector notation as $\mathbf{c} \in \mathbb{R}^{S\cdot T}$.
 Let the elements of $\hat{\mathbf{c}}$ be the maximum likelihood estimates, $\hat{c}_{st}$, as found via the framework above.
@@
 As mentioned above, the estimate of $c_5$ is infinite in this case.
-However, the EM algorithm will always provide a finite estimate for all parameters when it terminates.
+However, this EM algorithm will always provide a finite estimate for all parameters when it terminates.
 In this case, we set $\tau=10^{-5}$ and this caused the algorithm to terminate with $\hat c_5$ between $3.5$ and $4.0$.
@@
 A more recent method for testing prey preferences was developed by \citet{Agusti:2003}.
-This method attempts to test the hypothesis that individuals are eating prey at random using a test based on Monte Carlo simulation (more specificially the bootstrap).
+This method attempts to test the hypothesis that individuals are eating prey at random using a test based on Monte Carlo simulation (more specifically the bootstrap).
 Unfortunately, the method described does not attempt to model the distribution of the number of prey items eaten, as we have done, and therefore we provide a methodological advance on the approach described previously.
@@
 When human-made traps are relatively and consistently more efficient at catching prey than are the predators, $c_{st}$ will be relatively smaller.
-Conversely, when the traps are relatively and consistenly less efficient, $c_{st}$ will be relatively larger.
+Conversely, when the traps are relatively and consistently less efficient, $c_{st}$ will be relatively larger.
 Unfortunately, these factors can be difficult to control, practically, or otherwise, and are to some degree unavoidable.
@@
 The ratio $c_{st} = \lambda_{st} / \gamma_{st}$ implicitly treats $\gamma_{st}$ as an index proportional to true abundance.
-This assumption is valid if the traps maintain efficiency across all occassions, but otherwise is incorrect.
+This assumption is valid if the traps maintain efficiency across all occasions, but otherwise is incorrect.
 Rewriting $c_{st}$ as $k_{st}d_t$, such that $\sum_{s=1}^S k_{st} = 1$ for all $t$, allows for trap efficiency to vary across time.

### 2016-01-25 paper/EEST/Second_Submission/Biographies/EEST-D-15-00056_biographies.tex
plots to grey-scale

--- 2015-12-13 paper/EEST/First_Submission/Biographies/EEST-D-15-00056_biographies.tex
@@
 
-\textbf{Edward Roualdes} is a swell guy who is more than capable of writing his own bio.
+\textbf{Edward A.
+Roualdes} is an Assistant Professor of Statistics in the Department of Mathematics and Statistics at California State University, Chico, Chico, California, USA (eroualdes@csuchico.edu).
+His research interests are shared between Bayesian hierarchical models and advancing statistical methods for ecologists.
+This work was completed while Edward was a PhD student in the Department of Statistics at the University of Kentucky.
 
@@
 \end{document}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2016-01-25 paper/EEST/Second_Submission/Response/EEST-D-15-0056R1_response.tex
plots to grey-scale

--- 2015-12-09 paper/EEST/First_Submission/Response/EEST-D-15-00056_response.tex
@@
   \begin{Large}
-    \textbf{Response to Reviews of EEST-15-00056:}\\
+    \textbf{Response to Reviews of EEST-15-00056R1:}\\
     \textsc{Formal Modelling of Predator Preferences using Molecular Gut-Content
@@
 
-\section{Associate Editor:}
+\section{Reviewer \#1:}
+\begin{enumerate}
+\item  \textit{The authors responded everything I suggested at the first reviewing. I have no additional comment and recommendation, except one typing error; P4 L33 (L137) Yjst to Yist}
 
-\begin{enumerate}
-\item \textit{ ... called the probability mass function.}
-
-Thank you for your attention to detail.  Changes throughout the manuscript have been made accordingly.
+Fixed.
 
 \end{enumerate}
-\section{Reviewer \#1:}
-\begin{enumerate}
-\item  \textit{ ... commend the authors on an excellent contribution.}
+\end{document}
 
-  Thank you.
-
-\end{enumerate}
-
-\section{Reviewer \#2:}
-
-\subsection{Major Comments}
-
-\begin{enumerate}
-\item \textit{ The method estimates the ratio between lambda and gamma, not lambda and gamma themselves.
-The method does not estimate true abundances of preys, and I am not sure why without estimating the true abundance, why the ratio can be estimated so accurately as shown in the Figures?
-Could you explain some essence (trick?) of the success?}
-
-% In fact, we estimate gamma and the ratio of lambda and gamma, denoted by $c$, directly.
-Estimates of lambda are found via these two quantities and the invariance property of maximum likelihood.
-In the case of fully observed data, there is no trick.
-We use maximum likelihood methods to estimate parameters of interest, as stated in line 121.
-In the case of unobserved count data, this EM algorithm provides us a means to estimate the parameters of interest.
-When unobserved counts are present we discuss times at which the EM algorithm will fail.
-Though, this discussion itself has been reduced as per Minor Point 4.
-
-This is an important point that we overlooked in our initial submission.
-We had considered $\gamma_{st}$ as a measure of absolute abundance of prey species $s$ on sampling occasion $t$, and you are absolutely correct that this is not the case.
-Instead, it represents the mean number of individuals from this prey species caught in the traps.
-To relate this to absolute abundance would require extra information on the efficiency of the traps that could be obtained from a mark-recapture study, for example, but this information is not available for our particular study and probably not available for many other studies either.
-
-However, we can interpret $\gamma_{st}$ as a relative measure of abundance under some assumptions so that changes in $\gamma_{s_1t}/\gamma_{s_2t}$ represent changes in the relative abundance of species $s_1$ and $s_2$.
-This is the case if the capture efficiency of the traps is assumed to be constant for all species over all occasions.
-Under weaker assumptions we can interpret $\gamma_{st}$ as a relative measure of the availability of prey in the environment, rather than abundance, and $c_{st}$ as a relative measure of preference given abundance.
-We might assume that trap efficiency for a prey species depends on how much of the time individuals in that species are trappable.
-In this case, trappability might be interpreted as availability to the prey.
-In our application, prey species that are actively foraging in the leaf litter for longer periods may have higher trappability and also be more available to the spiders.
-In this case, we can interpret $\gamma_{s_1t}/\gamma_{s_2t}$ as a relative measure of the availability of prey and $c_{st}$ as a relative measure of preference given availability.
-
-With respect to the figures, the apparent trick is simply a result of our incorrect interpretation of $\gamma_{st}$.
-We generated data for the simulations by simulating the number of individuals from each prey species caught in traps given $\gamma_{st}$.
-Our model is able to estimate the ratio of $\lambda_{st}$ and $\gamma_{st}$ accurately, but it was incorrect of us to interpret this as estimating absolute abundance accurately.
-
-\item \textit{ The ratio is between the preference observed in gut-content and captured preys in the traps produced by human.
-The denominator is not the true abundance, but expresses how much that prey species are easily captured by that trapping system.
-What are ecological implications of the ratio between observed gut contents and the "efficiency" of human-produced traps?
-Of course, this could be unavoidable for population dynamics of animal species and for food-web studies, and I think that a few sentences about this point should be added as far as a journal is not specified to animal ecology.}
-
-Following from our response to the last point, we have changed the text so it is clear that $\gamma_{st}$ is a measure of relative, not absolute, abundance.
-We have also added two paragraphs to the Discussion commenting on the issues of trap efficiency and the assumptions needed to interpret $\gamma_{st}$ as a relative measure of abundance.
-
-The second paragraph we have added to the Discussion also introduces a new parameterization that helps to address these concerns.
-The new parametrization applies when the efficiency of traps cannot be assumed constant and explicitly separates changes in efficiency over time.
-Appendix A provides the mathematical details behind the new paramterization.
-
-Recognizing $\gamma_{st}$ as a measure of relative abundance also led us to remove the first of our hypotheses about the way in which $c_{st}$ might vary, namely $c_{st} = 1$.
-This leaves only four hypotheses on the way $c_{st}$ might change.
-As in our previous response, the relationship between the samples collected from the predators and the traps depends on the traps' efficiency and may be affected by, among other things, how big the traps are.
-Suppose that the experiment is repeated on two occasions with exactly the same properties that the traps double in size on the second occasion.
-If $c_{st}=1$ on the first occasion then you would expect $c_{st}=.5$ on the second occasion, but the biological process has not changed.
-This means that the hypothesis $c_{st} = 1$ does not have the interpretation in terms of absolute abundance that we previously thought.
-
-\end{enumerate}
-
-\subsection{Minor Points:}
-
-\begin{enumerate}
-\item  $Y_{jst} => Y_{ist}$ Change made.
-
-\item  Means now presented instead of total counts.
-This was not as simple as divide by the number of traps, since traps were left out for different numbers of days in different months.
-Nonetheless, this was a good point.
-
-\item  We doubled checked the calculations and stand by our results.
-The correct intervals for November 2011 are (2.7, 7.1) for Diptera and (2.0, 3.9) for Collembola.
-The overlap of the intervals is due to large variation in the point estimates, and hence we fail to see the difference between the two point estimates.
-
-\item  ... such long descriptions are not needed.
-
-A paragraph was removed, namely that which was the last paragraph in Section 3 Simulation Study.
-\end{enumerate}
-\end{document}
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2016-01-25 paper/main.tex
plots to grey-scale

@@
 % !143, 146
-The mass function of $Y_{jst}$ is exactly as in Section~\ref{sec:count} and so we focus on deriving the joint probability mass function of $X_{jst}$ and $Z_{jst}$.
+The mass function of $Y_{ist}$ is exactly as in Section~\ref{sec:count} and so we focus on deriving the joint probability mass function of $X_{jst}$ and $Z_{jst}$.
 With the distribution of $Z_{jst}$ given above, we can compute $f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda})$ by noting that $X_{jst}=0$ with probability 1 if $Z_{jst}=0$, and that $[X_{jst}|Z_{jst}=0]$ has a truncated Poisson distribution with mass function

### 2016-03-01 paper/main.tex
attempted to unify syntax; added citation to vignette; removed some useless parameters; changed my email address

@@
 \institute{E. A. Roualdes 
-        \at Department of Mathematics and Statistics, California State University, Chico, Chico, CA 95929, USA \\ \email{eroualdes@csuchico.edu}
+        \at Department of Statistics, University of Kentucky, Lexington, KY 40506, USA \emph{Present address: Department of Mathematics and Statistics, California State University, Chico, Chico, CA 95929, USA} \\ \email{eroualdes@csuchico.edu}
         \and S. J. Bonner
-        \at Department of Statistical and Actuarial Sciences, University of Western Ontario, 1151 Richmond Street, London, Ontario, N6A 5B7, Canada
+        \at Department of Statistics, University of Kentucky, Lexington, KY 40506, USA \emph{Present address: Department of Statistical and Actuarial Sciences and Department of Biology, University of Western Ontario, 1151 Richmond Street, London, Ontario, N6A 5B7, Canada}
         \and T.D. Whitney 

### 2016-03-01 R/contrasts.R

@@ -98,29 +98,29 @@ testC <- function(x, b, mu = 0, alternative = c("two.sided", "less", "greater"),
     alpha <- 1-conf.level
     a2 <- alpha/2
     alternative <- match.arg(alternative)
-    if ( alternative == 'two.sided' ) {
+    if ( alternative == "two.sided" ) {
         pval <- 2*pnorm(-abs(stat), mean=mu, sd=stdev)
         interval <- qnorm(c(a2, 1-a2), mean=stat, sd=stdev)
-    } else if ( alternative == 'less' ) {
+    } else if ( alternative == "less" ) {
         pval <- pnorm(stat, mean=mu, sd=stdev)
         interval <- c(-Inf, qnorm(conf.level, mean=stat, sd=stdev))
-    } else if ( alternative == 'greater' ) {
+    } else if ( alternative == "greater" ) {
         pval <- pnorm(stat, mean=mu, sd=stdev, lower.tail=F)
         interval <- c(qnorm(alpha, mean=stat, sd=stdev), Inf)
     }
 
     ## info to fill in print method of class htest
-    names(stat) <- 'b^t * C'
-    names(mu) <- 'linear contrast'
-    method <- paste('Linear Contrast: ', paste(t(b), collapse=' '), sep='')
-    attr(interval, 'conf.level') <- conf.level
+    names(stat) <- "b^t * C"
+    names(mu) <- "linear contrast"
+    method <- paste("Linear Contrast: ", paste(t(b), collapse=" "), sep="")
+    attr(interval, "conf.level") <- conf.level
     est <- as.vector(C)
-    names(est) <- paste('mean of c_', lc, sep='')
+    names(est) <- paste("mean of c_", lc, sep="")
 
     out <- list(statistic = stat, p.value = pval,
                 conf.int = interval, null.value = mu,
                 alternative = alternative, method = method,
                 estimate = est, data.name=x$data.name)
-    class(out) <- 'htest'
+    class(out) <- "htest"
     out
 }

### 2016-03-01 R/estimate1.R

@@ -42,9 +42,8 @@ est1 <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
             ## limit iterations
             em_iter <- em_iter+1
             if ( em_iter > em_maxiter ) {
-                stop(sprintf('est1: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
+                stop(sprintf("est1: max EM iterations, %d, reached. Please adjust accordingly.", em_maxiter))
             }
-
         }
 
         ## calc standard error with est params
@@ -59,8 +58,8 @@ est1 <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
         ## calc loglik with est params
         loglik <- llEM(Xdst, Ydst, gammaHat, gammaHat, J, I)
         
-        list('lambda' = as.matrix(gammaHat), 'gamma' = as.matrix(gammaHat),
-             'll' = loglik, 'var' = var)
+        list(lambda=as.matrix(gammaHat), gamma=as.matrix(gammaHat),
+             ll=loglik, var=var)
         
     } else {
         
@@ -73,7 +72,7 @@ est1 <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
         ## calc loglik with est params
         loglik <- ll(Xdst, Ydst, gammaHat, gammaHat, J, I)
 
-        list('lambda' = as.matrix(gammaHat), 'gamma' = as.matrix(gammaHat),
-             'll' = loglik, 'var' = solve(Info))        
+        list(lambda=as.matrix(gammaHat), gamma=as.matrix(gammaHat),
+             ll=loglik, var=solve(Info))        
     }
 }

### 2016-03-01 R/estimateC.R

@@ -55,7 +55,7 @@ estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
             
             ## limit iterations
             if ( em_iter > em_maxiter ) {
-                stop(sprintf('estC: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
+                stop(sprintf("estC: max EM iterations, %d, reached. Please adjust accordingly.", em_maxiter))
             }
             
         }
@@ -82,8 +82,8 @@ estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
         ## calc log-lik with est params
         loglik <- llEM(Xdst, Ydst, NA, gammaHat, J, I, as.numeric(cHat))
         
-        list('c' = cHat, 'gamma' = as.matrix(gammaHat), 'em_iters' = em_iter,
-             'll' = loglik, 'var' = var)
+        list(c=cHat, gamma=as.matrix(gammaHat), em_iters=em_iter,
+             ll=loglik, var=var)
     } else {
         if (BALANCED) {
 
@@ -107,8 +107,8 @@ estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
             ## calc log-lik with est params
             loglik <- ll(Xdst, Ydst, NA, gammaHat, J, I, cHat)
             
-            list('gamma' = as.matrix(gammaHat), 'c' = cHat,
-                 'll' = loglik, 'var' = solve(Info))
+            list(gamma=as.matrix(gammaHat), c=cHat,
+                 ll=loglik, var=solve(Info))
         } else {
             
             ## some numbers
@@ -149,7 +149,7 @@ estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
                 iter <- iter+1
                 
                 if ( iter > maxiter ) {
-                    stop(sprintf('estC: %d not sufficient iterations for simultaneous equations.', maxiter))
+                    stop(sprintf("estC: %d not sufficient iterations for simultaneous equations.", maxiter))
                 }
             }
             
@@ -167,8 +167,8 @@ estC <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
             ## calc log-lik with est params
             loglik <- ll(Xdst, Ydst, NA, gammaHat, J, I, cHat)
             
-            list('gamma' = as.matrix(gammaHat), 'c' = cHat, 'iters' = iter,
[…]

### 2016-03-01 R/estimateCs.R

@@ -65,7 +65,7 @@ estCs <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
             
             ## limit iterations
             if ( em_iter > em_maxiter ) {
-                stop(sprintf('estCs: max EM iterations, %d, reached. Please adjust accordingly.', em_maxiter))
+                stop(sprintf("estCs: max EM iterations, %d, reached. Please adjust accordingly.", em_maxiter))
             }
             
         }
@@ -95,8 +95,8 @@ estCs <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
         ## calc log-lik with est params
         loglik <- llEM(Xdst, Ydst, NA, gammaHat, J, I, cHat)
         
-        list('c' = cHat, 'gamma' = as.matrix(gammaHat), 'em_iters' = em_iter,
-             'll' = loglik, 'var' = var)
+        list(c=cHat, gamma=as.matrix(gammaHat), em_iters=em_iter,
+             ll=loglik, var=var)
     } else {
 
         XYdst <- Xdst + Ydst
@@ -132,7 +132,7 @@ estCs <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
             iter <- iter+1
             
             if ( iter > maxiter ) {
-                stop(sprintf('estCt: %d not sufficient iterations for simultaneous equations.', maxiter))
+                stop(sprintf("estCt: %d not sufficient iterations for simultaneous equations.", maxiter))
             }
             
         }
@@ -154,7 +154,7 @@ estCs <- function(Xdst, Ydst, J, I, EM, em_maxiter, BALANCED) {
         ## calc log-lik with est params
         loglik <- ll(Xdst, Ydst, NA, gammaHat, J, I, cHat)
             
-        list('gamma' = as.matrix(gammaHat), 'c' = cHat, 'iters' = iter,
-             'll' = loglik, 'var' = solve(Info))
+        list(gamma=as.matrix(gammaHat), c=cHat, iters=iter,
+             ll=loglik, var=solve(Info))
     }
 }

### 2016-03-01 R/summary.predPref.R

@@ -10,24 +10,24 @@
 summary.predPref <- function(object, ..., sig.level=0.05) {
     out <- list()
     if ( object$LRT$p.value > sig.level ) {
-        out[['estimates']] <- object$null
+        out[["estimates"]] <- object$null
     } else {
-        out[['estimates']] <- object$alt
+        out[["estimates"]] <- object$alt
     }
 
     ## indices of parameter estimates
-    out[['S']] <- ncol(out[['estimates']][['gamma']])
-    out[['s']] <- seq_len(out[['S']])
-    out[['T']] <- nrow(out[['estimates']][['gamma']])
-    out[['t']] <- seq_len(out[['T']])
+    out[["S"]] <- ncol(out[["estimates"]][["gamma"]])
+    out[["s"]] <- seq_len(out[["S"]])
+    out[["T"]] <- nrow(out[["estimates"]][["gamma"]])
+    out[["t"]] <- seq_len(out[["T"]])
     
-    out[['p.value']] <- object$LRT$p.value
-    out[['df']] <- object$LRT$df
-    out[['Lambda']] <- object$LRT$Lambda
-    out[['loglikH0']] <- object$loglikH0
-    out[['loglikH1']] <- object$loglikH1
-    out[['hypotheses']] <- object$hypotheses
-    class(out) <- 'summary.predPref'
+    out[["p.value"]] <- object$LRT$p.value
+    out[["df"]] <- object$LRT$df
+    out[["Lambda"]] <- object$LRT$Lambda
+    out[["loglikH0"]] <- object$loglikH0
+    out[["loglikH1"]] <- object$loglikH1
+    out[["hypotheses"]] <- object$hypotheses
+    class(out) <- "summary.predPref"
     out
 }
 
@@ -45,31 +45,31 @@ print.summary.predPref <- function(x, ...) {
     cat("Parameter Estimates:\n")
 
     ## indices of estimated parameters 
-    indices <- unlist(lapply(x[['s']],
-                             function(y) sapply(x[['t']],
-                                                function(x) paste(x, '_', y, sep=''))))
+    indices <- unlist(lapply(x[["s"]],
+                             function(y) sapply(x[["t"]],
+                                                function(x) paste(x, "_", y, sep=""))))
     
     if ( !is.null(x$estimates$c) ) {
         est <- cbind(c(as.vector(x$estimates$c),
                        as.vector(x$estimates$gamma)),
[…]

### 2016-03-01 R/util.R

@@ -2,15 +2,13 @@
 ##'
 ##' @param data a dataframe
 ##' @param vars column variables in data to sum over
-##' @param by extra variables to sum by
-getTimeCounts <- function(data, vars, by) {
-    if ( missing(by) ) {
-        by <- 'time'
-    } else {
-        by <- c('time', by)
-    }
-    ## ddply(data, by, function(dfr, idx) colSums(as.matrix(dfr[,idx])), vars)
-    ddply(data, by, colwise(sum, vars))
+getTimeCounts <- function(data, vars) {
+    ## if ( missing(by) ) {
+    ##     by <- 'time'
+    ## } else {
+    ##     by <- c('time', by)
+    ## }
+    ddply(data, "time", colwise(sum, vars))
 }
 
 ##' checks convergence of the parameters for the estimation functions
@@ -37,9 +35,9 @@ sumST <- function(mat) {
 }
 
 ## colors stolen from http://geography.uoregon.edu/datagraphics/color_scales.htm
-cols <- c('orange1' = "#FFBF80", 'orange2' = "#FF8000",
-          'yellow1' = "#FFFF99", 'yellow2' = "#FFFF33",
-          'green1' = "#B2FF8C", 'green2' = "#33FF00",
-          'blue1' = "#A6EDFF", 'blue2' = "#1AB2FF",
-          'purple1' = "#CCBFFF", 'puple2' = "#664CFF",
-          'red1' = "#FF99BF", 'red2' = "#E61A33")
+cols <- c(orange1="#FFBF80", orange2="#FF8000",
+          yellow1 = "#FFFF99", yellow2="#FFFF33",
+          green1 = "#B2FF8C", green2="#33FF00",
+          blue1 = "#A6EDFF", blue2="#1AB2FF",
+          purple1 = "#CCBFFF", puple2="#664CFF",
+          red1 = "#FF99BF", red2="#E61A33")
